### 2020-10-06 retro-concurrency.tex
Initial commit

\documentclass[sigplan,10pt,review,anonymous]{acmart}\settopmatter{printfolios=true,printccs=false,printacmref=false}

\begin{document}

\title{Retrofitting Concurrency onto OCaml}

\begin{abstract}
  A clear and well-documented \LaTeX\ document is presented as an
  article formatted for publication by ACM in a conference proceedings
  or journal publication. Based on the ``acmart'' document class, this
  article presents and explains many of the common variations, as well
  as many of the formatting elements an author may use in the
  preparation of the documentation of their work.
\end{abstract}

\maketitle

\section{Introduction}

\bibliographystyle{ACM-Reference-Format}
\bibliography{sample-base}

\end{document}

### 2020-10-26 retro-concurrency.tex
More semantics
More work on semantics
Semantics is completed
Low-level semantics wip. Better to use figures.
More stuff

@@
 \documentclass[sigplan,10pt,review,anonymous]{acmart}\settopmatter{printfolios=true,printccs=false,printacmref=false}
+
+%% Some recommended packages.
+\usepackage{booktabs}   %% For formal tables:
+                        %% http://ctan.org/pkg/booktabs
+\usepackage{subcaption} %% For complex figures with subfigures/subcaptions
+                        %% http://ctan.org/pkg/subcaption
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%% Custom definitions
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\usepackage{balance}
+\usepackage{url}
+\usepackage{enumitem}
+\usepackage{hyperref}
+\usepackage{listings}
+\usepackage{xspace}
+\usepackage{cleveref}
+\usepackage{multirow}
+\usepackage{microtype}
+\usepackage{soul,xcolor}
+\usepackage{mathpartir}
+\usepackage{amsmath,amsthm}
+\usepackage[utf8]{inputenc}
+\input{unicode-preamble.tex}
+\usepackage{mathtools}
+\usepackage{tikz}
+\usetikzlibrary{calc,decorations.pathmorphing,shapes}
+\usepackage{stmaryrd}
+
+\crefformat{section}{§#2#1#3}
+\newtheorem{theorem}{Theorem}
+\newtheorem{definition}[theorem]{Definition}
+
+\newenvironment{nop}{}{}
+\newenvironment{smathpar}{
+\begin{nop}\small\begin{mathpar}}{
+\end{mathpar}\end{nop}\ignorespacesafterend}
+
+\newcommand{\kc}[1]{{\color{red} {\it [KC says: #1]}}}
+\newcommand{\am}[1]{{\color{blue} {\it [AM says: #1]}}}
+
+\definecolor{Bittersweet}{rgb}{1.0, 0.44, 0.37}
+\definecolor{MidnightBlue}{rgb}{0.0, 0.2, 0.4}
+\definecolor{BrightBlue}{rgb}{0.0, 0.2, 0.7}
+\definecolor{byzantine}{rgb}{0.74, 0.2, 0.64}
+\definecolor{caribbeangreen}{rgb}{0.0, 0.8, 0.6}
+
+\lstset{
+			language=[Objective]caml,
+      basicstyle=\ttfamily\small,
+      flexiblecolumns=false,
+			tabsize=2,
+      escapechar=',
+      %basewidth={0.5em,0.45em},
+      %aboveskip={3pt},
+      %belowskip={3pt},
+			breaklines = true,
+      breakatwhitespace = true,
+      keywordstyle=\color{Bittersweet}\bfseries,
+      commentstyle=\color{blue}\itshape,
+      stringstyle=\color{MidnightBlue},
+			keywords=[1]{effect,perform,continue,discontinue},
+			keywordstyle=[1]\color{Bittersweet}\bfseries,
+			keywords=[2]{},
+			keywordstyle=[2]\color{byzantine}\bfseries,
+			keywords=[3]{int,string}, %types%
+			keywordstyle=[3]\color{caribbeangreen},
+			classoffset=1,
+			upquote=true,
+			keywordstyle=\color{byzantine}\bfseries,
+			classoffset=0,
+			mathescape=true,
+			numberstyle=\tiny\color{gray},
+			numbersep=2pt
+}
+
+\lstnewenvironment{code}[1][]
+    { % \centering
+      \lstset{
+				basicstyle=\ttfamily\footnotesize,
+				#1}%
+      \csname lst@setfirstlabel\endcsname}
+    { %\centering
+      \csname lst@savefirstlabel\endcsname}
+
+\lstMakeShortInline[columns=fullflexible]|
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%% Semantics definitions
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\newcommand{\olam}[2]{\lambda^o #1. #2}
+\newcommand{\llambda}{\lambda\!\!\lambda}
+\newcommand{\clam}[2]{\lambda^c #1. #2}
+\newcommand{\lam}[2]{\Lambda #1. #2}
+
+\newcommand{\env}{\epsilon}
+\newcommand{\envext}[3]{#1[#2 \mapsto #3]}
+
+\newcommand{\oclos}[3]{\llparenthesis \olam{#1}{#2}, #3 \rrparenthesis}
+\newcommand{\cclos}[3]{\llparenthesis \clam{#1}{#2}, #3 \rrparenthesis}
+\newcommand{\clos}[3]{\llparenthesis \lam{#1}{#2}, #3 \rrparenthesis}
+\newcommand{\kw}[1]{\text{\bf #1}}
+\newcommand{\effval}[2]{\textsf{do} ~#1 ~#2}
+\newcommand{\exnval}[1]{\textsf{exn} ~#1}
+
+\newcommand{\handle}[2]{\kw{match} ~#1 ~\kw{with} ~#2}
+\newcommand{\throw}[2]{\kw{raise} ~#1 ~#2}
+\newcommand{\perform}[2]{\kw{perform} ~#1 ~#2}
+
+\newcommand{\caseval}[2]{\kw{return}~#1 \mapsto #2}
+\newcommand{\caseexn}[3]{\kw{exception} ~#1 ~#2 \mapsto #3}
+\newcommand{\caseeff}[4]{\kw{effect} ~#1 ~#2 ~#3 \mapsto #4}
+
+\newcommand{\farg}[2]{\langle #1 ~#2 \rangle_a}
+\newcommand{\ffun}[1]{\langle #1 \rangle_f}
+\newcommand{\faritha}[3]{\langle #1 ~#2 ~#3 \rangle_{b1}}
+\newcommand{\farithb}[2]{\langle #1 ~#2 \rangle_{b2}}
+
+\newcommand{\kcons}{\lhd}
+\newcommand{\fiber}{\varphi}
+\newcommand{\fl}{\psi} % Frame List
+\newcommand{\hc}{\eta} % Handler Closure
+
+\newcommand{\cstack}{\gamma} % C Stack
+\newcommand{\ostack}{\omega} % OCaml Stack
+\newcommand{\cstacka}[2]{\big \lceil #1, #2 \big \rceil_c} % C stack with arguments
+\newcommand{\ostacka}[2]{\big \lceil #1, #2 \big \rceil_o} % OCaml stack with arguments
+\newcommand{\ostackemp}{\bullet}
+\newcommand{\stack}{\sigma}
+
+\newcommand{\term}{\tau}
+\newcommand{\config}{\mathfrak{C}}
+\newcommand{\configa}[3]{\|#1,#2,#3\|}
+
+\newcommand{\ostep}{\xrightarrow{o}}
+\newcommand{\cstep}{\xrightarrow{c}}
+\newcommand{\step}{\rightarrow}
 
@@
 
-\title{Retrofitting Concurrency onto OCaml}
+\title{Retrofitting Effect Handlers onto OCaml}
 
 \begin{abstract}
-  A clear and well-documented \LaTeX\ document is presented as an
-  article formatted for publication by ACM in a conference proceedings
-  or journal publication. Based on the ``acmart'' document class, this
-  article presents and explains many of the common variations, as well
-  as many of the formatting elements an author may use in the
-  preparation of the documentation of their work.
+	Effect handlers have gained immense interest in the recent years as a means
+	for modular programming with user-defined effects. The runtime semantics of
+	effect handlers is particularly attractive as they allow advanced control
+	flow mechanisms such as first-class continuations, generators, async/await,
+	coroutines to be composably expressed. In this paper, we present the design,
+	a full-fledged implementation and evaluation of effect handlers for OCaml, an
+	industrial-strength, multi-paradigm programming language. OCaml tends to be
+	widely used for systems programming, and is compatible with program analysis
+	tools that inspect the stack such as debuggers and profilers. Retrofitting
+	effect handlers onto OCaml is challenging since OCaml does not have any
+	non-local control flow mechanisms other than exceptions. Our implementation
+	of effect handlers for OCaml imposes negligible overhead on code that does
+	not use effect handlers, and remains compatible with program analysis tools
+	that inspect the stack.
 \end{abstract}
@@
 
+Effect Handlers~\cite{Plotkin09} provide a modular foundation for user-defined
+effects. The key idea is to separate the definition of the effectful
+\emph{operations} from their interpretations, which are defined by
+\emph{handlers} of the effects. For example, |effect Read : in_channel -> string|
+declares an effect |Read|, which is parametrised with an input channel of type
+|in_channel|, which when \emph{performed} returns a |string| value. A
+computation can perform the |Read| effect without knowing how the |Read| effect
+is implemented. This computation may be enclosed by different handlers that
+interpret |Read| differently. For example, |Read| may be implemented by
+performing a blocking read on the input channel or performing the read
+asynchronously by offloading it to an event loop such as |libuv|. Thanks to the
+separation of effectful operations from their interpretations, effect handlers
+enable new approaches to modular programming. A computation may perform many
+different effects, which are interpreted by one or more enclosing handlers.
+
+What makes effect handlers quite powerful is that they provide a structured
+form of delimited continuations, which enables advanced non-local control-flow
+mechanisms such as resumable exceptions, coroutines, generators and
+asynchronous I/O to be composably expressed. As an example, consider the task
+of handling division-by-zero case in a calculator application. Different
+calculators handle division-by-zero differently. In the iOS 13 Calculator App,
+division-by-zero returns |Error|. In the Google Search Calculator App |0/0|
+returns to |Error|, and for any other |n| not equal to |0|, |n/0| returns
+|Infinity| with the same sign as |n|. In the OCaml top-level, division-by-zero
+raises |Division_by_zero| exception. How can we capture these different
+behaviours without changing the evaluator?
+
+Common Lisp Conditions~\cite{Conditions} are an elegant way to solve this
+problem. Unlike exceptions where the library decides whether an error is fatal
+or not, conditions allow the client of the library to make that decision.
+Common Lisp Conditions provide a restart mechanism to resume execution from the
+point where the condition was signalled. On division-by-zero, the calculator
+signals the handler and lets the handler decide the course of action based on
+the desired semantics. We can implement this mechanism using effect handlers as
+shown in Figure~\ref{fig:}.
+
+The search for an expressive effect system which guarantees that all the
+effects performed in the program are handled (\emph{effect safety}) in the
+presence of advanced features such as polymorphism, modularity and generativity
+is an active area of research~\cite{Leijen14, Biernacki19, Biernacki20,
+Hillerstrom20}. We do not focus on this question in this paper, and the
+implementation of effect handlers in Multicore OCaml does not guarantee effect
+safety.
+
+\section{Background: OCaml Stacks}
+
+\section{Handlers by example}
+
+\section{Semantics}
+
+The initial configuration for an expression $e$ is
+$\configa{e}{\emptyset}{\cstacka{[]}{\bullet}}$. We define {\sf continue} and
+{\sf discontinue} as follows:
+
+\begin{smathpar}
+	\begin{array}{rcl}
+		\textsf{continue} ~k ~e & = & k ~(\olam{x}{x}) ~e \\
+		\textsf{discontinue} ~k ~l ~e & = & k ~(\olam{x}{\throw{l}{x}}) ~e
+	\end{array}
+\end{smathpar}
+
+\begin{figure*}
+	\begin{minipage}[t]{0.49\linewidth}
+	\begin{smathpar}
+	\begin{array}{rrcl}
+		\text{Constants} & n & \coloneqq & ℤ \\
+		\text{Abstractions} & \Lambda & \coloneqq & \lambda^o \mid \lambda^c \\
+		\text{Expressions} & e & \coloneqq & n \mid x \mid e ~e \mid  \lam{x}{e} \mid e \odot e \mid \throw{l}{e} \\
+											 &   & \mid      & \handle{e}{h} \mid \perform{l}{e} \\
+		\text{Handlers} & h & \coloneqq & \{\caseval{x}{e}\} \mid \{\caseexn{l}{x}{e}\} \uplus h \\
+		               &   & \mid & \{\caseeff{l}{x}{k}{e}\} \uplus h \\
+		\text{Values} & v & \coloneqq & n \mid k \mid \clos{x}{e}{\env} \mid \effval{l}{k} \mid \exnval{l} \\
+		\text{Frames} & r & \coloneqq & \farg{e}{\env} \mid \ffun{v} \mid \faritha{\odot}{e}{\env} \mid \farithb{\odot}{ℕ} \\
+		\text{Environments} & \env & \coloneqq & \emptyset \mid \envext{\env}{x}{v} \\
+	\end{array}
+	\end{smathpar}
+	\end{minipage}
+	\begin{minipage}[t]{0.49\linewidth}
+	\begin{smathpar}
+	\begin{array}{rrcl}
+		\text{Handler Closures} & \hc & \coloneqq & (h,\env) \\
+		\text{Frame List} & \fl & \coloneqq & [] \mid r :: \fl \\
+		\text{Fibers} & \fiber & \coloneqq & (\fl, \hc) \\
+		\text{Continuations} & k & \coloneqq & [] \mid \fiber \lhd k \\
+		\text{C stacks} & \cstack & \coloneqq & \cstacka{\fl}{\ostack}\\
+		\text{OCaml stacks} & \ostack & \coloneqq & \ostacka{k}{\cstack} \mid \ostackemp \\
+		\text{Stacks} & \stack & \coloneqq & \cstack \mid \ostack \\
+		\text{Terms} & \term & \coloneqq & e \mid v \\
+		\text{Configurations} & \config & \coloneqq & \configa{\tau}{\env}{\stack}
+	\end{array}
+	\end{smathpar}
+	\end{minipage}
+	\caption{Term Syntax}
+	\label{sem:syntax}
+\end{figure*}
+
+\begin{figure*}
+	\begin{smathpar}
+		\begin{array}{rrcl}
+			\textsc{Var}     & (x,\env, \fl) & \rightsquigarrow
+			                 & (\env(x), \env, \fl) \\
+			\textsc{Arith1}  & (e_1 \odot e_2, \env, \fl) & \rightsquigarrow
+			                 & (e_1,\env, \faritha{\odot}{e_2}{\env}::\fl) \\
+			\textsc{Arith2}  & (n_1, \_, \faritha{\odot}{e_2}{\env}::\fl) & \rightsquigarrow
+			                 & (e_2, \env, \farithb{\odot}{n_1}::\fl) \\
+			\textsc{Arith3}  & (n_2, \env, \farithb{\odot}{n_1}::\fl) & \rightsquigarrow
+			                 & (\llbracket n_1 \odot n_2 \rrbracket, \env, \fl) \\
+			\textsc{App1}    & (e_1 ~e_2, \env, \fl) & \rightsquigarrow
+			                 & (e_1, \env, \farg{e_2}{\env}::\fl) \\
+			\textsc{App2}    & (\lam{x}{e}, \env, \fl) & \rightsquigarrow
+			                 & (\clos{x}{e}{\env}, \env, \fl) \\
+			\textsc{App3}    & (\clos{x}{e_1}{\env_1}, \_, \farg{e_2}{\env_2}::\fl) & \rightsquigarrow
+			                 & (e_2, \env_2, \ffun{\clos{x}{e_1}{\env_1}}::\fl) \\
+			\textsc{Resume1} & (k,\_,\farg{e_1}{\env_1}::\farg{e_2}{\env_2}::\fl) & \rightsquigarrow
+			                 & (e_1, \env_1, \ffun{k}::\farg{e_2}{\env_2}::\fl) \\
+			\textsc{Resume2} & (\clos{x}{e_1}{\env_1}, \_, \ffun{k}::\farg{e_2}{\env_2}::\fl) & \rightsquigarrow
+			                 & (e_2, \env_2, \ffun{k}::\ffun{\clos{x}{e_1}{\env_1}}::\fl) \\
+			\textsc{Perform} & (\perform{l}{e}, \env, \fl) & \rightsquigarrow
+			                 & (e, \env, \ffun{\effval{l}{[[],(\{\caseval{x}{x}\},\emptyset)]}}::\fl) \\
+			\textsc{Raise}   & (\throw{l}{e}, \env, \fl) & \rightsquigarrow
+											 & (e, \env, \ffun{\exnval{l}}::\fl)
+		\end{array}
+	\end{smathpar}
+	\caption{Local Reductions -- $(\term, \env, \fl) \rightsquigarrow (\tau, \env, \fl)$.}
+	\label{sem:step}
+\end{figure*}
+
+\begin{figure*}
+	\begin{smathpar}
+		\begin{array}{rrcll}
+			\textsc{LocalO}   & (\term,\env,(\fl,\hc) \kcons k,\cstack) & \ostep
+												& \configa{\term'}{\env'}{\ostacka{(\fl',\hc) \kcons k}{\cstack}}
+												& \text{if } (\term,\env,\fl) \rightsquigarrow (\term',\env',\fl') \\
+			\textsc{RetC}   & (v, \_, [([],(h,\env))], \cstack) & \ostep
+												& \configa{e}{\envext{\env}{x}{v}}{\cstack}
+												& \text{if } \{\caseval{x}{e}\} \in h \\
+			\textsc{RetFib} & (v, \_, ([],(h,\env)) \kcons k, \cstack) & \ostep
+												& \configa{e}{\envext{\env}{x}{v}}{\ostacka{k}{\cstack}}
+												& \text{if } \{\caseval{x}{e}\} \in h \\
+			\textsc{CallO}    & (v, \_, (\oclos{x}{e}{\env}::\fl,\hc) \kcons k, \cstack) & \ostep
+												& \configa{e}{\envext{\env}{x}{v}}{\ostacka{(\fl,\hc) \kcons k}{\cstack}} \\
+			\textsc{ExtCall}  & (v, \_, (\cclos{x}{e}{\env}::\fl,\hc) \kcons k, \cstack) & \ostep
+												& \configa{e}{\envext{\env}{x}{v}}{\cstacka{[]}{\ostacka{(\fl,\hc) \kcons k}{\cstack}}} \\
+			\textsc{Resume}   & (v, \_, (\ffun{k}::\ffun{\oclos{x}{e}{\env}}::\fl,\hc) \kcons k, \cstack) & \ostep
+												& \configa{e}{\envext{\env}{x}{v}}{\ostacka{k ~@~ (\fl,\hc) \kcons k'}{\cstack}} \\
+			\textsc{Handle}   & (\handle{e}{h}, \env, k, \cstack) & \ostep
+												& \configa{e}{\env}{\ostacka{([],(h,\env)) \kcons k}{\cstack}} \\
+			\textsc{EffUnHn}  & (v, \_, [\ffun{\effval{l}{k}}::\fl,(h,\env)], \cstack) & \ostep
+												& \configa{e}{\emptyset}{\ostacka{k ~@~ [(\fl,(h,\env))]}{\cstack}}
+												& \text{if } \{\caseeff{l}{\_}{\_}{\_}\} \notin h \\
+									& & & & \text{and } {e = \throw{\textsf{unhandled}}{0}} \\
+			\textsc{EffHn}    & (v, \_, (\ffun{\effval{l}{k}}::\fl,(h,\env)) \kcons k', \cstack) & \ostep
+												& \configa{e}{\env[r \mapsto k''][x \mapsto v]}{\ostacka{k'}{\cstack}}
+												& \text{if } \{\caseeff{l}{x}{r}{e}\} \in h \\
+									& & & & \text{and } k'' = k ~@~ [(\fl,(h,\env))] \\
+			\textsc{EffFwd}   & (v, \env', (\ffun{\effval{l}{k}}::\fl,(h,\env)) \kcons (\fl',\hc') \kcons k', \cstack) & \ostep
+												& \configa{v}{\env'}{\ostacka{(\ffun{\effval{l}{k''}}::\fl',\hc') \kcons k'}{\cstack}}
+												& \text{if } \{\caseeff{l}{\_}{\_}{\_}\} \notin h \\
+								  & & & & \text{and } k'' = k ~@~ [(\fl,(h,\env))] \\
+			\textsc{ExFwdC}   & (v, \env, [\ffun{\exnval{l}}::\_,(h,\_)], \cstacka{\fl'}{\ostack}) & \ostep
+												& \configa{v}{\env}{\cstacka{\ffun{\exnval{l}}::\fl'}{\ostack}}
+												& \text{if } \{\caseexn{l}{\_}{\_}\} \notin h \\
+			\textsc{ExHn}     & (v, \_, (\ffun{\exnval{l}}::\_,(h,\env)) \kcons k', \cstack) & \ostep
+												& \configa{e}{\envext{\env}{x}{v}}{\ostacka{k'}{\cstack}}
+												& \text{if } \{\caseexn{l}{x}{e}\} \in h \\
+			\textsc{ExFwdFib} & (v, \env, (\ffun{\exnval{l}}::\_,(h,\_)) \kcons (\fl',\hc') \kcons k', \cstack) & \ostep
+												& \configa{v}{\env}{\ostacka{(\ffun{\exnval{l}}::\fl',\hc') \kcons k'}{\cstack}}
+												& \text{if } \{\caseexn{l}{\_}{\_}\} \notin h \\
+		\end{array}
+	\end{smathpar}
+	\caption{OCaml Reductions -- $(\term, \env, k, \cstack) \ostep \config$.}
+	\label{sem:ostep}
+\end{figure*}
+
+\begin{figure*}
+	\begin{smathpar}
+		\begin{array}{rrcll}
+			\textsc{LocalC}   & (\term,\env,\fl,\ostack) & \cstep
+												& \configa{\term'}{\env'}{\cstacka{\fl'}{\ostack}}
+												& \text{if } (\term,\env,\fl) \rightsquigarrow (\term',\env',\fl') \\
+			\textsc{RetToO}   & (v,\env,[],\ostacka{k}{\cstack}) & \cstep
+			                  & \configa{v}{\env}{\ostacka{k}{\cstack}} \\
+			\textsc{CallC}    & (v, \_, \cclos{x}{e}{\env}::\fl, \ostack) & \cstep
+												& \configa{e}{\envext{\env}{x}{v}}{\cstacka{\fl}{\ostack}} \\
+			\textsc{Callback} & (v, \_, \oclos{x}{e}{\env}::\fl, \ostack) & \cstep
+												& \configa{e}{\envext{\env}{x}{v}}{\ostacka{k}{\cstacka{\fl}{\ostack}}}
+												& \text{if } k = [[],(\{\caseval{x}{x}\},\emptyset)] \\
+			\textsc{ExnFwdO}  & (v, \env, \ffun{\exnval{l}}::\_, \ostacka{(\fl,\hc) \kcons k}{\cstack}) & \cstep
+												& \configa{v}{\env}{\ostacka{(\ffun{\exnval{l}}::\fl,\hc) \kcons k}{\cstack}}
+		\end{array}
+	\end{smathpar}
+	\caption{C Reductions -- $(\term, \env, \fl, \ostack) \cstep \config$.}
+	\label{sem:cstep}
+\end{figure*}
+
+\begin{figure*}
+	\begin{mathpar}
+		\begin{array}{rc}
+
+			\textsc{StepC} & \inferrule
+			                    {(\term, \env, \fl, \ostack) \cstep \config}
+													{\configa{\term}{\env}{\cstacka{\fl}{\ostack}} \step \config} \\ \\
+			\textsc{StepO} & \inferrule
+													{(\term, \env, k, \cstack) \ostep \config}
+													{\configa{\term}{\env}{\ostacka{k}{\cstack}} \step \config}
+		\end{array}
+	\end{mathpar}
+	\caption{Top-level Reductions -- $\config \step \config$.}
+	\label{sem:top}
+\end{figure*}
+
+\section{Implementation}
+
+\section{Evaluation}
+
+\section{Related Work}
+
+\section{Conclusions and Future Work}
+
 \bibliographystyle{ACM-Reference-Format}

### 2020-10-27 retro-concurrency.tex
Added most of intro

@@
 \lstset{
-			language=[Objective]caml,
-      basicstyle=\ttfamily\small,
+			language=caml,
+      basicstyle=\ttfamily\footnotesize,
       flexiblecolumns=false,
 			tabsize=2,
-      escapechar=',
+			escapechar={<@}{@>},
       %basewidth={0.5em,0.45em},
@@
       stringstyle=\color{MidnightBlue},
-			keywords=[1]{effect,perform,continue,discontinue},
-			keywordstyle=[1]\color{Bittersweet}\bfseries,
 			keywords=[2]{},
 			keywordstyle=[2]\color{byzantine}\bfseries,
-			keywords=[3]{int,string}, %types%
-			keywordstyle=[3]\color{caribbeangreen},
+			%keywords=[3]{int,string}, %types%
+			%keywordstyle=[3]\color{caribbeangreen},
+			keywords=[4]{effect,perform,continue,discontinue},
+			keywordstyle=[4]\color{Bittersweet}\bfseries,
 			classoffset=1,
@@
 			numberstyle=\tiny\color{gray},
-			numbersep=2pt
+			numbersep=5pt
 }
@@
 \emph{operations} from their interpretations, which are defined by
-\emph{handlers} of the effects. For example, |effect Read : in_channel -> string|
-declares an effect |Read|, which is parametrised with an input channel of type
+\emph{handlers} of the effects. For example,
+%
+\begin{lstlisting}
+effect Read : in_channel -> string
+\end{lstlisting}
+%
+declares an effect |Read|, which is parameterised with an input channel of type
 |in_channel|, which when \emph{performed} returns a |string| value. A
@@
 
+\subsection{An effectful calculator}
+
 What makes effect handlers quite powerful is that they provide a structured
-form of delimited continuations, which enables advanced non-local control-flow
-mechanisms such as resumable exceptions, coroutines, generators and
-asynchronous I/O to be composably expressed. As an example, consider the task
-of handling division-by-zero case in a calculator application. Different
-calculators handle division-by-zero differently. In the iOS 13 Calculator App,
-division-by-zero returns |Error|. In the Google Search Calculator App |0/0|
-returns to |Error|, and for any other |n| not equal to |0|, |n/0| returns
-|Infinity| with the same sign as |n|. In the OCaml top-level, division-by-zero
-raises |Division_by_zero| exception. How can we capture these different
-behaviours without changing the evaluator?
+form of delimited continuations~\cite{}, which enables advanced non-local
+control-flow mechanisms such as resumable exceptions, coroutines, generators
+and asynchronous I/O to be composably expressed. As an example, consider the
+task of handling division-by-zero case in a calculator application. Different
+calculators handle division-by-zero differently. In the iOS 13 Calculator App
+(|iOS|), division-by-zero returns |Error|. In the Google search calculator App
+(|gsearch|) |0/0| returns to |Error|, and for any other |n| not equal to |0|,
+|n/0| returns |Infinity| with the same sign as |n|. In the OCaml top-level,
+division-by-zero raises |Division_by_zero| exception. How can we capture these
+different behaviours without changing the evaluator?
 
@@
 problem. Unlike exceptions where the library decides whether an error is fatal
-or not, conditions allow the client of the library to make that decision.
-Common Lisp Conditions provide a restart mechanism to resume execution from the
-point where the condition was signalled. On division-by-zero, the calculator
-signals the handler and lets the handler decide the course of action based on
-the desired semantics. We can implement this mechanism using effect handlers as
-shown in Figure~\ref{fig:}.
+or not, conditions allow the client of the library to make that decision. If
+the error is non-fatal, Conditions provide a restart mechanism to
+resume execution from the point where the condition was signalled. On
+division-by-zero, the calculator signals the handler and lets the handler
+decide the course of action based on the desired semantics. We can implement
+this mechanism using effect handlers as shown below:
+
+\begin{lstlisting}[numbers=left]
+type sign = P | N
+type val = Z of int | Error | Inf of sign
+type exp = Val of val | ... | Div of exp*exp
+effect DivBy0 : int -> val
+
+let rec eval e =
+	match e with
+  | Val v -> v
+	| ...
+  | Div (e1,e2) ->
+    match eval e1, eval e2 with
+    | Z v1, Z v2 ->
+        if v2 = 0 then perform (DivBy0 v1)
+        else Z (v1/v2)
+	  | Error, _ -> Error
+		| Inf s, _ -> Inf s
+  	| ...
+
+let (/) n d = Div (Val (Z n), Val (Z d))
+\end{lstlisting}
+
+On line 4, we declare an effect |DivBy0| which is parameterised by an integer,
+which when performed returns a value. While interpreting the division
+expression, if the denominator is 0, then we perform |DivBy0| effect with the
+numerator value (line 13) so that the context may decide whether the error is
+fatal or not, and resume with an appropriate value if the error is not fatal.
+We can get the |iOS| semantics by wrapping the evaluator computation with the
+following handler:
+
+\begin{lstlisting}
+let ios e = match eval e with
+| v -> v (* value case *)
+| effect (DivBy0 _) k -> (* effect case *)
+    continue k Error
+\end{lstlisting}
+
+If the evaluator returns a value, we simply return that. In case the evaluator
+performs the |DivBy0| effect, we catch the effect in the effect case along with
+the continuation |k| of the corresponding |perform| \emph{delimited} by this
+handler. In the case of |iOS|, we |continue| the suspended computation with
+|Error| value. Hence, the computation |ios((1/0)/2) | returns |Error|. On the
+other hand, we can get |gsearch| semantics by enclosing the evaluator in the
+following handler:
+
+\begin{lstlisting}
+let gsearch e = match eval e with
+| v -> v
+| effect (DivBy0 v) k ->
+  if v = 0 then continue k Error
+	else if v > 0 then continue k (Inf P)
+	else continue k (Inf N)
+\end{lstlisting}
+
+The computation |gsearch((1/0)/2)| returns positive infinity (|Inf P|). We can
+extend the calculator modularly with the ability to read user input using the
+|Read| effect defined earlier as follows:
+
+\begin{lstlisting}
+type exp = ... | Inp
+effect Read : in_channel -> string
+
+let rec eval e = match e with
+| ...
+| Inp -> int_of_string(perform(Read stdin))
+\end{lstlisting}
+
+The |Read| effect can then be handled by another handler:
+
+\begin{lstlisting}
+let run e = match gsearch e with
+| v -> v
+| effect (Read chan) k ->
+	register_async_input_line (chan, k);
+	let ui = get_ui_k () in
+	continue ui ()
+\end{lstlisting}
+
+\noindent which registers the read operation to be performed asynchronously
+with the continuation as the callback. Subsequently, it resumes the
+user-interface continuation so that the application remains responsive. Observe
+that neither of the handlers needs to be aware of the other.
+
+\subsection{Retrofitting effect handlers}
+
+There are many research languages and libraries built around effect
+handlers~\cite{Koka,Links,Pyro,Frank,Eff}. Unlike these efforts, our goal is to
+retrofit effect handlers onto the OCaml programming language, which has been in
+continuous use for the past 25 years in large codebases including verification
+tools~\cite{FStar,Coq}, mission critical software systems~\cite{astree} and
+latency sensitive networked applications~\cite{JS,Docker,MirageOS}. OCaml is
+particularly favoured for its competitive yet predictable performance, with
+fast foreign-function interface (FFI) and excellent compatibility with program
+analysis such as debuggers and profilers that utilise DWARF stack unwind tables
+to obtain a backtrace. Our primary motivation for adding effect handlers to
+OCaml is to support (1) highly concurrent applications such as web servers to
+be expressed naturally in \emph{direct-style} as opposed to using callbacks and
+(2) efficient data and task parallel programs in the vein of OpenMP~\cite{} and
+Intel TBB~\cite{}.
+
+OCaml, however, does not support any non-local control flow mechanisms other
+than exceptions. This makes it particularly challenging for implement delimited
+continuations necessary for effect handlers without sacrificing the desirable
+properties of OCaml. One standard way of implementing continuations is to use
+continuation-passing style (CPS) in the compiler's intermediate representation
+(IR)~\cite{Koka}. If the compiler were to use CPS, then reifying continuations
+in the source language is relatively straight-forward. But OCaml does not use
+CPS IR, and changing the OCaml compiler to utilise CPS IR would be an enormous
+undertaking. CPS IR would also affect the performance profile of OCaml
+applications due to increased memory allocations as continuation closures get
+allocated on the heap~\cite{Folklore}. Moreover, with CPS, stack unwinding in
+meaningless and hence, we would lose compatibility with tools that inspect the
+program stack. Hence, we choose not to CPS translation.
+
+The alternative is to use the call stack for implementing continuation.
+Farvardin et al~\cite{Folklore} show that there is a spectrum of implementation
+strategies for continuation with different trade-offs. We motivate our
+implementation strategy based on the following ideal requirements:
+
+\begin{enumerate}[label=R\arabic*]
+	\item \textbf{Backwards compatibility} A well behaved OCaml program does not
+		break under OCaml extended with effect handlers. OCaml code that does not
+		use effect handlers will pay negligible performance cost.
+	\item \textbf{Tool compatibility} OCaml with effect handlers produce
+		well-formed backtraces, and remains compatible with program analysis tools
+		such as debuggers and profilers that inspect the stack using DWARF unwind
+		tables.
+	\item \textbf{Effect handler efficiency} In order to support
+		highly-concurrent applications, the program must accommodate millions of
+		continuations in the heap. Installing effect handlers, capturing and
+		resuming continuations must be fast.
+	\item \textbf{Effect handler expressivity} While making continuations
+		resumable at-most once (\emph{affine}) is a standard way to make
+		continuation capture fast, we must still support use cases where the
+		continuations are resumed more than once.
+\end{enumerate}
+
+The need to host millions of active continuations in the heap rules out the use
+of large contiguous stack space for continuations as in C. Instead, we resort
+to using small initial stacks and growing the stacks on demand. As a result,
+OCaml functions, irrespective of whether they use effect handlers, need to
+perform stack overflow checks and external C functions, which do not have stack
+overflow checks, must be performed on a separate stack. Additionally, we must
+generate DWARF stack unwind tables for stacks that may be non-contiguous and
+contain frames from different languages. In this work, we develop a series of
+techniques to implement efficient effect handlers for OCaml with the help of
+compiler and runtime support that satisfy the requirements above.
 
@@
 is an active area of research~\cite{Leijen14, Biernacki19, Biernacki20,
-Hillerstrom20}. We do not focus on this question in this paper, and the
-implementation of effect handlers in Multicore OCaml does not guarantee effect
-safety.
+Hillerstrom20}. We do not focus on this question in this paper, and our
+implementation of effect handlers in OCaml does not guarantee effect safety.
+For example, running |perform (Read stdin)| without an enclosing handler raises
+exception |Unhandled|. We leave the question of effect safety for future work.
+
+\subsection{Contributions}
+
+Our contributions are to present:
+
+%\begin{itemize}
+%	\item the design and implementation of affine effect handlers for OCaml with
+%		negligible performance impact on code that does not use effect handlers.
+%	\item
 

### 2020-10-27 retro-concurrency.tex
Updates

@@
 (|gsearch|) |0/0| returns to |Error|, and for any other |n| not equal to |0|,
-|n/0| returns |Infinity| with the same sign as |n|. In the OCaml top-level,
-division-by-zero raises |Division_by_zero| exception. How can we capture these
+|n/0| returns |Infinity| with the same sign as |n|. How can we capture these
 different behaviours without changing the evaluator?
@@
 handler. In the case of |iOS|, we |continue| the suspended computation with
-|Error| value. Hence, the computation |ios((1/0)/2) | returns |Error|. On the
+|Error| value. Hence, the computation |ios(1/0)| returns |Error|. On the
 other hand, we can get |gsearch| semantics by enclosing the evaluator in the
@@
 
-The computation |gsearch((1/0)/2)| returns positive infinity (|Inf P|). We can
+The computation |gsearch(1/0)| returns positive infinity (|Inf P|). We can
 extend the calculator modularly with the ability to read user input using the
@@
 \subsection{Retrofitting effect handlers}
+\label{sec:req}
 
@@
 generate DWARF stack unwind tables for stacks that may be non-contiguous and
-contain frames from different languages. In this work, we develop a series of
-techniques to implement efficient effect handlers for OCaml with the help of
-compiler and runtime support that satisfy the requirements above.
+contain frames from different languages. In this work, we develop the compiler
+and runtime support for implementing efficient effect handlers for OCaml.
 
-The search for an expressive effect system which guarantees that all the
-effects performed in the program are handled (\emph{effect safety}) in the
-presence of advanced features such as polymorphism, modularity and generativity
-is an active area of research~\cite{Leijen14, Biernacki19, Biernacki20,
+Our work is also timely. WebAssembly~\ref{} community group is actively
+considering effect handlers as one of the mechanism for supporting concurrency,
+asynchronous I/O and generators~\ref{WasmPropoal}. Project Loom~\cite{} is an
+OpenJDK project which adds virtual threads and delimited continuations to Java.
+We believe that our design choices will inform similar choices to be made in
+the other languages.
+
+The search for an expressive effect system that guarantees that all the effects
+performed in the program are handled (\emph{effect safety}) in the presence of
+advanced features such as polymorphism, modularity and generativity is an
+active area of research~\cite{Leijen14, Biernacki19, Biernacki20,
 Hillerstrom20}. We do not focus on this question in this paper, and our
@@
 
-%\begin{itemize}
-%	\item the design and implementation of affine effect handlers for OCaml with
-%		negligible performance impact on code that does not use effect handlers.
-%	\item
+\begin{itemize}
+	\item the design and implementation of affine effect handlers for OCaml with
+		negligible performance impact on code that does not use effect handlers.
+		Our design lets us retain OCaml's compatibility with program analysis tools
+		that inspect the stack.
+	\item a formal operation semantics for the effect handler implementation in
+		OCaml.
+	\item extensive evaluation which shows that our implementation has negligible
+		impact on code that does not use effect handlers, and serves as an
+		efficient foundation for scalable concurrent and parallel programming.
+\end{itemize}
+
+We have implemented effect handlers in a multicore extension of the OCaml
+programming language. We call our implementation ``Multicoe OCaml'' to
+distinguish it from stock OCaml. Multicore OCaml uses effect handlers as the
+mechanism to express user-level concurrency.
+
+The rest of the paper is organized as follows. We describe the details of the
+OCaml program stack in the next section, and highlight the features that the
+effect handler implementation must retain in order to provide backwards
+compatibility. In \S\ref{sec:handlers}, we motivate the design choice that we
+have made for effect handlers in Multicore OCaml with the help of examples.
+\S\ref{sec:semantics} presents the formal operational semantics for Multicore
+OCaml effect handlers. \S\ref{sec:impl} discusses the compiler and the runtime
+system support for implementing effect handlers. \S\ref{sec:eval} presents
+extensive performance evaluation of effect handlers against the goals set in
+\S\ref{sec:req}. \S\ref{sec:related} and \S\ref{sec:conc} present the related
+work and the conclusions.
 
 \section{Background: OCaml Stacks}
+\label{sec:stack}
 
 \section{Handlers by example}
+\label{sec:handlers}
 
 \section{Semantics}
+\label{sec:semantics}
 
@@
 \section{Implementation}
+\label{sec:impl}
 
 \section{Evaluation}
+\label{sec:eval}
 
 \section{Related Work}
+\label{sec:related}
 
 \section{Conclusions and Future Work}
+\label{sec:conc}
 

### 2020-10-28 code/meander_.c
Added code for meander

@@ -1,0 +1,7 @@
+#include <caml/mlvalues.h>
+#include <caml/callback.h>
+
+value ocaml_to_c (value unit) {
+  caml_callback(*caml_named_value("c_to_ocaml"), Val_unit);
+  return Val_int(0);
+}

### 2020-10-28 retro-concurrency.tex
minor typo readability edits

@@
 latency sensitive networked applications~\cite{JS,Docker,MirageOS}. OCaml is
-particularly favoured for its competitive yet predictable performance, with
-fast foreign-function interface (FFI) and excellent compatibility with program
-analysis such as debuggers and profilers that utilise DWARF stack unwind tables
-to obtain a backtrace. Our primary motivation for adding effect handlers to
-OCaml is to support (1) highly concurrent applications such as web servers to
+particularly favoured for its competitive yet predictable performance, with fast
+foreign-function interface (FFI) and excellent compatibility with program
+analysis tools such as debuggers and profilers that utilise DWARF stack unwind
+tables to obtain a backtrace. Our primary motivation for adding effect handlers
+to OCaml is to support (1) highly concurrent applications such as web servers to
 be expressed naturally in \emph{direct-style} as opposed to using callbacks and
@@
 OCaml, however, does not support any non-local control flow mechanisms other
-than exceptions. This makes it particularly challenging for implement delimited
-continuations necessary for effect handlers without sacrificing the desirable
-properties of OCaml. One standard way of implementing continuations is to use
-continuation-passing style (CPS) in the compiler's intermediate representation
-(IR)~\cite{Koka}. If the compiler were to use CPS, then reifying continuations
-in the source language is relatively straight-forward. But OCaml does not use
-CPS IR, and changing the OCaml compiler to utilise CPS IR would be an enormous
-undertaking. CPS IR would also affect the performance profile of OCaml
-applications due to increased memory allocations as continuation closures get
-allocated on the heap~\cite{Folklore}. Moreover, with CPS, stack unwinding in
-meaningless and hence, we would lose compatibility with tools that inspect the
-program stack. Hence, we choose not to CPS translation.
+than exceptions. This makes it particularly challenging to implement the
+delimited continuations necessary for effect handlers without sacrificing the
+desirable properties of OCaml. One standard way of implementing continuations is
+to use continuation-passing style (CPS) in the compiler's intermediate
+representation (IR)~\cite{Koka}. If the compiler were to use CPS, then reifying
+continuations in the source language is relatively straight-forward. But OCaml
+does not use CPS IR, and changing the OCaml compiler to utilise CPS IR would be
+an enormous undertaking. CPS IR would also affect the performance profile of
+OCaml applications due to increased memory allocations as continuation closures
+get allocated on the heap~\cite{Folklore}. Moreover, with CPS, stack unwinding
+is meaningless and hence, we would lose compatibility with tools that inspect
+the program stack. Hence, we choose not to CPS translation.
 
@@
 We have implemented effect handlers in a multicore extension of the OCaml
-programming language. We call our implementation ``Multicoe OCaml'' to
+programming language. We call our implementation ``Multicore OCaml'' to
 distinguish it from stock OCaml. Multicore OCaml uses effect handlers as the

### 2020-10-28 retro-concurrency.tex
Section 2

@@
 			keywordstyle=[2]\color{byzantine}\bfseries,
-			%keywords=[3]{int,string}, %types%
-			%keywordstyle=[3]\color{caribbeangreen},
-			keywords=[4]{effect,perform,continue,discontinue},
+			keywords=[3]{effect,perform,continue,discontinue},
+			keywordstyle=[3]\color{caribbeangreen}\bfseries,
+			keywords=[4]{call,return,external},
 			keywordstyle=[4]\color{Bittersweet}\bfseries,
@@
 latency sensitive networked applications~\cite{JS,Docker,MirageOS}. OCaml is
-particularly favoured for its competitive yet predictable performance, with fast
-foreign-function interface (FFI) and excellent compatibility with program
+particularly favoured for its competitive yet predictable performance, with
+fast foreign-function interface (FFI) and excellent compatibility with program
 analysis tools such as debuggers and profilers that utilise DWARF stack unwind
 tables to obtain a backtrace. Our primary motivation for adding effect handlers
-to OCaml is to support (1) highly concurrent applications such as web servers to
-be expressed naturally in \emph{direct-style} as opposed to using callbacks and
-(2) efficient data and task parallel programs in the vein of OpenMP~\cite{} and
-Intel TBB~\cite{}.
+to OCaml is to support (1) highly concurrent applications such as web servers
+to be expressed naturally in \emph{direct-style} as opposed to using callbacks
+and (2) efficient data and task parallel programs in the vein of Cilk~\cite{},
+OpenMP~\cite{} and Intel TBB~\cite{}.
 
@@
 \label{sec:stack}
+
+\begin{figure*}
+\begin{minipage}{0.59\linewidth}
+	\begin{minipage}{\linewidth}
+		\begin{lstlisting}
+external ocaml_to_c : unit -> int = "ocaml_to_c"
+exception E1
+exception E2
+
+let c_to_ocaml () = raise E1
+let _ = Callback.register "c_to_ocaml" c_to_ocaml
+
+let omain () =
+  try (* h1 *)
+    try (* h2 *) ocaml_to_c () with E2 -> 0
+  with E1 -> 42;;
+
+let _ = assert (omain () = 42)
+		\end{lstlisting}
+    \subcaption{\texttt{meander.ml}}
+    \label{code:meander_ml}
+	\end{minipage}
+	\begin{minipage}{\linewidth}
+		\begin{lstlisting}[language=c]
+#include <caml/mlvalues.h>
+#include <caml/callback.h>
+
+value ocaml_to_c (value unit) {
+  caml_callback(*caml_named_value("c_to_ocaml"), Val_unit);
+  return Val_int(0);
+}
+		\end{lstlisting}
+    \subcaption{\texttt{meander.c}}
+    \label{code:meander_c}
+	\end{minipage}
+\end{minipage}
+  \begin{minipage}{0.39\linewidth}
+    \centering
+    \includegraphics[scale=0.5]{figures/stock_stack}
+    \subcaption{Stack layout before raising exception.}
+    \label{fig:stock_stack}
+  \end{minipage}
+\end{figure*}
+
+The main challenge in implementing effect handlers in Multicore OCaml is
+managing the program stack. To this end, in this section, we provide an
+overview of the program stack and related mechanisms for stock OCaml.
+
+In order to illustrate the layout of the stock OCaml stack, consider the
+program in Figures~\ref{code:meander_ml} and~\ref{code:meander_c}. The OCaml
+main function |omain| installs two exception handlers |h1| and |h2| to handle
+exceptions |E1| and |E2|. |omain| calls the external C function |ocaml_to_c|,
+which in turn calls back into OCaml function |c_to_ocaml|, which raises the
+exception |E1|. OCaml supports raising exceptions in C as well as throwing
+exceptions across external calls. Hence, the exception |E1| gets caught in the
+handler |h1|, and |omain| returns |42|. The layout of the stack in native code
+backend just before raising the exception in |c_to_ocaml| is illustrated in
+Figure~\ref{fig:stock_stack}. The stack grows downwards.
+
+OCaml uses the hardware stack as C, and hence the program stack has alternating
+sequences of C and OCaml frames. Unlike C, OCaml does not create pointers into
+OCaml frames. OCaml uses the hardware support for |call| and |return|
+instructions for function calls and returns. OCaml does not perform explicit
+stack overflow checks in code, and, just like C, relies on the guard page at
+the end of the stack region to detect stack overflow. Stack overflow is
+detected by a memory fault, and |Stack_overflow| exception is raised to unwind
+the stack.
+
+OCaml does not use the C calling convention. In particular, there are no
+callee-saved registers in OCaml. On the |amd64| backend, a register each is
+pinned for the \emph{allocation pointer} into the minor heap used for bump
+pointer allocation, and the |Caml_state|, a table of global variables used by
+the runtime. These registers are chosen to be C callee saved registers. This
+makes external calls extremely fast in OCaml. If the external function does not
+touch the OCaml heap, then the external functions can be called directly, and
+no bookkeeping is necessary. For external functions that touch the OCaml heap,
+the cached allocation pointer is saved to a field in the |Caml_state|, and
+loaded back on return. Similarly, callbacks into OCaml are also cheap.
+Callbacks into OCaml involves loading the arguments in the right registers and
+calling the OCaml function. OCaml callbacks are relatively common as
+finalisation functions are executed as OCaml callbacks.
+
+The lack of callee-saved registers also makes exception handling fast as there
+are no callee-saved registers to be restored when the exception handler runs.
+Hence, installing an exception handler simply pushes the program counter (pc)
+of the handler and the current exception pointer (a field in |Caml_state|). The
+exception pointer is also updated to be the current stack pointer (sp). This
+creates a linked-list of exception handler frames on the stack as shown in
+Figure~\ref{fig:stock_stack}. The saved exception pointer is restored when an
+exception is handled or the handled computation returns normally with a value.
+
+In order to forward exceptions across C frames, the C stub function
+|caml_start_program|, which is the entry point for both the main OCaml code and
+callbacks into OCaml, pushes an exception handler frame that forwards the
+exception the innermost OCaml exception handler (|raised_exn_c| in
+Figure~\ref{fig:stock_stack}), or prints a fatal error (|fatal_error|) if there
+are no enclosing handlers. Exceptions are so cheap in OCaml that it is very
+common to use them for \emph{local} control flow.
+
+\kc{gdb backtrace}
 

### 2020-10-28 retro-concurrency.tex
Updates

@@
     \centering
-    \includegraphics[scale=0.5]{figures/stock_stack}
+    \includegraphics[scale=0.46]{figures/stock_stack}
     \subcaption{Stack layout before raising exception.}
@@
   \end{minipage}
+	\caption{Program stack on stock OCaml.}
 \end{figure*}
@@
 
+\subsection{External calls and callbacks}
+
 OCaml does not use the C calling convention. In particular, there are no
@@
 
+\subsection{Exception handlers}
+
 The lack of callee-saved registers also makes exception handling fast as there
@@
 exception the innermost OCaml exception handler (|raised_exn_c| in
-Figure~\ref{fig:stock_stack}), or prints a fatal error (|fatal_error|) if there
+Figure~\ref{fig:stock_stack}), or prints a fatal error (|fatal_uncaught|) if there
 are no enclosing handlers. Exceptions are so cheap in OCaml that it is very

### 2020-10-29 retro-concurrency.tex
Section 3

@@
 \usepackage{balance}
-\usepackage{url}
+%\usepackage{url}
 \usepackage{enumitem}
@@
 			keywordstyle=[2]\color{byzantine}\bfseries,
-			keywords=[3]{effect,perform,continue,discontinue},
+			keywords=[3]{effect,perform,continue,discontinue,clone_continuation},
 			keywordstyle=[3]\color{caribbeangreen}\bfseries,
@@
 
-The computation |gsearch(1/0)| returns positive infinity (|Inf P|). We can
-extend the calculator modularly with the ability to read user input using the
-|Read| effect defined earlier as follows:
+The computation |gsearch(1/0)| returns positive infinity (|Inf P|).
+
+\subsection{Asynchronous I/O}
+\label{sec:aio}
+
+We can extend the calculator modularly with the ability to read user input
+using the |Read| effect defined earlier as follows:
 
@@
 delimited continuations necessary for effect handlers without sacrificing the
-desirable properties of OCaml. One standard way of implementing continuations is
-to use continuation-passing style (CPS) in the compiler's intermediate
+desirable properties of OCaml. One standard way of implementing continuations
+is to use continuation-passing style (CPS) in the compiler's intermediate
 representation (IR)~\cite{Koka}. If the compiler were to use CPS, then reifying
@@
 is meaningless and hence, we would lose compatibility with tools that inspect
-the program stack. Hence, we choose not to CPS translation.
+the program stack. Hence, we choose not to CPS translation. The alternative is
+to use the call stack for implementing continuation. Farvardin et
+al~\cite{Folklore} show that there is a spectrum of implementation strategies
+for continuation with different trade-offs.
 
-The alternative is to use the call stack for implementing continuation.
-Farvardin et al~\cite{Folklore} show that there is a spectrum of implementation
-strategies for continuation with different trade-offs. We motivate our
-implementation strategy based on the following ideal requirements:
+The search for an expressive effect system that guarantees that all the effects
+performed in the program are handled (\emph{effect safety}) in the presence of
+advanced features such as polymorphism, modularity and generativity is an
+active area of research~\cite{Leijen14, Biernacki19, Biernacki20,
+Hillerstrom20}. We do not focus on this question in this paper, and our
+implementation of effect handlers in OCaml does not guarantee effect safety.
+For example, running |perform (Read stdin)| without an enclosing handler raises
+exception |Unhandled| at the point of |perform|. We leave the question of
+effect safety for future work.
+
+\subsection{Requirements}
+\label{sec:requirements}
+
+We motivate our implementation strategy based on the following ideal
+requirements:
 
@@
 		continuations are resumed more than once.
+	\item \textbf{Forwards compatibility} While this work does not present an
+		effect system, the design must be compatible with a future effect system
+		for OCaml. As a cornerstone of modularity, we also want blocking I/O code
+		to \emph{transparently} be made asynchronous with the help of effect
+		handlers.
 \end{enumerate}
@@
 The need to host millions of active continuations in the heap rules out the use
-of large contiguous stack space for continuations as in C. Instead, we resort
+of large contiguous stack space as in C for continuations. Instead, we resort
 to using small initial stacks and growing the stacks on demand. As a result,
@@
 the other languages.
-
-The search for an expressive effect system that guarantees that all the effects
-performed in the program are handled (\emph{effect safety}) in the presence of
-advanced features such as polymorphism, modularity and generativity is an
-active area of research~\cite{Leijen14, Biernacki19, Biernacki20,
-Hillerstrom20}. We do not focus on this question in this paper, and our
-implementation of effect handlers in OCaml does not guarantee effect safety.
-For example, running |perform (Read stdin)| without an enclosing handler raises
-exception |Unhandled|. We leave the question of effect safety for future work.
 
@@
 effect handler implementation must retain in order to provide backwards
-compatibility. In \S\ref{sec:handlers}, we motivate the design choice that we
-have made for effect handlers in Multicore OCaml with the help of examples.
-\S\ref{sec:semantics} presents the formal operational semantics for Multicore
-OCaml effect handlers. \S\ref{sec:impl} discusses the compiler and the runtime
-system support for implementing effect handlers. \S\ref{sec:eval} presents
-extensive performance evaluation of effect handlers against the goals set in
-\S\ref{sec:req}. \S\ref{sec:related} and \S\ref{sec:conc} present the related
-work and the conclusions.
+compatibility. In \S\ref{sec:refine}, we refine the design further
+focussing on the issues that come up when integrating effect handlers into a
+mainstream language. \S\ref{sec:semantics} presents the formal operational
+semantics for Multicore OCaml effect handlers. \S\ref{sec:impl} discusses the
+compiler and the runtime system support for implementing effect handlers.
+\S\ref{sec:eval} presents extensive performance evaluation of effect handlers
+against the goals set in \S\ref{sec:req}. \S\ref{sec:related} and
+\S\ref{sec:conc} present the related work and the conclusions.
 
@@
 	\begin{minipage}{\linewidth}
-		\begin{lstlisting}
+		\begin{lstlisting}[numbers=left]
 external ocaml_to_c : unit -> int = "ocaml_to_c"
@@
 	\begin{minipage}{\linewidth}
-		\begin{lstlisting}[language=c]
+		\begin{lstlisting}[language=c,numbers=left]
 #include <caml/mlvalues.h>
@@
 OCaml does not use the C calling convention. In particular, there are no
-callee-saved registers in OCaml. On the |amd64| backend, a register each is
+callee-saved registers in OCaml. On the |x86-64| backend, a register each is
 pinned for the \emph{allocation pointer} into the minor heap used for bump
@@
 makes external calls extremely fast in OCaml. If the external function does not
-touch the OCaml heap, then the external functions can be called directly, and
-no bookkeeping is necessary. For external functions that touch the OCaml heap,
-the cached allocation pointer is saved to a field in the |Caml_state|, and
-loaded back on return. Similarly, callbacks into OCaml are also cheap.
-Callbacks into OCaml involves loading the arguments in the right registers and
-calling the OCaml function. OCaml callbacks are relatively common as
-finalisation functions are executed as OCaml callbacks.
+allocate in the OCaml heap, then the external functions can be called directly,
+and no bookkeeping is necessary. For external functions which allocate in the
+OCaml heap, the cached allocation pointer is saved to a field in the
+|Caml_state| before the external call, and loaded back on return. Similarly,
+callbacks into OCaml are also cheap. Callbacks into OCaml involves loading the
+arguments in the right registers and calling the OCaml function. OCaml
+callbacks are relatively common as the garbage collector (GC) finalisation
+functions are executed as OCaml callbacks.
 
@@
 Hence, installing an exception handler simply pushes the program counter (pc)
-of the handler and the current exception pointer (a field in |Caml_state|). The
-exception pointer is also updated to be the current stack pointer (sp). This
-creates a linked-list of exception handler frames on the stack as shown in
-Figure~\ref{fig:stock_stack}. The saved exception pointer is restored when an
-exception is handled or the handled computation returns normally with a value.
+of the handler and the current exception pointer (|exn_ptr| -- a field in
+|Caml_state|). The exception pointer is also updated to be the current stack
+pointer (|sp|). This creates a linked-list of exception handler frames on the
+stack as shown in Figure~\ref{fig:stock_stack}. Raising an exception simply
+sets |sp| to |exn_ptr|, loads the saved |exn_ptr|, and jumps to the |pc| of the
+handler.
 
@@
 exception the innermost OCaml exception handler (|raised_exn_c| in
-Figure~\ref{fig:stock_stack}), or prints a fatal error (|fatal_uncaught|) if there
-are no enclosing handlers. Exceptions are so cheap in OCaml that it is very
-common to use them for \emph{local} control flow.
+Figure~\ref{fig:stock_stack}), or prints a fatal error (|fatal_uncaught|) if
+there are no enclosing handlers. Exceptions are so cheap in OCaml that it is
+very common to use them for \emph{local} control flow.
 
-\kc{gdb backtrace}
+\subsection{DWARF unwind tables}
 
-\section{Handlers by example}
-\label{sec:handlers}
+OCaml also generates precise DWARF unwind information for OCaml, thanks to
+which debuggers such as |gdb| and |lldb|, and profilers such as |perf| work
+out-of-the-box. For example, for the program in Figures~\ref{code:meander_ml}
+and~\ref{code:meander_c}, one could set a break point in |gdb| at
+|caml_raise_exn| to get the following backtrace.
+
+\begin{lstlisting}[language=c,basicstyle=\ttfamily\footnotesize]
+% gdb ./meander.exe
+(gdb) break caml_raise_exn
+Breakpoint 1 at 0x3e5dc
+(gdb) r
+Starting program: meander.exe
+Breakpoint 1, 0x925dc in caml_raise_exn ()
+(gdb) bt
+#0  0x925dc in caml_raise_exn ()
+#1  0x6fd3e in camlMeander__c_to_ocaml_83 () at meander.ml:5
+#2  0x925a4 in caml_start_program ()
+#3  0x8a84a in caml_callback_exn (...) at callback.c:145
+#4  caml_callback (...) at callback.c:199
+#5  0x76e0a in ocaml_to_c (unit=1) at meander_.c:5
+#6  0x6fd77 in camlMeander__omain_88 () at meander.ml:10
+#7  0x6fe92 in camlMeander__entry () at meander.ml:14
+#8  0x6f719 in caml_program ()
+#9  0x925a4 in caml_start_program ()
+#10 0x92e4c in caml_startup_common (...) at startup_nat.c:162
+#11 0x92eab in caml_startup_exn (...) at startup_nat.c:167
+#12 caml_startup (...) at startup_nat.c:172
+#13 0x6f55c in main (...) at main.c:44
+\end{lstlisting}
+
+\noindent which corresponds to the stack layout in
+Figure~\ref{fig:stock_stack}.
+
+The same backtrace can also be obtained by using \emph{frame pointers} instead
+of DWARF unwind tables. OCaml allows compiling code with frame pointers, but
+they are not enabled by default. OCaml stack tends to be deep with small frames
+due to the pervasive use of recursive functions, not all of which is
+tail-recursive. Hence, the addition of frame pointers can significantly
+increase the size of the
+stack~\footnote{https://github.com/ocaml/ocaml/issues/5721\#issuecomment-472965549}.
+Moreover, not using frame pointers saves two instructions in the function
+prologue and epilogue, and makes an extra register (|rbp| on x86) available.
+
+In summary, the design of effect handlers in OCaml must accommodate these
+desirable properties of OCaml with respect to the program stack.
+
+\section{Refining the design}
+\label{sec:refine}
+
+In this section, we refine the effect handler design for the OCaml use case.
+The primary motivation for adding effect handlers to OCaml is to support
+non-local control flow mechanism to implement asynchronous I/O in direct style,
+generators and enable task parallelism. These use cases do not resume the
+continuations more than once. Hence, our continuations are one-shot, and
+resuming the continuation more than once raises |Invalid_argument _| exception.
+It is well-known that one-shot continuations can be implemented
+efficiently~\cite{Bruggeman96RepresentingControl}.
+
+While OCaml permits throwing exceptions across C frames, we do not allow
+effects to propagate across C frames as the C frames would become part of the
+captured continuation. Managing C frames as part of the continuation is a
+complex endeavour~\cite{Daan17Aplas}, and we find that the complexity budget
+outweighs the relatively fewer mechanisms enabled by this addition in our
+setting.
+
+\subsection{Discontinuing a continuation}
+
+The interaction of non-local control flow with systems programming is quite
+subtle~\cite{}. Consider the following function that uses blocking I/O
+functions from the OCaml standard library to copy data from the input channel
+|ic| to the output channel |oc|:
+
+\begin{lstlisting}
+let rec copy ic oc =
+  try
+    let l = input_line ic in
+    output_string oc (l ^ "\n");
+    copy ic oc
+  with _ -> (close_in ic; close_out oc)
+\end{lstlisting}
+
+|input_line| raises |End_of_file| exception on reaching the end of input.
+Although not apparent, the code is written in a defensive style to handle other
+exceptional cases such as the channels being closed externally. Both
+|input_line| and |output_string| raise |Sys_error| exception if the channel is
+closed. This function cleanly closes the channels on exceptional cases;
+|close_*| functions do nothing if the channel is already closed.
+
+One of our goals(\S\ref{sec:requirements}) is to make this code transparently
+asynchronous. We can define effects for performing the I/O operations:
+\begin{lstlisting}
+effect In_line : in_channel -> string
+effect Out_str : out_channel -> string -> unit
+\end{lstlisting}
+\noindent and wrap them up in functions with the same signature as the one from
+the standard library.
+\begin{lstlisting}
+let input_line ic = perform (In_line ic)
+let output_string oc = perform (Out_str oc)
+\end{lstlisting}
+
+We can then use a top-level I/O handler similar to the introductory example
+(\S\ref{sec:aio}) to discharge the I/O operations asynchronously. This handles
+value return cases, but what about the exceptional cases |End_of_file| and
+|Sys_error|? We introduce a |discontinue| primitive to resume a continuation by
+raising an exception. In this example, on reaching the end of file, we would
+discontinue the continuation of the |copy| function with |discontinue k End_of_file|,
+which raises the exception at |input_line|, and the open channels
+will be closed.
+
+Program that use resources such as channels are usually written defensively
+with the assumption that calling a function will return \emph{exactly once},
+either normally or exceptionally. Since effect handlers in Multicore OCaml do
+not ensure that all the effects are handled, if the function performs an effect
+with no matching handler, then the function \emph{will not return at all}. To
+remedy this, when such an effect bubbles up to the top-level, we discontinue
+the continuation with |Unhandled| exception so that the exception handlers may
+run and clean up the resources.
+
+\subsection{Multi-shot continuations}
+
+Though our primary goal is to enable non-local control flow mechanisms, effect
+handlers have been gaining increasing attention for Bayesian machine learning
+and probabilisitic programming~\cite{PyroJMLR,MooreEdward2Google}, which resume
+the continuations more than once for performing inference over the
+probabilistic program. To enable these use cases, we provide a
+|clone_continuation| primitive that makes a copy of the given continuation.
+Care must be taken when using multi-shot continuations as they do not mix well
+with code that uses resources. With multi-shot continuations, a function
+\emph{may return more than once}. Consider the code below:
+
+\begin{lstlisting}
+let foo m = ...;perform E; ...
+let bar () =
+  let m = malloc 64 in
+	begin try foo m with _ -> () end;
+	free m
+let _ = match bar () with
+| () -> ()
+| effect E k ->
+    let k' = clone_continuation k in
+		continue k (); continue k' ()
+\end{lstlisting}
+
+The function |bar| expects |foo| to return exactly once, ignoring any exceptions. However,
+due to continuation cloning, the function returns twice, leading to double-free
+error.
 

### 2020-10-29 retro-concurrency.tex
Updates

@@
 
-The |Read| effect can then be handled by another handler:
+The |Read| effect can then be handled by another handler enclosing the
+division-by-zero handler:
 
@@
 		resuming continuations must be fast.
+\if{0}
 	\item \textbf{Effect handler expressivity} While making continuations
@@
 		continuations are resumed more than once.
+\fi
 	\item \textbf{Forwards compatibility} While this work does not present an
@@
 \begin{itemize}
-	\item the design and implementation of affine effect handlers for OCaml with
-		negligible performance impact on code that does not use effect handlers.
-		Our design lets us retain OCaml's compatibility with program analysis tools
-		that inspect the stack.
+	\item the design and implementation of affine effect handlers for OCaml. Our
+		design retains OCaml's compatibility with program analysis tools that
+		inspect the stack using DWARF unwind tables. We have validated our DWARF
+		unwind tables using an automated validator tool~\cite{}.
 	\item a formal operation semantics for the effect handler implementation in
@@
 
-\subsection{Discontinuing a continuation}
-
 The interaction of non-local control flow with systems programming is quite
-subtle~\cite{}. Consider the following function that uses blocking I/O
+subtle~\cite{TFP17}. Consider the following function that uses blocking I/O
 functions from the OCaml standard library to copy data from the input channel
@@
 
+\if{0}
 \subsection{Multi-shot continuations}
@@
 error.
+\fi
 

### 2020-10-30 retro-concurrency.tex
Minor fix
Added static semantics

@@
 			keywordstyle=[2]\color{byzantine}\bfseries,
-			keywords=[3]{effect,perform,continue,discontinue,clone_continuation},
+			keywords=[3]{eff,effect,perform,continue,discontinue,continuation,
+			             clone_continuation},
 			keywordstyle=[3]\color{caribbeangreen}\bfseries,
@@
 
-\lstMakeShortInline[columns=fullflexible]|
+\lstMakeShortInline[columns=fullflexible,keepspaces]|
 
@@
 		resuming continuations must be fast.
-\if{0}
-	\item \textbf{Effect handler expressivity} While making continuations
-		resumable at-most once (\emph{affine}) is a standard way to make
-		continuation capture fast, we must still support use cases where the
-		continuations are resumed more than once.
-\fi
 	\item \textbf{Forwards compatibility} While this work does not present an
@@
 they are not enabled by default. OCaml stack tends to be deep with small frames
-due to the pervasive use of recursive functions, not all of which is
+due to the pervasive use of recursive functions, not all of which are
 tail-recursive. Hence, the addition of frame pointers can significantly
@@
 
-\if{0}
-\subsection{Multi-shot continuations}
+\section{Semantics}
+\label{sec:semantics}
 
-Though our primary goal is to enable non-local control flow mechanisms, effect
-handlers have been gaining increasing attention for Bayesian machine learning
-and probabilisitic programming~\cite{PyroJMLR,MooreEdward2Google}, which resume
-the continuations more than once for performing inference over the
-probabilistic program. To enable these use cases, we provide a
-|clone_continuation| primitive that makes a copy of the given continuation.
-Care must be taken when using multi-shot continuations as they do not mix well
-with code that uses resources. With multi-shot continuations, a function
-\emph{may return more than once}. Consider the code below:
+In this section, we formalise the effect handler design in Multicore OCaml.
 
+\subsection{Static semantics}
+
+As mentioned earlier, effect handlers in Multicore OCaml do not guarantee
+effect safety, but only guarantee type safety. Programs without matching effect
+handlers are well-typed Multicore OCaml programs. As a result, our static
+semantics is simpler than languages that ensure effect
+safety~\cite{Links,Koka,Effekt,Frank,HeliumBiernacki}. This is important for
+backwards compatibility as our goal is to retrofit effect handlers to a
+language with large legacy codebases. Programs that do not use effects remain
+well-typed, and those that do compose well with those that don't.
+
+The static semantics of effect handlers in OCaml is captured succinctly by its
+API:
 \begin{lstlisting}
-let foo m = ...;perform E; ...
-let bar () =
-  let m = malloc 64 in
-	begin try foo m with _ -> () end;
-	free m
-let _ = match bar () with
-| () -> ()
-| effect E k ->
-    let k' = clone_continuation k in
-		continue k (); continue k' ()
+type 'a eff
+type ('a,'b) continuation
+val perform: 'a eff -> 'a
+val continue: ('a,'b) continuation -> 'a -> 'b
+val discontinue: ('a,'b) continuation -> exn -> 'b
+
+type 'a computation = unit -> 'a
+type ('a,'b) handler =
+{ret: 'a -> 'b;
+ eff_1: 'c1.'c1 eff -> ('c1,'b)continuation -> 'b;
+ ...
+ eff_n: 'cn.'cn eff -> ('cn,'b)continuation -> 'b}
+val match_with: 'a computation -> ('a,'b) handler
+                               -> 'b
 \end{lstlisting}
 
-The function |bar| expects |foo| to return exactly once, ignoring any exceptions. However,
-due to continuation cloning, the function returns twice, leading to double-free
-error.
-\fi
+We introduce the type |'a eff| of effect values which when performed using the
+|perform| primitive returns a |'a| value. Constructors for the value of type
+|'a eff| are declared using the effect declarations. For example, with the
+declaration |effect E : string -> int|, the effect value |E "42"| has the type
+|int eff|. We introduce the type |('a,'b) continuation| of delimited
+continuations which expect a |'a| value for resumption and return |'b| value.
+The continuations may be |continue|d with a suitably typed value or
+|discontinue|d with an exception.
 
-\section{Semantics}
-\label{sec:semantics}
+For handling the effects, our implementation extends OCaml's |match ... with|
+syntax with effect patterns. For the sake of exposition, we introduce a
+|('a,'b) handler| type. This handler handles a |'a computation| that returns a
+|'a| value, and itself returns a |'b| value. The handler has a single return
+case of type |'a -> 'b|, and one or more effect cases. Each effect case,
+handles a effect of type |'c eff|, whose continuation is of type
+|('c,'b) continuation|, and returns a value of type |'b|. For example, in the
+following code:
+\begin{lstlisting}
+effect E : string -> int
+effect F : float -> int
+let r = match perform (E "42") with
+| v -> v == 0
+| effect (E s) k ->
+    not (continue k (int_of_string s))
+| effect (F f) k ->
+    not (continue k (int_of_float f))
+\end{lstlisting}
+\noindent |r| has type |bool|.
+
+\subsection{Dynamic semantics}
 

### 2020-10-30 retro-concurrency.tex
Added semantics section

@@
 \newcommand{\kw}[1]{\text{\bf #1}}
-\newcommand{\effval}[2]{\textsf{do} ~#1 ~#2}
+\newcommand{\effval}[2]{\textsf{eff} ~#1 ~#2}
 \newcommand{\exnval}[1]{\textsf{exn} ~#1}
@@
 
-The initial configuration for an expression $e$ is
-$\configa{e}{\emptyset}{\cstacka{[]}{\bullet}}$. We define {\sf continue} and
-{\sf discontinue} as follows:
-
-\begin{smathpar}
-	\begin{array}{rcl}
-		\textsf{continue} ~k ~e & = & k ~(\olam{x}{x}) ~e \\
-		\textsf{discontinue} ~k ~l ~e & = & k ~(\olam{x}{\throw{l}{x}}) ~e
-	\end{array}
-\end{smathpar}
-
 \begin{figure*}
@@
 		\text{Fibers} & \fiber & \coloneqq & (\fl, \hc) \\
-		\text{Continuations} & k & \coloneqq & [] \mid \fiber \lhd k \\
+		\text{Continuations} & k & \coloneqq & [] \mid \fiber \kcons k \\
 		\text{C stacks} & \cstack & \coloneqq & \cstacka{\fl}{\ostack}\\
@@
 	\end{minipage}
-	\caption{Term Syntax}
+	\caption{Syntax of expressions and configurations}
 	\label{sem:syntax}
+\end{figure*}
+
+We present the operational semantics for a core language of effect handlers in
+Multicore OCaml. The operational semantics intends to faithfully capture the
+semantics of Multicore OCaml effect handlers, and to inform an implementation.
+We will highlight the differences whenever we differ from the semantics of the
+implementation. We have included an executable version of the dynamic semantics
+as an OCaml program in the supplementary material \kc{include hl-semantics.ml}.
+
+\subsubsection{Syntax}
+
+Figure~\ref{sem:syntax} presents the syntax of expressions and configurations.
+Our expressions consists of integer constants ($n$), variables ($x$),
+abstraction ($\lam{x}{x}$), application ($e~e$), arithmetic expressions ($e
+\odot e$) where $\odot$ ranges over $\{+,-,*,/\}$, raising exceptions
+($\throw{l}{e}$), performing effects ($\perform{l}{e}$), and handling effects
+($\handle{e}{h}$). Abstractions come in two forms: OCaml abstractions
+($\lambda^o$) and C abstractions ($\lambda^c$). The handler consists of a
+return case ($\caseval{x}{e}$), zero or more exception cases
+($\caseexn{l}{x}{e}$) with label $l$, parameter $x$ and body $e$, and zero or
+more effect cases ($\caseeff{l}{x}{k}{e}$) with label $l$, parameter $x$,
+continuation $k$ and body $e$.
+
+The operational semantics is an extension of a CEK machine
+semantics~\cite{Felleisen} for effect handlers, following the abstract machine
+semantics of Hillerstrom et al.~\cite{Hillerstrom2020}. The crucial difference
+from Hillerstrom et al. is that our stacks are composed of alternating sequence
+of OCaml and C stack segments. The program state is captured as configuration
+$\config \coloneqq \configa{\tau}{\env}{\stack}$ with the current term $\tau$
+under evaluation, its environment $\env$ and the current stack $\stack$. The
+term is either an expression $e$ or a value $v$. The value are integer
+constants $n$, continuations $k$, closure $\clos{x}{e}{\env}$, effect begin
+performed $\effval{l}{k}$ and exception being raised $\exnval{l}$. The
+environment is a map from variables to values.
+
+The stack $\stack$ is either a C stack ($\cstack$) or an OCaml stack
+($\ostack$). Every C stack $\cstacka{\fl}{\ostack}$ consists of a list of
+frames $\fl$, and, importantly, the OCaml stack $\ostack$ under it. Every OCaml
+stack is either an empty stack $\ostackemp$ or non-empty $\ostacka{k}{\cstack}$
+with the C stack $\cstack$ under it. Thus the stack is an alternating sequence
+of C and OCaml stacks terminating with an empty OCaml stack $\ostackemp$. The
+frame list $\fl$ is composed of individual frames $r$, which is one of an
+argument frame $\farg{e}{\env}$ with the expression $e$ at the argument
+position of an application with its environment $\env$, a function frame
+$\ffun{v}$ with the value at the function position of an application, and
+frames for evaluating the arguments of an arithmetic expression.
+
+Continuations $k$ are either empty $[]$ or non-empty sequence of \emph{fibers}.
+A fiber $\fiber \coloneqq (\fl,\hc)$ is a list of frames $\fl$ and a handler
+closure $\hc \coloneqq (h,\env)$, which is a pair of handler $h$ and its
+environment $\env$.
+
+\subsubsection{Top-level reductions}
+
+\begin{figure}
+	\begin{mathpar}
+		\begin{array}{rc}
+
+			\textsc{StepC} & \inferrule
+			                    {(\term, \env, \fl, \ostack) \cstep \config}
+													{\configa{\term}{\env}{\cstacka{\fl}{\ostack}} \step \config} \\ \\
+			\textsc{StepO} & \inferrule
+													{(\term, \env, k, \cstack) \ostep \config}
+													{\configa{\term}{\env}{\ostacka{k}{\cstack}} \step \config}
+		\end{array}
+	\end{mathpar}
+	\caption{Top-level Reductions -- $\config \step \config$.}
+	\label{sem:top}
+\end{figure}
+
+The initial configuration for an expression $e$ is
+$\configa{e}{\emptyset}{\cstacka{[]}{\bullet}}$, where the environment and the
+stack are empty. The top-level reductions (Figure~\ref{sem:top}) can be
+performed by either by taking a C step $\cstep$ or an OCaml step $\ostep$.
+Intuitively, this splits the world into executing C code and OCaml code.
+
+\begin{figure*}
+	\begin{smathpar}
+		\begin{array}{rrcll}
+			\textsc{AdminC}   & (\term,\env,\fl,\ostack) & \cstep
+												& \configa{\term'}{\env'}{\cstacka{\fl'}{\ostack}}
+												& \text{if } (\term,\env,\fl) \rightsquigarrow (\term',\env',\fl') \\
+			\textsc{RetToO}   & (v,\env,[],\ostacka{k}{\cstack}) & \cstep
+			                  & \configa{v}{\env}{\ostacka{k}{\cstack}} \\
+			\textsc{CallC}    & (v, \_, \cclos{x}{e}{\env}::\fl, \ostack) & \cstep
+												& \configa{e}{\envext{\env}{x}{v}}{\cstacka{\fl}{\ostack}} \\
+			\textsc{Callback} & (v, \_, \oclos{x}{e}{\env}::\fl, \ostack) & \cstep
+												& \configa{e}{\envext{\env}{x}{v}}{\ostacka{k}{\cstacka{\fl}{\ostack}}}
+												& \text{if } k = [[],(\{\caseval{x}{x}\},\emptyset)] \\
+			\textsc{ExnFwdO}  & (v, \env, \ffun{\exnval{l}}::\_, \ostacka{(\fl,\hc) \kcons k}{\cstack}) & \cstep
+												& \configa{v}{\env}{\ostacka{(\ffun{\exnval{l}}::\fl,\hc) \kcons k}{\cstack}}
+		\end{array}
+	\end{smathpar}
+	\caption{C Reductions -- $(\term, \env, \fl, \ostack) \cstep \config$.}
+	\label{sem:cstep}
 \end{figure*}
@@
 	\end{smathpar}
-	\caption{Local Reductions -- $(\term, \env, \fl) \rightsquigarrow (\tau, \env, \fl)$.}
+	\caption{Administrative Reductions -- $(\term, \env, \fl) \rightsquigarrow (\tau, \env, \fl)$.}
 	\label{sem:step}
 \end{figure*}
+
+\subsubsection{C reductions}
+
+The C reduction steps are given in Figure~\ref{sem:cstep}. We can take a C step
+by taking an administrative reduction step $\rightsquigarrow$. The
+administrative reductions are common to both C and OCaml. The rules
+\textsc{Var}, \textsc{Arith1}, \textsc{Arith2}, \textsc{App1}, \textsc{App2}
+and \textsc{App3} are standard. \textsc{Arith3} performs the arithmetic
+operation on the integers ($\llbracket n_1 \odot n_2 \rrbracket$).
+\textsc{Raise} pushes an function frame with exception value to indicate that
+an exception is being raised. Similarly, \textsc{Perform} pushes a function
+frame with an effect value with an empty continuation
+$[[],(\{\caseval{x}{x}\},\emptyset)]$ with no captured frames and an empty
+handler with an identity return case alone. \textsc{Resume1} and
+\textsc{Resume2} are used for resuming continuations, and we shall defer their
+explanation till later.
+
+Continuing with the rest of the C reduction steps, \textsc{CallC} captures the
+behaviour of calling a C function. Since the program is currently executing C,
+indicated by the current stack being a C stack $\cstacka{\fl}{\ostack}$, we can
+perform the call on the current stack. In case the abstraction is an OCaml
+abstraction (\textsc{Callback}), we create an OCaml stack with the C stack as
+its tail, with the current continuation being empty. This captures the
+behaviour of calling back into OCaml from C. \textsc{RetToO} returns a value to
+the enclosing OCaml stack. \textsc{ExnFwdO} forwards a raised exception to the
+enclosing OCaml stack, ignoring the unwinding the rest of the frames. This
+captures the semantics of raising OCaml exceptions from C.
+
+\subsubsection{OCaml Reductions}
 
@@
 		\begin{array}{rrcll}
-			\textsc{LocalO}   & (\term,\env,(\fl,\hc) \kcons k,\cstack) & \ostep
+			\textsc{AdminO}   & (\term,\env,(\fl,\hc) \kcons k,\cstack) & \ostep
 												& \configa{\term'}{\env'}{\ostacka{(\fl',\hc) \kcons k}{\cstack}}
 												& \text{if } (\term,\env,\fl) \rightsquigarrow (\term',\env',\fl') \\
-			\textsc{RetC}   & (v, \_, [([],(h,\env))], \cstack) & \ostep
-												& \configa{e}{\envext{\env}{x}{v}}{\cstack}
-												& \text{if } \{\caseval{x}{e}\} \in h \\
+			\textsc{RetToC}   & (v, \_, [([],(h,\emptyset))], \cstack) & \ostep
+												& \configa{v}{\env}{\cstack}
+												& \text{if } h = \{\caseval{x}{x}\} \\
 			\textsc{RetFib} & (v, \_, ([],(h,\env)) \kcons k, \cstack) & \ostep
@@
 			\textsc{Resume}   & (v, \_, (\ffun{k}::\ffun{\oclos{x}{e}{\env}}::\fl,\hc) \kcons k, \cstack) & \ostep
-												& \configa{e}{\envext{\env}{x}{v}}{\ostacka{k ~@~ (\fl,\hc) \kcons k'}{\cstack}} \\
+												& \configa{e}{\envext{\env}{x}{v}}{\ostacka{k ~@~ ((\fl,\hc) \kcons k')}{\cstack}} \\
 			\textsc{Handle}   & (\handle{e}{h}, \env, k, \cstack) & \ostep
@@
 
-\begin{figure*}
+OCaml reductions are given in Figure~\ref{sem:ostep}. In OCaml, reductions
+always occur on the top-most fiber in the current stack. \textsc{AdminO}
+performs administrative reductions. \textsc{CallO} evaluates an OCaml function
+on the current stack. \textsc{ExtCall} captures the behaviour of external
+calls. On an external call, we create a C stack with no frames, whose tail is
+the current OCaml stack, and evaluate the C functions body in the new stack.
+\textsc{RetToC} returns a value to the enclosing C stack. We have exactly one
+fiber on the stack, and this was created in the rule \textsc{Callback}, whose
+handler has identity return case alone and the environment is empty.
+\textsc{RetFib} returns value from a fiber to the previous one, evaluating the
+body of the return case.
+
+The rule \textsc{Handle} installs a handler by pushing a fiber with no frames
+and the given handler. The rule \textsc{ExHn} handles an exception, if the
+current handler has a matching exception case. Observe that fiber on which the
+exception was raise has been unwound. The rule \textsc{ExFwdC} forwards the
+exception to C. Here, there is exactly one fiber on the current stack, and the
+handler does not have a matching exception case. The rule \textsc{ExFwdFib}
+forwards the exception to the next fiber if the current handler does not handle
+it.
+
+The rule \textsc{EffHn} captures the handling of effects when the current
+handler has a matching effect case. We evaluate the body of the matching case,
+and bind the continuation parameter |r| to the captured continuation. Observe
+that the captured continuation |k''| includes the current handler. Intuitively,
+the handler wraps around captured continuation. This gives Multicore OCaml
+effect handlers deep handler semantics~\cite{}. \textsc{EffFwd} forwards the
+effect to the outer fiber, and extends the captured continuation |k''| in the
+process. Recall that we do not capture C frames as part of a continuation.
+\textsc{EffUnHn} models unhandled effect. If the effect bubbles up to the top
+fiber, which we know does not have an effect case (see \textsc{Callback} rule),
+we raise \textsc{unhandled} exception at the point where the corresponding
+effect was performed. This is achieved by appending the captured continuation
+to the front of the current continuation.
+
+Observe that |continue| and |discontinue| are not part of the expressions. They
+are encoded as follows:
 \begin{smathpar}
-		\begin{array}{rrcll}
-			\textsc{LocalC}   & (\term,\env,\fl,\ostack) & \cstep
-												& \configa{\term'}{\env'}{\cstacka{\fl'}{\ostack}}
-												& \text{if } (\term,\env,\fl) \rightsquigarrow (\term',\env',\fl') \\
-			\textsc{RetToO}   & (v,\env,[],\ostacka{k}{\cstack}) & \cstep
-			                  & \configa{v}{\env}{\ostacka{k}{\cstack}} \\
-			\textsc{CallC}    & (v, \_, \cclos{x}{e}{\env}::\fl, \ostack) & \cstep
-												& \configa{e}{\envext{\env}{x}{v}}{\cstacka{\fl}{\ostack}} \\
-			\textsc{Callback} & (v, \_, \oclos{x}{e}{\env}::\fl, \ostack) & \cstep
-												& \configa{e}{\envext{\env}{x}{v}}{\ostacka{k}{\cstacka{\fl}{\ostack}}}
-												& \text{if } k = [[],(\{\caseval{x}{x}\},\emptyset)] \\
-			\textsc{ExnFwdO}  & (v, \env, \ffun{\exnval{l}}::\_, \ostacka{(\fl,\hc) \kcons k}{\cstack}) & \cstep
-												& \configa{v}{\env}{\ostacka{(\ffun{\exnval{l}}::\fl,\hc) \kcons k}{\cstack}}
+	\begin{array}{rcl}
+		\textsf{continue} ~k ~e & = & k ~(\olam{x}{x}) ~e \\
+		\textsf{discontinue} ~k ~l ~e & = & k ~(\olam{x}{\throw{l}{x}}) ~e
 	\end{array}
 \end{smathpar}
-	\caption{C Reductions -- $(\term, \env, \fl, \ostack) \cstep \config$.}
-	\label{sem:cstep}
-\end{figure*}
 
-\begin{figure*}
-	\begin{mathpar}
-		\begin{array}{rc}
-
-			\textsc{StepC} & \inferrule
-			                    {(\term, \env, \fl, \ostack) \cstep \config}
-													{\configa{\term}{\env}{\cstacka{\fl}{\ostack}} \step \config} \\ \\
-			\textsc{StepO} & \inferrule
-													{(\term, \env, k, \cstack) \ostep \config}
-													{\configa{\term}{\env}{\ostacka{k}{\cstack}} \step \config}
-		\end{array}
-	\end{mathpar}
-	\caption{Top-level Reductions -- $\config \step \config$.}
-	\label{sem:top}
-\end{figure*}
+Intuitively, resuming a continuation in both the cases involves evaluating the
+appropriate abstraction on top of the continuation. We perform the
+administrative reductions \textsc{Resume1} and \textsc{Resume2} to evaluate the
+arguments to |continue| and |discontinue|. The rule \textsc{Resume} appends the
+given continuation to the front of the current continuation, and evaluates the
+body of the closure.
 

### 2020-10-31 retro-concurrency.tex
Updates
tabs to spaces
update static semantics

@@
 OpenJDK project which adds virtual threads and delimited continuations to Java.
-We believe that our design choices will inform similar choices to be made in
-the other languages.
+Swift concurrency roadmap~\cite{} includes direct-style asynchronous
+programming and structured concurrency. We believe that our design choices will
+inform similar choices to be made in the other languages.
 
@@
 \begin{lstlisting}
-type 'a eff
+type 'a eff = ..
 type ('a,'b) continuation
@@
 
-type 'a computation = unit -> 'a
+(* Internal API *)
+type 'a comp = unit -> 'a
 type ('a,'b) handler =
-{ret: 'a -> 'b;
- eff_1: 'c1.'c1 eff -> ('c1,'b)continuation -> 'b;
- ...
- eff_n: 'cn.'cn eff -> ('cn,'b)continuation -> 'b}
-val match_with: 'a computation -> ('a,'b) handler
-                               -> 'b
-\end{lstlisting}
-
-We introduce the type |'a eff| of effect values which when performed using the
-|perform| primitive returns a |'a| value. Constructors for the value of type
-|'a eff| are declared using the effect declarations. For example, with the
-declaration |effect E : string -> int|, the effect value |E "42"| has the type
-|int eff|. We introduce the type |('a,'b) continuation| of delimited
-continuations which expect a |'a| value for resumption and return |'b| value.
-The continuations may be |continue|d with a suitably typed value or
-|discontinue|d with an exception.
+{retc: 'a -> 'b;
+ effc: 'c.'c eff -> ('c,'b)continuation -> 'b; }
+val match_with: 'a comp -> ('a,'b) handler -> 'b
+\end{lstlisting}
+
+We introduce an extensible variant type~\cite{} |'a eff| of effect values,
+which when performed using the |perform| primitive returns a |'a| value.
+Constructors for the value of type |'a eff| are declared using the effect
+declarations. For example, the declaration |effect E : string -> int| is
+syntactic sugar for adding a new constructor to the variant type
+|type _ eff += E : string -> int eff|. We introduce the type
+|('a,'b) continuation| of delimited continuations which expect a |'a| value for
+resumption and return |'b| value. The continuations may be |continue|d with a
+suitably typed value or |discontinue|d with an exception.
 
 For handling the effects, our implementation extends OCaml's |match ... with|
-syntax with effect patterns. For the sake of exposition, we introduce a
-|('a,'b) handler| type. This handler handles a |'a computation| that returns a
-|'a| value, and itself returns a |'b| value. The handler has a single return
-case of type |'a -> 'b|, and one or more effect cases. Each effect case,
-handles a effect of type |'c eff|, whose continuation is of type
-|('c,'b) continuation|, and returns a value of type |'b|. For example, in the
-following code:
+syntax with effect patterns. The expression
 \begin{lstlisting}
-effect E : string -> int
-effect F : float -> int
-let r = match perform (E "42") with
-| v -> v == 0
-| effect (E s) k ->
-    not (continue k (int_of_string s))
-| effect (F f) k ->
-    not (continue k (int_of_float f))
-\end{lstlisting}
-\noindent |r| has type |bool|.
+match e with
+| None -> false | Some b -> b
+| effect (E s) k1 -> e1 | effect (F f) k2 -> e2
+\end{lstlisting}
+\noindent is translated to the equivalent of
+\begin{lstlisting}
+match_with (fun () -> e)
+{ retc = (function None -> false | Some b -> b);
+  effc = (function
+    | (E s) -> (fun k1 -> e1)
+    | (F f) -> (fun k2 -> e2)
+    | e -> (fun k -> match perform e with
+            | v -> continue k v
+            | exception e -> discontinue k e)); }
+\end{lstlisting}
+For the sake of exposition, we introduce a |('a,'b) handler| type. This handler
+handles a |'a comp| that returns a |'a| value, and itself returns a |'b| value.
+The handler has a return case of type |'a -> 'b|. The effect case |effc|
+handles effects of type |'c eff| with |('c,'b) continuation| and returns a
+value of type |'b|. The last pattern in |effc| forwards any unhandled effect to
+the outer handler and returns the value and exceptions back.
 
@@
 
+\if{0}
 \begin{figure}
@@
     \begin{array}{rc}
-
       \textsc{StepC} & \inferrule
@@
 \end{figure}
+\fi
+
+\begin{figure}
+  \begin{mathpar}
+    \begin{array}{cc}
+      \textsc{StepC} ~
+      \inferrule{(\term, \env, \fl, \ostack) \cstep \config}
+                {\configa{\term}{\env}{\cstacka{\fl}{\ostack}} \step \config} &
+      \textsc{StepO} ~
+      \inferrule{(\term, \env, k, \cstack) \ostep \config}
+                {\configa{\term}{\env}{\ostacka{k}{\cstack}} \step \config}
+    \end{array}
+  \end{mathpar}
+  \caption{Top-level Reductions -- $\config \step \config$.}
+  \label{sem:top}
+\end{figure}
 
@@
 Intuitively, this splits the world into executing C code and OCaml code.
-
-\begin{figure*}
-	\begin{smathpar}
-		\begin{array}{rrcll}
-			\textsc{AdminC}   & (\term,\env,\fl,\ostack) & \cstep
-												& \configa{\term'}{\env'}{\cstacka{\fl'}{\ostack}}
-												& \text{if } (\term,\env,\fl) \rightsquigarrow (\term',\env',\fl') \\
-			\textsc{RetToO}   & (v,\env,[],\ostacka{k}{\cstack}) & \cstep
-			                  & \configa{v}{\env}{\ostacka{k}{\cstack}} \\
-			\textsc{CallC}    & (v, \_, \cclos{x}{e}{\env}::\fl, \ostack) & \cstep
-												& \configa{e}{\envext{\env}{x}{v}}{\cstacka{\fl}{\ostack}} \\
-			\textsc{Callback} & (v, \_, \oclos{x}{e}{\env}::\fl, \ostack) & \cstep
-												& \configa{e}{\envext{\env}{x}{v}}{\ostacka{k}{\cstacka{\fl}{\ostack}}}
-												& \text{if } k = [[],(\{\caseval{x}{x}\},\emptyset)] \\
-			\textsc{ExnFwdO}  & (v, \env, \ffun{\exnval{l}}::\_, \ostacka{(\fl,\hc) \kcons k}{\cstack}) & \cstep
-												& \configa{v}{\env}{\ostacka{(\ffun{\exnval{l}}::\fl,\hc) \kcons k}{\cstack}}
-		\end{array}
-	\end{smathpar}
-	\caption{C Reductions -- $(\term, \env, \fl, \ostack) \cstep \config$.}
-	\label{sem:cstep}
-\end{figure*}
 
@@
   \label{sem:step}
+\end{figure*}
+\begin{figure*}
+  \begin{smathpar}
+    \begin{array}{rrcll}
+      \textsc{AdminC}   & (\term,\env,\fl,\ostack) & \cstep
+                        & \configa{\term'}{\env'}{\cstacka{\fl'}{\ostack}}
+                        & \text{if } (\term,\env,\fl) \rightsquigarrow (\term',\env',\fl') \\
+      \textsc{RetToO}   & (v,\env,[],\ostacka{k}{\cstack}) & \cstep
+                        & \configa{v}{\env}{\ostacka{k}{\cstack}} \\
+      \textsc{CallC}    & (v, \_, \cclos{x}{e}{\env}::\fl, \ostack) & \cstep
+                        & \configa{e}{\envext{\env}{x}{v}}{\cstacka{\fl}{\ostack}} \\
+      \textsc{Callback} & (v, \_, \oclos{x}{e}{\env}::\fl, \ostack) & \cstep
+                        & \configa{e}{\envext{\env}{x}{v}}{\ostacka{k}{\cstacka{\fl}{\ostack}}}
+                        & \text{if } k = [[],(\{\caseval{x}{x}\},\emptyset)] \\
+      \textsc{ExnFwdO}  & (v, \env, \ffun{\exnval{l}}::\_, \ostacka{(\fl,\hc) \kcons k}{\cstack}) & \cstep
+                        & \configa{v}{\env}{\ostacka{(\ffun{\exnval{l}}::\fl,\hc) \kcons k}{\cstack}}
+    \end{array}
+  \end{smathpar}
+  \caption{C Reductions -- $(\term, \env, \fl, \ostack) \cstep \config$.}
+  \label{sem:cstep}
 \end{figure*}

### 2020-10-31 retro-concurrency.tex
Fix static semantics

@@
 handles effects of type |'c eff| with |('c,'b) continuation| and returns a
-value of type |'b|. The last pattern in |effc| forwards any unhandled effect to
+value of type |'b|. The last pattern in |effc| forwards any unmatched effect to
 the outer handler and returns the value and exceptions back.
@@
 semantics of Multicore OCaml effect handlers, and to inform an implementation.
-We will highlight the differences whenever we differ from the semantics of the
-implementation. We have included an executable version of the dynamic semantics
-as an OCaml program in the supplementary material \kc{include hl-semantics.ml}.
+We have included an executable version of the dynamic semantics as an OCaml
+program in the supplementary material \kc{include hl-semantics.ml}.
 
@@
 term is either an expression $e$ or a value $v$. The value are integer
-constants $n$, continuations $k$, closure $\clos{x}{e}{\env}$, effect begin
+constants $n$, continuations $k$, closure $\clos{x}{e}{\env}$, effect being
 performed $\effval{l}{k}$ and exception being raised $\exnval{l}$. The
@@
 ($\ostack$). Every C stack $\cstacka{\fl}{\ostack}$ consists of a list of
-frames $\fl$, and, importantly, the OCaml stack $\ostack$ under it. Every OCaml
-stack is either an empty stack $\ostackemp$ or non-empty $\ostacka{k}{\cstack}$
-with the C stack $\cstack$ under it. Thus the stack is an alternating sequence
+frames $\fl$, and the OCaml stack $\ostack$ under it. Every OCaml stack is
+either an empty stack $\ostackemp$ or non-empty $\ostacka{k}{\cstack}$ with the
+C stack $\cstack$ under it. Thus, the program stack is an alternating sequence
 of C and OCaml stacks terminating with an empty OCaml stack $\ostackemp$. The

### 2020-11-01 retro-concurrency.tex
Updates

@@
 \subsection{External calls and callbacks}
+\label{sec:external}
 
@@
 \subsection{Exception handlers}
+\label{sec:exn_handlers}
 
@@
 
+% Note |caml_call_ocaml| is |caml_start_program| in ocaml-multicore source
+
 In order to forward exceptions across C frames, the C stub function
-|caml_start_program|, which is the entry point for both the main OCaml code and
+|caml_call_ocaml|, which is the entry point for both the main OCaml code and
 callbacks into OCaml, pushes an exception handler frame that forwards the
@@
 
-\subsection{DWARF unwind tables}
+\subsection{Stack unwinding}
+
+OCaml generates \emph{stack maps} in order to accurately identify roots on the
+stack for assisting GC. For every call point in the program, OCaml compiler
+emits the size of the frame and the set of all live registers in the frame
+containing pointers into the heap. During a GC, the OCaml stack is walked and
+the roots are marked.
 
@@
 |caml_raise_exn| to get the following backtrace.
-
 \begin{lstlisting}[language=c,basicstyle=\ttfamily\footnotesize]
@@
 #1  0x6fd3e in camlMeander__c_to_ocaml_83 () at meander.ml:5
-#2  0x925a4 in caml_start_program ()
+#2  0x925a4 in caml_call_ocaml ()
 #3  0x8a84a in caml_callback_exn (...) at callback.c:145
@@
 #8  0x6f719 in caml_program ()
-#9  0x925a4 in caml_start_program ()
+#9  0x925a4 in caml_call_ocaml ()
 #10 0x92e4c in caml_startup_common (...) at startup_nat.c:162
@@
 \end{lstlisting}
-
 \noindent which corresponds to the stack layout in
@@
 prologue and epilogue, and makes an extra register (|rbp| on x86) available.
+Note that the DWARF unwind information is complementary to the information used
+by OCaml to walk the stack for GC.
 
@@
 \subsection{Static semantics}
+\label{sec:static_semantics}
 
@@
 value of type |'b|. The last pattern in |effc| forwards any unmatched effect to
-the outer handler and returns the value and exceptions back.
+the outer handler and returns the value and exceptions back to the original
+performer.
 
@@
 
-\if{0}
-\begin{figure}
-  \begin{mathpar}
-    \begin{array}{rc}
-      \textsc{StepC} & \inferrule
-                          {(\term, \env, \fl, \ostack) \cstep \config}
-                          {\configa{\term}{\env}{\cstacka{\fl}{\ostack}} \step \config} \\ \\
-      \textsc{StepO} & \inferrule
-                          {(\term, \env, k, \cstack) \ostep \config}
-                          {\configa{\term}{\env}{\ostacka{k}{\cstack}} \step \config}
-    \end{array}
-  \end{mathpar}
-  \caption{Top-level Reductions -- $\config \step \config$.}
-  \label{sem:top}
-\end{figure}
-\fi
-
-\begin{figure}
+The initial configuration for an expression $e$ is
+$\configa{e}{\emptyset}{\cstacka{[]}{\bullet}}$, where the environment and the
+stack are empty. The top-level reductions
   \begin{mathpar}
@@
   \end{mathpar}
-  \caption{Top-level Reductions -- $\config \step \config$.}
-  \label{sem:top}
-\end{figure}
-
-The initial configuration for an expression $e$ is
-$\configa{e}{\emptyset}{\cstacka{[]}{\bullet}}$, where the environment and the
-stack are empty. The top-level reductions (Figure~\ref{sem:top}) can be
-performed by either by taking a C step $\cstep$ or an OCaml step $\ostep$.
-Intuitively, this splits the world into executing C code and OCaml code.
+can be performed by either by taking a C step $\cstep$ or an OCaml step
+$\ostep$. Intuitively, this splits the world into executing C code and OCaml
+code.
 
@@
 the current OCaml stack, and evaluate the C functions body in the new stack.
-\textsc{RetToC} returns a value to the enclosing C stack. We have exactly one
+\textsc{RetToC} returns a value to the enclosing C stack. In this case, we have exactly one
 fiber on the stack, and this was created in the rule \textsc{Callback}, whose
@@
 and the given handler. The rule \textsc{ExHn} handles an exception, if the
-current handler has a matching exception case. Observe that fiber on which the
-exception was raise has been unwound. The rule \textsc{ExFwdC} forwards the
-exception to C. Here, there is exactly one fiber on the current stack, and the
-handler does not have a matching exception case. The rule \textsc{ExFwdFib}
-forwards the exception to the next fiber if the current handler does not handle
-it.
+current handler has a matching exception case, unwinding the current fiber. The
+rule \textsc{ExFwdC} forwards the exception to C. Here, there is exactly one
+fiber on the current stack, and the handler does not have a matching exception
+case, which we know is the case (see \textsc{Callback} rule). The rule
+\textsc{ExFwdFib} forwards the exception to the next fiber if the current
+handler does not handle it.
 
@@
 effect to the outer fiber, and extends the captured continuation |k''| in the
-process. Recall that we do not capture C frames as part of a continuation.
-\textsc{EffUnHn} models unhandled effect. If the effect bubbles up to the top
-fiber, which we know does not have an effect case (see \textsc{Callback} rule),
-we raise \textsc{unhandled} exception at the point where the corresponding
-effect was performed. This is achieved by appending the captured continuation
-to the front of the current continuation.
+process. Recall that we do not capture C frames as part of a continuation. To
+this end, \textsc{EffUnHn} models unhandled effect. If the effect bubbles up to
+the top fiber, which we know does not have an effect case (see
+\textsc{Callback} rule), we raise \textsc{unhandled} exception at the point
+where the corresponding effect was performed. This is achieved by appending the
+captured continuation to the front of the current continuation.
 
@@
   \begin{array}{rcl}
-    \textsf{continue} ~k ~e & = & k ~(\olam{x}{x}) ~e \\
-    \textsf{discontinue} ~k ~l ~e & = & k ~(\olam{x}{\throw{l}{x}}) ~e
+    \textsf{continue} ~k ~e & = & (k ~(\olam{x}{x})) ~e \\
+    \textsf{discontinue} ~k ~l ~e & = & (k ~(\olam{x}{\throw{l}{x}})) ~e
   \end{array}
 \end{smathpar}
-
 Intuitively, resuming a continuation in both the cases involves evaluating the
@@
 
+\kc{Include example reductions from hl-semantics.ml in Appendix}
+
 \section{Implementation}
@@
 
+In this section, we present the implementation details of effect handlers in
+Multicore OCaml. The implementation follows the operational semantics, but has
+a few key representational differences. Multicore OCaml currently only supports
+the x86-64 backend on Linux, but our design preclude other backends. The
+description in the sequel assumes we are on an x86-64 architecture.
+
+Unlike the operational semantics, handlers with just exception patterns
+(exception handlers) are implemented differently from effect handlers. As
+mentioned in \S~\ref{sec:exn_handlers}, exceptions are pervasive in OCaml and
+are so cheap that they are used for local control flow. Hence, we retain the
+linked exception handler frame implementation of stock OCaml exception handlers
+in Multicore OCaml for performance backwards compatibility. This differs from
+other research languages with effect handlers~\cite{Links,Frank,Eff}, which
+implement exceptions using effects (by ignoring the continuation argument in
+the handler).
+
+\begin{figure}
+  \includegraphics[width=0.7\linewidth]{figures/fiber}
+  \caption{Layout of Multicore OCaml fiber}
+  \label{fig:fiber}
+\end{figure}
+
+\subsection{Heap-allocated fibers}
+
+In the operational semantics, the continuations may be resumed more than once.
+Captured continuations are copies of the original fibers and resuming the
+continuation copies the fibers and leaves the continuation as it is. In our use
+case, continuations will be resumed at-most once, and copying fibers is
+unnecessary and inefficient. Instead, Multicore OCaml allocates fibers on the
+heap. Figure~\ref{fig:fiber} shows the layout of a suspended fiber in Multicore
+OCaml. A fiber is said to be suspended if it is not the current (top-most)
+fiber during OCaml execution.
+
+At the bottom of the stack, we have the |handler_info|, which contains the
+pointer to parent fiber, and the closures for the value, exception and effect
+cases. The closures are created by the translation described in
+\S\ref{sec:static_semantics}; Multicore OCaml supports exception patterns in
+addition to effect patterns in the same handler. This is followed by a block to
+support DWARF unwinding, whose details we will return to later. Then, there is
+a top-level handler that invokes the exception handler closure |clos_hexn|
+after switching to the parent fiber. This is followed by the |pc| of the code
+that invokes value handler after switching to the parent stack. The stack is
+laid out such that when the computation in the handler returns, control goes to
+value handler.
+
+We have the variable-sized area for the OCaml frames next. Since we want the
+handlers to be as small as possible, this area is initially 16 words in length.
+When the stack pointer |rsp| goes less than stack threshold (a field in the
+|Caml_state| table), the stack is said to have overflown. On stack overflow, we
+copy the whole fiber to a new area with double the size. In Multicore OCaml, we
+introduce stack overflow checks in function prologue of OCaml functions. We
+observed that most function calls are to leaf functions with small frame size.
+In order to take advantage of this fact, we introduce a small red zone at the
+top of the stack. The compiler elides the stack overflow check for leaf
+functions whose frame size is less than the size of the red zone.
+
+Finally, we have the saved exception pointer, which points to the top-most
+exception frame, and the saved stack pointer, which points to the top of the
+stack. Switching between fiber only involves saving the exception and the stack
+pointer of the current stack and loading the same on the target stack. We
+allocate the space for the fiber using |malloc| and |free| it when no longer
+required. We do not use a \emph{stack cache} of recently freed stacks for
+speeding up fiber allocation. In our experiments, we observed that directly
+using |mimalloc| outperforms having the stack cache.
+
+\subsection{External calls and callbacks}
+
+Since C functions do not have stack overflow checks, we have to execute the
+external calls in the system stack. Calling a C function from OCaml involves
+saving the exception pointer and the stack pointer in the |fiber_info| of the
+current fiber, and the allocation pointer in the |Caml_state|, updating |rsp|
+to the top of system stack (maintained in |Caml_state|), and calling the C
+function. The actions are reversed when returning from the external call. One
+subtlety is that for C functions that take arguments on the stack, the
+arguments must be copied to the C stack from the OCaml stack where they are
+laid out. For vast majority of external calls, we perform only a few additional
+instructions stock OCaml in order to to set up the right stack.
+
+\begin{figure}
+  \includegraphics[width=0.69\linewidth]{figures/multicore_stack}
+	\caption{Layout of Multicore OCaml stack on \texttt{meander} example from
+	\S\ref{sec:stack}.}
+  \label{fig:mcstack}
+\end{figure}
+
+When we first enter OCaml from C, a new fiber is allocated for the main OCaml
+stack. Since callbacks may be frequent in OCaml programs that use finalisers,
+we run the callbacks on the same fiber as the current one. For example, the
+layout of the Multicore OCaml stack at |caml_raise_exn| in the |meander|
+example from \S\ref{sec:stack} is shown in Figure~\ref{fig:mcstack}. The
+functions |caml_call_c| and |caml_call_ocaml| switch the stacks, and hence are
+show in both the system stack and the fiber. The details of the exception
+frames have been elided in the figure. Thanks to the fiber representation,
+external calls and callbacks remain fast.
+
 \section{Evaluation}

### 2020-11-01 retro-concurrency.tex
Updates

@@
 effect DivBy0 : int -> val
-
 let rec eval e =
@@
         else Z (v1/v2)
-    | Error, _ -> Error
-    | Inf s, _ -> Inf s
+    | Error, _ -> Error | Inf s, _ -> Inf s
     | ...
-
 let (/) n d = Div (Val (Z n), Val (Z d))
@@
 let ios e = match eval e with
-| v -> v (* value case *)
-| effect (DivBy0 _) k -> (* effect case *)
-    continue k Error
+| v -> v
+| effect (DivBy0 _) k -> continue k Error
 \end{lstlisting}
@@
 effect Read : in_channel -> string
-
 let rec eval e = match e with
@@
   register_async_input_line (chan, k);
-  let ui = get_ui_k () in
-  continue ui ()
+	continue (get_ui_k ()) ()
 \end{lstlisting}
@@
 \begin{lstlisting}[language=c,basicstyle=\ttfamily\footnotesize]
-% gdb ./meander.exe
-(gdb) break caml_raise_exn
-Breakpoint 1 at 0x3e5dc
-(gdb) r
-Starting program: meander.exe
-Breakpoint 1, 0x925dc in caml_raise_exn ()
-(gdb) bt
 #0  0x925dc in caml_raise_exn ()
@@
 #6  0x6fd77 in camlMeander__omain_88 () at meander.ml:10
-#7  0x6fe92 in camlMeander__entry () at meander.ml:14
+#7  0x6fe92 in camlMeander__entry () at meander.ml:13
 #8  0x6f719 in caml_program ()
@@
 |ic| to the output channel |oc|:
-
 \begin{lstlisting}
 let rec copy ic oc =
-  try
-    let l = input_line ic in
-    output_string oc (l ^ "\n");
+  try output_string oc ((input_line ic) ^ "\n");
       copy ic oc
@@
 \end{lstlisting}
-
 |input_line| raises |End_of_file| exception on reaching the end of input.
@@
 One of our goals(\S\ref{sec:requirements}) is to make this code transparently
-asynchronous. We can define effects for performing the I/O operations:
+asynchronous. We can define effects for performing the I/O operations  and wrap
+them up in functions with the same signature as the one from the standard
+library:
 \begin{lstlisting}
@@
 effect Out_str : out_channel -> string -> unit
-\end{lstlisting}
-\noindent and wrap them up in functions with the same signature as the one from
-the standard library.
-\begin{lstlisting}
 let input_line ic = perform (In_line ic)
@@
 handles effects of type |'c eff| with |('c,'b) continuation| and returns a
-value of type |'b|. The last pattern in |effc| forwards any unmatched effect to
+value of type |'b|. The last case in |effc| forwards any unmatched effect to
 the outer handler and returns the value and exceptions back to the original
@@
 stack. Switching between fiber only involves saving the exception and the stack
-pointer of the current stack and loading the same on the target stack. We
-allocate the space for the fiber using |malloc| and |free| it when no longer
-required. We do not use a \emph{stack cache} of recently freed stacks for
-speeding up fiber allocation. In our experiments, we observed that directly
-using |mimalloc| outperforms having the stack cache.
+pointer of the current stack and loading the same on the target stack. Since
+OCaml does not generate pointers into the stack, the two |fiber_info| fields
+are the only ones that need to be updated when fibers are moved. We allocate
+the space for the fiber using |malloc| and |free| it when no longer required.
+We do not use a \emph{stack cache} of recently freed stacks for speeding up
+fiber allocation. In our experiments, we observed that directly using
+|mimalloc| outperforms having the stack cache.
 
@@
 functions |caml_call_c| and |caml_call_ocaml| switch the stacks, and hence are
-show in both the system stack and the fiber. The details of the exception
+shown in both the system stack and the fiber. The details of the exception
 frames have been elided in the figure. Thanks to the fiber representation,
@@
 
+\subsection{Effect handlers}
+
+\begin{figure}
+	\includegraphics[width=0.6\linewidth]{figures/fiber_handler}
+	\caption{Layout of stack with effect handlers}
+	\label{fig:fiber_handler}
+\end{figure}
+
+In order to illustrate the implementation of handling effects, consider the
+example presented below:
+\begin{lstlisting}
+effect E : unit
+effect F : unit;;
+match (* comp_e *)
+  match (* comp_f *) (*p1*) perform E (*p3*)
+  with v -> v | effect F kf -> continue kf ()
+with v -> v | effect E ke -> (*p2*) continue ke ()
+\end{lstlisting}
+The code performs |E| which is handled in the outer-most handler, and is
+immediately resumed. The layout of the stack as the program executes is
+captured in Figure~\ref{fig:fiber_handler}. At position |p1|, |rsp| is at the
+top of the fiber |f|.
+
+When an effect is performed, we invoke the effect closure |clos_heff| in the
+fiber |f| with the current fiber pointer as the continuation. Since the first
+handler does not handle effect |E|, the effect is forwarded to outer handler by
+performing the effect closure of fiber |e|. This handles |E|. We allocate a one
+word continuation object |ke| on the heap that holds the pointer to fiber |e|
+which points to the fiber |f| (position |p2|). Thus, continuations are captured
+without copying frames. Since each handler closure is evaluated until a machine
+one is found, the take to handle an effect is linear in the number of handlers.
+
+When the continuation is resumed, we overwrite the value of |ke| to |NULL| to
+enforce at-most once semantics. Resuming a continuation involves traversing the
+linked-list of fibers and making the last fiber point to the current fiber.
+Just as in the operational semantics, the implementation invokes the
+appropriate closure to either |continue| or |discontinue| the continuation
+(position |p3|).
+
+\subsection{Stack unwinding}
+
 \section{Evaluation}

### 2020-11-02 retro-concurrency.tex
Implementation section is finished

@@
 analysis tools such as debuggers and profilers that utilise DWARF stack unwind
-tables to obtain a backtrace. Our primary motivation for adding effect handlers
-to OCaml is to support (1) highly concurrent applications such as web servers
-to be expressed naturally in \emph{direct-style} as opposed to using callbacks
-and (2) efficient data and task parallel programs in the vein of Cilk~\cite{},
-OpenMP~\cite{} and Intel TBB~\cite{}.
+tables~\cite{DWARF} to obtain a backtrace. Our primary motivation for adding
+effect handlers to OCaml is to support (1) highly concurrent applications such
+as web servers to be expressed naturally in \emph{direct-style} as opposed to
+using callbacks and (2) efficient data and task parallel programs in the vein
+of Cilk~\cite{}, OpenMP~\cite{} and Intel TBB~\cite{}.
 
@@
     \includegraphics[scale=0.46]{figures/stock_stack}
-    \subcaption{Stack layout before raising exception.}
+    \subcaption{Stack layout before raising exception \texttt{E1}.}
     \label{fig:stock_stack}
@@
 \subsection{Stack unwinding}
+\ref{sec:unwind}
 
@@
 immediately resumed. The layout of the stack as the program executes is
-captured in Figure~\ref{fig:fiber_handler}. At position |p1|, |rsp| is at the
-top of the fiber |f|.
+captured in Figure~\ref{fig:fiber_handler}. The arrows between rectangular
+fibers are parent pointers. At position |p1|, |rsp| is at the top of the fiber
+|f|.
 
@@
 appropriate closure to either |continue| or |discontinue| the continuation
-(position |p3|).
+(position |p3|). We perform standard tail-call optimization so that resumptions
+at tail positions do not build up stack.
 
 \subsection{Stack unwinding}
+
+For stack walking during GC, the only change necessary was to make the runtime
+aware of fibers layout and how to follow the parent pointers. The stack map
+generation and scanning each frame for roots remains the same. Due to the
+uniformity of fiber layout, suspended fibers are scanned the same way as the
+current active fiber. Previous work~\cite{RetroParallel} presents the
+challenges and solution for the interaction of fibers and the concurrent
+mark-and-sweep GC of Multicore OCaml.
+
+The challenge with DWARF stack unwinding is to make it aware of the
+non-contiguous stacks. While the complete details of DWARF stack unwinding is
+beyond the scope of the paper, it is beneficial to know how DWARF unwind tables
+are constructed in order to appreciate our solution. We refer the interested
+reader to Bastian et al.~\cite{} for a good overview of DWARF stack unwinding.
+
+Logically, DWARF maintains a large table which records for every machine
+instruction where the return address and callee-saved registers are stored. To
+avoid reifying this large table, DWARF represents the table using a compact,
+turing-complete, bytecode representation, which is interpreted on demand for
+building the unwind table. Within each function, DWARF maintains a
+\emph{canonical frame address} (CFA), which is invariant in the function body,
+and is traditionally the stack pointer before entering this function. Hence, on
+x86-64, where the return address is pushed on the stack on call, the return
+address is at |CFA - 8|.
+
+Our goal is to help DWARF identify the CFA when stacks are switched at external
+calls, callbacks and effect handlers. We have implemented short, hand-written
+DWARF bytecode at these three places, which has been validated with the DWARF
+unwind table validator tool~\cite{Bastian}. When starting to run on a new
+fiber, we insert DWARF bytecode to use the follow the |parent_fiber| pointer
+and dereference the |saved_sp| to get the CFA. During callback, we save the
+current system stack pointer in the DWARF block in Figure~\ref{fig:fiber} to
+identify the CFA in the C stack. DWARF unwinding for external calls is
+implemented similarly by saving the current OCaml stack pointer. With these
+changes, we get the same backtrace for the |meander.exe| program from
+\S\ref{sec:unwind}, modulo runtime system functions due to effect handlers.
 

### 2020-11-02 retro-concurrency.tex
Minor edits; TK comment on effects implementation

@@
 \newcommand{\am}[1]{{\color{blue} {\it [AM says: #1]}}}
+\newcommand{\tk}[1]{{\color{blue} {\it [TK says: #1]}}}
 
@@
 are no callee-saved registers to be restored when the exception handler runs.
-Hence, installing an exception handler simply pushes the program counter (pc)
+Hence, installing an exception handler simply pushes the program counter (|pc|)
 of the handler and the current exception pointer (|exn_ptr| -- a field in
@@
 a few key representational differences. Multicore OCaml currently only supports
-the x86-64 backend on Linux, but our design preclude other backends. The
-description in the sequel assumes we are on an x86-64 architecture.
+the x86-64 backend on Linux, but our design doesn't preclude other backends.
+The description that follows assumes a x86-64 architecture.
 
@@
 copy the whole fiber to a new area with double the size. In Multicore OCaml, we
-introduce stack overflow checks in function prologue of OCaml functions. We
+introduce stack overflow checks into the function prologue of OCaml functions. We
 observed that most function calls are to leaf functions with small frame size.
@@
 exception frame, and the saved stack pointer, which points to the top of the
-stack. Switching between fiber only involves saving the exception and the stack
+stack. Switching between fibers only involves saving the exception and the stack
 pointer of the current stack and loading the same on the target stack. Since
@@
 laid out. For vast majority of external calls, we perform only a few additional
-instructions stock OCaml in order to to set up the right stack.
+instructions over stock OCaml in order to to set up the right stack.
 
@@
 
+\tk{Something wasn't quite right with this: we allocate the continuation object
+at the point of the perform. Also should we describe how the first handler
+passes the effect on and adds the e onto the continuation that started with f?}
+
 When the continuation is resumed, we overwrite the value of |ke| to |NULL| to
@@
 For stack walking during GC, the only change necessary was to make the runtime
-aware of fibers layout and how to follow the parent pointers. The stack map
-generation and scanning each frame for roots remains the same. Due to the
-uniformity of fiber layout, suspended fibers are scanned the same way as the
-current active fiber. Previous work~\cite{RetroParallel} presents the
-challenges and solution for the interaction of fibers and the concurrent
-mark-and-sweep GC of Multicore OCaml.
+aware of fibers layout, the closures for the handlers and how to follow the
+parent pointers. The stack map generation and scanning each frame for roots
+remains the same. Due to the uniformity of fiber layout, suspended fibers are
+scanned the same way as the current active fiber. Previous
+work~\cite{RetroParallel} presents the challenges and solution for the
+interaction of fibers and the concurrent mark-and-sweep GC of Multicore OCaml.
 
@@
 calls, callbacks and effect handlers. We have implemented short, hand-written
-DWARF bytecode at these three places, which has been validated with the DWARF
+DWARF bytecode at these three places, which we have validated with the DWARF
 unwind table validator tool~\cite{Bastian}. When starting to run on a new
-fiber, we insert DWARF bytecode to use the follow the |parent_fiber| pointer
+fiber, we insert DWARF bytecode to follow the |parent_fiber| pointer
 and dereference the |saved_sp| to get the CFA. During callback, we save the

### 2020-11-02 retro-concurrency.tex
Added skeleton of what to present in the evaluation section

@@
 
+\subsection{The impact on code not using effects}
+
+There are several changes to the runtime to support fibers and effect handlers
+that will be executed by code not using effects: OCaml stacks now have dynamic
+checks in the function prologue and the FFI interface tracks stack switching.
+
+TODO:
+\begin{itemize}
+\item Aim: |R1| not altered existing code performance too much
+\item sandmark macro benchmarks to show sequential performance: 4.10.0+stock, 4.10.0+multicore,
+4.10.0+multicore no red-zone/leaf-optimization, 4.10.0+multicore
+\item do we want to include some micro-benchmarks focusing on the FFI callstack overhead?
+\end{itemize}
+
+\subsection{Measurements of overheads in our effects implementation}
+
+We have implemented effects within OCaml, here we present timing measurements
+for the various stages of stack creation (|a| to |b|), performing and handling
+an effect (|b| to |c|), continuation resumption (|c| to |d|) and returning from
+a computation and freeing the stack (|d| to |e|).
+
+\begin{lstlisting}
+let foo () =
+ (* a *) match
+ (* b *)   perform E
+ (* d *) with effect E k ->
+ (* c *)   continue k ()
+ (* e *)
+\end{lstlisting}
+
+TODO:
+\begin{itemize}
+\item Aim: |R3| to provide some timing information as to the size of the overhead and where the overhead is.
+\item use KC's measurement stubs as presented in WASM
+\end{itemize}
+
+\subsection{Effects benchmarks}
+
+We now look at the impact of the overhead in a variety of benchmarks comparing
+the cost of using our effects abstraction versus alternative methods of
+implementation.
+
+TODO:
+\begin{itemize}
+\item Aim: |R3| Do we have an efficient implementation of effects and are effects competitive vs alternative implementation strategies.
+\item we can show the sequential vs stack benchmarks as used by 'Folklore to fact' \tk{slightly weak story as the sequential ones will be much faster for these very simple benchmarks and effects don't really add anything to their implementation}
+\item general generator examples (using examples from effects-example)
+\item chamenos vs chamenos effects vs lwt (using examples from effects-example)
+\end{itemize}
+
+\subsection{Webserver benchmarks}
+
+TODO:
+\begin{itemize}
+\item Aim: |R3| Do we have an efficient implementation of effects.
+\item effects are a very nice way to program non-blocking IO for webservers
+\item show that using effects doesn't impose undue overhead and that we are competitive with LWT, Async, Go with the wrk web benchmark
+\end{itemize}
+
+\subsection{Using fibers for task based parallelism}
+
+TODO:
+\begin{itemize}
+\item Aim: |R3| Do we have an efficient implementation of effects.
+\item minor alteration to domainslib to launch tasks in their own stack
+\item effects are the right way to do some task base parallelism. Use the nqueens example and show the scaling we can achieve in parallel
+\item show that using fibers doesn't have large detrimental impact to existing parallel applications: Take sandmark: minilight, nbody, LU decomposition
+\end{itemize}
+
 \section{Related Work}

### 2020-11-03 retro-concurrency.tex
Updates

@@
 \documentclass[sigplan,10pt,review,anonymous]{acmart}\settopmatter{printfolios=true,printccs=false,printacmref=false}
+\settopmatter{printacmref=false}
+\setcopyright{none}
+\renewcommand\footnotetextcopyrightpermission[1]{}
+\pagestyle{plain}
 
@@
 
+\newcommand{\mycaption}[1]{\vspace{-4mm}\caption{#1}}
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 Effect Handlers~\cite{Plotkin09} provide a modular foundation for user-defined
-effects. The key idea is to separate the definition of the effectful
-\emph{operations} from their interpretations, which are defined by
-\emph{handlers} of the effects. For example,
+effects. The key idea is to separate the definition of the effectful operations
+from their interpretations, which are defined by \emph{handlers} of the
+effects. For example,
 %
@@
 %
-declares an effect |Read|, which is parameterised with an input channel of type
-|in_channel|, which when \emph{performed} returns a |string| value. A
+declares an \emph{effect} |Read|, which is parameterised with an input channel
+of type |in_channel|, which when \emph{performed} returns a |string| value. A
 computation can perform the |Read| effect without knowing how the |Read| effect
@@
 separation of effectful operations from their interpretations, effect handlers
-enable new approaches to modular programming. A computation may perform many
-different effects, which are interpreted by one or more enclosing handlers.
+enable new approaches to modular programming.
 
@@
 or not, conditions allow the client of the library to make that decision. If
-the error is non-fatal, Conditions provide a restart mechanism to
-resume execution from the point where the condition was signalled. On
+the error is non-fatal, Conditions provide a \emph{restart} mechanism to resume
+execution from the point where the condition was signalled. On
 division-by-zero, the calculator signals the handler and lets the handler
@@
 On line 4, we declare an effect |DivBy0| which is parameterised by an integer,
-which when performed returns a value. While interpreting the division
-expression, if the denominator is 0, then we perform |DivBy0| effect with the
-numerator value (line 13) so that the context may decide whether the error is
-fatal or not, and resume with an appropriate value if the error is not fatal.
-We can get the |iOS| semantics by wrapping the evaluator computation with the
-following handler:
-
+which when performed returns result of division. If the denominator is 0, then
+we perform |DivBy0| effect with the numerator value (line 12) so that the
+context may decide the return value. We can get the |iOS| semantics by wrapping
+the evaluator computation with the following handler:
 \begin{lstlisting}
@@
 \end{lstlisting}
-
 If the evaluator returns a value, we simply return that. In case the evaluator
@@
 following handler:
-
 \begin{lstlisting}
@@
 \end{lstlisting}
-
 The computation |gsearch(1/0)| returns positive infinity (|Inf P|).
@@
 using the |Read| effect defined earlier as follows:
-
 \begin{lstlisting}
@@
 \end{lstlisting}
-
 The |Read| effect can then be handled by another handler enclosing the
 division-by-zero handler:
-
 \begin{lstlisting}
@@
 \end{lstlisting}
-
 \noindent which registers the read operation to be performed asynchronously
@@
 is meaningless and hence, we would lose compatibility with tools that inspect
-the program stack. Hence, we choose not to CPS translation. The alternative is
-to use the call stack for implementing continuation. Farvardin et
-al~\cite{Folklore} show that there is a spectrum of implementation strategies
-for continuation with different trade-offs.
+the program stack. Hence, we choose not to use CPS translation and represent
+continuations as call stacks.
 
@@
 For example, running |perform (Read stdin)| without an enclosing handler raises
-exception |Unhandled| at the point of |perform|. We leave the question of
+|Unhandled| exception at the point of |perform|. We leave the question of
 effect safety for future work.
@@
 
-We motivate our implementation strategy based on the following ideal
+We motivate our effect handler design based on the following ideal
 requirements:
 
+\if{0}
 \begin{enumerate}[label=R\arabic*]
-  \item \textbf{Backwards compatibility} A well behaved OCaml program does not
+  \item \textbf{Backwards compatibility} A well-behaved OCaml program does not
     break under OCaml extended with effect handlers. OCaml code that does not
@@
     well-formed backtraces, and remains compatible with program analysis tools
-    such as debuggers and profilers that inspect the stack using DWARF unwind
-    tables.
+    such as debuggers and profilers that inspect the stack using DWARF stack
+    unwind tables.
   \item \textbf{Effect handler efficiency} In order to support
     highly-concurrent applications, the program must accommodate millions of
-    continuations in the heap. Installing effect handlers, capturing and
+    continuations at the same time. Installing effect handlers, capturing and
     resuming continuations must be fast.
@@
 \end{enumerate}
-
-The need to host millions of active continuations in the heap rules out the use
+\fi
+
+\noindent \textbf{R1: Backwards compatibility.} A well-behaved OCaml program does
+not break under OCaml extended with effect handlers. OCaml code that does not
+use effect handlers will pay negligible performance cost.
+
+\noindent \textbf{R2: Tool compatibility.} OCaml with effect handlers produce
+well-formed backtraces, and remains compatible with program analysis tools such
+as debuggers and profilers that inspect the stack using DWARF stack unwind
+tables.
+
+\noindent \textbf{R3: Effect handler efficiency.} In order to support highly-concurrent
+applications, the program must accommodate millions of continuations at the
+same time. Installing effect handlers, capturing and resuming continuations
+must be fast.
+
+\noindent \textbf{R4: Forwards compatibility.} While this work does not present an
+effect system, the design must be compatible with a future effect system for
+OCaml. As a cornerstone of modularity, we also want blocking I/O code to
+\emph{transparently} be made asynchronous with the help of effect handlers.
+
+The need to host millions of continuations at the same time rules out the use
 of large contiguous stack space as in C for continuations. Instead, we resort
@@
 perform stack overflow checks and external C functions, which do not have stack
-overflow checks, must be performed on a separate stack. Additionally, we must
-generate DWARF stack unwind tables for stacks that may be non-contiguous and
-contain frames from different languages. In this work, we develop the compiler
-and runtime support for implementing efficient effect handlers for OCaml.
+overflow checks, must be performed on a separate system stack. Additionally, we
+must generate DWARF stack unwind tables for stacks that may be non-contiguous.
+In this work, we develop the compiler and runtime support for implementing
+efficient effect handlers for OCaml that satisfy these requirements.
 
@@
 Swift concurrency roadmap~\cite{} includes direct-style asynchronous
-programming and structured concurrency. We believe that our design choices will
-inform similar choices to be made in the other languages.
+programming and structured concurrency as milestones. We believe that our
+design choices will inform similar choices to be made in the other
+industrial-strength languages.
 
@@
 \begin{itemize}
-  \item the design and implementation of affine effect handlers for OCaml. Our
-    design retains OCaml's compatibility with program analysis tools that
+  \item the design and implementation of one-shot effect handlers for OCaml.
+    Our design retains OCaml's compatibility with program analysis tools that
     inspect the stack using DWARF unwind tables. We have validated our DWARF
     unwind tables using an automated validator tool~\cite{}.
-  \item a formal operation semantics for the effect handler implementation in
+  \item a formal operational semantics for the effect handler implementation in
     OCaml.
@@
 programming language. We call our implementation ``Multicore OCaml'' to
-distinguish it from stock OCaml. Multicore OCaml uses effect handlers as the
-mechanism to express user-level concurrency.
+distinguish it from stock OCaml.
 
@@
 effect handler implementation must retain in order to provide backwards
-compatibility. In \S\ref{sec:refine}, we refine the design further
-focussing on the issues that come up when integrating effect handlers into a
-mainstream language. \S\ref{sec:semantics} presents the formal operational
+compatibility. In \S\ref{sec:refine}, we refine the design further focussing on
+the issues that come up when integrating effect handlers into a mainstream
+systems language. \S\ref{sec:semantics} presents the formal operational
 semantics for Multicore OCaml effect handlers. \S\ref{sec:impl} discusses the
@@
 \begin{figure*}
-\begin{minipage}{0.59\linewidth}
+\begin{minipage}{0.30\linewidth}
   \begin{minipage}{\linewidth}
     \begin{lstlisting}[numbers=left]
-external ocaml_to_c : unit -> int = "ocaml_to_c"
+external ocaml_to_c
+  : unit -> int = "ocaml_to_c"
 exception E1
 exception E2
-
 let c_to_ocaml () = raise E1
-let _ = Callback.register "c_to_ocaml" c_to_ocaml
-
+let _ = Callback.register
+    "c_to_ocaml" c_to_ocaml
 let omain () =
   try (* h1 *)
-    try (* h2 *) ocaml_to_c () with E2 -> 0
-  with E1 -> 42;;
-
+    try (* h2 *) ocaml_to_c ()
+    | with E2 -> 0
+  | with E1 -> 42;;
 let _ = assert (omain () = 42)
     \end{lstlisting}
+    \vspace{-2mm}
     \subcaption{\texttt{meander.ml}}
@@
 value ocaml_to_c (value unit) {
-  caml_callback(*caml_named_value("c_to_ocaml"), Val_unit);
+  caml_callback(*caml_named_value
+    ("c_to_ocaml"), Val_unit);
   return Val_int(0);
@@
     \end{lstlisting}
+    \vspace{-4mm}
     \subcaption{\texttt{meander.c}}
@@
 \end{minipage}
-  \begin{minipage}{0.39\linewidth}
+%
+\begin{minipage}{0.30\linewidth}
   \centering
   \includegraphics[scale=0.46]{figures/stock_stack}
-    \subcaption{Stack layout before raising exception \texttt{E1}.}
+  \subcaption{Stack layout before raise \texttt{E1}.}
   \label{fig:stock_stack}
 \end{minipage}
-  \caption{Program stack on stock OCaml.}
+%
+\begin{minipage}{0.39\linewidth}
+\begin{lstlisting}[language=c,basicstyle=\ttfamily\footnotesize]
+#0  0x925dc in caml_raise_exn ()
+#1  0x6fd3e in camlMeander__c_to_ocaml_83 () at meander.ml:5
+#2  0x925a4 in caml_call_ocaml ()
+#3  0x8a84a in caml_callback_exn (...) at callback.c:145
+#4  caml_callback (...) at callback.c:199
+#5  0x76e0a in ocaml_to_c (unit=1) at meander.c:5
+#6  0x6fd77 in camlMeander__omain_88 () at meander.ml:10
+#7  0x6fe92 in camlMeander__entry () at meander.ml:13
+#8  0x6f719 in caml_program ()
+#9  0x925a4 in caml_call_ocaml ()
+#10 0x92e4c in caml_startup_common (...) at startup_nat.c:162
+#11 0x92eab in caml_startup_exn (...) at startup_nat.c:167
+#12 caml_startup (...) at startup_nat.c:172
+#13 0x6f55c in main (...) at main.c:44
+\end{lstlisting}
+\vspace{-2mm}
+\subcaption{\texttt{gdb} backtrace before raise \texttt{E1}.}
+\label{code:gdb_backtrace}
+\end{minipage}
+  \mycaption{Program stack on stock OCaml.}
 \end{figure*}
@@
 managing the program stack. To this end, in this section, we provide an
-overview of the program stack and related mechanisms for stock OCaml.
+overview of the program stack and related mechanisms in stock OCaml.
 
@@
 backend just before raising the exception in |c_to_ocaml| is illustrated in
-Figure~\ref{fig:stock_stack}. The stack grows downwards.
-
-OCaml uses the hardware stack as C, and hence the program stack has alternating
-sequences of C and OCaml frames. Unlike C, OCaml does not create pointers into
-OCaml frames. OCaml uses the hardware support for |call| and |return|
-instructions for function calls and returns. OCaml does not perform explicit
-stack overflow checks in code, and, just like C, relies on the guard page at
-the end of the stack region to detect stack overflow. Stack overflow is
-detected by a memory fault, and |Stack_overflow| exception is raised to unwind
-the stack.
+Figure~\ref{fig:stock_stack}. Note that the stack grows downwards.
+
+OCaml uses the same hardware stack as C, and hence the program stack has
+alternating sequences of C and OCaml frames. However, unlike C, OCaml does not
+create pointers into OCaml frames. OCaml uses the hardware support for |call|
+and |return| instructions for function calls and returns. OCaml does not
+perform explicit stack overflow checks in code, and, just like C, relies on the
+guard page at the end of the stack region to detect stack overflow. Stack
+overflow is detected by a memory fault, and |Stack_overflow| exception is
+raised to unwind the stack.
 
@@
 OCaml does not use the C calling convention. In particular, there are no
-callee-saved registers in OCaml. On the |x86-64| backend, a register each is
-pinned for the \emph{allocation pointer} into the minor heap used for bump
-pointer allocation, and the |Caml_state|, a table of global variables used by
-the runtime. These registers are chosen to be C callee saved registers. This
-makes external calls extremely fast in OCaml. If the external function does not
-allocate in the OCaml heap, then the external functions can be called directly,
-and no bookkeeping is necessary. For external functions which allocate in the
-OCaml heap, the cached allocation pointer is saved to a field in the
-|Caml_state| before the external call, and loaded back on return. Similarly,
-callbacks into OCaml are also cheap. Callbacks into OCaml involves loading the
-arguments in the right registers and calling the OCaml function. OCaml
-callbacks are relatively common as the garbage collector (GC) finalisation
-functions are executed as OCaml callbacks.
+callee-saved registers in OCaml. On the |x86-64| backend, the C callee-saved
+registers, |r15| holds the \emph{allocation pointer} into the minor heap used
+for bump pointer allocation, and |r14| holds the |Caml_state|, a table of
+global variables used by the runtime. This makes external calls extremely fast
+in OCaml. If the external function does not allocate in the OCaml heap, then
+the external functions can be called directly, and no bookkeeping is necessary.
+For external functions which allocate in the OCaml heap, the cached allocation
+pointer is saved to a field in the |Caml_state| before the external call, and
+loaded back on return. Similarly, callbacks into OCaml are also cheap.
+Callbacks into OCaml involves loading the arguments in the right registers and
+calling the OCaml function. OCaml callbacks are relatively common as the
+garbage collector (GC) finalisation functions are executed as OCaml callbacks.
 
@@
 of the handler and the current exception pointer (|exn_ptr| -- a field in
-|Caml_state|). The exception pointer is also updated to be the current stack
-pointer (|sp|). This creates a linked-list of exception handler frames on the
-stack as shown in Figure~\ref{fig:stock_stack}. Raising an exception simply
-sets |sp| to |exn_ptr|, loads the saved |exn_ptr|, and jumps to the |pc| of the
-handler.
+|Caml_state|). At this point, the exception pointer is updated to be the
+current stack pointer (|rsp|). This creates a linked-list of exception handler
+frames on the stack as shown in Figure~\ref{fig:stock_stack}. Raising an
+exception simply sets |rsp| to |exn_ptr|, loads the saved |exn_ptr|, and jumps
+to the |pc| of the handler.
 
@@
 callbacks into OCaml, pushes an exception handler frame that forwards the
-exception the innermost OCaml exception handler (|raised_exn_c| in
+exception to the innermost OCaml exception handler (|raise_exn_c| in
 Figure~\ref{fig:stock_stack}), or prints a fatal error (|fatal_uncaught|) if
@@
 \subsection{Stack unwinding}
-\ref{sec:unwind}
+\label{sec:unwind}
 
@@
 containing pointers into the heap. During a GC, the OCaml stack is walked and
-the roots are marked.
+the roots are marked skipping over the C frames.
 
@@
 and~\ref{code:meander_c}, one could set a break point in |gdb| at
-|caml_raise_exn| to get the following backtrace.
-\begin{lstlisting}[language=c,basicstyle=\ttfamily\footnotesize]
-#0  0x925dc in caml_raise_exn ()
-#1  0x6fd3e in camlMeander__c_to_ocaml_83 () at meander.ml:5
-#2  0x925a4 in caml_call_ocaml ()
-#3  0x8a84a in caml_callback_exn (...) at callback.c:145
-#4  caml_callback (...) at callback.c:199
-#5  0x76e0a in ocaml_to_c (unit=1) at meander_.c:5
-#6  0x6fd77 in camlMeander__omain_88 () at meander.ml:10
-#7  0x6fe92 in camlMeander__entry () at meander.ml:13
-#8  0x6f719 in caml_program ()
-#9  0x925a4 in caml_call_ocaml ()
-#10 0x92e4c in caml_startup_common (...) at startup_nat.c:162
-#11 0x92eab in caml_startup_exn (...) at startup_nat.c:167
-#12 caml_startup (...) at startup_nat.c:172
-#13 0x6f55c in main (...) at main.c:44
-\end{lstlisting}
-\noindent which corresponds to the stack layout in
-Figure~\ref{fig:stock_stack}.
+|caml_raise_exn| to get the backtrace in Figuere~\ref{code:gdb_backtrace} which
+corresponds to the stack layout in Figure~\ref{fig:stock_stack}.
 
@@
 continuations more than once. Hence, our continuations are one-shot, and
-resuming the continuation more than once raises |Invalid_argument _| exception.
+resuming the continuation more than once raises |Invalid_argument| exception.
 It is well-known that one-shot continuations can be implemented
@@
 \end{lstlisting}
-|input_line| raises |End_of_file| exception on reaching the end of input.
+
+The function |input_line| raises |End_of_file| exception on reaching the end of input.
 Although not apparent, the code is written in a defensive style to handle other
@@
 We can then use a top-level I/O handler similar to the introductory example
-(\S\ref{sec:aio}) to discharge the I/O operations asynchronously. This handles
-value return cases, but what about the exceptional cases |End_of_file| and
-|Sys_error|? We introduce a |discontinue| primitive to resume a continuation by
-raising an exception. In this example, on reaching the end of file, we would
-discontinue the continuation of the |copy| function with |discontinue k End_of_file|,
-which raises the exception at |input_line|, and the open channels
-will be closed.
+(\S\ref{sec:aio}) to discharge the I/O operations asynchronously and resumes
+with the result. This handles value return cases, but what about the
+exceptional cases |End_of_file| and |Sys_error|? We introduce a |discontinue|
+primitive to resume a continuation by raising an exception. In this example, on
+reaching the end of file, we would discontinue the continuation of the |copy|
+function with |discontinue k End_of_file|, which raises the exception at
+|input_line|, and the open channels will be closed.
 
@@
 val discontinue: ('a,'b) continuation -> exn -> 'b
-
 (* Internal API *)
@@
 \end{lstlisting}
-
 We introduce an extensible variant type~\cite{} |'a eff| of effect values,
@@
 handles effects of type |'c eff| with |('c,'b) continuation| and returns a
-value of type |'b|. The last case in |effc| forwards any unmatched effect to
-the outer handler and returns the value and exceptions back to the original
-performer.
+value of type |'b|. The last case in |effc| \emph{reperforms} any unmatched
+effect to the outer handler and returns the value and exceptions back to the
+original performer. In the implementation, reperfom is implemented as a
+primitive to avoid executing code on the resumption path.
 
@@
   \end{minipage}
-  \caption{Syntax of expressions and configurations}
+  \mycaption{Syntax of expressions and configurations}
   \label{sem:syntax}
@@
 semantics of Multicore OCaml effect handlers, and to inform an implementation.
-We have included an executable version of the dynamic semantics as an OCaml
-program in the supplementary material \kc{include hl-semantics.ml}.
 
@@
 frames $\fl$, and the OCaml stack $\ostack$ under it. Every OCaml stack is
-either an empty stack $\ostackemp$ or non-empty $\ostacka{k}{\cstack}$ with the
+either empty $\ostackemp$ or non-empty $\ostacka{k}{\cstack}$ with the
 C stack $\cstack$ under it. Thus, the program stack is an alternating sequence
@@
   \end{smathpar}
-  \caption{Administrative Reductions -- $(\term, \env, \fl) \rightsquigarrow (\tau, \env, \fl)$.}
+  \vspace{-1mm}
+  \mycaption{Administrative Reductions -- $(\term, \env, \fl) \rightsquigarrow (\tau, \env, \fl)$.}
   \label{sem:step}
@@
   \end{smathpar}
-  \caption{C Reductions -- $(\term, \env, \fl, \ostack) \cstep \config$.}
+  \vspace{-1mm}
+  \mycaption{C Reductions -- $(\term, \env, \fl, \ostack) \cstep \config$.}
   \label{sem:cstep}
@@
 $[[],(\{\caseval{x}{x}\},\emptyset)]$ with no captured frames and an empty
-handler with an identity return case alone. \textsc{Resume1} and
-\textsc{Resume2} are used for resuming continuations, and we shall defer their
-explanation till later.
+handler with an identity return case alone. We shall return to \textsc{Resume1}
+and \textsc{Resume2} later.
 
@@
 behaviour of calling a C function. Since the program is currently executing C,
-indicated by the current stack being a C stack $\cstacka{\fl}{\ostack}$, we can
-perform the call on the current stack. In case the abstraction is an OCaml
-abstraction (\textsc{Callback}), we create an OCaml stack with the C stack as
-its tail, with the current continuation being empty. This captures the
+we can perform the call on the current stack. In case the abstraction is an
+OCaml abstraction (\textsc{Callback}), we create an OCaml stack with the C
+stack as its tail, with the current continuation being empty. This captures the
 behaviour of calling back into OCaml from C. \textsc{RetToO} returns a value to
 the enclosing OCaml stack. \textsc{ExnFwdO} forwards a raised exception to the
-enclosing OCaml stack, ignoring the unwinding the rest of the frames. This
-captures the semantics of raising OCaml exceptions from C.
-
-\subsubsection{OCaml Reductions}
+enclosing OCaml stack, unwinding the rest of the frames. This captures the
+semantics of raising OCaml exceptions from C.
+
+\subsubsection{OCaml reductions}
 
@@
   \end{smathpar}
-  \caption{OCaml Reductions -- $(\term, \env, k, \cstack) \ostep \config$.}
+  \vspace{-1mm}
+  \mycaption{OCaml Reductions -- $(\term, \env, k, \cstack) \ostep \config$.}
   \label{sem:ostep}
@@
 
-OCaml reductions are given in Figure~\ref{sem:ostep}. In OCaml, reductions
-always occur on the top-most fiber in the current stack. \textsc{AdminO}
-performs administrative reductions. \textsc{CallO} evaluates an OCaml function
-on the current stack. \textsc{ExtCall} captures the behaviour of external
-calls. On an external call, we create a C stack with no frames, whose tail is
-the current OCaml stack, and evaluate the C functions body in the new stack.
-\textsc{RetToC} returns a value to the enclosing C stack. In this case, we have exactly one
-fiber on the stack, and this was created in the rule \textsc{Callback}, whose
-handler has identity return case alone and the environment is empty.
-\textsc{RetFib} returns value from a fiber to the previous one, evaluating the
-body of the return case.
+In OCaml (Figure~\ref{sem:ostep}), reductions always occur on the top-most
+fiber in the current stack. \textsc{AdminO} performs administrative reductions.
+\textsc{CallO} evaluates an OCaml function on the current stack.
+\textsc{ExtCall} captures the behaviour of external calls, which are evaluated
+on an empty C stack. \textsc{RetToC} returns a value to the enclosing C stack.
+In this case, we have exactly one fiber on the stack, and this was created in
+the rule \textsc{Callback}, whose handler has identity return case alone and
+the environment is empty. \textsc{RetFib} returns value from a fiber to the
+previous one, evaluating the body of the return case.
 
@@
 Observe that |continue| and |discontinue| are not part of the expressions. They
-are encoded as follows:
-\begin{smathpar}
-  \begin{array}{rcl}
-    \textsf{continue} ~k ~e & = & (k ~(\olam{x}{x})) ~e \\
-    \textsf{discontinue} ~k ~l ~e & = & (k ~(\olam{x}{\throw{l}{x}})) ~e
-  \end{array}
-\end{smathpar}
+are encoded as follows: $\textsf{continue} ~k ~e = (k ~(\olam{x}{x})) ~e$ and
+$\textsf{discontinue} ~k ~l ~e = (k ~(\olam{x}{\throw{l}{x}})) ~e$.
 Intuitively, resuming a continuation in both the cases involves evaluating the
@@
 
-\kc{Include example reductions from hl-semantics.ml in Appendix}
+The traces of example programs are included in the supplementary material along
+with an executable version of the semantics implemented in OCaml.
 
@@
 a few key representational differences. Multicore OCaml currently only supports
-the x86-64 backend on Linux, but our design doesn't preclude other backends.
-The description that follows assumes a x86-64 architecture.
+the x86-64 backend on Linux, but our design does not preclude other backends.
+We assume x86-64 architecture in the sequel.
 
@@
 
-\begin{figure}
-  \includegraphics[width=0.7\linewidth]{figures/fiber}
-  \caption{Layout of Multicore OCaml fiber}
+\subsection{Heap-allocated fibers}
+
+\begin{figure*}
+\begin{minipage}{0.35\linewidth}
+  \centering
+  \includegraphics[scale=0.45]{figures/fiber}
+  \subcaption{Fiber layout}
   \label{fig:fiber}
-\end{figure}
-
-\subsection{Heap-allocated fibers}
+\end{minipage}
+\begin{minipage}{0.64\linewidth}
+  \begin{minipage}{\linewidth}
+    \centering
+    \includegraphics[scale=0.4]{figures/multicore_stack}
+    \subcaption{Stack layout for \texttt{meander} example from \S\ref{sec:stack}.}
+    \label{fig:mcstack}
+    \vspace{2mm}
+  \end{minipage}
+  %
+  \begin{minipage}{0.55\linewidth}
+    \begin{lstlisting}
+effect E : unit
+effect F : unit;;
+match (* comp_e *)
+  match (* comp_f *)
+    (*p1*) perform E (*p3*)
+  with | v -> v | effect F kf -> ...
+with | v -> v
+		 | effect E ke ->
+		     (*p2*) continue ke ()
+    \end{lstlisting}
+		\subcaption{Constructing continuation objects}
+		\label{code:effimpl}
+  \end{minipage}
+  %
+  \begin{minipage}{0.44\linewidth}
+    \centering
+    \includegraphics[scale=0.42]{figures/fiber_handler}
+		\subcaption{Program state for Figure~\ref{code:effimpl}}
+    \label{fig:fiber_handler}
+  \end{minipage}
+\end{minipage}
+\vspace{2mm}
+\mycaption{Layout of Multicore OCaml effect handlers.}
+\end{figure*}
 
@@
 continuation copies the fibers and leaves the continuation as it is. In our use
-case, continuations will be resumed at-most once, and copying fibers is
+cases, continuations will be resumed at-most once, and copying fibers is
 unnecessary and inefficient. Instead, Multicore OCaml allocates fibers on the
-heap. Figure~\ref{fig:fiber} shows the layout of a suspended fiber in Multicore
-OCaml. A fiber is said to be suspended if it is not the current (top-most)
-fiber during OCaml execution.
+heap. Figure~\ref{fig:fiber} shows the layout of a fiber in Multicore OCaml.
 
@@
 after switching to the parent fiber. This is followed by the |pc| of the code
-that invokes value handler after switching to the parent stack. The stack is
-laid out such that when the computation in the handler returns, control goes to
-value handler.
+that invokes value handler |clos_hval| after switching to the parent stack. The
+stack is laid out such that when the computation in the handler returns,
+control goes to value handler.
 
@@
 handlers to be as small as possible, this area is initially 16 words in length.
-When the stack pointer |rsp| goes less than stack threshold (a field in the
+When the stack pointer |rsp| goes less than stack threshold (maintained in the
 |Caml_state| table), the stack is said to have overflown. On stack overflow, we
 copy the whole fiber to a new area with double the size. In Multicore OCaml, we
-introduce stack overflow checks into the function prologue of OCaml functions. We
-observed that most function calls are to leaf functions with small frame size.
-In order to take advantage of this fact, we introduce a small red zone at the
-top of the stack. The compiler elides the stack overflow check for leaf
+introduce stack overflow checks into the function prologue of OCaml functions.
+We observed that most function calls are to leaf functions with small frame
+size. In order to take advantage of this fact, we introduce a small red zone at
+the top of the stack. The compiler elides the stack overflow check for leaf
 functions whose frame size is less than the size of the red zone.
@@
 exception frame, and the saved stack pointer, which points to the top of the
-stack. Switching between fibers only involves saving the exception and the stack
-pointer of the current stack and loading the same on the target stack. Since
-OCaml does not generate pointers into the stack, the two |fiber_info| fields
-are the only ones that need to be updated when fibers are moved. We allocate
-the space for the fiber using |malloc| and |free| it when no longer required.
-We do not use a \emph{stack cache} of recently freed stacks for speeding up
-fiber allocation. In our experiments, we observed that directly using
-|mimalloc| outperforms having the stack cache.
+stack. Switching between fibers only involves saving the exception and the
+stack pointer of the current stack and loading the same on the target stack.
+Since OCaml does not generate pointers into the stack, the two |fiber_info|
+fields are the only ones that need to be updated when fibers are moved. We
+allocate the space for the fiber using |malloc| and |free| it when no longer
+required. We do not use a \emph{stack cache} of recently freed stacks for
+speeding up fiber allocation~\ref{Farvardin}. In our experiments, we observed
+that directly using |mimalloc| outperforms having the stack cache.
 
@@
 external calls in the system stack. Calling a C function from OCaml involves
-saving the exception pointer and the stack pointer in the |fiber_info| of the
-current fiber, and the allocation pointer in the |Caml_state|, updating |rsp|
-to the top of system stack (maintained in |Caml_state|), and calling the C
-function. The actions are reversed when returning from the external call. One
-subtlety is that for C functions that take arguments on the stack, the
-arguments must be copied to the C stack from the OCaml stack where they are
-laid out. For vast majority of external calls, we perform only a few additional
-instructions over stock OCaml in order to to set up the right stack.
-
-\begin{figure}
-  \includegraphics[width=0.69\linewidth]{figures/multicore_stack}
-	\caption{Layout of Multicore OCaml stack on \texttt{meander} example from
-	\S\ref{sec:stack}.}
-  \label{fig:mcstack}
-\end{figure}
+saving the exception and stack pointers in the current fiber, saving the
+allocation pointer in the |Caml_state|, updating |rsp| to the top of system
+stack (maintained in |Caml_state|), and calling the C function. The actions are
+reversed when returning from the external call. One subtlety is that for C
+functions that take arguments on the stack, the arguments must be copied to the
+C stack from the OCaml stack where they are laid out. For vast majority of
+external calls, we perform only a few additional instructions over stock OCaml.
 
@@
 \subsection{Effect handlers}
-
-\begin{figure}
-	\includegraphics[width=0.6\linewidth]{figures/fiber_handler}
-	\caption{Layout of stack with effect handlers}
-	\label{fig:fiber_handler}
-\end{figure}
+\label{sec:effimpl}
 
 In order to illustrate the implementation of handling effects, consider the
-example presented below:
-\begin{lstlisting}
-effect E : unit
-effect F : unit;;
-match (* comp_e *)
-  match (* comp_f *) (*p1*) perform E (*p3*)
-  with v -> v | effect F kf -> continue kf ()
-with v -> v | effect E ke -> (*p2*) continue ke ()
-\end{lstlisting}
-The code performs |E| which is handled in the outer-most handler, and is
-immediately resumed. The layout of the stack as the program executes is
-captured in Figure~\ref{fig:fiber_handler}. The arrows between rectangular
-fibers are parent pointers. At position |p1|, |rsp| is at the top of the fiber
-|f|.
-
-When an effect is performed, we invoke the effect closure |clos_heff| in the
-fiber |f| with the current fiber pointer as the continuation. Since the first
-handler does not handle effect |E|, the effect is forwarded to outer handler by
-performing the effect closure of fiber |e|. This handles |E|. We allocate a one
-word continuation object |ke| on the heap that holds the pointer to fiber |e|
-which points to the fiber |f| (position |p2|). Thus, continuations are captured
-without copying frames. Since each handler closure is evaluated until a machine
-one is found, the take to handle an effect is linear in the number of handlers.
-
-\tk{Something wasn't quite right with this: we allocate the continuation object
-at the point of the perform. Also should we describe how the first handler
-passes the effect on and adds the e onto the continuation that started with f?}
+example presented in Figure~\ref{code:effimpl}. The layout of the program state
+as the program executes is captured in Figure~\ref{fig:fiber_handler}. The code
+performs |E| which is handled in the outer-most handler, and is immediately
+resumed. The arrows between the fibers are parent pointers. At position |p1|,
+|rsp| is at the top of the fiber |f|.
+
+When the effect |E| performed, we allocate a continuation object |ke| in the
+heap that points to the current fiber |f|, set fiber |f|'s parent pointer to
+|NULL|, and evaluate the continuation closure |clos_heffect| on the parent
+fiber |e| with the effect |E| and the continuation |ke| as arguments. Since the
+first handler does not handle effect |E|, the effect is \emph{reperformed}
+(\S\ref{sec:static_semantics}) by appending the fiber |e| to the tail of
+continuation |ke|, set fiber |e|'s parent pointer to |NULL|, and evaluate the
+current continuation closure on the parent fiber |main| with |E| and |ke| as
+arguments, which handles |E| (position |p2|). Thus, continuations are captured
+without copying frames. Since every handler closure is evaluated until a
+machine one is found, the time taken to handle an effect is linear in the
+number of handlers. We observe that the handler stack is shallow in real
+programs.
 
@@
 
-For stack walking during GC, the only change necessary was to make the runtime
-aware of fibers layout, the closures for the handlers and how to follow the
-parent pointers. The stack map generation and scanning each frame for roots
-remains the same. Due to the uniformity of fiber layout, suspended fibers are
-scanned the same way as the current active fiber. Previous
-work~\cite{RetroParallel} presents the challenges and solution for the
-interaction of fibers and the concurrent mark-and-sweep GC of Multicore OCaml.
-
 The challenge with DWARF stack unwinding is to make it aware of the
@@
 turing-complete, bytecode representation, which is interpreted on demand for
-building the unwind table. Within each function, DWARF maintains a
+building the stack unwind table. Within each function, DWARF maintains a
 \emph{canonical frame address} (CFA), which is invariant in the function body,
@@
 unwind table validator tool~\cite{Bastian}. When starting to run on a new
-fiber, we insert DWARF bytecode to follow the |parent_fiber| pointer
-and dereference the |saved_sp| to get the CFA. During callback, we save the
-current system stack pointer in the DWARF block in Figure~\ref{fig:fiber} to
-identify the CFA in the C stack. DWARF unwinding for external calls is
-implemented similarly by saving the current OCaml stack pointer. With these
-changes, we get the same backtrace for the |meander.exe| program from
-\S\ref{sec:unwind}, modulo runtime system functions due to effect handlers.
+fiber, we insert DWARF bytecode to follow the |parent_fiber| pointer and
+dereference the |saved_sp| to get the CFA (|saved_sp + 8|). During callback, we
+save the current system stack pointer in the |DWARF block| in
+Figure~\ref{fig:fiber} to identify the CFA in the C stack. DWARF unwinding for
+external calls is implemented similarly by saving the current OCaml stack
+pointer. With these changes, we get the same backtrace for the \texttt{meander}
+program from \S\ref{sec:unwind}, modulo runtime system functions due to effect
+handlers.
+
+\subsection{Garbage collection}
+
+Recall that OCaml programs are written with the expectation that function calls
+return exactly once (\S\ref{sec:refine}). Consider the scenario when a
+continuation is neither |continue|d nor |discontinue|d. Since fibers are
+|malloc|ed and are |free|d when the computation being handled returns a value
+or raises an exception, not resuming continuations leaks memory. In addition,
+unresumed continuations may also leak other resource such as open file
+descriptors.
+
+We consider well-behaved Multicore OCaml programs to resume their continuations
+\emph{exactly once}. This can be enforced by installing a finaliser on every
+captured continuation that |discontinue|s the continuation during finalisation
+and ignores the result:
+\begin{lstlisting}
+Gc.finalise (fun k ->
+  try ignore (discontinue k Unwind) | _ -> ()) k
+\end{lstlisting}
+Multicore OCaml does not enforce this behaviour in order to minimise the cost
+of using effect handlers.
+
+For stack walking during GC, the only change necessary was to make the runtime
+aware of fiber layout and how to follow the parent pointers. The stack map
+generation and scanning each frame for roots remains the same. Sivaramakrishnan
+et al.~\cite{RetroParallel} present the challenges and solutions for
+integrating fibers with the concurrent mark-and-sweep GC of Multicore OCaml.
 

### 2020-11-03 retro-concurrency.tex
minor edit
Updated the microbenchmarks for comparing stock vs effects: external, callbacks, handler return, handler + raise

@@
 \subsection{Retrofitting effect handlers}
-\label{sec:req}
 
@@
 \subsection{Requirements}
-\label{sec:requirements}
+\label{sec:req}
 
@@
 
-One of our goals(\S\ref{sec:requirements}) is to make this code transparently
+One of our goals(\S\ref{sec:req}) is to make this code transparently
 asynchronous. We can define effects for performing the I/O operations  and wrap
@@
 \item Aim: |R1| not altered existing code performance too much
+\item micro-benchmarks covering 1M external calls, 1M callbacks, 1M install handler return, 1M install handler + raise; 4.10.0+stock vs 4.10.0+multicore
 \item sandmark macro benchmarks to show sequential performance: 4.10.0+stock, 4.10.0+multicore,
 4.10.0+multicore no red-zone/leaf-optimization, 4.10.0+multicore
-\item do we want to include some micro-benchmarks focusing on the FFI callstack overhead?
 \end{itemize}

### 2020-11-04 retro-concurrency.tex

@@
 
+In this section, we evaluate the performance of Multicore OCaml effect handlers
+against the performance requirements set in \S\ref{sec:req}. Multicore OCaml is
+an extension of 4.10.0 OCaml compiler with support for shared memory
+parallelism and effect handlers. These results were obtained on 2-socket,
+Intel\textregistered Xeon\textregistered Gold 5120 x86-64 server, with 28
+physical cores (14 cores on each socket), and 2 hardware threads per core. Each
+core runs at 2.20GHz and has 32 KB of L1 data cache, 32 KB of L1 instruction
+cache and 1MB of L2 cache. The cores on a socket share a 19.25 MB L3 cache. The
+server has 64GB of main memory and runs Ubuntu 18.04.
+
+\subsection{No effects benchmaks}
+
+In this section, we measure the impact of the addition of effect handlers on
+code that does not use effect handlers.
+
+\section{Related Work}
+\label{sec:related}
+
+\section{Conclusions and Future Work}
+\label{sec:conc}
+
+\bibliographystyle{ACM-Reference-Format}
+\bibliography{sample-base}
+
+\newpage
+
+\section*{Supplementary Material}
+
+\section*{Evaluation Planning}
+
 \subsection{The impact on code not using effects}
@@
 
-\section{Related Work}
-\label{sec:related}
-
-\section{Conclusions and Future Work}
-\label{sec:conc}
-
-\bibliographystyle{ACM-Reference-Format}
-\bibliography{sample-base}
-
 \end{document}

### 2020-11-05 retro-concurrency.tex
Updates

@@
 \usepackage{stmaryrd}
+\usepackage{adjustbox}
+
+\newcommand{\rot}[1]{\adjustbox{angle=90}{\textbf{#1}}}
 
@@
 \newcommand{\mycaption}[1]{\vspace{-4mm}\caption{#1}}
+\setlength{\tabcolsep}{3pt}
 
@@
 functions |caml_call_c| and |caml_call_ocaml| switch the stacks, and hence are
-shown in both the system stack and the fiber. The details of the exception
-frames have been elided in the figure. Thanks to the fiber representation,
-external calls and callbacks remain fast.
+shown in both the system stack and the fiber. Since we are reusing the fiber
+for the callback, care must be taken to save and restore the |handler_info|
+before calling and after returning from |c_to_ocaml| function, respectively.
+The details of the exception frames have been elided in the figure. Thanks to
+the fiber representation, external calls and callbacks remain fast.
 
@@
 In this section, we measure the impact of the addition of effect handlers on
-code that does not use effect handlers.
+code that does not use effect handlers. Our macro benchmark suite consists of
+54 real OCaml workloads including verification tools (|Coq|, |Cubicle|,
+|AltErgo|), parsers (|menhir|,|yojson|), storage engines (|irmin|), utilities
+(|cpdf|,|decompress|), bioinformatics (|fasta|, |knucleotide|, |revcomp2|,
+|regexredux2|), numerical analysis (|grammatrix|,|LU_decomposition|) and
+simulations (|nbody|,|game_of_life|). In addition to Stock (|stock|) and
+Multicore OCaml (|MC|), we also ran the benchmarks on Multicore OCaml with no
+red zone (|MC+RedZone0|) in the fiber (all OCaml functions will have stack
+overflow check) and a red zone size of 32 words (|MC+RedZone32|). Recall that
+the default red zone size in Multicore OCaml is 16 words.
+
+The complete results are presented in the supplementary material, which we
+summarise here. On average (geometric mean of the normalized values against
+|stock| as the baseline), the multicore variants were \textbf{less than 1\%
+slower}. The outliers (on either ends) were due to the difference in the
+allocator and the GC between stock and Multicore OCaml. The biggest impact was
+the increase in the code size for OCaml code due to stack overflow checks.
+Compared to |stock|, the code size for OCaml code 19\% (30\%) more for |MC| and
+|MC+RedZone32| (|MC+RedZone0|). The result shows that our red zone is effective
+at reducing the code size, but further work is required to bring it closer to
+|stock|.
+
+\begin{table}
+\caption{Micro benchmarks without effects. Each entry shows percentage
+	difference for Multicore OCaml over stock OCaml.}
+\vspace{-5mm}
+{\small
+\begin{tabular}{r c c c c c c c c c c}
+	& \rot{exnval} & \rot{exnraise} & \rot{extcall} & \rot{callback} & \rot{ack}
+	& \rot{evenodd} & \rot{fib} & \rot{motzkin} & \rot{sudan} & \rot{tak} \\ \hline
+	\textbf{Time} & +0.0 & -1.9 & +17 & +65  & +5.3
+								& -24 & +2.2 & +10 & +0.0 & +4.2 \\
+	\textbf{Instr} & +0.0 & +0.0 & +10 & +72 & +16
+								 & +0.0 & +24 & +16 & +14 & +17 \\ \hline
+\end{tabular}
+}
+\label{tab:micro}
+\end{table}
+
+We also present micro benchmarks results in Table~\ref{tab:micro}. Since micro
+benchmarks magnify micro-architectural optimisations, we also report the number
+of instructions executed along with time. |exnval| perform 100 million
+iterations of installing exception handlers and returning with value.
+|exnraise| is similar, but raises an exception in each iteration. |extcall| and
+|callback| perform 100 million external calls and callbacks to identity
+functions. The other micro benchmarks are highly recursive programs and were
+taken from ~\ref{Folklore}.
+
+The results show that exceptions are no more expensive in |MC| compared to
+|stock|. In the other programs, |MC| executes more instructions due to stack
+overflow checks, except |evenodd| on which the check is eliminated by the
+compiler. Despite this, observe that |MC| is 24\% faster than stock on
+|evenodd|. At these scales, we observed that padding tightloops with a few
+|nop| instructions, which changes the loop alignment, makes the code upto 15\%
+faster. Hence, the difference in running times under 15\% is not statistically
+significant. External calls and callbacks are more expensive due to additional
+bookkeeping. Callbacks are significantly more expensive, but are also less of a
+concern compared to external calls which are far more numerous. As seen
+previously, these results however do not affect macro benchmark performance.
 
@@
 \section*{Supplementary Material}
+\setcounter{figure}{0}
+
+\subsection*{No effect macro benchmarks}
+
+\begin{figure*}
+	\begin{minipage}{\linewidth}
+	\includegraphics[width=\linewidth]{sandmark-notebook/sandmark_time}
+	\subcaption{Normalized time of macro benchmarks. Baseline is Stock OCaml, whose running time in seconds in given in parenthesis.}
+	\includegraphics[width=\linewidth]{sandmark-notebook/sandmark_codesize}
+	\subcaption{Normalized OCaml code size of macro benchmarks. Baseline is Stock OCaml, whose size in bytes given in parenthesis.}
+	\end{minipage}
+	\caption{Macro benchmark performance on code that does not use effects.}
+	\label{res:macro}
+\end{figure*}
+
+Figure~\ref{res:macro} presents the performance of macro benchmarks that do not
+use effects. The aim is to quantify the impact of the addition of effect
+handlers to the language on code that does not use effect handlers. The |MC| is
+Multicore OCaml with default red zone size (16 words). |MC+RedZone0| uses red
+zone size of 0; every OCaml function has a stack overflow check. |MC+RedZone32|
+uses a red zone size of 32 words.
 
 \section*{Evaluation Planning}
-
-\subsection{The impact on code not using effects}
-
-There are several changes to the runtime to support fibers and effect handlers
-that will be executed by code not using effects: OCaml stacks now have dynamic
-checks in the function prologue and the FFI interface tracks stack switching.
-
-TODO:
-\begin{itemize}
-\item Aim: |R1| not altered existing code performance too much
-\item micro-benchmarks covering 1M external calls, 1M callbacks, 1M install handler return, 1M install handler + raise; 4.10.0+stock vs 4.10.0+multicore
-\item sandmark macro benchmarks to show sequential performance: 4.10.0+stock, 4.10.0+multicore,
-4.10.0+multicore no red-zone/leaf-optimization, 4.10.0+multicore
-\end{itemize}
 

### 2020-11-05 retro-concurrency.tex

@@
 \vspace{-5mm}
-{\small
+{
 \begin{tabular}{r c c c c c c c c c c}
 	& \rot{exnval} & \rot{exnraise} & \rot{extcall} & \rot{callback} & \rot{ack}
-	& \rot{evenodd} & \rot{fib} & \rot{motzkin} & \rot{sudan} & \rot{tak} \\ \hline
+	& \rot{fib} & \rot{motzkin} & \rot{sudan} & \rot{tak} \\ \hline
 	\textbf{Time} & +0.0 & -1.9 & +17 & +65  & +5.3
-								& -24 & +2.2 & +10 & +0.0 & +4.2 \\
+								& +2.2 & +10 & +0.0 & +4.2 \\
 	\textbf{Instr} & +0.0 & +0.0 & +10 & +72 & +16
-								 & +0.0 & +24 & +16 & +14 & +17 \\ \hline
+								 & +24 & +16 & +14 & +17 \\ \hline
 \end{tabular}
@@
 functions. The other micro benchmarks are highly recursive programs and were
-taken from ~\ref{Folklore}.
+taken from ~\ref{Folklore}. For micro benchmarks, we observed that padding
+tightloops with a few |nop| instructions, which changes the loop alignment,
+makes the code upto 15\% faster. Hence, the difference in running times under
+15\% may not be statistically significant.
 
@@
 |stock|. In the other programs, |MC| executes more instructions due to stack
-overflow checks, except |evenodd| on which the check is eliminated by the
-compiler. Despite this, observe that |MC| is 24\% faster than stock on
-|evenodd|. At these scales, we observed that padding tightloops with a few
-|nop| instructions, which changes the loop alignment, makes the code upto 15\%
-faster. Hence, the difference in running times under 15\% is not statistically
-significant. External calls and callbacks are more expensive due to additional
-bookkeeping. Callbacks are significantly more expensive, but are also less of a
-concern compared to external calls which are far more numerous. As seen
-previously, these results however do not affect macro benchmark performance.
+overflow checks. External calls and callbacks are more expensive due to
+additional bookkeeping. Callbacks are significantly more expensive in |MC|.
+Since we reuse the current fiber stack for callbacks, we need to ensure it has
+enough room for inserting additional frames, while |stock| does not need to do
+this. Callback performance is less important than external calls, which are far
+more numerous. As seen previously, these results however do not affect macro
+benchmark performance.
 

### 2020-11-06 retro-concurrency.tex
No perform benchmarks

@@
 
-\newcommand{\rot}[1]{\adjustbox{angle=90}{\textbf{#1}}}
+\newcommand{\rot}[2][75]{\adjustbox{angle=#1}{\textbf{#2}}}
 
@@
 appropriate closure to either |continue| or |discontinue| the continuation
-(position |p3|). We perform standard tail-call optimization so that resumptions
+(position |p3|). We perform standard tail-call optimisation so that resumptions
 at tail positions do not build up stack.
@@
 
-\subsection{No effects benchmaks}
+\subsection{No effects benchmarks}
 
@@
 \begin{table}
-\caption{Micro benchmarks without effects. Each entry shows percentage
+\caption{Micro benchmarks without effects. Each entry is the percentage
 	difference for Multicore OCaml over stock OCaml.}
@@
 \end{tabular}
+\vspace{-3mm}
 }
-\label{tab:micro}
+\label{tab:micro_noeffect}
 \end{table}
 
-We also present micro benchmarks results in Table~\ref{tab:micro}. Since micro
-benchmarks magnify micro-architectural optimisations, we also report the number
-of instructions executed along with time. |exnval| perform 100 million
-iterations of installing exception handlers and returning with value.
-|exnraise| is similar, but raises an exception in each iteration. |extcall| and
-|callback| perform 100 million external calls and callbacks to identity
-functions. The other micro benchmarks are highly recursive programs and were
-taken from ~\ref{Folklore}. For micro benchmarks, we observed that padding
-tightloops with a few |nop| instructions, which changes the loop alignment,
-makes the code upto 15\% faster. Hence, the difference in running times under
-15\% may not be statistically significant.
+We also present micro benchmarks results in Table~\ref{tab:micro_noeffect}.
+Since micro benchmarks magnify micro-architectural optimisations, we also
+report the number of instructions executed (obtained using |perf|) along with
+time. |exnval| perform 100 million iterations of installing exception handlers
+and returning with value. |exnraise| is similar, but raises an exception in
+each iteration. |extcall| and |callback| perform 100 million external calls and
+callbacks to identity functions. The other micro benchmarks are highly
+recursive programs and were taken from ~\ref{Folklore}. For micro benchmarks,
+we observed that padding tightloops with a few |nop| instructions, which
+changes the loop alignment, makes the code upto 15\% faster. Hence, the
+difference in running times under 15\% may not be statistically significant.
 
@@
 benchmark performance.
+
+\subsection{No perform benchmarks}
+
+\begin{table}
+\caption{Micro benchmarks with handlers but no perform. Each entry is the
+	slowdown factor ($\times$ times) over its idiomatic implementation in stock OCaml.}
+\vspace{-3mm}
+{
+\begin{tabular}{r c c c c c}
+	& \textbf{ack} & \textbf{fib} & \textbf{motzkin} & \textbf{sudan} & \textbf{tak} \\ \hline
+	\texttt{MC} 	 	& 10.57 & 15.76 & 16.96 & 12.88 & 13.64 \\
+	\texttt{monad} 	& 42.79 & 69.77 & 348.69 & 33.29 & 39.24 \\ \hline
+\end{tabular}
+}
+\label{tab:micro_noperform}
+\end{table}
+
+Next we look at the cost of setting up an effect handler, but not |perform|ing
+an effect. The aim is to quantify the cost of setting up and tearing down
+effect handlers compared to a non-tail function call. To this end, we surround
+the non-tail calls in the recursive micro benchmarks with an effect handler. We
+also include a version of the same benchmarks implemented using concurrency
+monad~\cite{} (|monad|) as a proxy for CPS versions. Recall that OCaml compiler
+does not use CPS in its IR. In the |monad| version, we use a |fork| to invoke
+the non-tail call and use an |MVar| to collect its result.
+
+The results are presented in Table~\ref{tab:micro_noperform}. They show that
+using effect handlers (concurrency monad) is 13.77$\times$ (67.09$\times$) more
+expensive than the idiomatic implementation using non-tail calls. Concurrency
+monad suffers due to the heap allocation of continuation frames (which need to
+be GCed), where as effect handlers benefit from stack allocation of the frames.
+We note that a compiler that uses CPS IR will be faster than the concurrency
+monad implementation due to optimisations to reduce the heap allocation of
+continuation frames. But Farvardin et al.~\cite{} show that CPS with
+optimisations is still slower than using the call stack.
+
+\subsection{Concurrent benchmarks}
 

### 2020-11-06 retro-concurrency.tex
More evaluation

@@
 
+Next we look at benchmarks that utilize non-local control flow. First, we
+quantify the overhead individual operations in effect handling. Consider the
+following annotated code:
+\begin{lstlisting}
+effect E : unit
+(* a *) match (* b *) perform E (* d *) with
+| v -> (* e *) v
+| effect E k -> (* c *) continue k ()
+\end{lstlisting}
+The sequence |a-b| involves allocating a new fiber and switching to it. |b-c|
+is performing the effect and handling it. |c-d| is resuming the continuation.
+|d-e| is returning from the fiber with a value and freeing the fiber. We
+measured the time taken to execute these sequences using |perf| support for
+cycle-accurate tracing on modern Intel processors. We executed 10 iterations of
+the code, ignoring 3 warm-up runs. We observed that the sequences
+|a-b|, |b-c|, |c-d| and |d-e| took 48ns, 10ns, 13ns and 19ns, respectively. For
+reference, the idle memory latency for the local NUMA node is 93.2ns.
+
 \section{Related Work}

### 2020-11-06 retro-concurrency.tex
Add minor comment

@@
 be GCed), where as effect handlers benefit from stack allocation of the frames.
+Our concurrency monad, and indeed monadic concurrency libraries such as Lwt and
+Async, also has other downsides; due to the lack of stack, exceptions,
+backtraces, and DWARF unwinding are no longer useful.
+
 We note that a compiler that uses CPS IR will be faster than the concurrency
@@
 \end{lstlisting}
-The sequence |a-b| involves allocating a new fiber and switching to it. |b-c|
-is performing the effect and handling it. |c-d| is resuming the continuation.
-|d-e| is returning from the fiber with a value and freeing the fiber. We
-measured the time taken to execute these sequences using |perf| support for
-cycle-accurate tracing on modern Intel processors. We executed 10 iterations of
-the code, ignoring 3 warm-up runs. We observed that the sequences
-|a-b|, |b-c|, |c-d| and |d-e| took 48ns, 10ns, 13ns and 19ns, respectively. For
-reference, the idle memory latency for the local NUMA node is 93.2ns.
+The sequence \texttt{a-b} involves allocating a new fiber and switching to it.
+\texttt{b-c} is performing the effect and handling it. \texttt{c-d} is resuming
+the continuation. \texttt{d-e} is returning from the fiber with a value and
+freeing the fiber. We measured the time taken to execute these sequences using
+|perf| support for cycle-accurate tracing on modern Intel processors. We
+executed 10 iterations of the code, ignoring 3 warm-up runs. We observed that
+the sequences \texttt{a-b}, \texttt{b-c}, \texttt{c-d} and \texttt{d-e} took
+48ns, 10ns, 13ns and 19ns, respectively. For reference, the idle memory latency
+for the local NUMA node is 93.2ns.
 

### 2020-11-07 retro-concurrency.tex

@@
 Next we look at the cost of setting up an effect handler, but not |perform|ing
-an effect. The aim is to quantify the cost of setting up and tearing down
+an effect. The aim is to quantify the overhead of setting up and tearing down
 effect handlers compared to a non-tail function call. To this end, we surround
@@
 |perf| support for cycle-accurate tracing on modern Intel processors. We
-executed 10 iterations of the code, ignoring 3 warm-up runs. We observed that
-the sequences \texttt{a-b}, \texttt{b-c}, \texttt{c-d} and \texttt{d-e} took
-48ns, 10ns, 13ns and 19ns, respectively. For reference, the idle memory latency
-for the local NUMA node is 93.2ns.
+executed 10 iterations of the code, with 3 warm-up runs. We observed that the
+sequences \texttt{a-b}, \texttt{b-c}, \texttt{c-d} and \texttt{d-e} took \textbf{48ns,
+10ns, 13ns and 19ns}, respectively. For reference, the idle memory latency for
+the local NUMA node is 93.2ns as measured using the Intel MLC tool~\cite{}.
+
+\paragraph{Generators} Generators allow data structures to be traversed on
+demand. Many languages including JavaScript and Python provide generators as a
+built-in primitive. Using effect handlers, given any data structure (|'a t|)
+and its iterator (|val iter: 'a t -> ('a -> unit) -> unit|), we can derive its
+generator function (|val next : unit -> 'a option|) (|MC| version). The code is
+available in the supplementary material.
+
+We evaluate the performance of traversing a complete binary tree of depth 25
+using this generator. For comparison, we also implemented a hand-written,
+selective CPSed~\cite{}, defunctionalised~\cite{DefunctionalizationAtWork}
+version (|cps|) and a concurrency monad (|monad|) version of the generator for
+the tree. Both |cps| and |monad| versions are specialised to the binary tree,
+and come with the usual caveats of not using the stack for function calls. We
+observed that the |cps| version was the fastest thanks to specialisation and
+hand optimisation. |MC| version was only 2.76$\times$ slower than |cps| while
+being a generic solution, whereas the |monad| version was 8.69$\times$ slower
+than |cps|.
 
@@
 
-\subsection{Measurements of overheads in our effects implementation}
-
-We have implemented effects within OCaml, here we present timing measurements
-for the various stages of stack creation (|a| to |b|), performing and handling
-an effect (|b| to |c|), continuation resumption (|c| to |d|) and returning from
-a computation and freeing the stack (|d| to |e|).
-
-\begin{lstlisting}
-let foo () =
- (* a *) match
- (* b *)   perform E
- (* d *) with effect E k ->
- (* c *)   continue k ()
- (* e *)
-\end{lstlisting}
-
-TODO:
-\begin{itemize}
-\item Aim: |R3| to provide some timing information as to the size of the overhead and where the overhead is.
-\item use KC's measurement stubs as presented in WASM
-\end{itemize}
-
 \subsection{Effects benchmarks}

### 2020-11-07 retro-concurrency.tex
Updates

@@
 54 real OCaml workloads including verification tools (|Coq|, |Cubicle|,
-|AltErgo|), parsers (|menhir|,|yojson|), storage engines (|irmin|), utilities
-(|cpdf|,|decompress|), bioinformatics (|fasta|, |knucleotide|, |revcomp2|,
-|regexredux2|), numerical analysis (|grammatrix|,|LU_decomposition|) and
-simulations (|nbody|,|game_of_life|). In addition to Stock (|stock|) and
+|AltErgo|), parsers (|menhir|, |yojson|), storage engines (|irmin|), utilities
+(|cpdf|, |decompress|), bioinformatics (|fasta|, |knucleotide|, |revcomp2|,
+|regexredux2|), numerical analysis (|grammatrix|, |LU_decomposition|) and
+simulations (|nbody|, |game_of_life|). In addition to Stock (|stock|) and
 Multicore OCaml (|MC|), we also ran the benchmarks on Multicore OCaml with no
@@
 
-Next we look at the cost of setting up an effect handler, but not |perform|ing
-an effect. The aim is to quantify the overhead of setting up and tearing down
-effect handlers compared to a non-tail function call. To this end, we surround
-the non-tail calls in the recursive micro benchmarks with an effect handler. We
-also include a version of the same benchmarks implemented using concurrency
-monad~\cite{} (|monad|) as a proxy for CPS versions. Recall that OCaml compiler
-does not use CPS in its IR. In the |monad| version, we use a |fork| to invoke
-the non-tail call and use an |MVar| to collect its result.
+Next, we aim to quantify the overhead of setting up and tearing down effect
+handlers compared to a non-tail function call. To this end, we surround the
+non-tail calls in the recursive micro benchmarks with an effect handler. The
+code does not |perform| effects. We also include a version of the same
+benchmarks implemented using concurrency monad~\cite{} (|monad|) as a proxy for
+CPS versions. Recall that OCaml compiler does not use CPS in its IR. In the
+|monad| version, we use a |fork| to invoke the non-tail call and use an |MVar|
+to collect its result.
 
@@
 
-Next we look at benchmarks that utilize non-local control flow. First, we
-quantify the overhead individual operations in effect handling. Consider the
-following annotated code:
+Next we look at benchmarks that utilize non-local control flow using effect
+handlers. First, we quantify the overhead individual operations in effect
+handling. Consider the following annotated code:
 \begin{lstlisting}
@@
 executed 10 iterations of the code, with 3 warm-up runs. We observed that the
-sequences \texttt{a-b}, \texttt{b-c}, \texttt{c-d} and \texttt{d-e} took \textbf{48ns,
-10ns, 13ns and 19ns}, respectively. For reference, the idle memory latency for
-the local NUMA node is 93.2ns as measured using the Intel MLC tool~\cite{}.
+sequences \texttt{a-b}, \texttt{b-c}, \texttt{c-d} and \texttt{d-e} took
+\textbf{48ns, 10ns, 13ns and 19ns}, respectively. For reference, the idle
+memory latency for the local NUMA node is 93.2ns as measured using the Intel
+MLC tool~\cite{}.
 
@@
 demand. Many languages including JavaScript and Python provide generators as a
-built-in primitive. Using effect handlers, given any data structure (|'a t|)
-and its iterator (|val iter: 'a t -> ('a -> unit) -> unit|), we can derive its
-generator function (|val next : unit -> 'a option|) (|MC| version). The code is
-available in the supplementary material.
-
-We evaluate the performance of traversing a complete binary tree of depth 25
-using this generator. For comparison, we also implemented a hand-written,
-selective CPSed~\cite{}, defunctionalised~\cite{DefunctionalizationAtWork}
-version (|cps|) and a concurrency monad (|monad|) version of the generator for
-the tree. Both |cps| and |monad| versions are specialised to the binary tree,
-and come with the usual caveats of not using the stack for function calls. We
-observed that the |cps| version was the fastest thanks to specialisation and
-hand optimisation. |MC| version was only 2.76$\times$ slower than |cps| while
-being a generic solution, whereas the |monad| version was 8.69$\times$ slower
-than |cps|.
-
-\section{Related Work}
-\label{sec:related}
-
-\section{Conclusions and Future Work}
-\label{sec:conc}
+built-in primitive. Using effect handlers (|MC|), given any data structure
+(|'a t|) and its iterator (|val iter: 'a t -> ('a -> unit) -> unit|), we can
+derive its generator function (|val next : unit -> 'a option|). The code is
+available in the supplementary material. We evaluate the performance of
+traversing a complete binary tree of depth 25 using this generator. This
+involves $2^{26}$ stack switches in total. For comparison, we also implemented
+a hand-written, selective CPSed~\cite{},
+defunctionalised~\cite{DefunctionalizationAtWork} version (|cps|) and a
+concurrency monad (|monad|) version of the generator for the tree. Both |cps|
+and |monad| versions are specialised to the binary tree, and come with the
+usual caveats of not using the stack for function calls. We observed that the
+|cps| version was the fastest thanks to specialisation and hand optimisation.
+|MC| version was only 2.76$\times$ slower than |cps| while being a generic
+solution, whereas the |monad| version was 8.69$\times$ slower than |cps|.
+
+\paragraph{Chameneos} Chameneos~\cite{} is a concurrency game aimed at
+measuring context switching and synchronization overheads. Our implementation
+uses MVars~\cite{} for synchronization. We compare effect handler (|MC|),
+concurrency monad (|monad|) and Lwt, a widely used concurrency programming
+library for OCaml (|lwt|) versions. We observed that |MC| was the fastest, and
+|monad| (|lwt|) was 1.67$\times$ (4.29$\times$) slower than |MC|.
+
+\section{Related Work} \label{sec:related}
+
+\section{Conclusions and Future Work} \label{sec:conc}
 
@@
 
-\subsection{Effects benchmarks}
-
-We now look at the impact of the overhead in a variety of benchmarks comparing
-the cost of using our effects abstraction versus alternative methods of
-implementation.
-
-TODO:
-\begin{itemize}
-\item Aim: |R3| Do we have an efficient implementation of effects and are effects competitive vs alternative implementation strategies.
-\item we can show the sequential vs stack benchmarks as used by 'Folklore to fact' \tk{slightly weak story as the sequential ones will be much faster for these very simple benchmarks and effects don't really add anything to their implementation}
-\item general generator examples (using examples from effects-example)
-\item chamenos vs chamenos effects vs lwt (using examples from effects-example)
-\end{itemize}
-
 \subsection{Webserver benchmarks}

### 2020-11-09 retro-concurrency.tex
Added web server results

@@
 
+\paragraph{Web server} As one of our primary goals of adding effect handlers to
+OCaml is to support direct-style concurrency, we implement a full-fledged
+HTTP/1.1 web server using effect handlers (|MC|). We use effect handlers to
+implement a user-level threading library, which spawns a thread per request. We
+use \texttt{httpaf} library~\cite{} for HTTP handling, and
+\texttt{libev}~\cite{} for the eventloop. We compare our implementation against
+an Lwt version (|lwt|) which also uses \texttt{httpaf} and \texttt{libev}.
+Unlike using effect handlers, Lwt version is written in monadic style, and does
+not have a notion of thread per request. For comparison, we also include a Go
+version (|go|) that uses \texttt{net/http}~\cite{} package.
+
+\begin{figure}
+	\begin{minipage}{0.49\linewidth}
+		\centering
+		\includegraphics[width=\linewidth]{http-benchmarks/throughput.pdf}
+		\subcaption{Throughput}
+		\label{grf:throughput}
+	\end{minipage}
+	\begin{minipage}{0.49\linewidth}
+		\includegraphics[width=\linewidth]{http-benchmarks/latency.pdf}
+		\subcaption{Tail latency}
+		\label{grf:latency}
+	\end{minipage}
+	\mycaption{Web server performance.}
+	\vspace{-2mm}
+\end{figure}
+
+The client workload was generated with \texttt{wrk2}~\cite{}. We maintain 1k
+open connections and perform requests for a static web page at different rates,
+and record the request service rate and latency. The throughput and tail
+latency graphs are given in Figures~\ref{grf:throughput} and~\ref{grf:latency}.
+In all the versions, the throughput plateaus at around 30k requests per second.
+We measure the tail latencies at 2/3rd of this rate (at 20k requests per
+second), as would be the case in real deployments. We observe that both the
+OCaml versions remain competitive with |go|, and |MC| performs best in terms of
+tail latency.
+
+Using effect handlers in the web server also helps with debugging the system as
+a whole. Multicore OCaml supports getting the backtrace of a continuation,
+which is a minor extension to getting the current backtrace. We can use this
+feature to dump the backtrace of all the user-level threads in the web server.
+Such a feature is provided by Go~\cite{}, but not possible with OCaml
+concurrency libraries such as Async and Lwt which lack the notion of a thread.
+
 \section{Related Work} \label{sec:related}
@@
 
-\section*{Evaluation Planning}
-
-\subsection{Webserver benchmarks}
-
-TODO:
-\begin{itemize}
-\item Aim: |R3| Do we have an efficient implementation of effects.
-\item effects are a very nice way to program non-blocking IO for webservers
-\item show that using effects doesn't impose undue overhead and that we are competitive with LWT, Async, Go with the wrk web benchmark
-\end{itemize}
-
-\subsection{Using fibers for task based parallelism}
-
-TODO:
-\begin{itemize}
-\item Aim: |R3| Do we have an efficient implementation of effects.
-\item minor alteration to domainslib to launch tasks in their own stack
-\item effects are the right way to do some task base parallelism. Use the nqueens example and show the scaling we can achieve in parallel
-\item show that using fibers doesn't have large detrimental impact to existing parallel applications: Take sandmark: minilight, nbody, LU decomposition
-\end{itemize}
-
 \end{document}

### 2020-11-09 http-benchmarks/run_benchmarks.sh

@@ -1,0 +1,22 @@
+set -x
+
+export GOMAXPROCS=1
+export LD_PRELOAD=/usr/local/lib/libmimalloc.so
+
+mkdir -p output
+
+for cmd in "httpserv_go" "httpserv_lwt" "httpserv_effects"; do
+  for rps in 2500 5000 10000 15000 20000 25000 30000 35000 40000 45000 50000 55000 60000; do
+    for cons in 1000 5000 10000; do
+			if [ ! -s output/run-$cmd-$rps-$cons.txt ]
+			then
+				./$cmd &
+				running_pid=$!
+				sleep 2;
+				./wrk -t 24 -d60s -L -s json.lua -R $rps -c $cons http://localhost:8080 > output/run-$cmd-$rps-$cons.txt;
+				kill ${running_pid};
+				sleep 1;
+			fi
+    done
+  done
+done

### 2020-11-09 retro-concurrency.tex
Updates

@@
 What makes effect handlers quite powerful is that they provide a structured
-form of delimited continuations~\cite{}, which enables advanced non-local
-control-flow mechanisms such as resumable exceptions, coroutines, generators
-and asynchronous I/O to be composably expressed. As an example, consider the
-task of handling division-by-zero case in a calculator application. Different
-calculators handle division-by-zero differently. In the iOS 13 Calculator App
-(|iOS|), division-by-zero returns |Error|. In the Google search calculator App
-(|gsearch|) |0/0| returns to |Error|, and for any other |n| not equal to |0|,
-|n/0| returns |Infinity| with the same sign as |n|. How can we capture these
-different behaviours without changing the evaluator?
-
-Common Lisp Conditions~\cite{Conditions} are an elegant way to solve this
+form of delimited continuations~\cite{Danvy90}, which enables advanced
+non-local control-flow mechanisms such as resumable exceptions, coroutines,
+generators and asynchronous I/O to be composably expressed. As an example,
+consider the task of handling division-by-zero case in a calculator
+application. Different calculators handle division-by-zero differently. In the
+iOS 13 Calculator App (|iOS|), division-by-zero returns |Error|. In the Google
+search calculator App (|gsearch|) |0/0| returns to |Error|, and for any other
+|n| not equal to |0|, |n/0| returns |Infinity| with the same sign as |n|. How
+can we capture these different behaviours without changing the evaluator?
+
+Common Lisp Conditions~\cite{Steele90} are an elegant way to solve this
 problem. Unlike exceptions where the library decides whether an error is fatal
@@
 as web servers to be expressed naturally in \emph{direct-style} as opposed to
-using callbacks and (2) efficient data and task parallel programs in the vein
-of Cilk~\cite{}, OpenMP~\cite{} and Intel TBB~\cite{}.
+using callbacks and (2) efficient nested parallel programs in the vein of
+Cilk~\cite{Cilk}, OpenMP~\cite{}, Habenero Java~\cite{Imam14}, X10~\cite{},
+Intel TBB~\cite{IntelTBB}, parallel Haskell variants~\cite{} and parallel ML
+variants~\cite{}.
 
@@
 \bibliographystyle{ACM-Reference-Format}
-\bibliography{sample-base}
+\bibliography{retro-concurrency}
 

### 2020-11-10 retro-concurrency.tex
updates

@@
 There are many research languages and libraries built around effect
-handlers~\cite{Koka,Links,Pyro,Frank,Eff}. Unlike these efforts, our goal is to
+handlers~\cite{Leijen14,Hillerstrom20,Pyro,Frank,Eff}. Unlike these efforts, our goal is to
 retrofit effect handlers onto the OCaml programming language, which has been in
 continuous use for the past 25 years in large codebases including verification
-tools~\cite{FStar,Coq}, mission critical software systems~\cite{astree} and
-latency sensitive networked applications~\cite{JS,Docker,MirageOS}. OCaml is
+tools~\cite{everest,Coq}, mission critical software systems~\cite{astree} and
+latency sensitive networked applications~\cite{Madhavapeddy13}. OCaml is
 particularly favoured for its competitive yet predictable performance, with
@@
 using callbacks and (2) efficient nested parallel programs in the vein of
-Cilk~\cite{Cilk}, OpenMP~\cite{}, Habenero Java~\cite{Imam14}, X10~\cite{},
-Intel TBB~\cite{IntelTBB}, parallel Haskell variants~\cite{} and parallel ML
-variants~\cite{}.
+Cilk~\cite{Cilk}, OpenMP~\cite{OpenMP}, X10~\cite{Charles05}, Intel
+TBB~\cite{IntelTBB}, and variants of Parallel ML~\cite{MaPLe, Fluet10,
+Sivaramakrishnan14}.
 
@@
 is to use continuation-passing style (CPS) in the compiler's intermediate
-representation (IR)~\cite{Koka}. If the compiler were to use CPS, then reifying
+representation (IR)~\cite{Leijen14}. If the compiler were to use CPS, then reifying
 continuations in the source language is relatively straight-forward. But OCaml
@@
 OCaml applications due to increased memory allocations as continuation closures
-get allocated on the heap~\cite{Folklore}. Moreover, with CPS, stack unwinding
-is meaningless and hence, we would lose compatibility with tools that inspect
-the program stack. Hence, we choose not to use CPS translation and represent
-continuations as call stacks.
+get allocated on the heap~\cite{Farvardin20}. Moreover, with CPS, stack
+unwinding is meaningless and hence, we would lose compatibility with tools that
+inspect the program stack. Hence, we choose not to use CPS translation and
+represent continuations as call stacks.
 
@@
 
-Our work is also timely. WebAssembly~\ref{} community group is actively
+Our work is also timely. WebAssembly~\cite{Hass17} community group is actively
 considering effect handlers as one of the mechanism for supporting concurrency,
-asynchronous I/O and generators~\ref{WasmPropoal}. Project Loom~\cite{} is an
+asynchronous I/O and generators~\ref{WasmPropoal}. Project Loom~\cite{loom} is an
 OpenJDK project which adds virtual threads and delimited continuations to Java.
-Swift concurrency roadmap~\cite{} includes direct-style asynchronous
+Swift concurrency roadmap~\cite{swift} includes direct-style asynchronous
 programming and structured concurrency as milestones. We believe that our
@@
     inspect the stack using DWARF unwind tables. We have validated our DWARF
-    unwind tables using an automated validator tool~\cite{}.
+    unwind tables using an automated validator tool~\cite{Bastian19}.
   \item a formal operational semantics for the effect handler implementation in
@@
 It is well-known that one-shot continuations can be implemented
-efficiently~\cite{Bruggeman96RepresentingControl}.
+efficiently~\cite{Bruggeman96}.
 
@@
 captured continuation. Managing C frames as part of the continuation is a
-complex endeavour~\cite{Daan17Aplas}, and we find that the complexity budget
+complex endeavour~\cite{Leijen17}, and we find that the complexity budget
 outweighs the relatively fewer mechanisms enabled by this addition in our
@@
 semantics is simpler than languages that ensure effect
-safety~\cite{Links,Koka,Effekt,Frank,HeliumBiernacki}. This is important for
+safety~\cite{Hillerstrom20,Leijen14,Effekt,Frank,Biernacki19}. This is important for
 backwards compatibility as our goal is to retrofit effect handlers to a
@@
 \end{lstlisting}
-We introduce an extensible variant type~\cite{} |'a eff| of effect values,
-which when performed using the |perform| primitive returns a |'a| value.
-Constructors for the value of type |'a eff| are declared using the effect
-declarations. For example, the declaration |effect E : string -> int| is
-syntactic sugar for adding a new constructor to the variant type
+We introduce an extensible variant type~\cite{ExtVariants} |'a eff| of effect
+values, which when performed using the |perform| primitive returns a |'a|
+value. Constructors for the value of type |'a eff| are declared using the
+effect declarations. For example, the declaration |effect E : string -> int|
+is syntactic sugar for adding a new constructor to the variant type
 |type _ eff += E : string -> int eff|. We introduce the type
@@
 The operational semantics is an extension of a CEK machine
-semantics~\cite{Felleisen} for effect handlers, following the abstract machine
-semantics of Hillerstrom et al.~\cite{Hillerstrom2020}. The crucial difference
+semantics~\cite{Felleisen86} for effect handlers, following the abstract machine
+semantics of Hillerstrom et al.~\cite{Hillerstrom20}. The crucial difference
 from Hillerstrom et al. is that our stacks are composed of alternating sequence
@@
 the handler wraps around captured continuation. This gives Multicore OCaml
-effect handlers deep handler semantics~\cite{}. \textsc{EffFwd} forwards the
+effect handlers deep handler semantics~\cite{Hillerstrom20}. \textsc{EffFwd} forwards the
 effect to the outer fiber, and extends the captured continuation |k''| in the
@@
 in Multicore OCaml for performance backwards compatibility. This differs from
-other research languages with effect handlers~\cite{Links,Frank,Eff}, which
+other research languages with effect handlers~\cite{Hillerstrom20,Frank,Eff}, which
 implement exceptions using effects (by ignoring the continuation argument in
@@
     \includegraphics[scale=0.42]{figures/fiber_handler}
-		\subcaption{Program state for Figure~\ref{code:effimpl}}
+		\subcaption{Program state for code in~\ref{code:effimpl}}
     \label{fig:fiber_handler}
@@
 are constructed in order to appreciate our solution. We refer the interested
-reader to Bastian et al.~\cite{} for a good overview of DWARF stack unwinding.
+reader to Bastian et al.~\cite{Bastian19} for a good overview of DWARF stack
+unwinding.
 
@@
 DWARF bytecode at these three places, which we have validated with the DWARF
-unwind table validator tool~\cite{Bastian}. When starting to run on a new
+unwind table validator tool~\cite{Bastian19}. When starting to run on a new
 fiber, we insert DWARF bytecode to follow the |parent_fiber| pointer and
@@
 generation and scanning each frame for roots remains the same. Sivaramakrishnan
-et al.~\cite{RetroParallel} present the challenges and solutions for
+et al.~\cite{Sivaramakrishnan20} present the challenges and solutions for
 integrating fibers with the concurrent mark-and-sweep GC of Multicore OCaml.
@@
 code does not |perform| effects. We also include a version of the same
-benchmarks implemented using concurrency monad~\cite{} (|monad|) as a proxy for
-CPS versions. Recall that OCaml compiler does not use CPS in its IR. In the
-|monad| version, we use a |fork| to invoke the non-tail call and use an |MVar|
-to collect its result.
+benchmarks implemented using concurrency monad~\cite{Claessen99} (|monad|) as a
+proxy for CPS versions. Recall that OCaml compiler does not use CPS in its IR.
+In the |monad| version, we use a |fork| to invoke the non-tail call and use an
+|MVar| to collect its result.
 
@@
 be GCed), where as effect handlers benefit from stack allocation of the frames.
-Our concurrency monad, and indeed monadic concurrency libraries such as Lwt and
-Async, also has other downsides; due to the lack of stack, exceptions,
-backtraces, and DWARF unwinding are no longer useful.
+Our concurrency monad, and indeed monadic concurrency libraries such as
+Lwt~\cite{lwt} and Async~\cite{async}, also have other downsides; due to the
+lack of stack, exceptions, backtraces, and DWARF unwinding are no longer
+useful.
 
@@
 monad implementation due to optimisations to reduce the heap allocation of
-continuation frames. But Farvardin et al.~\cite{} show that CPS with
+continuation frames. But Farvardin et al.~\cite{Farvardin20} show that CPS with
 optimisations is still slower than using the call stack.
@@
 memory latency for the local NUMA node is 93.2ns as measured using the Intel
-MLC tool~\cite{}.
+MLC tool~\cite{mlc}.
 
@@
 involves $2^{26}$ stack switches in total. For comparison, we also implemented
-a hand-written, selective CPSed~\cite{},
-defunctionalised~\cite{DefunctionalizationAtWork} version (|cps|) and a
+a hand-written, selective CPSed~\cite{Nielson01},
+defunctionalised~\cite{Danvy01} version (|cps|) and a
 concurrency monad (|monad|) version of the generator for the tree. Both |cps|
@@
 
-\paragraph{Chameneos} Chameneos~\cite{} is a concurrency game aimed at
+\paragraph{Chameneos} Chameneos~\cite{Chameneos} is a concurrency game aimed at
 measuring context switching and synchronization overheads. Our implementation
-uses MVars~\cite{} for synchronization. We compare effect handler (|MC|),
-concurrency monad (|monad|) and Lwt, a widely used concurrency programming
-library for OCaml (|lwt|) versions. We observed that |MC| was the fastest, and
-|monad| (|lwt|) was 1.67$\times$ (4.29$\times$) slower than |MC|.
+uses |MVars| for synchronization. We compare effect handler (|MC|), concurrency
+monad (|monad|) and Lwt, a widely used concurrency programming library for
+OCaml (|lwt|) versions. We observed that |MC| was the fastest, and |monad|
+(|lwt|) was 1.67$\times$ (4.29$\times$) slower than |MC|.
 
@@
 implement a user-level threading library, which spawns a thread per request. We
-use \texttt{httpaf} library~\cite{} for HTTP handling, and
-\texttt{libev}~\cite{} for the eventloop. We compare our implementation against
-an Lwt version (|lwt|) which also uses \texttt{httpaf} and \texttt{libev}.
-Unlike using effect handlers, Lwt version is written in monadic style, and does
-not have a notion of thread per request. For comparison, we also include a Go
-version (|go|) that uses \texttt{net/http}~\cite{} package.
+use \texttt{httpaf} library~\cite{httpaf} for HTTP handling, and
+\texttt{libev}~\cite{libev} for the eventloop. We compare our implementation
+against an Lwt version (|lwt|) which also uses \texttt{httpaf} and
+\texttt{libev}. Unlike using effect handlers, Lwt version is written in monadic
+style, and does not have a notion of thread per request. For comparison, we
+also include a Go version (|go|) that uses \texttt{net/http}~\cite{nethttp}
+package.
 
@@
 
-The client workload was generated with \texttt{wrk2}~\cite{}. We maintain 1k
-open connections and perform requests for a static web page at different rates,
-and record the request service rate and latency. The throughput and tail
+The client workload was generated with \texttt{wrk2}~\cite{wrk2}. We maintain
+1k open connections and perform requests for a static web page at different
+rates, and record the request service rate and latency. The throughput and tail
 latency graphs are given in Figures~\ref{grf:throughput} and~\ref{grf:latency}.
 In all the versions, the throughput plateaus at around 30k requests per second.
-We measure the tail latencies at 2/3rd of this rate (at 20k requests per
-second), as would be the case in real deployments. We observe that both the
-OCaml versions remain competitive with |go|, and |MC| performs best in terms of
-tail latency.
+We measure the tail latencies at 2/3rd of this rate (20k requests per second)
+to simulate optimal load. We observe that both of the OCaml versions remain
+competitive with |go|, and |MC| performs best in terms of tail latency.
 
@@
 feature to dump the backtrace of all the user-level threads in the web server.
-Such a feature is provided by Go~\cite{}, but not possible with OCaml
+Such a feature is provided by Go~\cite{gopprof}, but not possible with OCaml
 concurrency libraries such as Async and Lwt which lack the notion of a thread.
@@
 \section{Conclusions and Future Work} \label{sec:conc}
+
+\clearpage
 
@@
 
-\newpage
+\clearpage
 

### 2020-11-10 retro-concurrency.tex
Added related work

@@
 \end{minipage}
-\vspace{2mm}
 \mycaption{Layout of Multicore OCaml effect handlers.}
@@
 speeding up fiber allocation~\ref{Farvardin}. In our experiments, we observed
-that directly using |mimalloc| outperforms having the stack cache.
+that directly using |mimalloc|~\cite{Leijen19} outperforms having the stack
+cache.
 
@@
 	\mycaption{Web server performance.}
-	\vspace{-2mm}
+	\vspace{-4mm}
 \end{figure}
@@
 
-\section{Related Work} \label{sec:related}
-
-\section{Conclusions and Future Work} \label{sec:conc}
+\section{Related Work}
+\label{sec:related}
+
+There are several strategies for implementing effect handlers. Eff~\cite{Eff},
+Helium~\cite{Biernacki20}, Frank~\cite{Frank}, Links server
+backend~\cite{Hillerstrom20} use an interpreter similar to our operational
+semantics to implement effect handlers. Effekt~\cite{Effekt}, Links JavaScript
+backend~\cite{Hillerstrom20} and Koka~\cite{Leijen17} use type-directed
+selective CPS translation. These language are equipped with an effect system,
+which allows compiling pure code in direct-style and effectful code in
+continuation passing style. Leijen et al.~\cite{Leijen14} implements effect
+handlers as a library in C using stack copying. Since C allows pointers into
+the stack, care is taken to ensure that when continuations are resumed, the
+constituent frames are restored to the same memory addresses as at the time of
+capture. Kiselyov et al.~\cite{Kiselyov18} use an implementation of
+multi-prompt delimited continuations as an OCaml library~\cite{Kiselyov12} to
+embed Eff language in OCaml. Indeed, Forester et al.~\cite{Forster19} showed
+that in an untyped setting, effect handlers, monadic reflection and delimited
+control can macro-express each other.
+
+Multicore OCaml effect handlers are similar to~\cite{Leijen14, Kiselyov12}, but
+are more efficient; since we only allow our continuations to be resumed at-most
+once, there is no need for copying frames. Bruggeman et al.~\cite{Bruggeman96}
+show how to implement one-shot continuations efficiently using segmented stacks
+in Scheme. Farvardin et al.~\cite{Farvardin20} perform a comprehensive
+evaluation of various implementation strategies for continuations on modern
+hardware. Multicore OCaml stacks do not neatly fit the description of one of
+these implementation strategies. They best described as using the resize
+strategy from Farvardin et al. for each of the fibers, which are in turn linked
+to represent the current stack and the captured continuations.
+
+Kawahara et al.~\cite{Kawahara20} implement one-shot effect handlers using
+coroutines (a non-local control flow mechanism) as a macro-expressible
+translation, and present an embedding in Lua and Ruby. Lua provides asymmetric
+coroutines~\cite{Lua} where each coroutine uses its own stack similar to how
+each handled computation runs in its own fiber in Multicore OCaml. Similar to
+Multicore OCaml, Lua does not allow coroutines to include C frames. It is
+useful to note that Lua is an interpreted language unlike OCaml and DWARF
+unwind tables are not a concern in Lua.
+
+The interaction on non-local control flow and resources have been studied
+extensively. Scheme uses |dynamic-wind| to ensure de-allocation (and
+re-allocation) of resources when the non-local control leaves and enters back
+into a context. |dynamic-wind| is known to be complex and not quite what one
+wants for resource clean up~\cite{Kiselyov,Sitaram03}. The main challenge is distinguishing
+non-returning exits from ones that would, and clean up to occur only on
+non-returning exit. In Multicore OCaml, we assume that continuations are used
+linearly, and hence, no clean up occurs when effects are performed but only when
+an exception is thrown, perhaps from the use of |discontinue|. This obviates
+the need for |dynamic-wind|. Leijen~\cite{Leijen18} explicitly extends effect
+handlers with |finally| clause. Instead, Multicore OCaml |discontinue|
+primitive builds on extant defensive coding practices against exceptions to
+clean up resource.
+
+Languages with effect handlers typically allow continuations to be resumed more
+than once to enable use cases such as backtracking search and probabilistic
+programming~\cite{Pyro}. Multicore OCaml has an experimental feature for
+explicitly cloning continuations, which makes a copy of each of its constituent
+fibers. Multi-shot continuations are difficult to reason about in the presence
+of linear resources such as file descriptors. Recall that OCaml programs are
+written with the expectation that function calls return \emph{exactly-once}
+either normally or exceptionally. With multi-shot continuations, they may
+return more than once, leading to errors such as double-freeing malloc'ed
+memory. Since our goal is to retrofit effect handlers onto OCaml, we relegate
+continuation cloning to the OCaml |Obj| module, which contains unsafe features.
+
+\section{Conclusions}
+\label{sec:conc}
+
+We presented the motivation, design and implementation of effect handlers
+specifically aimed at retrofitting them onto OCaml with tool compatibility,
+negligible overheads for existing OCaml programs and good performance for those
+that use effect handlers. Our evaluation shows that this is indeed the case.
 

### 2020-11-11 retro-concurrency.tex
Minor update
Updates

@@
       copy ic oc
-  with _ -> (close_in ic; close_out oc)
-\end{lstlisting}
-
-The function |input_line| raises |End_of_file| exception on reaching the end of input.
-Although not apparent, the code is written in a defensive style to handle other
-exceptional cases such as the channels being closed externally. Both
-|input_line| and |output_string| raise |Sys_error| exception if the channel is
-closed. This function cleanly closes the channels on exceptional cases;
-|close_*| functions do nothing if the channel is already closed.
+  with End_of_file -> close_in ic; close_out oc
+	   | e -> close_in ic; close_out oc; raise e
+\end{lstlisting}
+The function |input_line| raises |End_of_file| exception on reaching the end of
+input, which is handled by the exception handler which closes the channels. The
+|close_*| functions do nothing if the channel is already closed. The code is
+written in a defensive style to handle other exceptional cases such as the
+channels being closed externally. Both |input_line| and |output_string| raise
+|Sys_error| exception if the channel is closed. In this case, the channels are
+closed and the exception is reraised to communicate the exceptional behaviour
+to the caller.
 
@@
 stack are empty. The top-level reductions
-  \begin{mathpar}
+\begin{smathpar}
 	\begin{array}{cc}
@@
 	\end{array}
-  \end{mathpar}
+\end{smathpar}
 can be performed by either by taking a C step $\cstep$ or an OCaml step
@@
 \end{minipage}
-\mycaption{Layout of Multicore OCaml effect handlers.}
+\vspace{-2mm}
+\caption{Layout of Multicore OCaml effect handlers.}
+\vspace{-3mm}
 \end{figure*}
@@
 callbacks to identity functions. The other micro benchmarks are highly
-recursive programs and were taken from ~\ref{Folklore}. For micro benchmarks,
+recursive programs and were taken from ~\cite{Farvardin20}. For micro benchmarks,
 we observed that padding tightloops with a few |nop| instructions, which
@@
 \label{tab:micro_noperform}
+	\vspace{-3mm}
 \end{table}
@@
 
-\paragraph{Generators} Generators allow data structures to be traversed on
+\textbf{\textit{Generators.}} Generators allow data structures to be traversed on
 demand. Many languages including JavaScript and Python provide generators as a
@@
 
-\paragraph{Chameneos} Chameneos~\cite{Chameneos} is a concurrency game aimed at
+\textbf{\textit{Chameneos.}} Chameneos~\cite{Chameneos} is a concurrency game aimed at
 measuring context switching and synchronization overheads. Our implementation
@@
 
-\paragraph{Web server} As one of our primary goals of adding effect handlers to
+\textbf{\textit{Web server.}} As one of our primary goals of adding effect handlers to
 OCaml is to support direct-style concurrency, we implement a full-fledged
@@
 	\end{minipage}
-	\mycaption{Web server performance.}
+	\vspace{-3mm}
+	\caption{Web server performance.}
 	\vspace{-4mm}
@@
 which allows compiling pure code in direct-style and effectful code in
-continuation passing style. Leijen et al.~\cite{Leijen14} implements effect
+continuation passing style. Leijen et al.~\cite{Leijen14} implement effect
 handlers as a library in C using stack copying. Since C allows pointers into
@@
 Multicore OCaml, Lua does not allow coroutines to include C frames. It is
-useful to note that Lua is an interpreted language unlike OCaml and DWARF
+useful to note that Lua is an interpreted language, unlike OCaml, and DWARF
 unwind tables are not a concern in Lua.
@@
 re-allocation) of resources when the non-local control leaves and enters back
-into a context. |dynamic-wind| is known to be complex and not quite what one
-wants for resource clean up~\cite{Kiselyov,Sitaram03}. The main challenge is distinguishing
-non-returning exits from ones that would, and clean up to occur only on
+into a context. |dynamic-wind| is considered to be complex and not quite what
+one wants for resource clean up~\cite{Kiselyov,Sitaram03}. The main challenge
+is distinguishing returning exits from a computation closing over resources
+from exits that would not return. Resource clean up needs to occur only during
 non-returning exit. In Multicore OCaml, we assume that continuations are used
-linearly, and hence, no clean up occurs when effects are performed but only when
-an exception is thrown, perhaps from the use of |discontinue|. This obviates
-the need for |dynamic-wind|. Leijen~\cite{Leijen18} explicitly extends effect
-handlers with |finally| clause. Instead, Multicore OCaml |discontinue|
-primitive builds on extant defensive coding practices against exceptions to
-clean up resource.
+linearly, and hence, no clean up occurs when effects are performed but only
+when an exception is thrown, perhaps from the use of |discontinue|. This
+obviates the need for |dynamic-wind|.
+
+OCaml does not have a |try/finally| construct available in many programming
+languages. In OCaml, alternative standard libraries such as Base~\cite{} and
+Core~\cite{} provide mechanisms analogous to Common Lisp |unwind-protect|,
+which are in turn implemented using OCaml exceptions. Leijen~\cite{Leijen18}
+explicitly extends effect handlers with |finally| clause in Koka. Instead,
+Multicore OCaml |discontinue| primitive builds on extant defensive coding
+practices against exceptions to clean up resources. This design choice is
+crucial to ensure backwards compatibility of legacy systems code with non-local
+control flow introduced by effect handlers. Sivaramakrishnan et
+al.~\cite{TFP17} describe the interaction of effect handlers and asynchronous
+exceptions. This is orthogonal to the contributions of this paper which
+focusses on compiler and runtime system support for implementing effect
+handlers. Ahman et al.~\cite{Ahman20} present a effect handler core calculus
+with asynchrony which might help lay the formal foundations for reconciling
+asynchrony in systems and effect handlers.
 
@@
 either normally or exceptionally. With multi-shot continuations, they may
-return more than once, leading to errors such as double-freeing malloc'ed
-memory. Since our goal is to retrofit effect handlers onto OCaml, we relegate
-continuation cloning to the OCaml |Obj| module, which contains unsafe features.
+return \emph{more than once}, leading to errors such as double-freeing
+malloc'ed memory. Since our goal is to retrofit effect handlers onto OCaml, we
+relegate continuation cloning to the OCaml |Obj| module, which contains unsafe
+features.
 

### 2020-11-11 retro-concurrency.tex
Updates

@@
 
-\newcommand{\rot}[2][75]{\adjustbox{angle=#1}{\textbf{#2}}}
+\newcommand{\rot}[2][70]{\adjustbox{angle=#1}{\textbf{#2}}}
 
@@
 considering effect handlers as one of the mechanism for supporting concurrency,
-asynchronous I/O and generators~\ref{WasmPropoal}. Project Loom~\cite{loom} is an
+asynchronous I/O and generators~\cite{WasmPropoal}. Project Loom~\cite{loom} is an
 OpenJDK project which adds virtual threads and delimited continuations to Java.
@@
 		impact on code that does not use effect handlers, and serves as an
-    efficient foundation for scalable concurrent and parallel programming.
+		efficient foundation for scalable concurrent programming.
 \end{itemize}
@@
 The main challenge in implementing effect handlers in Multicore OCaml is
-managing the program stack. To this end, in this section, we provide an
-overview of the program stack and related mechanisms in stock OCaml.
+managing the program stack. In this section, we provide an overview of the
+program stack and related mechanisms in stock OCaml.
 
@@
 callee-saved registers in OCaml. On the |x86-64| backend, the C callee-saved
-registers, |r15| holds the \emph{allocation pointer} into the minor heap used
-for bump pointer allocation, and |r14| holds the |Caml_state|, a table of
-global variables used by the runtime. This makes external calls extremely fast
-in OCaml. If the external function does not allocate in the OCaml heap, then
-the external functions can be called directly, and no bookkeeping is necessary.
-For external functions which allocate in the OCaml heap, the cached allocation
-pointer is saved to a field in the |Caml_state| before the external call, and
-loaded back on return. Similarly, callbacks into OCaml are also cheap.
-Callbacks into OCaml involves loading the arguments in the right registers and
-calling the OCaml function. OCaml callbacks are relatively common as the
-garbage collector (GC) finalisation functions are executed as OCaml callbacks.
+registers  |r15| holds the \emph{allocation pointer} into the minor heap used
+for bump pointer allocation, and |r14| holds a reference to the |Caml_state|
+table of global variables used by the runtime. This makes external calls
+extremely fast in OCaml. If the external function does not allocate in the
+OCaml heap, then the external functions can be called directly, and no
+bookkeeping is necessary. For external functions which allocate in the OCaml
+heap, the cached allocation pointer is saved to a field in the |Caml_state|
+before the external call, and loaded back on return. Similarly, callbacks into
+OCaml are also cheap. Callbacks into OCaml involves loading the arguments in
+the right registers and calling the OCaml function. OCaml callbacks are
+relatively common as the garbage collector (GC) finalisation functions are
+executed as OCaml callbacks.
 
@@
   \end{minipage}
-  \mycaption{Syntax of expressions and configurations}
+	\vspace{-3mm}
+  \caption{Syntax of expressions and configurations}
   \label{sem:syntax}
+	\vspace{-3mm}
 \end{figure*}
-
-We present the operational semantics for a core language of effect handlers in
-Multicore OCaml. The operational semantics intends to faithfully capture the
-semantics of Multicore OCaml effect handlers, and to inform an implementation.
-
-\subsubsection{Syntax}
-
-Figure~\ref{sem:syntax} presents the syntax of expressions and configurations.
-Our expressions consists of integer constants ($n$), variables ($x$),
-abstraction ($\lam{x}{x}$), application ($e~e$), arithmetic expressions ($e
-\odot e$) where $\odot$ ranges over $\{+,-,*,/\}$, raising exceptions
-($\throw{l}{e}$), performing effects ($\perform{l}{e}$), and handling effects
-($\handle{e}{h}$). Abstractions come in two forms: OCaml abstractions
-($\lambda^o$) and C abstractions ($\lambda^c$). The handler consists of a
-return case ($\caseval{x}{e}$), zero or more exception cases
-($\caseexn{l}{x}{e}$) with label $l$, parameter $x$ and body $e$, and zero or
-more effect cases ($\caseeff{l}{x}{k}{e}$) with label $l$, parameter $x$,
-continuation $k$ and body $e$.
-
-The operational semantics is an extension of a CEK machine
-semantics~\cite{Felleisen86} for effect handlers, following the abstract machine
-semantics of Hillerstrom et al.~\cite{Hillerstrom20}. The crucial difference
-from Hillerstrom et al. is that our stacks are composed of alternating sequence
-of OCaml and C stack segments. The program state is captured as configuration
-$\config \coloneqq \configa{\tau}{\env}{\stack}$ with the current term $\tau$
-under evaluation, its environment $\env$ and the current stack $\stack$. The
-term is either an expression $e$ or a value $v$. The value are integer
-constants $n$, continuations $k$, closure $\clos{x}{e}{\env}$, effect being
-performed $\effval{l}{k}$ and exception being raised $\exnval{l}$. The
-environment is a map from variables to values.
-
-The stack $\stack$ is either a C stack ($\cstack$) or an OCaml stack
-($\ostack$). Every C stack $\cstacka{\fl}{\ostack}$ consists of a list of
-frames $\fl$, and the OCaml stack $\ostack$ under it. Every OCaml stack is
-either empty $\ostackemp$ or non-empty $\ostacka{k}{\cstack}$ with the
-C stack $\cstack$ under it. Thus, the program stack is an alternating sequence
-of C and OCaml stacks terminating with an empty OCaml stack $\ostackemp$. The
-frame list $\fl$ is composed of individual frames $r$, which is one of an
-argument frame $\farg{e}{\env}$ with the expression $e$ at the argument
-position of an application with its environment $\env$, a function frame
-$\ffun{v}$ with the value at the function position of an application, and
-frames for evaluating the arguments of an arithmetic expression.
-
-Continuations $k$ are either empty $[]$ or non-empty sequence of \emph{fibers}.
-A fiber $\fiber \coloneqq (\fl,\hc)$ is a list of frames $\fl$ and a handler
-closure $\hc \coloneqq (h,\env)$, which is a pair of handler $h$ and its
-environment $\env$.
-
-\subsubsection{Top-level reductions}
-
-The initial configuration for an expression $e$ is
-$\configa{e}{\emptyset}{\cstacka{[]}{\bullet}}$, where the environment and the
-stack are empty. The top-level reductions
+\begin{figure*}
 	\begin{smathpar}
@@
 	\end{smathpar}
-can be performed by either by taking a C step $\cstep$ or an OCaml step
-$\ostep$. Intuitively, this splits the world into executing C code and OCaml
-code.
-
+	\mycaption{Top-level reductions}
+	\label{sem:toplevel}
+	\vspace{-3mm}
+\end{figure*}
 \begin{figure*}
@@
   \end{smathpar}
-  \vspace{-1mm}
   \mycaption{Administrative Reductions -- $(\term, \env, \fl) \rightsquigarrow (\tau, \env, \fl)$.}
   \label{sem:step}
+	\vspace{-3mm}
 \end{figure*}
@@
   \end{smathpar}
-  \vspace{-1mm}
   \mycaption{C Reductions -- $(\term, \env, \fl, \ostack) \cstep \config$.}
   \label{sem:cstep}
+	\vspace{-3mm}
 \end{figure*}
+
+We present the operational semantics for a core language of effect handlers in
+Multicore OCaml. The operational semantics intends to faithfully capture the
+semantics of Multicore OCaml effect handlers, and to inform an implementation.
+
+\subsubsection{Syntax}
+
+Figure~\ref{sem:syntax} presents the syntax of expressions and configurations.
+Our expressions consists of integer constants ($n$), variables ($x$),
+abstraction ($\lam{x}{x}$), application ($e~e$), arithmetic expressions ($e
+\odot e$) where $\odot$ ranges over $\{+,-,*,/\}$, raising exceptions
+($\throw{l}{e}$), performing effects ($\perform{l}{e}$), and handling effects
+($\handle{e}{h}$). Abstractions come in two forms: OCaml abstractions
+($\lambda^o$) and C abstractions ($\lambda^c$). The handler consists of a
+return case ($\caseval{x}{e}$), zero or more exception cases
+($\caseexn{l}{x}{e}$) with label $l$, parameter $x$ and body $e$, and zero or
+more effect cases ($\caseeff{l}{x}{k}{e}$) with label $l$, parameter $x$,
+continuation $k$ and body $e$.
+
+The operational semantics is an extension of a CEK machine
+semantics~\cite{Felleisen86} for effect handlers, following the abstract
+machine semantics of Hillerstrom et al.~\cite{Hillerstrom20}. The key
+difference from Hillerstrom et al. is that our stacks are composed of
+alternating sequence of OCaml and C stack segments. The program state is
+captured as configuration $\config \coloneqq \configa{\tau}{\env}{\stack}$ with
+the current term $\tau$ under evaluation, its environment $\env$ and the
+current stack $\stack$. The term is either an expression $e$ or a value $v$.
+The value are integer constants $n$, continuations $k$, closure
+$\clos{x}{e}{\env}$, effect being performed $\effval{l}{k}$ and exception being
+raised $\exnval{l}$. The environment is a map from variables to values.
+
+The stack $\stack$ is either a C stack ($\cstack$) or an OCaml stack
+($\ostack$). Every C stack $\cstacka{\fl}{\ostack}$ consists of a list of
+frames $\fl$, and the OCaml stack $\ostack$ under it. Every OCaml stack is
+either empty $\ostackemp$ or non-empty $\ostacka{k}{\cstack}$ with the
+C stack $\cstack$ under it. Thus, the program stack is an alternating sequence
+of C and OCaml stacks terminating with an empty OCaml stack $\ostackemp$. The
+frame list $\fl$ is composed of individual frames $r$, which is one of an
+argument frame $\farg{e}{\env}$ with the expression $e$ at the argument
+position of an application with its environment $\env$, a function frame
+$\ffun{v}$ with the value at the function position of an application, and
+frames for evaluating the arguments of an arithmetic expression.
+
+Continuations $k$ are either empty $[]$ or non-empty sequence of \emph{fibers}.
+A fiber $\fiber \coloneqq (\fl,\hc)$ is a list of frames $\fl$ and a handler
+closure $\hc \coloneqq (h,\env)$, which is a pair of handler $h$ and its
+environment $\env$.
+
+\subsubsection{Top-level reductions}
+
+The initial configuration for an expression $e$ is
+$\configa{e}{\emptyset}{\cstacka{[]}{\bullet}}$, where the environment and the
+stack are empty. The top-level reductions (Figure ~\ref{sem:toplevel}) can be
+performed by either by taking a C step $\cstep$ or an OCaml step $\ostep$.
+Intuitively, this splits the world into executing C code and OCaml code.
 
@@
 in Multicore OCaml for performance backwards compatibility. This differs from
-other research languages with effect handlers~\cite{Hillerstrom20,Frank,Eff}, which
-implement exceptions using effects (by ignoring the continuation argument in
-the handler).
-
+other research languages with effect handlers~\cite{Hillerstrom20,Frank,Eff},
+which implement exceptions using effects (by ignoring the continuation argument
+in the handler).
+
+\vspace{-2mm}
 \subsection{Heap-allocated fibers}
@@
 before calling and after returning from |c_to_ocaml| function, respectively.
-The details of the exception frames have been elided in the figure. Thanks to
-the fiber representation, external calls and callbacks remain fast.
-
+Thanks to the fiber representation, external calls and callbacks remain
+competitive with stock OCaml.
+
+\vspace{-2mm}
 \subsection{Effect handlers}
@@
 \label{tab:micro_noperform}
-	\vspace{-3mm}
+	\vspace{-4mm}
 \end{table}
@@
 Kawahara et al.~\cite{Kawahara20} implement one-shot effect handlers using
-coroutines (a non-local control flow mechanism) as a macro-expressible
-translation, and present an embedding in Lua and Ruby. Lua provides asymmetric
-coroutines~\cite{Lua} where each coroutine uses its own stack similar to how
-each handled computation runs in its own fiber in Multicore OCaml. Similar to
-Multicore OCaml, Lua does not allow coroutines to include C frames. It is
-useful to note that Lua is an interpreted language, unlike OCaml, and DWARF
-unwind tables are not a concern in Lua.
+coroutines as a macro-expressible translation, and present an embedding in Lua
+and Ruby. Lua provides asymmetric coroutines~\cite{Lua} where each coroutine
+uses its own stack similar to how each handled computation runs in its own
+fiber in Multicore OCaml. Similar to Multicore OCaml, Lua does not allow
+coroutines to include C frames. It is useful to note that Lua is an interpreted
+language, unlike OCaml, and DWARF unwind tables are not a concern in Lua.
 
@@
 is distinguishing returning exits from a computation closing over resources
-from exits that would not return. Resource clean up needs to occur only during
+from non-returning ones. Resource clean up needs to occur only during
 non-returning exit. In Multicore OCaml, we assume that continuations are used
@@
 
-OCaml does not have a |try/finally| construct available in many programming
-languages. In OCaml, alternative standard libraries such as Base~\cite{} and
-Core~\cite{} provide mechanisms analogous to Common Lisp |unwind-protect|,
-which are in turn implemented using OCaml exceptions. Leijen~\cite{Leijen18}
-explicitly extends effect handlers with |finally| clause in Koka. Instead,
-Multicore OCaml |discontinue| primitive builds on extant defensive coding
-practices against exceptions to clean up resources. This design choice is
-crucial to ensure backwards compatibility of legacy systems code with non-local
-control flow introduced by effect handlers. Sivaramakrishnan et
-al.~\cite{TFP17} describe the interaction of effect handlers and asynchronous
-exceptions. This is orthogonal to the contributions of this paper which
-focusses on compiler and runtime system support for implementing effect
-handlers. Ahman et al.~\cite{Ahman20} present a effect handler core calculus
-with asynchrony which might help lay the formal foundations for reconciling
-asynchrony in systems and effect handlers.
+OCaml does not have a \texttt{\footnotesize try/finally} construct available in
+many programming languages. In OCaml, alternative standard libraries such as
+Base~\cite{} and Core~\cite{} provide mechanisms analogous to Common Lisp
+|unwind-protect|, which are in turn implemented using OCaml exceptions.
+Leijen~\cite{Leijen18} explicitly extends effect handlers with |finally| clause
+in Koka. Instead, Multicore OCaml |discontinue| primitive builds on extant
+defensive coding practices against exceptions to clean up resources. This
+design choice is crucial to ensure backwards compatibility of legacy systems
+code with non-local control flow introduced by effect handlers.
+Sivaramakrishnan et al.~\cite{TFP17} describe the interaction of effect
+handlers and asynchronous exceptions. This is orthogonal to the contributions
+of this paper which focusses on compiler and runtime system support for
+implementing effect handlers. Ahman et al.~\cite{Ahman20} present a effect
+handler core calculus with asynchrony which might help lay the formal
+foundations for reconciling asynchrony in systems and effect handlers.
 

### 2020-11-11 retro-concurrency.tex
Updates

@@
 considering effect handlers as one of the mechanism for supporting concurrency,
-asynchronous I/O and generators~\cite{WasmPropoal}. Project Loom~\cite{loom} is an
+asynchronous I/O and generators~\cite{WasmProposal}. Project Loom~\cite{loom} is an
 OpenJDK project which adds virtual threads and delimited continuations to Java.
@@
 required. We do not use a \emph{stack cache} of recently freed stacks for
-speeding up fiber allocation~\ref{Farvardin}. In our experiments, we observed
+speeding up fiber allocation~\cite{Farvardin20}. In our experiments, we observed
 that directly using |mimalloc|~\cite{Leijen19} outperforms having the stack
@@
 \end{lstlisting}
-Multicore OCaml does not enforce this behaviour in order to minimise the cost
-of using effect handlers.
+Multicore OCaml does not enforce this behaviour in order to avoid installing an
+effect handler every time a continuation is captured.
 
@@
 
-The interaction on non-local control flow and resources have been studied
-extensively. Scheme uses |dynamic-wind| to ensure de-allocation (and
-re-allocation) of resources when the non-local control leaves and enters back
-into a context. |dynamic-wind| is considered to be complex and not quite what
-one wants for resource clean up~\cite{Kiselyov,Sitaram03}. The main challenge
-is distinguishing returning exits from a computation closing over resources
-from non-returning ones. Resource clean up needs to occur only during
-non-returning exit. In Multicore OCaml, we assume that continuations are used
-linearly, and hence, no clean up occurs when effects are performed but only
-when an exception is thrown, perhaps from the use of |discontinue|. This
-obviates the need for |dynamic-wind|.
-
-OCaml does not have a \texttt{\footnotesize try/finally} construct available in
-many programming languages. In OCaml, alternative standard libraries such as
-Base~\cite{} and Core~\cite{} provide mechanisms analogous to Common Lisp
-|unwind-protect|, which are in turn implemented using OCaml exceptions.
-Leijen~\cite{Leijen18} explicitly extends effect handlers with |finally| clause
-in Koka. Instead, Multicore OCaml |discontinue| primitive builds on extant
-defensive coding practices against exceptions to clean up resources. This
-design choice is crucial to ensure backwards compatibility of legacy systems
-code with non-local control flow introduced by effect handlers.
-Sivaramakrishnan et al.~\cite{TFP17} describe the interaction of effect
-handlers and asynchronous exceptions. This is orthogonal to the contributions
-of this paper which focusses on compiler and runtime system support for
-implementing effect handlers. Ahman et al.~\cite{Ahman20} present a effect
-handler core calculus with asynchrony which might help lay the formal
-foundations for reconciling asynchrony in systems and effect handlers.
+The interaction of non-local control flow and resources has been studied
+extensively. Scheme uses |dynamic-wind|~\cite{}, which is a generalisation of
+Common Lisp |unwind-protect|~\cite{}, which ensures de-allocation (and
+re-allocation) of resources every time the non-local control leaves and enters
+back into a context. |dynamic-wind| is considered to be complex and not quite
+what one wants for resource clean up~\cite{Kiselyov,Sitaram03}. The main
+challenge is distinguishing returning exits from non-returning ones. Resource
+clean up needs to occur only during non-returning exit.
+
+In Multicore OCaml, we build on the existing defensive coding practices against
+exceptions to clean up resources on non-returning exits. Our continuations are
+assumed to be resumed exactly once using |continue| or |discontinue|. As a
+result, when a computation performs an effect, we are guaranteed that the
+control will return. For the non-returning cases (value and exceptional
+return), the code already handles resource cleanup. OCaml does not have a
+\texttt{\footnotesize try/finally} construct commonly used for resource cleanup
+in many programming languages. Alternative standard libraries such as
+Base~\cite{BaseProtect} and Core~\cite{CoreProtect} provide mechanisms
+analogous to |unwind-protect|, which are in turn implemented using exceptions.
+Thus, the linear use of continuations enabled by the |discontinue| primitive
+ensures backwards compatibility of legacy OCaml systems code with non-local
+control flow introduced by effect handlers.
+
+Leijen~\cite{Leijen18} explicitly extends effect handlers with |initially| and
+|finally| clauses in Koka for resource safety. Sivaramakrishnan et
+al.~\cite{TFP17} describe the interaction of effect handlers and asynchronous
+exceptions. This is orthogonal to the contributions of this paper. Our focus i
+the compiler and runtime system support for implementing effect handlers.
 
@@
 negligible overheads for existing OCaml programs and good performance for those
-that use effect handlers. Our evaluation shows that this is indeed the case.
+that use effect handlers. Our evaluation shows that this is indeed the case. We
+plan to upstream the compiler and runtime support for effect handlers followed
+by an effect system.
 

### 2020-11-11 retro-concurrency.tex
Updates

@@
 |('a,'b) continuation| of delimited continuations which expect a |'a| value for
-resumption and return |'b| value. The continuations may be |continue|d with a
-suitably typed value or |discontinue|d with an exception.
+resumption and return |'b| value. The continuations may be continued with a
+suitably typed value or discontinued with an exception.
 
@@
 
-We present the operational semantics for a core language of effect handlers in
-Multicore OCaml. The operational semantics intends to faithfully capture the
-semantics of Multicore OCaml effect handlers, and to inform an implementation.
+We present the operational semantics for a core language of effect handlers.
+The operational semantics intends to faithfully capture the semantics of
+Multicore OCaml effect handlers, and to inform an implementation.
 
@@
 
-Figure~\ref{sem:syntax} presents the syntax of expressions and configurations.
-Our expressions consists of integer constants ($n$), variables ($x$),
-abstraction ($\lam{x}{x}$), application ($e~e$), arithmetic expressions ($e
-\odot e$) where $\odot$ ranges over $\{+,-,*,/\}$, raising exceptions
-($\throw{l}{e}$), performing effects ($\perform{l}{e}$), and handling effects
-($\handle{e}{h}$). Abstractions come in two forms: OCaml abstractions
-($\lambda^o$) and C abstractions ($\lambda^c$). The handler consists of a
-return case ($\caseval{x}{e}$), zero or more exception cases
+Our expressions (Figure~\ref{sem:syntax}) consists of integer constants ($n$),
+variables ($x$), abstraction ($\lam{x}{x}$), application ($e~e$), arithmetic
+expressions ($e \odot e$) where $\odot$ ranges over $\{+,-,*,/\}$, raising
+exceptions ($\throw{l}{e}$), performing effects ($\perform{l}{e}$), and
+handling effects ($\handle{e}{h}$). Abstractions come in two forms: OCaml
+abstractions ($\lambda^o$) and C abstractions ($\lambda^c$). The handler
+consists of a return case ($\caseval{x}{e}$), zero or more exception cases
 ($\caseexn{l}{x}{e}$) with label $l$, parameter $x$ and body $e$, and zero or
@@
 The value are integer constants $n$, continuations $k$, closure
-$\clos{x}{e}{\env}$, effect being performed $\effval{l}{k}$ and exception being
-raised $\exnval{l}$. The environment is a map from variables to values.
+$\clos{x}{e}{\env}$, effect being performed ($\effval{l}{k}$) and exception
+being raised ($\exnval{l}$). The environment is a map from variables to values.
 
@@
 frames $\fl$, and the OCaml stack $\ostack$ under it. Every OCaml stack is
-either empty $\ostackemp$ or non-empty $\ostacka{k}{\cstack}$ with the
-C stack $\cstack$ under it. Thus, the program stack is an alternating sequence
-of C and OCaml stacks terminating with an empty OCaml stack $\ostackemp$. The
-frame list $\fl$ is composed of individual frames $r$, which is one of an
-argument frame $\farg{e}{\env}$ with the expression $e$ at the argument
-position of an application with its environment $\env$, a function frame
-$\ffun{v}$ with the value at the function position of an application, and
-frames for evaluating the arguments of an arithmetic expression.
+either empty $\ostackemp$ or non-empty $\ostacka{k}{\cstack}$ with the C stack
+$\cstack$ under it. Thus, the program stack is an alternating sequence of C and
+OCaml stacks terminating with an empty OCaml stack $\ostackemp$. The frame list
+$\fl$ is composed of individual frames $r$, which is one of an argument frame
+$\farg{e}{\env}$ with the expression $e$ at the argument position of an
+application with its environment $\env$, a function frame $\ffun{v}$ with the
+value $v$ at the function position of an application, and frames for evaluating
+the arguments of an arithmetic expression.
 
@@
 performed by either by taking a C step $\cstep$ or an OCaml step $\ostep$.
-Intuitively, this splits the world into executing C code and OCaml code.
 
@@
 
-The C reduction steps are given in Figure~\ref{sem:cstep}. We can take a C step
-by taking an administrative reduction step $\rightsquigarrow$. The
-administrative reductions are common to both C and OCaml. The rules
-\textsc{Var}, \textsc{Arith1}, \textsc{Arith2}, \textsc{App1}, \textsc{App2}
-and \textsc{App3} are standard. \textsc{Arith3} performs the arithmetic
-operation on the integers ($\llbracket n_1 \odot n_2 \rrbracket$).
-\textsc{Raise} pushes an function frame with exception value to indicate that
-an exception is being raised. Similarly, \textsc{Perform} pushes a function
-frame with an effect value with an empty continuation
+We can take a C step (Figure~\ref{sem:cstep}) by taking an administrative
+reduction step $\rightsquigarrow$. The administrative reductions are common to
+both C and OCaml. The rules \textsc{Var}, \textsc{Arith1}, \textsc{Arith2},
+\textsc{App1}, \textsc{App2} and \textsc{App3} are standard. \textsc{Arith3}
+performs the arithmetic operation on the integers ($\llbracket n_1 \odot n_2
+\rrbracket$). \textsc{Raise} pushes an function frame with exception value to
+indicate that an exception is being raised. Similarly, \textsc{Perform} pushes
+a function frame with an effect value with an empty continuation
 $[[],(\{\caseval{x}{x}\},\emptyset)]$ with no captured frames and an empty
@@
 required. We do not use a \emph{stack cache} of recently freed stacks for
-speeding up fiber allocation~\cite{Farvardin20}. In our experiments, we observed
-that directly using |mimalloc|~\cite{Leijen19} outperforms having the stack
+speeding up fiber allocation~\cite{Farvardin20}. In our experiments, we
+observed that directly using |mimalloc|~\cite{Leijen19} outperforms the stack
 cache.
@@
 return exactly once (\S\ref{sec:refine}). Consider the scenario when a
-continuation is neither |continue|d nor |discontinue|d. Since fibers are
-|malloc|ed and are |free|d when the computation being handled returns a value
+continuation is neither continued nor discontinued. Since fibers are
+malloced and are |free|d when the computation being handled returns a value
 or raises an exception, not resuming continuations leaks memory. In addition,
@@
 
-Multicore OCaml effect handlers are similar to~\cite{Leijen14, Kiselyov12}, but
-are more efficient; since we only allow our continuations to be resumed at-most
-once, there is no need for copying frames. Bruggeman et al.~\cite{Bruggeman96}
-show how to implement one-shot continuations efficiently using segmented stacks
-in Scheme. Farvardin et al.~\cite{Farvardin20} perform a comprehensive
-evaluation of various implementation strategies for continuations on modern
-hardware. Multicore OCaml stacks do not neatly fit the description of one of
-these implementation strategies. They best described as using the resize
+Similar to~\cite{Leijen14, Kiselyov12}, Multicore OCaml uses the call stack for
+implementing continuations, but our continuations are one-shot. Bruggeman et
+al.~\cite{Bruggeman96} show how to implement one-shot continuations efficiently
+using segmented stacks in Scheme. Farvardin et al.~\cite{Farvardin20} perform a
+comprehensive evaluation of various implementation strategies for continuations
+on modern hardware. Multicore OCaml stacks do not neatly fit the description of
+one of these implementation strategies. They best described as using the resize
 strategy from Farvardin et al. for each of the fibers, which are in turn linked
@@
 
+Multicore OCaml is not the first language to support stack inspection in the
+presence of non-local control operators. Chez Scheme supports continuation
+marks~\cite{Flatt20} which permit stack inspection to implement dynamic
+binding, exceptions, profilers, debuggers, etc, in the presence of first-class
+continuations. As the authors note, continuation marks can be implemented using
+effect handlers, but direct support for continuation marks leads to better
+performance. In this work, we focus on retaining the support for stack
+inspection through DWARF unwind tables in the presence of effect handlers.
+
 The interaction of non-local control flow and resources has been studied
-extensively. Scheme uses |dynamic-wind|~\cite{}, which is a generalisation of
-Common Lisp |unwind-protect|~\cite{}, which ensures de-allocation (and
+extensively. Scheme uses |dynamic-wind|~\cite{R5RS}, which is a generalisation of
+Common Lisp |unwind-protect|~\cite{Steele90}, which ensures de-allocation (and
 re-allocation) of resources every time the non-local control leaves and enters
-back into a context. |dynamic-wind| is considered to be complex and not quite
-what one wants for resource clean up~\cite{Kiselyov,Sitaram03}. The main
-challenge is distinguishing returning exits from non-returning ones. Resource
-clean up needs to occur only during non-returning exit.
+back into a context. |dynamic-wind| is not quite what one wants for resource
+clean up~\cite{Kiselyov,Sitaram03}. The main challenge is distinguishing
+returning exits from non-returning ones. Resource clean up needs to occur only
+during non-returning exit.
 
@@
 al.~\cite{TFP17} describe the interaction of effect handlers and asynchronous
-exceptions. This is orthogonal to the contributions of this paper. Our focus i
+exceptions. This is orthogonal to the contributions of this paper. Our focus is
 the compiler and runtime system support for implementing effect handlers.
@@
 return \emph{more than once}, leading to errors such as double-freeing
-malloc'ed memory. Since our goal is to retrofit effect handlers onto OCaml, we
+malloced memory. Since our goal is to retrofit effect handlers onto OCaml, we
 relegate continuation cloning to the OCaml |Obj| module, which contains unsafe

### 2020-11-11 retro-concurrency.tex
Minor rewords and tweaks: sections 1-3

--- 2020-11-11 retro-concurrency.tex
@@
 for modular programming with user-defined effects. The runtime semantics of
-  effect handlers is particularly attractive as they allow advanced control
-  flow mechanisms such as first-class continuations, generators, async/await,
-  coroutines to be composably expressed. In this paper, we present the design,
-  a full-fledged implementation and evaluation of effect handlers for OCaml, an
-  industrial-strength, multi-paradigm programming language. OCaml tends to be
-  widely used for systems programming, and is compatible with program analysis
-  tools that inspect the stack such as debuggers and profilers. Retrofitting
+effect handlers is particularly attractive as they allow advanced control flow
+mechanisms such as first-class continuations, generators, async/await,
+coroutines to be composably expressed. This paper presents a design, a
+full-fledged implementation and an evaluation of effect handlers for OCaml.
+OCaml is industrial-strength multi-paradigm programming language that tends to
+be widely used for systems programming, our implementation strives to maintain
+the existing performance profile and toolchain compatibility. Retrofitting
 effect handlers onto OCaml is challenging since OCaml does not have any
-  non-local control flow mechanisms other than exceptions. Our implementation
-  of effect handlers for OCaml imposes negligible overhead on code that does
-  not use effect handlers, and remains compatible with program analysis tools
-  that inspect the stack.
+non-local control flow mechanisms other than exceptions. Our implementation of
+effect handlers for OCaml imposes negligible overhead on code that does not use
+effect handlers, remains compatible with program analysis tools that inspect the
+stack and is efficient for code that makes use of effect handlers.
 \end{abstract}
@@
 
-On line 4, we declare an effect |DivBy0| which is parameterised by an integer,
-which when performed returns result of division. If the denominator is 0, then
-we perform |DivBy0| effect with the numerator value (line 12) so that the
-context may decide the return value. We can get the |iOS| semantics by wrapping
-the evaluator computation with the following handler:
+On line 4, we declare an effect |DivBy0| that is parameterised by an integer,
+which when performed returns the result of division-by-zero. If the denominator
+is 0, then we perform |DivBy0| effect with the numerator value (line 12) so that
+the context may decide the return value. We can get the |iOS| semantics by
+wrapping the evaluator with the following handler:
 \begin{lstlisting}
@@
 \end{lstlisting}
-If the evaluator returns a value, we simply return that. In case the evaluator
+If the evaluator returns a value, we simply return that. If the evaluator
 performs the |DivBy0| effect, we catch the effect in the effect case along with
@@
 |Error| value. Hence, the computation |ios(1/0)| returns |Error|. On the
-other hand, we can get |gsearch| semantics by enclosing the evaluator in the
+other hand, we can get the |gsearch| semantics by enclosing the evaluator in the
 following handler:
@@
 
-We can extend the calculator modularly with the ability to read user input
-using the |Read| effect defined earlier as follows:
+We can modularly extend the calculator to read user input using the |Read|
+effect defined earlier as follows:
 \begin{lstlisting}
@@
 There are many research languages and libraries built around effect
-handlers~\cite{Leijen14,Hillerstrom20,Pyro,Frank,Eff}. Unlike these efforts, our goal is to
-retrofit effect handlers onto the OCaml programming language, which has been in
-continuous use for the past 25 years in large codebases including verification
-tools~\cite{everest,Coq}, mission critical software systems~\cite{astree} and
-latency sensitive networked applications~\cite{Madhavapeddy13}. OCaml is
-particularly favoured for its competitive yet predictable performance, with
-fast foreign-function interface (FFI) and excellent compatibility with program
-analysis tools such as debuggers and profilers that utilise DWARF stack unwind
-tables~\cite{DWARF} to obtain a backtrace. Our primary motivation for adding
-effect handlers to OCaml is to support (1) highly concurrent applications such
-as web servers to be expressed naturally in \emph{direct-style} as opposed to
-using callbacks and (2) efficient nested parallel programs in the vein of
-Cilk~\cite{Cilk}, OpenMP~\cite{OpenMP}, X10~\cite{Charles05}, Intel
-TBB~\cite{IntelTBB}, and variants of Parallel ML~\cite{MaPLe, Fluet10,
-Sivaramakrishnan14}.
+handlers~\cite{Leijen14,Hillerstrom20,Pyro,Frank,Eff}. Unlike these efforts, our
+goal is to retrofit effect handlers onto the OCaml programming language, which
+has been in continuous use for the past 25 years in large codebases including
+verification tools~\cite{everest,Coq}, mission critical software
+systems~\cite{astree} and latency sensitive networked
+applications~\cite{Madhavapeddy13}. OCaml is particularly favoured for its
+competitive yet predictable performance, with fast foreign-function interface
+(FFI) and excellent compatibility with program analysis tools such as debuggers
+and profilers that utilise DWARF stack unwind tables~\cite{DWARF} to obtain a
+backtrace. Our primary motivation for adding effect handlers to OCaml is to
+support (1) highly concurrent applications such as web servers using a
+natural \emph{direct-style} as opposed to using callbacks and (2) efficient
+nested parallel programs in the vein of Cilk~\cite{Cilk}, OpenMP~\cite{OpenMP},
+X10~\cite{Charles05}, Intel TBB~\cite{IntelTBB}, and variants of Parallel
+ML~\cite{MaPLe, Fluet10, Sivaramakrishnan14}.
 
@@
 delimited continuations necessary for effect handlers without sacrificing the
-desirable properties of OCaml. One standard way of implementing continuations
-is to use continuation-passing style (CPS) in the compiler's intermediate
-representation (IR)~\cite{Leijen14}. If the compiler were to use CPS, then reifying
-continuations in the source language is relatively straight-forward. But OCaml
-does not use CPS IR, and changing the OCaml compiler to utilise CPS IR would be
-an enormous undertaking. CPS IR would also affect the performance profile of
-OCaml applications due to increased memory allocations as continuation closures
-get allocated on the heap~\cite{Farvardin20}. Moreover, with CPS, stack
-unwinding is meaningless and hence, we would lose compatibility with tools that
-inspect the program stack. Hence, we choose not to use CPS translation and
-represent continuations as call stacks.
+desirable properties of OCaml. One standard way of implementing continuations is
+to use continuation-passing style (CPS) in the compiler's intermediate
+representation (IR)~\cite{Leijen14}. If the compiler were to use CPS, then
+reifying continuations in the source language is relatively straight-forward.
+But OCaml does not use CPS IR, and changing the OCaml compiler to utilise CPS IR
+would be an enormous undertaking. Utilizing a CPS IR would also affect the
+performance profile of OCaml applications due to increased memory allocations as
+continuation closures get allocated on the heap~\cite{Farvardin20}. Moreover,
+with CPS, stack unwinding is meaningless and hence, we would lose compatibility
+with tools that inspect the program stack. Hence, we choose not to use CPS
+translation and represent continuations as call stacks.
 
@@
 
-\noindent \textbf{R3: Effect handler efficiency.} In order to support highly-concurrent
-applications, the program must accommodate millions of continuations at the
-same time. Installing effect handlers, capturing and resuming continuations
-must be fast.
+\noindent \textbf{R3: Effect handler efficiency.} In order to support
+highly-concurrent applications, the program must accommodate millions of
+continuations at the same time. Installing effect handlers, capturing and
+resuming continuations must be fast.
 
@@
 The need to host millions of continuations at the same time rules out the use
-of large contiguous stack space as in C for continuations. Instead, we resort
+of a large contiguous stack space as in C for continuations. Instead, we resort
 to using small initial stacks and growing the stacks on demand. As a result,
@@
 
-Our work is also timely. WebAssembly~\cite{Hass17} community group is actively
-considering effect handlers as one of the mechanism for supporting concurrency,
-asynchronous I/O and generators~\cite{WasmProposal}. Project Loom~\cite{loom} is an
-OpenJDK project which adds virtual threads and delimited continuations to Java.
-Swift concurrency roadmap~\cite{swift} includes direct-style asynchronous
-programming and structured concurrency as milestones. We believe that our
-design choices will inform similar choices to be made in the other
-industrial-strength languages.
+Our work is also timely. The WebAssembly~\cite{Hass17} community group is
+actively considering effect handlers as one of the mechanism for supporting
+concurrency, asynchronous I/O and generators~\cite{WasmProposal}. Project
+Loom~\cite{loom} is an OpenJDK project which adds virtual threads and delimited
+continuations to Java. Swift concurrency roadmap~\cite{swift} includes
+direct-style asynchronous programming and structured concurrency as milestones.
+We believe that our design choices will inform similar choices to be made in the
+other industrial-strength languages.
 
@@
     inspect the stack using DWARF unwind tables. We have validated our DWARF
-    unwind tables using an automated validator tool~\cite{Bastian19}.
+    unwind tables with the assistance of an automated validator
+    tool~\cite{Bastian19}.
   \item a formal operational semantics for the effect handler implementation in
@@
 
-In order to illustrate the layout of the stock OCaml stack, consider the
+Consider the layout of the stock OCaml stack for the
 program in Figures~\ref{code:meander_ml} and~\ref{code:meander_c}. The OCaml
 main function |omain| installs two exception handlers |h1| and |h2| to handle
-exceptions |E1| and |E2|. |omain| calls the external C function |ocaml_to_c|,
-which in turn calls back into OCaml function |c_to_ocaml|, which raises the
+the exceptions |E1| and |E2|. |omain| calls the external C function |ocaml_to_c|,
+which in turn calls back into the OCaml function |c_to_ocaml|, which raises the
 exception |E1|. OCaml supports raising exceptions in C as well as throwing
 exceptions across external calls. Hence, the exception |E1| gets caught in the
-handler |h1|, and |omain| returns |42|. The layout of the stack in native code
+handler |h1|, and |omain| returns |42|. The layout of the stack in the native code
 backend just before raising the exception in |c_to_ocaml| is illustrated in
@@
 
-OCaml uses the same hardware stack as C, and hence the program stack has
+OCaml uses the same program stack as C, and hence the stack has
 alternating sequences of C and OCaml frames. However, unlike C, OCaml does not
@@
 guard page at the end of the stack region to detect stack overflow. Stack
-overflow is detected by a memory fault, and |Stack_overflow| exception is
+overflow is detected by a memory fault and a |Stack_overflow| exception is
 raised to unwind the stack.
@@
 OCaml does not use the C calling convention. In particular, there are no
-callee-saved registers in OCaml. On the |x86-64| backend, the C callee-saved
-registers  |r15| holds the \emph{allocation pointer} into the minor heap used
-for bump pointer allocation, and |r14| holds a reference to the |Caml_state|
-table of global variables used by the runtime. This makes external calls
-extremely fast in OCaml. If the external function does not allocate in the
-OCaml heap, then the external functions can be called directly, and no
+callee-saved registers in OCaml. On the |x86-64| backend, the OCaml runtime
+makes use of C callee-saved registers: |r15| holds the \emph{allocation pointer}
+into the minor heap used for bump pointer allocation, and |r14| holds a
+reference to the |Caml_state| table of global variables used by the runtime.
+This makes external calls extremely fast in OCaml. If the external function does
+not allocate in the OCaml heap, then it can be called directly and no
 bookkeeping is necessary. For external functions which allocate in the OCaml
-heap, the cached allocation pointer is saved to a field in the |Caml_state|
-before the external call, and loaded back on return. Similarly, callbacks into
-OCaml are also cheap. Callbacks into OCaml involves loading the arguments in
-the right registers and calling the OCaml function. OCaml callbacks are
-relatively common as the garbage collector (GC) finalisation functions are
-executed as OCaml callbacks.
+heap, the cached allocation pointer is saved to |Caml_state| before the external
+call and it is restored on return. Similarly, callbacks into OCaml are also
+cheap: these involve loading the arguments in the right registers and calling
+the OCaml function. OCaml callbacks are relatively common as the garbage
+collector (GC) finalisation functions are executed as OCaml callbacks.
 
@@
 are no callee-saved registers to be restored when the exception handler runs.
-Hence, installing an exception handler simply pushes the program counter (|pc|)
-of the handler and the current exception pointer (|exn_ptr| -- a field in
-|Caml_state|). At this point, the exception pointer is updated to be the
+Installing an exception handler simply pushes the program counter (|pc|) of the
+handler and the current exception pointer (|exn_ptr| -- a field in |Caml_state|)
+to the stack. At this point, the current exception pointer is updated to be the
 current stack pointer (|rsp|). This creates a linked-list of exception handler
@@
 In order to forward exceptions across C frames, the C stub function
-|caml_call_ocaml|, which is the entry point for both the main OCaml code and
-callbacks into OCaml, pushes an exception handler frame that forwards the
+|caml_call_ocaml|, pushes an exception handler frame that forwards the
 exception to the innermost OCaml exception handler (|raise_exn_c| in
-Figure~\ref{fig:stock_stack}), or prints a fatal error (|fatal_uncaught|) if
+Figure~\ref{fig:stock_stack}) or prints a fatal error (|fatal_uncaught|) if
 there are no enclosing handlers. Exceptions are so cheap in OCaml that it is
-very common to use them for \emph{local} control flow.
+common to use them for \emph{local} control flow.
 
@@
 OCaml generates \emph{stack maps} in order to accurately identify roots on the
-stack for assisting GC. For every call point in the program, OCaml compiler
+stack for assisting the GC. For every call point in the program, the OCaml compiler
 emits the size of the frame and the set of all live registers in the frame
@@
 and~\ref{code:meander_c}, one could set a break point in |gdb| at
-|caml_raise_exn| to get the backtrace in Figuere~\ref{code:gdb_backtrace} which
-corresponds to the stack layout in Figure~\ref{fig:stock_stack}.
+|caml_raise_exn| to get the backtrace in Figure~\ref{code:gdb_backtrace} which
+corresponds to the stack in Figure~\ref{fig:stock_stack}.
 
@@
 
-In this section, we refine the effect handler design for the OCaml use case.
-The primary motivation for adding effect handlers to OCaml is to support
-non-local control flow mechanism to implement asynchronous I/O in direct style,
-generators and enable task parallelism. These use cases do not resume the
-continuations more than once. Hence, our continuations are one-shot, and
-resuming the continuation more than once raises |Invalid_argument| exception.
-It is well-known that one-shot continuations can be implemented
+In this section, we refine the effect handler design for the OCaml use case. The
+primary motivation for adding effect handlers to OCaml is to support a non-local
+control flow mechanism to implement asynchronous I/O in direct style, generators
+and enable task parallelism. These use cases do not resume the continuations
+more than once. Hence, our continuations are one-shot, and resuming the
+continuation more than once raises |Invalid_argument| exception. It is
+well-known that one-shot continuations can be implemented
 efficiently~\cite{Bruggeman96}.
@@
 \end{lstlisting}
-The function |input_line| raises |End_of_file| exception on reaching the end of
+The function |input_line| raises an |End_of_file| exception on reaching the end of
 input, which is handled by the exception handler which closes the channels. The
@@
 
-One of our goals(\S\ref{sec:req}) is to make this code transparently
+One of our goals (\S\ref{sec:req}) is to make this code transparently
 asynchronous. We can define effects for performing the I/O operations and wrap
@@
 
-Program that use resources such as channels are usually written defensively
+Programs that use resources such as channels are usually written defensively
 with the assumption that calling a function will return \emph{exactly once},
@@
 with no matching handler, then the function \emph{will not return at all}. To
-remedy this, when such an effect bubbles up to the top-level, we discontinue
-the continuation with |Unhandled| exception so that the exception handlers may
+remedy this, when such an effect bubbles up to the top-level, we discontinue the
+continuation with an |Unhandled| exception so that the exception handlers may
 run and clean up the resources.

### 2020-11-11 retro-concurrency.tex
Some changes to Section5: minor redrafts; more information around the DWARF verification

@@
 support DWARF unwinding, whose details we will return to later. Then, there is
-a top-level handler that invokes the exception handler closure |clos_hexn|
+a top-level handler that invokes the exception handler |clos_hexn|
 after switching to the parent fiber. This is followed by the |pc| of the code
-that invokes value handler |clos_hval| after switching to the parent stack. The
+that invokes the value handler |clos_hval| after switching to the parent stack. The
 stack is laid out such that when the computation in the handler returns,
-control goes to value handler.
+control goes to the value handler.
 
@@
 handlers to be as small as possible, this area is initially 16 words in length.
-When the stack pointer |rsp| goes less than stack threshold (maintained in the
+When the stack pointer |rsp| becomes less than the stack threshold (maintained in the
 |Caml_state| table), the stack is said to have overflown. On stack overflow, we
@@
 speeding up fiber allocation~\cite{Farvardin20}. In our experiments, we
-observed that directly using |mimalloc|~\cite{Leijen19} outperforms the stack
+observed that directly using |mimalloc|~\cite{Leijen19} outperforms a stack
 cache.
@@
 external calls in the system stack. Calling a C function from OCaml involves
-saving the exception and stack pointers in the current fiber, saving the
+saving the stack pointer in the current fiber, saving the
 allocation pointer in the |Caml_state|, updating |rsp| to the top of system
 stack (maintained in |Caml_state|), and calling the C function. The actions are
-reversed when returning from the external call. One subtlety is that for C
+reversed when returning from the external call. For C
 functions that take arguments on the stack, the arguments must be copied to the
-C stack from the OCaml stack where they are laid out. For vast majority of
+C stack from the OCaml stack where they are laid out. In the vast majority of
 external calls, we perform only a few additional instructions over stock OCaml.
@@
 
-Logically, DWARF maintains a large table which records for every machine
-instruction where the return address and callee-saved registers are stored. To
-avoid reifying this large table, DWARF represents the table using a compact,
-turing-complete, bytecode representation, which is interpreted on demand for
-building the stack unwind table. Within each function, DWARF maintains a
-\emph{canonical frame address} (CFA), which is invariant in the function body,
-and is traditionally the stack pointer before entering this function. Hence, on
-x86-64, where the return address is pushed on the stack on call, the return
-address is at |CFA - 8|.
-
-Our goal is to help DWARF identify the CFA when stacks are switched at external
-calls, callbacks and effect handlers. We have implemented short, hand-written
-DWARF bytecode at these three places, which we have validated with the DWARF
-unwind table validator tool~\cite{Bastian19}. When starting to run on a new
+Logically, DWARF call-frame information maintains a large table which records
+for every machine instruction where the return address and callee-saved
+registers are stored. To avoid reifying this large table, DWARF directives
+represent the table using a compact bytecode representation. This representation
+is interpreted on demand for building the stack unwind table. Within each
+function, DWARF maintains a \emph{canonical frame address} (CFA) and is
+traditionally the stack pointer before entering this function. Hence, on x86-64,
+where the return address is pushed on the stack on call, the return address is
+at |CFA - 8|.
+
+One challenge is to encode in DWARF directives the CFA when stacks are switched
+at effect handlers, callbacks and external calls. When starting to run on a new
 fiber, we insert DWARF bytecode to follow the |parent_fiber| pointer and
-dereference the |saved_sp| to get the CFA (|saved_sp + 8|). During callback, we
-save the current system stack pointer in the |DWARF block| in
+dereference the |saved_sp| to get the CFA (|saved_sp + 8|). During callbacks
+into OCaml, we save the current system stack pointer in the |DWARF block| in
 Figure~\ref{fig:fiber} to identify the CFA in the C stack. DWARF unwinding for
-external calls is implemented similarly by saving the current OCaml stack
+external calls is implemented by following a link to the current OCaml stack
 pointer. With these changes, we get the same backtrace for the \texttt{meander}
 program from \S\ref{sec:unwind}, modulo runtime system functions due to effect
-handlers.
+handlers. We have utilized a DWARF verification tool~\cite{Bastian19} with
+extensions to handle PIE and additional DWARF primitives to flag possible errors
+in the hand written DWARF directives where we have altered the OCaml assembler
+to implement effects. In three places additional manual verification was
+required; two were where the stack is intentionally switched between fibers and
+one was where the return to the parent stack on child stack completion could not
+track stack resizes and intermediate effect handler frames.
 
@@
 continuation is neither continued nor discontinued. Since fibers are
-malloced and are |free|d when the computation being handled returns a value
+|malloc|ed and are |free|d when the computation being handled returns a value
 or raises an exception, not resuming continuations leaks memory. In addition,

### 2020-11-12 retro-concurrency.tex
Updates

@@
       escapechar={<@}{@>},
-      %basewidth={0.5em,0.45em},
-      %aboveskip={3pt},
-      %belowskip={3pt},
+      basewidth={0.5em,0.45em},
+      aboveskip={3pt},
+      belowskip={3pt},
       breaklines = true,
@@
 
-\newcommand{\mycaption}[1]{\vspace{-4mm}\caption{#1}}
 \setlength{\tabcolsep}{3pt}
@@
 | ...
-| Inp -> int_of_string(perform(Read stdin))
+| Inp -> int_of_string (perform (Read stdin))
 \end{lstlisting}
@@
 We have implemented effect handlers in a multicore extension of the OCaml
-programming language. We call our implementation ``Multicore OCaml'' to
+programming language. We call our implementation \textbf{Multicore OCaml} to
 distinguish it from stock OCaml.
 
-The rest of the paper is organized as follows. We describe the details of the
-OCaml program stack in the next section, and highlight the features that the
-effect handler implementation must retain in order to provide backwards
-compatibility. In \S\ref{sec:refine}, we refine the design further focussing on
-the issues that come up when integrating effect handlers into a mainstream
-systems language. \S\ref{sec:semantics} presents the formal operational
-semantics for Multicore OCaml effect handlers. \S\ref{sec:impl} discusses the
-compiler and the runtime system support for implementing effect handlers.
-\S\ref{sec:eval} presents extensive performance evaluation of effect handlers
-against the goals set in \S\ref{sec:req}. \S\ref{sec:related} and
-\S\ref{sec:conc} present the related work and the conclusions.
+The rest of the paper is organized as follows. The next section describes the
+details of stock OCaml program stack. In \S\ref{sec:refine}, we refine the
+design further focussing on retrofitting effect handlers into a mainstream
+\emph{systems} language. \S\ref{sec:semantics} presents the compilation scheme
+and operational semantics semantics for Multicore OCaml effect handlers.
+\S\ref{sec:impl} discusses the compiler and the runtime system support for
+implementing effect handlers. \S\ref{sec:eval} presents extensive performance
+evaluation of effect handlers against the goals set in \S\ref{sec:req}.
+\S\ref{sec:related} and \S\ref{sec:conc} present the related work and the
+conclusions.
 
@@
   \centering
-  \includegraphics[scale=0.46]{figures/stock_stack}
+  \includegraphics[scale=0.44]{figures/stock_stack}
   \subcaption{Stack layout before raise \texttt{E1}.}
@@
 \end{minipage}
-  \mycaption{Program stack on stock OCaml.}
+\vspace{-2mm}
+\caption{Program stack on stock OCaml.}
+\vspace{-2mm}
 \end{figure*}
@@
 The main challenge in implementing effect handlers in Multicore OCaml is
-managing the program stack. In this section, we provide an overview of the
-program stack and related mechanisms in stock OCaml.
+managing the program stack and preserving its desirable properties. In this
+section, we provide an overview of the program stack and related mechanisms in
+stock OCaml.
 
@@
 OCaml does not use the C calling convention. In particular, there are no
-callee-saved registers in OCaml. On the |x86-64| backend, the OCaml runtime
-makes use of C callee-saved registers: |r15| holds the \emph{allocation pointer}
-into the minor heap used for bump pointer allocation, and |r14| holds a
-reference to the |Caml_state| table of global variables used by the runtime.
-This makes external calls extremely fast in OCaml. If the external function does
-not allocate in the OCaml heap, then it can be called directly and no
-bookkeeping is necessary. For external functions which allocate in the OCaml
-heap, the cached allocation pointer is saved to |Caml_state| before the external
-call and it is restored on return. Similarly, callbacks into OCaml are also
-cheap: these involve loading the arguments in the right registers and calling
-the OCaml function. OCaml callbacks are relatively common as the garbage
-collector (GC) finalisation functions are executed as OCaml callbacks.
+callee-saved registers in OCaml. In the |x86-64| backend, the OCaml runtime
+makes use of two C callee-saved registers for supporting OCaml execution. The
+register |r15| holds the \emph{allocation pointer} into the minor heap used for
+bump pointer allocation, and |r14| holds a reference to the |Caml_state| table
+of global variables used by the runtime. This makes external calls extremely
+fast in OCaml. If the external function does not allocate in the OCaml heap,
+then it can be called directly and no bookkeeping is necessary. For external
+functions which allocate in the OCaml heap, the cached allocation pointer is
+saved to |Caml_state| before the external call and it is restored on return.
+Similarly, callbacks into OCaml are also cheap: these involve loading the
+arguments in the right registers and calling the OCaml function. OCaml
+callbacks are relatively common as the garbage collector (GC) finalisation
+functions are executed as OCaml callbacks.
 
@@
 Installing an exception handler simply pushes the program counter (|pc|) of the
-handler and the current exception pointer (|exn_ptr| -- a field in |Caml_state|)
-to the stack. At this point, the current exception pointer is updated to be the
-current stack pointer (|rsp|). This creates a linked-list of exception handler
-frames on the stack as shown in Figure~\ref{fig:stock_stack}. Raising an
-exception simply sets |rsp| to |exn_ptr|, loads the saved |exn_ptr|, and jumps
-to the |pc| of the handler.
+handler and the current exception pointer (|exn_ptr| -- a field in
+|Caml_state|) to the stack. After this, the current exception pointer is
+updated to be the current stack pointer (|rsp|). This creates a linked-list of
+exception handler frames on the stack as shown in Figure~\ref{fig:stock_stack}.
+Raising an exception simply sets |rsp| to |exn_ptr|, loads the saved |exn_ptr|,
+and jumps to the |pc| of the handler.
 
@@
 OCaml generates \emph{stack maps} in order to accurately identify roots on the
-stack for assisting the GC. For every call point in the program, the OCaml compiler
-emits the size of the frame and the set of all live registers in the frame
-containing pointers into the heap. During a GC, the OCaml stack is walked and
-the roots are marked skipping over the C frames.
+stack for assisting the GC. For every call point in the program, the OCaml
+compiler emits the size of the frame and the set of all live registers in the
+frame containing pointers into the heap. During a GC, the OCaml stack is walked
+and the roots are marked skipping over the C frames.
 
@@
 increase the size of the
-stack~\footnote{https://github.com/ocaml/ocaml/issues/5721\#issuecomment-472965549}.
+stack\footnote{https://github.com/ocaml/ocaml/issues/5721\#issuecomment-472965549}.
 Moreover, not using frame pointers saves two instructions in the function
@@
 by OCaml to walk the stack for GC.
-
-In summary, the design of effect handlers in OCaml must accommodate these
-desirable properties of OCaml with respect to the program stack.
 
@@
 
+\vspace{-2mm}
 \section{Semantics}
 \label{sec:semantics}
-
-In this section, we formalise the effect handler design in Multicore OCaml.
 
@@
 {retc: 'a -> 'b;
- effc: 'c.'c eff -> ('c,'b)continuation -> 'b; }
+ effc: 'c.'c eff -> ('c,'b) continuation -> 'b; }
 val match_with: 'a comp -> ('a,'b) handler -> 'b
@@
 
-\subsection{Dynamic semantics}
-
 \begin{figure*}
+\begin{minipage}{\linewidth}
   \begin{minipage}[t]{0.49\linewidth}
@@
 	\end{minipage}
-	\vspace{-3mm}
-  \caption{Syntax of expressions and configurations}
+  \subcaption{Syntax of expressions and configurations}
   \label{sem:syntax}
-	\vspace{-3mm}
-\end{figure*}
-\begin{figure*}
+\end{minipage}
+%
+\begin{minipage}{\linewidth}
 	\begin{smathpar}
@@
 	\end{smathpar}
-	\mycaption{Top-level reductions}
+	\subcaption{Top-level reductions}
 	\label{sem:toplevel}
-	\vspace{-3mm}
-\end{figure*}
-\begin{figure*}
+\end{minipage}
+%
+\begin{minipage}{\linewidth}
   \begin{smathpar}
@@
   \end{smathpar}
-  \mycaption{Administrative Reductions -- $(\term, \env, \fl) \rightsquigarrow (\tau, \env, \fl)$.}
+  \subcaption{Administrative Reductions -- $(\term, \env, \fl) \rightsquigarrow (\tau, \env, \fl)$.}
   \label{sem:step}
-	\vspace{-3mm}
-\end{figure*}
-\begin{figure*}
+\end{minipage}
+%
+\begin{minipage}{\linewidth}
   \begin{smathpar}
@@
   \end{smathpar}
-  \mycaption{C Reductions -- $(\term, \env, \fl, \ostack) \cstep \config$.}
+  \subcaption{C Reductions -- $(\term, \env, \fl, \ostack) \cstep \config$.}
   \label{sem:cstep}
-	\vspace{-3mm}
-\end{figure*}
-
-We present the operational semantics for a core language of effect handlers.
-The operational semantics intends to faithfully capture the semantics of
-Multicore OCaml effect handlers, and to inform an implementation.
-
-\subsubsection{Syntax}
-
-Our expressions (Figure~\ref{sem:syntax}) consists of integer constants ($n$),
-variables ($x$), abstraction ($\lam{x}{x}$), application ($e~e$), arithmetic
-expressions ($e \odot e$) where $\odot$ ranges over $\{+,-,*,/\}$, raising
-exceptions ($\throw{l}{e}$), performing effects ($\perform{l}{e}$), and
-handling effects ($\handle{e}{h}$). Abstractions come in two forms: OCaml
-abstractions ($\lambda^o$) and C abstractions ($\lambda^c$). The handler
-consists of a return case ($\caseval{x}{e}$), zero or more exception cases
-($\caseexn{l}{x}{e}$) with label $l$, parameter $x$ and body $e$, and zero or
-more effect cases ($\caseeff{l}{x}{k}{e}$) with label $l$, parameter $x$,
-continuation $k$ and body $e$.
-
-The operational semantics is an extension of a CEK machine
-semantics~\cite{Felleisen86} for effect handlers, following the abstract
-machine semantics of Hillerstrom et al.~\cite{Hillerstrom20}. The key
-difference from Hillerstrom et al. is that our stacks are composed of
-alternating sequence of OCaml and C stack segments. The program state is
-captured as configuration $\config \coloneqq \configa{\tau}{\env}{\stack}$ with
-the current term $\tau$ under evaluation, its environment $\env$ and the
-current stack $\stack$. The term is either an expression $e$ or a value $v$.
-The value are integer constants $n$, continuations $k$, closure
-$\clos{x}{e}{\env}$, effect being performed ($\effval{l}{k}$) and exception
-being raised ($\exnval{l}$). The environment is a map from variables to values.
-
-The stack $\stack$ is either a C stack ($\cstack$) or an OCaml stack
-($\ostack$). Every C stack $\cstacka{\fl}{\ostack}$ consists of a list of
-frames $\fl$, and the OCaml stack $\ostack$ under it. Every OCaml stack is
-either empty $\ostackemp$ or non-empty $\ostacka{k}{\cstack}$ with the C stack
-$\cstack$ under it. Thus, the program stack is an alternating sequence of C and
-OCaml stacks terminating with an empty OCaml stack $\ostackemp$. The frame list
-$\fl$ is composed of individual frames $r$, which is one of an argument frame
-$\farg{e}{\env}$ with the expression $e$ at the argument position of an
-application with its environment $\env$, a function frame $\ffun{v}$ with the
-value $v$ at the function position of an application, and frames for evaluating
-the arguments of an arithmetic expression.
-
-Continuations $k$ are either empty $[]$ or non-empty sequence of \emph{fibers}.
-A fiber $\fiber \coloneqq (\fl,\hc)$ is a list of frames $\fl$ and a handler
-closure $\hc \coloneqq (h,\env)$, which is a pair of handler $h$ and its
-environment $\env$.
-
-\subsubsection{Top-level reductions}
-
-The initial configuration for an expression $e$ is
-$\configa{e}{\emptyset}{\cstacka{[]}{\bullet}}$, where the environment and the
-stack are empty. The top-level reductions (Figure ~\ref{sem:toplevel}) can be
-performed by either by taking a C step $\cstep$ or an OCaml step $\ostep$.
-
-\subsubsection{C reductions}
-
-We can take a C step (Figure~\ref{sem:cstep}) by taking an administrative
-reduction step $\rightsquigarrow$. The administrative reductions are common to
-both C and OCaml. The rules \textsc{Var}, \textsc{Arith1}, \textsc{Arith2},
-\textsc{App1}, \textsc{App2} and \textsc{App3} are standard. \textsc{Arith3}
-performs the arithmetic operation on the integers ($\llbracket n_1 \odot n_2
-\rrbracket$). \textsc{Raise} pushes an function frame with exception value to
-indicate that an exception is being raised. Similarly, \textsc{Perform} pushes
-a function frame with an effect value with an empty continuation
-$[[],(\{\caseval{x}{x}\},\emptyset)]$ with no captured frames and an empty
-handler with an identity return case alone. We shall return to \textsc{Resume1}
-and \textsc{Resume2} later.
-
-Continuing with the rest of the C reduction steps, \textsc{CallC} captures the
-behaviour of calling a C function. Since the program is currently executing C,
-we can perform the call on the current stack. In case the abstraction is an
-OCaml abstraction (\textsc{Callback}), we create an OCaml stack with the C
-stack as its tail, with the current continuation being empty. This captures the
-behaviour of calling back into OCaml from C. \textsc{RetToO} returns a value to
-the enclosing OCaml stack. \textsc{ExnFwdO} forwards a raised exception to the
-enclosing OCaml stack, unwinding the rest of the frames. This captures the
-semantics of raising OCaml exceptions from C.
-
-\subsubsection{OCaml reductions}
-
-\begin{figure*}
+\end{minipage}
+%
+\begin{minipage}{\linewidth}
   \begin{smathpar}
@@
   \end{smathpar}
-  \vspace{-1mm}
-  \mycaption{OCaml Reductions -- $(\term, \env, k, \cstack) \ostep \config$.}
+  \subcaption{OCaml Reductions -- $(\term, \env, k, \cstack) \ostep \config$.}
   \label{sem:ostep}
+\end{minipage}
+\caption{Operational semantics of Multicore OCaml effect handlers.}
 \end{figure*}
 
-In OCaml (Figure~\ref{sem:ostep}), reductions always occur on the top-most
-fiber in the current stack. \textsc{AdminO} performs administrative reductions.
-\textsc{CallO} evaluates an OCaml function on the current stack.
-\textsc{ExtCall} captures the behaviour of external calls, which are evaluated
-on an empty C stack. \textsc{RetToC} returns a value to the enclosing C stack.
-In this case, we have exactly one fiber on the stack, and this was created in
-the rule \textsc{Callback}, whose handler has identity return case alone and
-the environment is empty. \textsc{RetFib} returns value from a fiber to the
-previous one, evaluating the body of the return case.
+\subsection{Dynamic semantics}
+
+We present an operational semantics for a core language of effect handlers that
+faithfully captures the semantics of the Multicore OCaml implementation.
+
+\textbf{\textit{Syntax.}} Our expressions (Figure~\ref{sem:syntax}) consists of
+integer constants ($n$), variables ($x$), abstraction ($\lam{x}{x}$),
+application ($e~e$), arithmetic expressions ($e \odot e$) where $\odot$ ranges
+over $\{+,-,*,/\}$, raising exceptions ($\throw{l}{e}$), performing effects
+($\perform{l}{e}$), and handling effects ($\handle{e}{h}$). Abstractions come
+in two forms: OCaml abstractions ($\lambda^o$) and C abstractions
+($\lambda^c$). The handler consists of a return case ($\caseval{x}{e}$), zero
+or more exception cases ($\caseexn{l}{x}{e}$) with label $l$, parameter $x$ and
+body $e$, and zero or more effect cases ($\caseeff{l}{x}{k}{e}$) with label
+$l$, parameter $x$, continuation $k$ and body $e$.
+
+The operational semantics is an extension of a CEK machine
+semantics~\cite{Felleisen86} for effect handlers, following the abstract
+machine semantics of Hillerstrom et al.~\cite{Hillerstrom20}. The key
+difference from Hillerstrom et al. is that our stacks are composed of
+alternating sequence of OCaml and C stack segments. The program state is
+captured as configuration $\config \coloneqq \configa{\tau}{\env}{\stack}$ with
+the current term $\tau$ under evaluation, its environment $\env$ and the
+current stack $\stack$. The term is either an expression $e$ or a value $v$.
+The value are integer constants $n$, continuations $k$, closure
+$\clos{x}{e}{\env}$, effect being performed ($\effval{l}{k}$) and exception
+being raised ($\exnval{l}$). The environment is a map from variables to values.
+
+The stack $\stack$ is either a C stack ($\cstack$) or an OCaml stack
+($\ostack$). Every C stack $\cstacka{\fl}{\ostack}$ consists of a list of
+frames $\fl$, and the OCaml stack $\ostack$ under it. Every OCaml stack is
+either empty $\ostackemp$ or non-empty $\ostacka{k}{\cstack}$ with the C stack
+$\cstack$ under it. Thus, the program stack is an alternating sequence of C and
+OCaml stacks terminating with an empty OCaml stack $\ostackemp$. The frame list
+$\fl$ is composed of individual frames $r$, which is one of an argument frame
+$\farg{e}{\env}$ with the expression $e$ at the argument position of an
+application with its environment $\env$, a function frame $\ffun{v}$ with the
+value $v$ at the function position of an application, and frames for evaluating
+the arguments of an arithmetic expression.
+
+Continuations $k$ are either empty $[]$ or non-empty sequence of \emph{fibers}.
+A fiber $\fiber \coloneqq (\fl,\hc)$ is a list of frames $\fl$ and a handler
+closure $\hc \coloneqq (h,\env)$, which is a pair of handler $h$ and its
+environment $\env$.
+
+\textbf{\textit{Top-level reductions.}} The initial configuration for an expression
+$e$ is $\configa{e}{\emptyset}{\cstacka{[]}{\bullet}}$, where the environment
+and the stack are empty. The top-level reductions (Figure ~\ref{sem:toplevel})
+can be performed by either by taking a C step $\cstep$ or an OCaml step
+$\ostep$.
+
+\textbf{\textit{C reductions.}} We can take a C step (Figure~\ref{sem:cstep}) by
+taking an administrative reduction step $\rightsquigarrow$. The administrative
+reductions are common to both C and OCaml. The rules \textsc{Var},
+\textsc{Arith1}, \textsc{Arith2}, \textsc{App1}, \textsc{App2} and
+\textsc{App3} are standard. \textsc{Arith3} performs the arithmetic operation
+on the integers ($\llbracket n_1 \odot n_2 \rrbracket$). \textsc{Raise} pushes
+an function frame with exception value to indicate that an exception is being
+raised. Similarly, \textsc{Perform} pushes a function frame with an effect
+value with an empty continuation $[[],(\{\caseval{x}{x}\},\emptyset)]$ with no
+captured frames and an empty handler with an identity return case alone. We
+shall return to \textsc{Resume1} and \textsc{Resume2} later.
+
+Continuing with the rest of the C reduction steps, \textsc{CallC} captures the
+behaviour of calling a C function. Since the program is currently executing C,
+we can perform the call on the current stack. In case the abstraction is an
+OCaml abstraction (\textsc{Callback}), we create an OCaml stack with the C
+stack as its tail, with the current continuation being empty. This captures the
+behaviour of calling back into OCaml from C. \textsc{RetToO} returns a value to
+the enclosing OCaml stack. \textsc{ExnFwdO} forwards a raised exception to the
+enclosing OCaml stack, unwinding the rest of the frames. This captures the
+semantics of raising OCaml exceptions from C.
+
+\textbf{\textit{OCaml reductions.}} In OCaml (Figure~\ref{sem:ostep}), reductions
+always occur on the top-most fiber in the current stack. \textsc{AdminO}
+performs administrative reductions. \textsc{CallO} evaluates an OCaml function
+on the current stack. \textsc{ExtCall} captures the behaviour of external
+calls, which are evaluated on an empty C stack. \textsc{RetToC} returns a value
+to the enclosing C stack. In this case, we have exactly one fiber on the stack,
+and this was created in the rule \textsc{Callback}, whose handler has identity
+return case alone and the environment is empty. \textsc{RetFib} returns value
+from a fiber to the previous one, evaluating the body of the return case.
 
@@
     (*p1*) perform E (*p3*)
-  with | v -> v | effect F kf -> ...
+	with | v -> v | effect F kf -> ()
 with | v -> v
-		 | effect E ke ->
-		     (*p2*) continue ke ()
+		 | effect E ke -> (*p2*) continue ke ()
     \end{lstlisting}
@@
 selective CPS translation. These language are equipped with an effect system,
-which allows compiling pure code in direct-style and effectful code in
-continuation passing style. Leijen et al.~\cite{Leijen14} implement effect
-handlers as a library in C using stack copying. Since C allows pointers into
-the stack, care is taken to ensure that when continuations are resumed, the
-constituent frames are restored to the same memory addresses as at the time of
-capture. Kiselyov et al.~\cite{Kiselyov18} use an implementation of
-multi-prompt delimited continuations as an OCaml library~\cite{Kiselyov12} to
-embed Eff language in OCaml. Indeed, Forester et al.~\cite{Forster19} showed
-that in an untyped setting, effect handlers, monadic reflection and delimited
-control can macro-express each other.
+which allows compiling pure code in direct-style and effectful code in CPS.
+Leijen~\cite{Leijen14} implements effect handlers as a library in C using stack
+copying. Since C allows pointers into the stack, care is taken to ensure that
+when continuations are resumed, the constituent frames are restored to the same
+memory addresses as at the time of capture. Kiselyov et al.~\cite{Kiselyov18}
+use an implementation of multi-prompt delimited continuations as an OCaml
+library~\cite{Kiselyov12} to embed Eff language in OCaml. Indeed, Forester et
+al.~\cite{Forster19} showed that in an untyped setting, effect handlers,
+monadic reflection and delimited control can macro-express each other.
 
@@
 presence of non-local control operators. Chez Scheme supports continuation
-marks~\cite{Flatt20} which permit stack inspection to implement dynamic
-binding, exceptions, profilers, debuggers, etc, in the presence of first-class
-continuations. As the authors note, continuation marks can be implemented using
-effect handlers, but direct support for continuation marks leads to better
-performance. In this work, we focus on retaining the support for stack
-inspection through DWARF unwind tables in the presence of effect handlers.
+marks~\cite{Flatt20} which permit stack inspection as a language feature. This
+enables implementation of dynamic binding, exceptions, profilers, debuggers,
+etc, in the presence of first-class continuations. As the authors note,
+continuation marks can be implemented using effect handlers, but direct support
+for continuation marks leads to better performance. In this work, we focus on
+retaining the support for stack inspection through DWARF unwind tables in the
+presence of effect handlers.
 
@@
 Thus, the linear use of continuations enabled by the |discontinue| primitive
-ensures backwards compatibility of legacy OCaml systems code with non-local
-control flow introduced by effect handlers.
-
-Leijen~\cite{Leijen18} explicitly extends effect handlers with |initially| and
-|finally| clauses in Koka for resource safety. Sivaramakrishnan et
-al.~\cite{TFP17} describe the interaction of effect handlers and asynchronous
-exceptions. This is orthogonal to the contributions of this paper. Our focus is
-the compiler and runtime system support for implementing effect handlers.
-
-Languages with effect handlers typically allow continuations to be resumed more
-than once to enable use cases such as backtracking search and probabilistic
-programming~\cite{Pyro}. Multicore OCaml has an experimental feature for
-explicitly cloning continuations, which makes a copy of each of its constituent
-fibers. Multi-shot continuations are difficult to reason about in the presence
-of linear resources such as file descriptors. Recall that OCaml programs are
-written with the expectation that function calls return \emph{exactly-once}
-either normally or exceptionally. With multi-shot continuations, they may
-return \emph{more than once}, leading to errors such as double-freeing
-malloced memory. Since our goal is to retrofit effect handlers onto OCaml, we
-relegate continuation cloning to the OCaml |Obj| module, which contains unsafe
-features.
+ensures backwards compatibility of legacy OCaml systems code under non-local
+control flow introduced by effect handlers. Leijen~\cite{Leijen18} explicitly
+extends effect handlers with |initially| and |finally| clauses in Koka for
+resource safety.
+
+Sivaramakrishnan et al.~\cite{TFP17} describe the interaction of effect
+handlers and asynchronous exceptions. This is orthogonal to the contributions
+of this paper. Our focus is the compiler and runtime system support for
+implementing effect handlers.
 
@@
 
+\subsection*{Experimental support for multi-shot continuations}
+
+Languages with effect handlers typically allow continuations to be resumed more
+than once to enable use cases such as backtracking search and probabilistic
+programming (Uber's Pyro langauge~\cite{Pyro}). Multicore OCaml has an
+experimental feature for explicitly cloning continuations, which makes a copy
+of each of its constituent fibers. Multi-shot continuations are difficult to
+reason about in the presence of linear resources such as file descriptors.
+Recall that OCaml programs are written with the expectation that function calls
+return \emph{exactly-once} either normally or exceptionally. With multi-shot
+continuations, they may return \emph{more than once}, leading to errors such as
+double-freeing malloced memory. Since our goal is to retrofit effect handlers
+onto OCaml, we relegate continuation cloning to the OCaml |Obj| module, which
+contains unsafe features.
+
 \subsection*{No effect macro benchmarks}

### 2020-11-12 retro-concurrency.tex
Updates

@@
 
-\vspace{-2mm}
 \section{Semantics}
@@
 
+\vspace{-5mm}
 \subsection{Dynamic semantics}
@@
 
+\vspace{-3mm}
 \subsection{External calls and callbacks}
@@
 
+\vspace{-2mm}
 \subsection{Stack unwinding}
@@
 program from \S\ref{sec:unwind}, modulo runtime system functions due to effect
-handlers. We have utilized a DWARF verification tool~\cite{Bastian19} with
-extensions to handle PIE and additional DWARF primitives to flag possible errors
-in the hand written DWARF directives where we have altered the OCaml assembler
-to implement effects. In three places additional manual verification was
-required; two were where the stack is intentionally switched between fibers and
-one was where the return to the parent stack on child stack completion could not
-track stack resizes and intermediate effect handler frames.
-
+handlers.
+
+We have utilized a DWARF verification tool~\cite{Bastian19} with extensions to
+handle PIE and additional DWARF primitives to flag possible errors in the hand
+written DWARF directives where we have altered the OCaml assembler to implement
+effects. In three places additional manual verification was required; two were
+where the stack is intentionally switched between fibers and one was where the
+return to the parent stack on child stack completion could not track stack
+resizes and intermediate effect handler frames.
+
+Even with correct DWARF unwind information, using DWARF to record call stack
+information in |perf| only captures the call stack of the current fiber in
+Multicore OCaml. Since stack unwinding using DWARF is slow due to bytecode
+interpretation overhead, |perf| dumps the (user) call stack when
+sampled~\cite{Bastian19}. This only includes the frames from the current fiber.
+This is a limitation of |perf| and not of our stack layout.
+
+\vspace{-2mm}
 \subsection{Garbage collection}
@@
 
+\vspace{-3mm}
 \section{Evaluation}
@@
 
+\vspace{-3mm}
 \subsection{No effects benchmarks}
@@
 
+\vspace{-3mm}
 \subsection{No perform benchmarks}
@@
 
+\vspace{-3mm}
 \subsection{Concurrent benchmarks}
@@
 
+\vspace{-3mm}
 \section{Related Work}
@@
 negligible overheads for existing OCaml programs and good performance for those
-that use effect handlers. Our evaluation shows that this is indeed the case. We
-plan to upstream the compiler and runtime support for effect handlers followed
-by an effect system.
+that use effect handlers. Our evaluation shows that this is indeed the case.
 
@@
 
+In addition, multi-shot continuations currently are not safe under compiler
+optimisations. For example consider the following program:
+\begin{lstlisting}
+effect E : unit
+
+let foo () = perform E
+
+let bar f =
+  let r = ref 0 in
+	f ();
+  r := !r + 1;
+  !r
+
+let _ = match bar foo with
+| v -> v
+| effect E k ->
+    let k' = Obj.clone_continuation k in
+    let r1 = continue k () in
+    let r2 = continue k' () in
+    Printf.printf "%d %d\n" r1 r2;
+		0
+\end{lstlisting}
+This program is expected to print |1 2|. However, OCaml optimizes |bar| by
+converting the heap allocation of |r| to a stack allocation. As a result, each
+continuation ends up with a copy of |r|, and the program prints |1 1|. In the
+presence of multi-shot continuations, compiler optimisations need to be made
+aware of the effects of a function. We leave proper support for multi-shot
+continuations to future work.
+
 \subsection*{No effect macro benchmarks}

### 2020-11-12 retro-concurrency.tex
abstract

@@
 \begin{abstract}
-  Effect handlers have gained immense interest in the recent years as a means
+Effect handlers have been gathering momentum as a mechanism
 for modular programming with user-defined effects. The runtime semantics of
-effect handlers is particularly attractive as they allow advanced control flow
-mechanisms such as first-class continuations, generators, async/await,
-coroutines to be composably expressed. This paper presents a design, a
-full-fledged implementation and an evaluation of effect handlers for OCaml.
-OCaml is industrial-strength multi-paradigm programming language that tends to
-be widely used for systems programming, our implementation strives to maintain
-the existing performance profile and toolchain compatibility. Retrofitting
-effect handlers onto OCaml is challenging since OCaml does not have any
-non-local control flow mechanisms other than exceptions. Our implementation of
-effect handlers for OCaml imposes negligible overhead on code that does not use
-effect handlers, remains compatible with program analysis tools that inspect the
-stack and is efficient for code that makes use of effect handlers.
+effect handlers allow for control flow
+mechanisms such as first-class continuations, generators, async/await and coroutines to be composably expressed. We present a design and evaluate a
+full-fledged efficient implementation of effect handlers for OCaml, an industrial-strength multi-paradigm programming language.
+Our implementation strives to maintain the existing performance profile of conventional OCaml code, particularly in the domain of systems and toolchain programming.
+Retrofitting effect handlers onto OCaml is challenging since OCaml does currently not have any non-local control flow mechanisms other than exceptions.
+Our implementation of
+effect handlers for OCaml: {\em (i)} imposes negligible overhead on code that does not use
+effect handlers; {\em (ii)} remains compatible with program analysis tools that inspect the
+stack; and {\em (iii)} is efficient for new code that makes use of effect handlers.
 \end{abstract}

### 2020-11-12 retro-concurrency.tex
iintro edits

@@
 
-Effect Handlers~\cite{Plotkin09} provide a modular foundation for user-defined
+Effect handlers~\cite{Plotkin09} provide a modular foundation for user-defined
 effects. The key idea is to separate the definition of the effectful operations
@@
 
-What makes effect handlers quite powerful is that they provide a structured
-form of delimited continuations~\cite{Danvy90}, which enables advanced
+Effect handlers provide a structured
+form of delimited continuations~\cite{Danvy90} that enables
 non-local control-flow mechanisms such as resumable exceptions, coroutines,
-generators and asynchronous I/O to be composably expressed. As an example,
-consider the task of handling division-by-zero case in a calculator
+generators and asynchronous I/O to be composably expressed. For example,
+consider the task of handling division-by-zero in a calculator
 application. Different calculators handle division-by-zero differently. In the
-iOS 13 Calculator App (|iOS|), division-by-zero returns |Error|. In the Google
-search calculator App (|gsearch|) |0/0| returns to |Error|, and for any other
-|n| not equal to |0|, |n/0| returns |Infinity| with the same sign as |n|. How
-can we capture these different behaviours without changing the evaluator?
+iOS 13 Calculator (|iOS|), |0/0| returns |Error|. In the Google
+search calculator (|gsearch|) |0/0| returns |Error|, but for
+|n!=0| the |n/0| returns |Infinity| with the same sign as |n|.
+How do we capture all these different behaviours without changing the evaluator?
 
@@
 problem. Unlike exceptions where the library decides whether an error is fatal
-or not, conditions allow the client of the library to make that decision. If
-the error is non-fatal, Conditions provide a \emph{restart} mechanism to resume
+or not, conditions allow the client of the library to make that decision.
+Conditions provide a \emph{restart} mechanism for non-fatal errors to resume
 execution from the point where the condition was signalled. On
 division-by-zero, the calculator signals the handler and lets the handler
-decide the course of action based on the desired semantics. We can implement
-this mechanism using effect handlers as shown below:
+decide the course of action based on the desired semantics. We now implement
+this using effect handlers in OCaml syntax:
 
@@
 
-On line 4, we declare an effect |DivBy0| that is parameterised by an integer,
+Line 4 declares an effect |DivBy0| that is parameterised by an integer,
 which when performed returns the result of division-by-zero. If the denominator
-is 0, then we perform |DivBy0| effect with the numerator value (line 12) so that
-the context may decide the return value. We can get the |iOS| semantics by
-wrapping the evaluator with the following handler:
+is 0, then we perform the |DivBy0| effect with the numerator value (line 12) so that
+the context may decide the return value. We obtain the |iOS| semantics by
+wrapping the evaluator with this handler:
 \begin{lstlisting}
@@
 If the evaluator returns a value, we simply return that. If the evaluator
-performs the |DivBy0| effect, we catch the effect in the effect case along with
+performs the |DivBy0| effect, we catch the effect in the pattern match along with
 the continuation |k| of the corresponding |perform| \emph{delimited} by this
-handler. In the case of |iOS|, we |continue| the suspended computation with
-|Error| value. Hence, the computation |ios(1/0)| returns |Error|. On the
-other hand, we can get the |gsearch| semantics by enclosing the evaluator in the
-following handler:
+handler. For |iOS|, we |continue| the suspended computation with an
+|Error| value and the computation |ios(1/0)| returns |Error|.
+We can get the |gsearch| semantics by enclosing the evaluator in this handler:
 \begin{lstlisting}
@@
 \end{lstlisting}
-The computation |gsearch(1/0)| returns positive infinity (|Inf P|).
+Executing |gsearch(1/0)| returns positive infinity (|Inf P|).
 

### 2020-11-12 retro-concurrency.tex
sec1

@@
 applications~\cite{Madhavapeddy13}. OCaml is particularly favoured for its
-competitive yet predictable performance, with fast foreign-function interface
-(FFI) and excellent compatibility with program analysis tools such as debuggers
+competitive yet predictable performance, with a fast foreign-function interface
+(FFI). It has excellent compatibility with program analysis tools such as debuggers
 and profilers that utilise DWARF stack unwind tables~\cite{DWARF} to obtain a
 backtrace. Our primary motivation for adding effect handlers to OCaml is to
-support (1) highly concurrent applications such as web servers using a
-natural \emph{direct-style} as opposed to using callbacks and (2) efficient
+support: {\em (i)} highly concurrent applications such as web servers using a
+natural \emph{direct-style} as opposed to using callbacks; and {\em (ii)} efficient
 nested parallel programs in the vein of Cilk~\cite{Cilk}, OpenMP~\cite{OpenMP},
@@
 
-OCaml, however, does not support any non-local control flow mechanisms other
+OCaml currently does not support any non-local control flow mechanisms other
 than exceptions. This makes it particularly challenging to implement the
@@
 reifying continuations in the source language is relatively straight-forward.
-But OCaml does not use CPS IR, and changing the OCaml compiler to utilise CPS IR
-would be an enormous undertaking. Utilizing a CPS IR would also affect the
-performance profile of OCaml applications due to increased memory allocations as
+But OCaml does not use a CPS IR, and changing the OCaml compiler to utilise a CPS IR
+would be an enormous undertaking that would affect the
+performance profile of existing OCaml applications due to increased memory allocations as
 continuation closures get allocated on the heap~\cite{Farvardin20}. Moreover,

### 2020-11-12 retro-concurrency.tex
Updates

@@
 \S\ref{sec:static_semantics}; Multicore OCaml supports exception patterns in
-addition to effect patterns in the same handler. This is followed by a block to
-support DWARF unwinding, whose details we will return to later. Then, there is
-a top-level handler that invokes the exception handler |clos_hexn|
-after switching to the parent fiber. This is followed by the |pc| of the code
-that invokes the value handler |clos_hval| after switching to the parent stack. The
+addition to effect patterns in the same handler. This is followed by a context
+block needed for DWARF and GC bookkeeping with callbacks. Then, there is a
+top-level handler that invokes the exception handler |clos_hexn| after
+switching to the parent fiber. This is followed by the |pc| of the code that
+invokes the value handler |clos_hval| after switching to the parent stack. The
 stack is laid out such that when the computation in the handler returns,
@@
 dereference the |saved_sp| to get the CFA (|saved_sp + 8|). During callbacks
-into OCaml, we save the current system stack pointer in the |DWARF block| in
+into OCaml, we save the current system stack pointer in the |context block| in
 Figure~\ref{fig:fiber} to identify the CFA in the C stack. DWARF unwinding for
@@
 also include a Go version (|go|) that uses \texttt{net/http}~\cite{nethttp}
-package.
+package. As both OCaml versions are single threaded, the Go benchmark is run with
+\texttt{GOMAXPROCS=1}.
 

### 2020-11-12 retro-concurrency.tex
add note about GOMAXPROCS
include go version number
Updates

@@
 
-\vspace{-5mm}
 \subsection{Dynamic semantics}
@@
 
-\vspace{-2mm}
 \subsection{Heap-allocated fibers}
@@
 
-\vspace{-3mm}
 \subsection{External calls and callbacks}
@@
 
-\vspace{-2mm}
 \subsection{Effect handlers}
@@
 
-\vspace{-2mm}
+\vspace{-2.5mm}
 \subsection{Stack unwinding}
@@
 
-\vspace{-2mm}
 \subsection{Garbage collection}
@@
 
-\vspace{-3mm}
 \section{Evaluation}
@@
 
-\textbf{\textit{Web server.}} As one of our primary goals of adding effect handlers to
-OCaml is to support direct-style concurrency, we implement a full-fledged
-HTTP/1.1 web server using effect handlers (|MC|). We use effect handlers to
-implement a user-level threading library, which spawns a thread per request. We
-use \texttt{httpaf} library~\cite{httpaf} for HTTP handling, and
+\textbf{\textit{Web server.}} As one of our primary goals of adding effect
+handlers to OCaml is to support direct-style concurrency, we implement a
+full-fledged HTTP/1.1 web server using effect handlers (|MC|). We use effect
+handlers to implement a user-level threading library, which spawns a thread per
+request. We use \texttt{httpaf} library~\cite{httpaf} for HTTP handling, and
 \texttt{libev}~\cite{libev} for the eventloop. We compare our implementation
@@
 style, and does not have a notion of thread per request. For comparison, we
-also include a Go version (|go|) that uses \texttt{net/http}~\cite{nethttp}
-package. As both OCaml versions are single threaded, the Go benchmark is run with
-\texttt{GOMAXPROCS=1}.
+also include a Go 1.13 version (|go|) that uses
+\texttt{net/http}~\cite{nethttp} package. As both the OCaml versions are single
+threaded, the Go benchmark is run with \texttt{GOMAXPROCS=1}.
 
@@
 optimisations. For example consider the following program:
+
+\begin{minipage}{\linewidth}
 \begin{lstlisting}
 effect E : unit
-
 let foo () = perform E
-
 let bar f =
@@
   !r
-
-let _ = match bar foo with
+;;
+match bar foo with
 | v -> v
@@
 \end{lstlisting}
+\end{minipage}
+
 This program is expected to print |1 2|. However, OCaml optimizes |bar| by
 converting the heap allocation of |r| to a stack allocation. As a result, each
-continuation ends up with a copy of |r|, and the program prints |1 1|. In the
-presence of multi-shot continuations, compiler optimisations need to be made
-aware of the effects of a function. We leave proper support for multi-shot
+continuation ends up with a copy of |r|, and the program prints |1 1|. Thus, in
+the presence of multi-shot continuations, compiler optimisations need to be
+made aware of the effects of a function. We leave proper support for multi-shot
 continuations to future work.

### 2020-11-13 retro-concurrency.tex
updates

@@
 In this section, we present the implementation details of effect handlers in
-Multicore OCaml. The implementation follows the operational semantics, but has
-a few key representational differences. Multicore OCaml currently only supports
-the x86-64 backend on Linux, but our design does not preclude other backends.
-We assume x86-64 architecture in the sequel.
-
-Unlike the operational semantics, handlers with just exception patterns
-(exception handlers) are implemented differently from effect handlers. As
-mentioned in \S~\ref{sec:exn_handlers}, exceptions are pervasive in OCaml and
-are so cheap that they are used for local control flow. Hence, we retain the
-linked exception handler frame implementation of stock OCaml exception handlers
-in Multicore OCaml for performance backwards compatibility. This differs from
-other research languages with effect handlers~\cite{Hillerstrom20,Frank,Eff},
-which implement exceptions using effects (by ignoring the continuation argument
-in the handler).
+Multicore OCaml. Multicore OCaml currently only supports the x86-64 backend on
+Linux, but our design does not preclude other backends. We assume x86-64
+architecture in the sequel.
+
+\subsection{Exceptions}
+
+The implementation follows the operational semantics, but has a few key
+representational differences. Unlike the operational semantics, handlers with
+just exception patterns (exception handlers) are implemented differently than
+effect handlers. As mentioned in \S~\ref{sec:exn_handlers}, exceptions are
+pervasive in OCaml and are so cheap that they are used for local control flow.
+Hence, we retain the linked exception handler frame implementation of stock
+OCaml exception handlers in Multicore OCaml for performance backwards
+compatibility. This differs from other research languages with effect
+handlers~\cite{Hillerstrom20,Frank,Eff}, which implement exceptions using
+effects (by ignoring the continuation argument in the handler).
 
@@
 cases, continuations will be resumed at-most once, and copying fibers is
-unnecessary and inefficient. Instead, Multicore OCaml allocates fibers on the
-heap. Figure~\ref{fig:fiber} shows the layout of a fiber in Multicore OCaml.
-
-At the bottom of the stack, we have the |handler_info|, which contains the
-pointer to parent fiber, and the closures for the value, exception and effect
-cases. The closures are created by the translation described in
+unnecessary and inefficient. Instead, Multicore OCaml optimises fibers for
+one-shot continuations.
+
+Fibers are allocated on the heap using |malloc| and are |free|d when no longer
+necessary. We do not use a \emph{stack cache} of recently freed stacks for
+speeding up fiber allocation~\cite{Farvardin20}. In our experiments, we
+observed that directly using |mimalloc|~\cite{Leijen19} outperforms a stack
+cache.
+
+Figure~\ref{fig:fiber} shows the layout of a fiber in Multicore
+OCaml. At the bottom of the stack, we have the |handler_info|, which contains
+the pointer to parent fiber, and the closures for the value, exception and
+effect cases. The closures are created by the translation described in
 \S\ref{sec:static_semantics}; Multicore OCaml supports exception patterns in
@@
 We have the variable-sized area for the OCaml frames next. Since we want the
-handlers to be as small as possible, this area is initially 16 words in length.
-When the stack pointer |rsp| becomes less than the stack threshold (maintained in the
-|Caml_state| table), the stack is said to have overflown. On stack overflow, we
-copy the whole fiber to a new area with double the size. In Multicore OCaml, we
-introduce stack overflow checks into the function prologue of OCaml functions.
-We observed that most function calls are to leaf functions with small frame
-size. In order to take advantage of this fact, we introduce a small red zone at
-the top of the stack. The compiler elides the stack overflow check for leaf
-functions whose frame size is less than the size of the red zone.
+fibers to be as small as possible, this area is initially 16 words in length.
+When the stack pointer |rsp| becomes less than the stack threshold (maintained
+in the |Caml_state| table), the stack is said to have overflown. On stack
+overflow, we copy the whole fiber to a new area with double the size. In
+Multicore OCaml, we introduce stack overflow checks into the function prologue
+of OCaml functions. In our evaluation of real world OCaml programs, we observed
+that most function calls are to leaf functions with small frame size. In order
+to take advantage of this fact, we introduce a small red zone at the top of the
+stack. The compiler elides the stack overflow check for leaf functions whose
+frame size is less than the size of the red zone.
 
@@
 Since OCaml does not generate pointers into the stack, the two |fiber_info|
-fields are the only ones that need to be updated when fibers are moved. We
-allocate the space for the fiber using |malloc| and |free| it when no longer
-required. We do not use a \emph{stack cache} of recently freed stacks for
-speeding up fiber allocation~\cite{Farvardin20}. In our experiments, we
-observed that directly using |mimalloc|~\cite{Leijen19} outperforms a stack
-cache.
+fields are the only ones that need to be updated when fibers are moved.
 
@@
 external calls in the system stack. Calling a C function from OCaml involves
-saving the stack pointer in the current fiber, saving the
-allocation pointer in the |Caml_state|, updating |rsp| to the top of system
-stack (maintained in |Caml_state|), and calling the C function. The actions are
-reversed when returning from the external call. For C
-functions that take arguments on the stack, the arguments must be copied to the
-C stack from the OCaml stack where they are laid out. In the vast majority of
-external calls, we perform only a few additional instructions over stock OCaml.
+saving the stack pointer in the current fiber, saving the allocation pointer in
+the |Caml_state|, updating |rsp| to the top of system stack (maintained in
+|Caml_state|), and calling the C function. The actions are reversed when
+returning from the external call. For C functions that take arguments on the
+stack, the arguments must be copied to the C stack from the OCaml stack where
+they are laid out. In the vast majority of external calls, we perform only a
+few additional instructions over stock OCaml.
 
@@
 current continuation closure on the parent fiber |main| with |E| and |ke| as
-arguments, which handles |E| (position |p2|). Thus, continuations are captured
-without copying frames. Since every handler closure is evaluated until a
-machine one is found, the time taken to handle an effect is linear in the
-number of handlers. We observe that the handler stack is shallow in real
+arguments, which handles |E| (position |p2|). Thus, \emph{continuations are
+captured without copying frames}. Since every handler closure is evaluated
+until a matching one is found, the time taken to handle an effect is linear in
+the number of handlers. We observe that the handler stack is shallow in real
 programs.
@@
 
-\vspace{-2.5mm}
 \subsection{Stack unwinding}
@@
 	difference for Multicore OCaml over stock OCaml.}
-\vspace{-5mm}
+\vspace{-7mm}
 {
@@
 \end{tabular}
-\vspace{-3mm}
+\vspace{-5mm}
 }
@@
 \label{tab:micro_noperform}
-	\vspace{-4mm}
+	\vspace{-5mm}
 \end{table}
@@
 \textbf{48ns, 10ns, 13ns and 19ns}, respectively. For reference, the idle
-memory latency for the local NUMA node is 93.2ns as measured using the Intel
-MLC tool~\cite{mlc}.
+memory load latency for the local NUMA node is 93.2ns as measured using the
+Intel MLC tool~\cite{mlc}.
 
@@
 		\centering
-		\includegraphics[width=\linewidth]{http-benchmarks/throughput.pdf}
+		\includegraphics[width=0.9\linewidth]{http-benchmarks/throughput.pdf}
 		\subcaption{Throughput}
@@
 	\begin{minipage}{0.49\linewidth}
-		\includegraphics[width=\linewidth]{http-benchmarks/latency.pdf}
+		\includegraphics[width=0.9\linewidth]{http-benchmarks/latency.pdf}
 		\subcaption{Tail latency}
@@
 	\caption{Web server performance.}
-	\vspace{-4mm}
+	\vspace{-5mm}
 \end{figure}
@@
 1k open connections and perform requests for a static web page at different
-rates, and record the request service rate and latency. The throughput and tail
-latency graphs are given in Figures~\ref{grf:throughput} and~\ref{grf:latency}.
-In all the versions, the throughput plateaus at around 30k requests per second.
-We measure the tail latencies at 2/3rd of this rate (20k requests per second)
-to simulate optimal load. We observe that both of the OCaml versions remain
+rates, and record the service rate and latency. The throughput and tail latency
+graphs are given in Figures~\ref{grf:throughput} and~\ref{grf:latency}. In all
+the versions, the throughput plateaus at around 30k requests per second. We
+measure the tail latencies at 2/3rd of this rate (20k requests per second) to
+simulate optimal load. We observe that both of the OCaml versions remain
 competitive with |go|, and |MC| performs best in terms of tail latency.
@@
 
-\vspace{-3mm}
+\vspace{-3.5mm}
 \section{Related Work}

### 2020-11-13 retro-concurrency.tex
sec2

@@
   \item \textbf{Backwards compatibility} A well-behaved OCaml program does not
-    break under OCaml extended with effect handlers. OCaml code that does not
+    break under an OCaml extended with effect handlers. OCaml code that does not
     use effect handlers will pay negligible performance cost.
@@
 \noindent \textbf{R1: Backwards compatibility.} A well-behaved OCaml program does
-not break under OCaml extended with effect handlers. OCaml code that does not
-use effect handlers will pay negligible performance cost.
-
-\noindent \textbf{R2: Tool compatibility.} OCaml with effect handlers produce
-well-formed backtraces, and remains compatible with program analysis tools such
-as debuggers and profilers that inspect the stack using DWARF stack unwind
+not break under an OCaml that is extended with effect handlers. OCaml code that does not
+use effect handlers will pay negligible performance and memory cost.
+
+\noindent \textbf{R2: Tool compatibility.} OCaml programs with effect handlers produce
+well-formed backtraces and remain compatible with program analysis tools such
+as debuggers and profilers that inspect the stack using DWARF unwind
 tables.
 
-\noindent \textbf{R3: Effect handler efficiency.} In order to support
-highly-concurrent applications, the program must accommodate millions of
-continuations at the same time. Installing effect handlers, capturing and
+\noindent \textbf{R3: Effect handler efficiency.} The program must accommodate millions of
+continuations at the same time to support
+highly-concurrent applications. Installing effect handlers, capturing and
 resuming continuations must be fast.

### 2020-11-13 retro-concurrency.tex
Fix typo in code
Update abstract

@@
 \begin{abstract}
-Effect handlers have been gathering momentum as a mechanism
-for modular programming with user-defined effects. The runtime semantics of
-effect handlers allow for control flow
-mechanisms such as first-class continuations, generators, async/await and coroutines to be composably expressed. We present a design and evaluate a
-full-fledged efficient implementation of effect handlers for OCaml, an industrial-strength multi-paradigm programming language.
-Our implementation strives to maintain the existing performance profile of conventional OCaml code, particularly in the domain of systems and toolchain programming.
-Retrofitting effect handlers onto OCaml is challenging since OCaml does currently not have any non-local control flow mechanisms other than exceptions.
-Our implementation of
-effect handlers for OCaml: {\em (i)} imposes negligible overhead on code that does not use
-effect handlers; {\em (ii)} remains compatible with program analysis tools that inspect the
-stack; and {\em (iii)} is efficient for new code that makes use of effect handlers.
+Effect handlers have been gathering momentum as a mechanism for modular
+	programming with user-defined effects. Effect handlers allow for non-local
+	control flow mechanisms such as generators, async/await, lightweight threads
+	and coroutines to be composably expressed. We present a design and evaluate a
+	full-fledged efficient implementation of effect handlers for OCaml, an
+	industrial-strength multi-paradigm programming language. Our implementation
+	strives to maintain the backwards compatibility and existing performance
+	profile of conventional OCaml code. Retrofitting effect handlers onto OCaml
+	is challenging since OCaml does currently not have any non-local control flow
+	mechanisms other than exceptions. Our implementation of effect handlers for
+	OCaml: {\em (i)} imposes negligible overhead on code that does not use effect
+	handlers; {\em (ii)} remains compatible with program analysis tools that
+	inspect the stack; and {\em (iii)} is efficient for new code that makes use
+	of effect handlers.
 \end{abstract}
@@
 Gc.finalise (fun k ->
-  try ignore (discontinue k Unwind) | _ -> ()) k
+  try ignore (discontinue k Unwind) with _ -> ()) k
 \end{lstlisting}

### 2020-11-13 retro-concurrency.tex
tweak

--- 2020-11-13 retro-concurrency.tex
@@
 OCaml functions, irrespective of whether they use effect handlers, need to
-perform stack overflow checks and external C functions, which do not have stack
-overflow checks, must be performed on a separate system stack. Additionally, we
+perform stack overflow checks and external C functions (which do not have stack
+overflow checks) must be performed on a separate system stack. Additionally, we
 must generate DWARF stack unwind tables for stacks that may be non-contiguous.

### 2020-11-13 retro-concurrency.tex
cps

@@
 to use continuation-passing style (CPS) in the compiler's intermediate
-representation (IR)~\cite{Leijen14}. If the compiler were to use CPS, then
-reifying continuations in the source language is relatively straight-forward.
-But OCaml does not use a CPS IR, and changing the OCaml compiler to utilise a CPS IR
+representation (IR)~\cite{Leijen14}. 
+%If the compiler were to use CPS, then
+%reifying continuations in the source language is relatively straight-forward.
+OCaml does not use a CPS IR, and changing the compiler to utilise a CPS IR
 would be an enormous undertaking that would affect the
@@
 Our work is also timely. The WebAssembly~\cite{Hass17} community group is
-actively considering effect handlers as one of the mechanism for supporting
+considering effect handlers as one of the mechanisms for supporting
 concurrency, asynchronous I/O and generators~\cite{WasmProposal}. Project
-Loom~\cite{loom} is an OpenJDK project which adds virtual threads and delimited
-continuations to Java. Swift concurrency roadmap~\cite{swift} includes
+Loom~\cite{loom} is an OpenJDK project that adds virtual threads and delimited
+continuations to Java. The Swift roadmap~\cite{swift} includes
 direct-style asynchronous programming and structured concurrency as milestones.

### 2020-11-13 retro-concurrency.tex
tweaks

@@
 of DWARF unwind tables. OCaml allows compiling code with frame pointers, but
-they are not enabled by default. OCaml stack tends to be deep with small frames
+they are not enabled by default. The OCaml stack tends to be deep with small frames
 due to the pervasive use of recursive functions, not all of which are
@@
 Moreover, not using frame pointers saves two instructions in the function
-prologue and epilogue, and makes an extra register (|rbp| on x86) available.
+prologue and epilogue, and makes an extra register (|rbp| on x86\_64) available.
 Note that the DWARF unwind information is complementary to the information used
@@
 more than once. Hence, our continuations are one-shot, and resuming the
-continuation more than once raises |Invalid_argument| exception. It is
+continuation more than once raises an |Invalid_argument| exception. It is
 well-known that one-shot continuations can be implemented
@@
 just exception patterns (exception handlers) are implemented differently than
-effect handlers. As mentioned in \S~\ref{sec:exn_handlers}, exceptions are
+effect handlers. As mentioned in \S\ref{sec:exn_handlers}, exceptions are
 pervasive in OCaml and are so cheap that they are used for local control flow.

### 2020-11-15 retro-concurrency.tex
Edits
Updates

@@
 \label{sec:semantics}
+
+In this section, we formalise the effect handler design for Multicore OCaml.
 
@@
 continuation is neither continued nor discontinued. Since fibers are
-|malloc|ed and are |free|d when the computation being handled returns a value
+|malloc|ed, and are |free|d when the computation being handled returns a value
 or raises an exception, not resuming continuations leaks memory. In addition,
@@
 \end{lstlisting}
-Multicore OCaml does not enforce this behaviour in order to avoid installing an
-effect handler every time a continuation is captured.
+However, Multicore OCaml does not enforce this behaviour in order to avoid
+installing a finaliser every time a continuation is captured.
 

### 2020-11-16 retro-concurrency.tex
edits

@@
 	industrial-strength multi-paradigm programming language. Our implementation
-	strives to maintain the backwards compatibility and existing performance
-	profile of conventional OCaml code. Retrofitting effect handlers onto OCaml
-	is challenging since OCaml does currently not have any non-local control flow
-	mechanisms other than exceptions. Our implementation of effect handlers for
-	OCaml: {\em (i)} imposes negligible overhead on code that does not use effect
-	handlers; {\em (ii)} remains compatible with program analysis tools that
-	inspect the stack; and {\em (iii)} is efficient for new code that makes use
-	of effect handlers.
+	strives to maintain the backwards compatibility and performance profile of
+	existing OCaml code. Retrofitting effect handlers onto OCaml is challenging
+	since OCaml does not currently have any non-local control flow mechanisms
+	other than exceptions. Our implementation of effect handlers for OCaml: {\em
+	(i)} imposes negligible overhead on code that does not use effect handlers;
+	{\em (ii)} remains compatible with program analysis tools that inspect the
+	stack; and {\em (iii)} is efficient for new code that makes use of effect
+	handlers.
 \end{abstract}
@@
 effects. The key idea is to separate the definition of the effectful operations
-from their interpretations, which are defined by \emph{handlers} of the
-effects. For example,
+from their interpretations, which are given by \emph{handlers} of the effects.
+For example,
 %
@@
 
-Effect handlers provide a structured
-form of delimited continuations~\cite{Danvy90} that enables
-non-local control-flow mechanisms such as resumable exceptions, coroutines,
-generators and asynchronous I/O to be composably expressed. For example,
-consider the task of handling division-by-zero in a calculator
-application. Different calculators handle division-by-zero differently. In the
-iOS 13 Calculator (|iOS|), |0/0| returns |Error|. In the Google
-search calculator (|gsearch|) |0/0| returns |Error|, but for
-|n!=0| the |n/0| returns |Infinity| with the same sign as |n|.
-How do we capture all these different behaviours without changing the evaluator?
+Effect handlers provide a structured form of delimited
+continuations~\cite{Danvy90} that enables non-local control-flow mechanisms
+such as resumable exceptions, coroutines, generators and asynchronous I/O to be
+composably expressed. For example, consider the task of handling
+division-by-zero in a calculator application. Different calculators handle
+division-by-zero differently. In the iOS 13 Calculator (|iOS|), |0/0| returns
+|Error|. In the Google search calculator (|gsearch|) |0/0| returns |Error|, but
+for |n!=0| the |n/0| returns |Infinity| with the same sign as |n|. How do we
+capture all these different behaviours without changing the evaluator?
 
@@
 
-Line 4 declares an effect |DivBy0| that is parameterised by an integer,
-which when performed returns the result of division-by-zero. If the denominator
-is 0, then we perform the |DivBy0| effect with the numerator value (line 12) so that
-the context may decide the return value. We obtain the |iOS| semantics by
-wrapping the evaluator with this handler:
+Line 4 declares an effect |DivBy0| that is parameterised by an integer, which
+when performed returns the result of division-by-zero. If the denominator of
+the division is 0, then we perform the |DivBy0| effect with the numerator value
+(line 12) so that the context may decide the return value. We obtain the |iOS|
+semantics by wrapping the evaluator with this handler:
 \begin{lstlisting}
 let ios e = match eval e with
-| v -> v
-| effect (DivBy0 _) k -> continue k Error
+| v -> v (* value case *)
+| effect (DivBy0 _) k -> continue k Error (* effect case *)
 \end{lstlisting}
@@
 OCaml functions, irrespective of whether they use effect handlers, need to
-perform stack overflow checks and external C functions (which do not have stack
+perform stack overflow checks, and external C functions (which do not have stack
 overflow checks) must be performed on a separate system stack. Additionally, we
@@
 In order to forward exceptions across C frames, the C stub function
-|caml_call_ocaml|, pushes an exception handler frame that forwards the
+|caml_call_ocaml|, pushes an exception handler frame that either forwards the
 exception to the innermost OCaml exception handler (|raise_exn_c| in
@@
 frame containing pointers into the heap. During a GC, the OCaml stack is walked
-and the roots are marked skipping over the C frames.
+and the roots are marked, skipping over the C frames.
 
@@
 \end{lstlisting}
-The function |input_line| raises an |End_of_file| exception on reaching the end of
-input, which is handled by the exception handler which closes the channels. The
-|close_*| functions do nothing if the channel is already closed. The code is
-written in a defensive style to handle other exceptional cases such as the
+The function |input_line| raises an |End_of_file| exception on reaching the end
+of input, which is handled by the exception handler which closes the channels.
+The |close_*| functions do nothing if the channel is already closed. The code
+is written in a defensive style to handle other exceptional cases such as the
 channels being closed externally. Both |input_line| and |output_string| raise
-|Sys_error| exception if the channel is closed. In this case, the channels are
-closed and the exception is reraised to communicate the exceptional behaviour
-to the caller.
+|Sys_error| exception if the channel is closed. In this case, the catch-all
+exception handler closes the channels and reraises the exception to communicate
+the exceptional behaviour to the caller.
 
@@
 
-Programs that use resources such as channels are usually written defensively
+OCaml programs that use resources such as channels are usually written defensively
 with the assumption that calling a function will return \emph{exactly once},
@@
 semantics is simpler than languages that ensure effect
-safety~\cite{Hillerstrom20,Leijen14,Effekt,Frank,Biernacki19}. This is important for
-backwards compatibility as our goal is to retrofit effect handlers to a
-language with large legacy codebases. Programs that do not use effects remain
-well-typed, and those that do compose well with those that don't.
+safety~\cite{Eff,Hillerstrom20,Leijen14,Effekt,Frank,Biernacki19}. This is
+important for backwards compatibility as our goal is to retrofit effect
+handlers to a language with large legacy codebases. Programs that do not use
+effects remain well-typed, and those that do compose well with those that
+don't.
 
@@
 handles a |'a comp| that returns a |'a| value, and itself returns a |'b| value.
-The handler has a return case of type |'a -> 'b|. The effect case |effc|
+The handler has a return case |retc| of type |'a -> 'b|. The effect case |effc|
 handles effects of type |'c eff| with |('c,'b) continuation| and returns a
@@
                         & \text{if } (\term,\env,\fl) \rightsquigarrow (\term',\env',\fl') \\
-      \textsc{RetToO}   & (v,\env,[],\ostacka{k}{\cstack}) & \cstep
-                        & \configa{v}{\env}{\ostacka{k}{\cstack}} \\
       \textsc{CallC}    & (v, \_, \cclos{x}{e}{\env}::\fl, \ostack) & \cstep
@@
                         & \text{if } k = [[],(\{\caseval{x}{x}\},\emptyset)] \\
+      \textsc{RetToO}   & (v,\env,[],\ostacka{k}{\cstack}) & \cstep
+                        & \configa{v}{\env}{\ostacka{k}{\cstack}} \\
       \textsc{ExnFwdO}  & (v, \env, \ffun{\exnval{l}}::\_, \ostacka{(\fl,\hc) \kcons k}{\cstack}) & \cstep
@@
                         & \text{if } (\term,\env,\fl) \rightsquigarrow (\term',\env',\fl') \\
+      \textsc{CallO}    & (v, \_, (\oclos{x}{e}{\env}::\fl,\hc) \kcons k, \cstack) & \ostep
+                        & \configa{e}{\envext{\env}{x}{v}}{\ostacka{(\fl,\hc) \kcons k}{\cstack}} \\
+      \textsc{ExtCall}  & (v, \_, (\cclos{x}{e}{\env}::\fl,\hc) \kcons k, \cstack) & \ostep
+                        & \configa{e}{\envext{\env}{x}{v}}{\cstacka{[]}{\ostacka{(\fl,\hc) \kcons k}{\cstack}}} \\
       \textsc{RetToC}   & (v, \_, [([],(h,\emptyset))], \cstack) & \ostep
@@
                         & \text{if } \{\caseval{x}{e}\} \in h \\
-      \textsc{CallO}    & (v, \_, (\oclos{x}{e}{\env}::\fl,\hc) \kcons k, \cstack) & \ostep
-                        & \configa{e}{\envext{\env}{x}{v}}{\ostacka{(\fl,\hc) \kcons k}{\cstack}} \\
-      \textsc{ExtCall}  & (v, \_, (\cclos{x}{e}{\env}::\fl,\hc) \kcons k, \cstack) & \ostep
-                        & \configa{e}{\envext{\env}{x}{v}}{\cstacka{[]}{\ostacka{(\fl,\hc) \kcons k}{\cstack}}} \\
-      \textsc{Resume}   & (v, \_, (\ffun{k}::\ffun{\oclos{x}{e}{\env}}::\fl,\hc) \kcons k, \cstack) & \ostep
-                        & \configa{e}{\envext{\env}{x}{v}}{\ostacka{k ~@~ ((\fl,\hc) \kcons k')}{\cstack}} \\
+
       \textsc{Handle}   & (\handle{e}{h}, \env, k, \cstack) & \ostep
                         & \configa{e}{\env}{\ostacka{([],(h,\env)) \kcons k}{\cstack}} \\
-      \textsc{EffUnHn}  & (v, \_, [\ffun{\effval{l}{k}}::\fl,(h,\env)], \cstack) & \ostep
-                        & \configa{e}{\emptyset}{\ostacka{k ~@~ [(\fl,(h,\env))]}{\cstack}}
-                        & \text{if } \{\caseeff{l}{\_}{\_}{\_}\} \notin h \\
-                  & & & & \text{and } {e = \throw{\textsf{unhandled}}{0}} \\
+      \textsc{ExHn}     & (v, \_, (\ffun{\exnval{l}}::\_,(h,\env)) \kcons k', \cstack) & \ostep
+                        & \configa{e}{\envext{\env}{x}{v}}{\ostacka{k'}{\cstack}}
+                        & \text{if } \{\caseexn{l}{x}{e}\} \in h \\
+      \textsc{ExFwdC}   & (v, \env, [\ffun{\exnval{l}}::\_,(h,\_)], \cstacka{\fl'}{\ostack}) & \ostep
+                        & \configa{v}{\env}{\cstacka{\ffun{\exnval{l}}::\fl'}{\ostack}}
+                        & \text{if } \{\caseexn{l}{\_}{\_}\} \notin h \\
+      \textsc{ExFwdFib} & (v, \env, (\ffun{\exnval{l}}::\_,(h,\_)) \kcons (\fl',\hc') \kcons k', \cstack) & \ostep
+                        & \configa{v}{\env}{\ostacka{(\ffun{\exnval{l}}::\fl',\hc') \kcons k'}{\cstack}}
+                        & \text{if } \{\caseexn{l}{\_}{\_}\} \notin h \\
+
       \textsc{EffHn}    & (v, \_, (\ffun{\effval{l}{k}}::\fl,(h,\env)) \kcons k', \cstack) & \ostep
@@
                   & & & & \text{and } k'' = k ~@~ [(\fl,(h,\env))] \\
-      \textsc{ExFwdC}   & (v, \env, [\ffun{\exnval{l}}::\_,(h,\_)], \cstacka{\fl'}{\ostack}) & \ostep
-                        & \configa{v}{\env}{\cstacka{\ffun{\exnval{l}}::\fl'}{\ostack}}
-                        & \text{if } \{\caseexn{l}{\_}{\_}\} \notin h \\
-      \textsc{ExHn}     & (v, \_, (\ffun{\exnval{l}}::\_,(h,\env)) \kcons k', \cstack) & \ostep
-                        & \configa{e}{\envext{\env}{x}{v}}{\ostacka{k'}{\cstack}}
-                        & \text{if } \{\caseexn{l}{x}{e}\} \in h \\
-      \textsc{ExFwdFib} & (v, \env, (\ffun{\exnval{l}}::\_,(h,\_)) \kcons (\fl',\hc') \kcons k', \cstack) & \ostep
-                        & \configa{v}{\env}{\ostacka{(\ffun{\exnval{l}}::\fl',\hc') \kcons k'}{\cstack}}
-                        & \text{if } \{\caseexn{l}{\_}{\_}\} \notin h \\
+
+      \textsc{EffUnHn}  & (v, \_, [\ffun{\effval{l}{k}}::\fl,(h,\env)], \cstack) & \ostep
+                        & \configa{e}{\emptyset}{\ostacka{k ~@~ [(\fl,(h,\env))]}{\cstack}}
+                        & \text{if } \{\caseeff{l}{\_}{\_}{\_}\} \notin h \\
+                  & & & & \text{and } {e = \throw{\textsf{unhandled}}{0}} \\
+      \textsc{Resume}   & (v, \_, (\ffun{k}::\ffun{\oclos{x}{e}{\env}}::\fl,\hc) \kcons k', \cstack) & \ostep
+                        & \configa{e}{\envext{\env}{x}{v}}{\ostacka{k ~@~ ((\fl,\hc) \kcons k')}{\cstack}} \\
     \end{array}
@@
 current stack $\stack$. The term is either an expression $e$ or a value $v$.
-The value are integer constants $n$, continuations $k$, closure
-$\clos{x}{e}{\env}$, effect being performed ($\effval{l}{k}$) and exception
+The values are integer constants $n$, continuations $k$, closures
+$\clos{x}{e}{\env}$, effects being performed ($\effval{l}{k}$) and exceptions
 being raised ($\exnval{l}$). The environment is a map from variables to values.
@@
 frames $\fl$, and the OCaml stack $\ostack$ under it. Every OCaml stack is
-either empty $\ostackemp$ or non-empty $\ostacka{k}{\cstack}$ with the C stack
-$\cstack$ under it. Thus, the program stack is an alternating sequence of C and
-OCaml stacks terminating with an empty OCaml stack $\ostackemp$. The frame list
-$\fl$ is composed of individual frames $r$, which is one of an argument frame
-$\farg{e}{\env}$ with the expression $e$ at the argument position of an
-application with its environment $\env$, a function frame $\ffun{v}$ with the
-value $v$ at the function position of an application, and frames for evaluating
-the arguments of an arithmetic expression.
+either empty $\ostackemp$ or non-empty $\ostacka{k}{\cstack}$ with the current
+continuation $k$ and the C stack $\cstack$ under it. Thus, the program stack is
+an alternating sequence of C and OCaml stacks terminating with an empty OCaml
+stack $\ostackemp$. The frame list $\fl$ is composed of individual frames $r$,
+which is one of an argument frame $\farg{e}{\env}$ with the expression $e$ at
+the argument position of an application with its environment $\env$, a function
+frame $\ffun{v}$ with the value $v$ at the function position of an application,
+and frames for evaluating the arguments of an arithmetic expression.
 
@@
 
-\textbf{\textit{OCaml reductions.}} In OCaml (Figure~\ref{sem:ostep}), reductions
-always occur on the top-most fiber in the current stack. \textsc{AdminO}
-performs administrative reductions. \textsc{CallO} evaluates an OCaml function
-on the current stack. \textsc{ExtCall} captures the behaviour of external
-calls, which are evaluated on an empty C stack. \textsc{RetToC} returns a value
-to the enclosing C stack. In this case, we have exactly one fiber on the stack,
-and this was created in the rule \textsc{Callback}, whose handler has identity
-return case alone and the environment is empty. \textsc{RetFib} returns value
-from a fiber to the previous one, evaluating the body of the return case.
+\textbf{\textit{OCaml reductions.}} In OCaml (Figure~\ref{sem:ostep}),
+reductions always occur on the top-most fiber in the current stack.
+\textsc{AdminO} performs administrative reductions. \textsc{CallO} evaluates an
+OCaml function on the current stack. \textsc{ExtCall} captures the behaviour of
+external calls, which are evaluated on an empty C stack with the current OCaml
+stack as its tail. \textsc{RetToC} returns a value to the enclosing C stack. In
+this case, we have exactly one fiber on the stack, and this was created in the
+rule \textsc{Callback}, whose handler has identity return case alone and the
+environment is empty. \textsc{RetFib} returns value from a fiber to the
+previous one, evaluating the body of the return case.
 
@@
 handler has a matching effect case. We evaluate the body of the matching case,
-and bind the continuation parameter |r| to the captured continuation. Observe
-that the captured continuation |k''| includes the current handler. Intuitively,
-the handler wraps around captured continuation. This gives Multicore OCaml
-effect handlers deep handler semantics~\cite{Hillerstrom20}. \textsc{EffFwd} forwards the
-effect to the outer fiber, and extends the captured continuation |k''| in the
-process. Recall that we do not capture C frames as part of a continuation. To
-this end, \textsc{EffUnHn} models unhandled effect. If the effect bubbles up to
-the top fiber, which we know does not have an effect case (see
-\textsc{Callback} rule), we raise \textsc{unhandled} exception at the point
-where the corresponding effect was performed. This is achieved by appending the
-captured continuation to the front of the current continuation.
+and bind the continuation parameter |r| to the captured continuation |k''|.
+Observe that the captured continuation |k''| includes the current handler.
+Intuitively, the handler wraps around captured continuation. This gives
+Multicore OCaml effect handlers deep handler semantics~\cite{Hillerstrom20}.
+\textsc{EffFwd} forwards the effect to the outer fiber, and extends the
+captured continuation |k''| in the process. Recall that we do not capture C
+frames as part of a continuation. To this end, \textsc{EffUnHn} models
+unhandled effect. If the effect bubbles up to the top fiber, which we know does
+not have an effect case (see \textsc{Callback} rule), we raise
+\textsc{unhandled} exception at the point where the corresponding effect was
+performed. This is achieved by appending the captured continuation to the front
+of the current continuation.
 
@@
 
+\vspace{-2mm}
 \subsection{Exceptions}
@@
 
+\vspace{-2mm}
 \subsection{Heap-allocated fibers}
@@
 
-Figure~\ref{fig:fiber} shows the layout of a fiber in Multicore
-OCaml. At the bottom of the stack, we have the |handler_info|, which contains
-the pointer to parent fiber, and the closures for the value, exception and
-effect cases. The closures are created by the translation described in
+Figure~\ref{fig:fiber} shows the layout of a fiber in Multicore OCaml. At the
+bottom of the stack, we have the |handler_info|, which contains the pointer to
+parent fiber, and the closures for the value, exception and effect cases. The
+closures are created by the translation described in
 \S\ref{sec:static_semantics}; Multicore OCaml supports exception patterns in
@@
 block needed for DWARF and GC bookkeeping with callbacks. Then, there is a
-top-level handler that invokes the exception handler |clos_hexn| after
-switching to the parent fiber. This is followed by the |pc| of the code that
-invokes the value handler |clos_hval| after switching to the parent stack. The
-stack is laid out such that when the computation in the handler returns,
-control goes to the value handler.
-
-We have the variable-sized area for the OCaml frames next. Since we want the
-fibers to be as small as possible, this area is initially 16 words in length.
-When the stack pointer |rsp| becomes less than the stack threshold (maintained
-in the |Caml_state| table), the stack is said to have overflown. On stack
-overflow, we copy the whole fiber to a new area with double the size. In
-Multicore OCaml, we introduce stack overflow checks into the function prologue
-of OCaml functions. In our evaluation of real world OCaml programs, we observed
-that most function calls are to leaf functions with small frame size. In order
-to take advantage of this fact, we introduce a small red zone at the top of the
-stack. The compiler elides the stack overflow check for leaf functions whose
-frame size is less than the size of the red zone.
+top-level exception handler frame that forwards exceptions to the parent fiber.
+When the exceptions are caught by this handler, the control switches to the
+parent stack, and the exception handler closure |clos_hexn| is invoked. This is
+followed by the |pc| of the code that returns values to the parent
+fiber. This stack is laid out such that when the handled computation returns,
+the control switches to the parent fiber and the value handler |clos_hval|
+is invoked.
+
+Next, we have the variable-sized area for the OCaml frames. In order to keep
+fibers small, this area is initially 16 words in length. When the stack pointer
+|rsp| becomes less than the stack threshold (maintained in the |Caml_state|
+table), the stack is said to have overflown. On stack overflow, we copy the
+whole fiber to a new area with double the size. In Multicore OCaml, we
+introduce stack overflow checks into the function prologue of OCaml functions.
+In our evaluation of real world OCaml programs, we observed that most function
+calls are to leaf functions with small frame size. In order to take advantage
+of this fact, we introduce a small red zone at the top of the stack. The
+compiler elides the stack overflow check for leaf functions whose frame size is
+less than the size of the red zone.
 
@@
 
+\vspace{-2mm}
 \subsection{External calls and callbacks}
@@
 external calls in the system stack. Calling a C function from OCaml involves
-saving the stack pointer in the current fiber, saving the allocation pointer in
-the |Caml_state|, updating |rsp| to the top of system stack (maintained in
-|Caml_state|), and calling the C function. The actions are reversed when
-returning from the external call. For C functions that take arguments on the
-stack, the arguments must be copied to the C stack from the OCaml stack where
-they are laid out. In the vast majority of external calls, we perform only a
-few additional instructions over stock OCaml.
+saving the stack pointer in the current fiber, saving the allocation pointer
+value in |r15| in the |Caml_state|, updating |rsp| to the top of system stack
+(maintained in |Caml_state|), and calling the C function. The actions are
+reversed when returning from the external call. For C functions that take
+arguments on the stack, the arguments must be copied to the C stack from the
+OCaml stack.
 
@@
 
+\vspace{-2mm}
 \subsection{Effect handlers}
@@
 as the program executes is captured in Figure~\ref{fig:fiber_handler}. The code
-performs |E| which is handled in the outer-most handler, and is immediately
-resumed. The arrows between the fibers are parent pointers. At position |p1|,
-|rsp| is at the top of the fiber |f|.
+performs effect |E| which is handled in the outer-most handler, and is
+immediately resumed. The arrows between the fibers are parent pointers. At
+position |p1|, |rsp| is at the top of the fiber |f|.
 
@@
 
-One challenge is to encode in DWARF directives the CFA when stacks are switched
-at effect handlers, callbacks and external calls. When starting to run on a new
-fiber, we insert DWARF bytecode to follow the |parent_fiber| pointer and
-dereference the |saved_sp| to get the CFA (|saved_sp + 8|). During callbacks
-into OCaml, we save the current system stack pointer in the |context block| in
-Figure~\ref{fig:fiber} to identify the CFA in the C stack. DWARF unwinding for
-external calls is implemented by following a link to the current OCaml stack
-pointer. With these changes, we get the same backtrace for the \texttt{meander}
-program from \S\ref{sec:unwind}, modulo runtime system functions due to effect
-handlers.
-
-We have utilized a DWARF verification tool~\cite{Bastian19} with extensions to
-handle PIE and additional DWARF primitives to flag possible errors in the hand
-written DWARF directives where we have altered the OCaml assembler to implement
-effects. In three places additional manual verification was required; two were
-where the stack is intentionally switched between fibers and one was where the
-return to the parent stack on child stack completion could not track stack
-resizes and intermediate effect handler frames.
+The goal is to encode the CFA using DWARF directives when stacks are switched.
+Recall that stack switching occurs in effect handlers, external calls and
+callbacks. When starting to run on a new fiber, we insert DWARF bytecode to
+follow the |parent_fiber| pointer and dereference the |saved_sp| to get the CFA
+(|saved_sp + 8|). During callbacks into OCaml, we save the current system stack
+pointer in the |context block| in Figure~\ref{fig:fiber} to identify the CFA in
+the C stack. DWARF unwinding for external calls is implemented by following a
+link to the current OCaml stack pointer. With these changes, we get the same
+backtrace for the \texttt{meander} program from \S\ref{sec:unwind}, modulo
+runtime system functions due to effect handlers.
+
+We have utilized the DWARF verification tool from~\cite{Bastian19} to flag
+possible errors in the hand written DWARF directives where we have altered the
+OCaml assembler to implement effect handlers. In three places additional manual
+verification was required; two were where the stack is intentionally switched
+between fibers and one was where the control returns to the parent fiber which
+happened to be resized due to effect handler invocation, which the tool could
+not track.
 
@@
 \end{lstlisting}
-However, Multicore OCaml does not enforce this behaviour in order to avoid
-installing a finaliser every time a continuation is captured.
+However, Multicore OCaml does not enforce this behaviour in order to avoid the
+costs associated with installing a finaliser every time a continuation is
+captured.
 
@@
 |stock| as the baseline), the multicore variants were \textbf{less than 1\%
-slower}. The outliers (on either ends) were due to the difference in the
-allocator and the GC between stock and Multicore OCaml. The biggest impact was
-the increase in the code size for OCaml code due to stack overflow checks.
-Compared to |stock|, the code size for OCaml code 19\% (30\%) more for |MC| and
-|MC+RedZone32| (|MC+RedZone0|). The result shows that our red zone is effective
-at reducing the code size, but further work is required to bring it closer to
-|stock|.
+slower} than |stock|. The outliers (on either ends) were due to the difference
+in the allocator and the GC between stock and Multicore OCaml. The biggest
+impact was the increase in the code size for OCaml code due to stack overflow
+checks. Compared to |stock|, the code size for OCaml code 19\% (30\%) more for
+|MC| and |MC+RedZone32| (|MC+RedZone0|). The result shows that our red zone is
+effective at reducing the code size, but further work is required to bring it
+closer to |stock|.
 
@@
 a hand-written, selective CPSed~\cite{Nielson01},
-defunctionalised~\cite{Danvy01} version (|cps|) and a
-concurrency monad (|monad|) version of the generator for the tree. Both |cps|
-and |monad| versions are specialised to the binary tree, and come with the
-usual caveats of not using the stack for function calls. We observed that the
-|cps| version was the fastest thanks to specialisation and hand optimisation.
-|MC| version was only 2.76$\times$ slower than |cps| while being a generic
+defunctionalised~\cite{Danvy01} version (|cps|) and a concurrency monad
+(|monad|) version of the generator for the tree. Both |cps| and |monad|
+versions are specialised to the binary tree, and come with the usual caveats of
+not using the stack for function calls. We observed that the |cps| version was
+the fastest thanks to specialisation and hand optimisation. |MC| version was
+only 2.76$\times$ slower than |cps| while being a generic
 solution, whereas the |monad| version was 8.69$\times$ slower than |cps|.
 
-\textbf{\textit{Chameneos.}} Chameneos~\cite{Chameneos} is a concurrency game aimed at
-measuring context switching and synchronization overheads. Our implementation
-uses |MVars| for synchronization. We compare effect handler (|MC|), concurrency
-monad (|monad|) and Lwt, a widely used concurrency programming library for
-OCaml (|lwt|) versions. We observed that |MC| was the fastest, and |monad|
-(|lwt|) was 1.67$\times$ (4.29$\times$) slower than |MC|.
+\textbf{\textit{Chameneos.}} Chameneos~\cite{Chameneos} is a concurrency game
+aimed at measuring context switching and synchronization overheads. Our
+implementation uses |MVars| for synchronization. We compare effect handler
+(|MC|), concurrency monad (|monad|) and Lwt, a widely used concurrency
+programming library for OCaml (|lwt|) versions. We observed that |MC| was the
+fastest, and |monad| (|lwt|) was 1.67$\times$ (4.29$\times$) slower than |MC|.
 

### 2020-11-17 retro-concurrency.tex
WIP trace in supplement

@@
 \newtheorem{definition}[theorem]{Definition}
+\newtheorem{example}[theorem]{Example}
 
@@
 \newcommand{\step}{\rightarrow}
+\newcommand{\multistep}{\rightarrow\!\!*}
 
@@
                         & \text{if } \{\caseeff{l}{\_}{\_}{\_}\} \notin h \\
-                  & & & & \text{and } {e = \throw{\textsf{unhandled}}{0}} \\
+                  & & & & \text{and } {e = \throw{\textsf{Unhandled}}{0}} \\
       \textsc{Resume}   & (v, \_, (\ffun{k}::\ffun{\oclos{x}{e}{\env}}::\fl,\hc) \kcons k', \cstack) & \ostep
@@
 not have an effect case (see \textsc{Callback} rule), we raise
-\textsc{unhandled} exception at the point where the corresponding effect was
+\textsc{Unhandled} exception at the point where the corresponding effect was
 performed. This is achieved by appending the captured continuation to the front
@@
 \setcounter{figure}{0}
+
+\subsection*{Dynamic Semantics -- Examples}
+
+\begin{definition}[Initial configuration]
+	Given an expression $e$, we define the initial configuration $\textsf{init}(e)$
+	to be $\configa{e}{\emptyset}{\cstacka{[]}{\bullet}}$.
+\end{definition}
+
+Intuitively, the initial environment is empty and the initial stack is an empty
+C stack.
+
+\begin{definition}[Final configuration]
+	Given a configuration $\config$, we say that the configuration is final
+	(written $\textsf{final}(\config)$) if ~$\config =
+	\configa{v}{\env}{\cstacka{[]}{\bullet}}$~ for some value $v$ and environment
+	$\env$.
+\end{definition}
+
+Given a configuration $\config = \configa{\term}{\env}{\stack}$, we define
+$\textsf{term}(\config)$ to be $\tau$. We define the relation $\multistep$ as
+the transitive reflexive closure of the top-level reduction relation $\step$.
+
+\subsubsection*{Example 1}
+
+Let
+
+\begin{smathpar}
+	\begin{array}{rcll}
+		e & = & (\olam{x}{ & \kw{match} ~\perform{E}{~0}~ \kw{with} \\
+			&   &            & \mid \caseexn{\textsf{Unhandled}}{x}{42} \\
+			&   &            & \mid \caseval{v}{v}})~0
+	\end{array}
+\end{smathpar}
+
+then $init(e) \multistep ~\config \wedge final(\config) \wedge term(\config) =
+42$.
+
+\begin{proof}
+	$init(e) = \configa{e}{\emptyset}{\cstacka{[]}{\bullet}}$
+
+	By \textsc{StepC}, \textsc{AdminC}, \textsc{App1},
+
+\begin{array}{rcll}
+	configa{e}{\emptyset}{\cstacka{[]}{\bullet}} \\
+	\step \\
+
+\end{proof}
+
+\subsection*{No effect macro benchmarks}
+
+\begin{figure*}
+	\begin{minipage}{\linewidth}
+	\includegraphics[width=\linewidth]{sandmark-notebook/sandmark_time}
+	\subcaption{Normalized time of macro benchmarks. Baseline is Stock OCaml,
+		whose running time in seconds in given in parenthesis.}
+	\includegraphics[width=\linewidth]{sandmark-notebook/sandmark_codesize}
+	\subcaption{Normalized OCaml code size of macro benchmarks. Baseline is Stock
+		OCaml, whose size in bytes given in parenthesis.}
+	\end{minipage}
+	\caption{Macro benchmark performance on code that does not use effects.}
+	\label{res:macro}
+\end{figure*}
+
+Figure~\ref{res:macro} presents the performance of macro benchmarks that do not
+use effects. The aim is to quantify the impact of the addition of effect
+handlers to the language on code that does not use effect handlers. The |MC| is
+Multicore OCaml with default red zone size (16 words). |MC+RedZone0| uses red
+zone size of 0; every OCaml function has a stack overflow check. |MC+RedZone32|
+uses a red zone size of 32 words.
 
@@
 
-\subsection*{No effect macro benchmarks}
-
-\begin{figure*}
-	\begin{minipage}{\linewidth}
-	\includegraphics[width=\linewidth]{sandmark-notebook/sandmark_time}
-	\subcaption{Normalized time of macro benchmarks. Baseline is Stock OCaml, whose running time in seconds in given in parenthesis.}
-	\includegraphics[width=\linewidth]{sandmark-notebook/sandmark_codesize}
-	\subcaption{Normalized OCaml code size of macro benchmarks. Baseline is Stock OCaml, whose size in bytes given in parenthesis.}
-	\end{minipage}
-	\caption{Macro benchmark performance on code that does not use effects.}
-	\label{res:macro}
-\end{figure*}
-
-Figure~\ref{res:macro} presents the performance of macro benchmarks that do not
-use effects. The aim is to quantify the impact of the addition of effect
-handlers to the language on code that does not use effect handlers. The |MC| is
-Multicore OCaml with default red zone size (16 words). |MC+RedZone0| uses red
-zone size of 0; every OCaml function has a stack overflow check. |MC+RedZone32|
-uses a red zone size of 32 words.
-
 \end{document}

### 2020-11-17 retro-concurrency.tex
Deleting semantics (in git history), added supplement

--- 2020-11-16 retro-concurrency.tex
@@
 We present an operational semantics for a core language of effect handlers that
-faithfully captures the semantics of the Multicore OCaml implementation.
+faithfully captures the semantics of the Multicore OCaml implementation. An
+executable version of the semantics implemented as an OCaml interpreter along
+with examples is included in the supplementary material.
 
@@
 body of the closure.
-
-The traces of example programs are included in the supplementary material along
-with an executable version of the semantics implemented in OCaml.
 
@@
 \setcounter{figure}{0}
+
+\subsection*{No effect macro benchmarks}
+
+\begin{figure*}
+	\begin{minipage}{\linewidth}
+	\includegraphics[width=\linewidth]{sandmark-notebook/sandmark_time}
+	\subcaption{Normalized time of macro benchmarks. Baseline is Stock OCaml,
+		whose running time in seconds in given in parenthesis.}
+	\includegraphics[width=\linewidth]{sandmark-notebook/sandmark_codesize}
+	\subcaption{Normalized OCaml code size of macro benchmarks. Baseline is Stock
+		OCaml, whose size in bytes given in parenthesis.}
+	\end{minipage}
+	\caption{Macro benchmark performance on code that does not use effects.}
+	\label{res:macro}
+\end{figure*}
+
+Figure~\ref{res:macro} presents the performance of macro benchmarks that do not
+use effects. The aim is to quantify the impact of the addition of effect
+handlers to the language on code that does not use effect handlers. The |MC| is
+Multicore OCaml with default red zone size (16 words). |MC+RedZone0| uses red
+zone size of 0; every OCaml function has a stack overflow check. |MC+RedZone32|
+uses a red zone size of 32 words.
 

### 2020-11-17 retro-concurrency.tex
Updates

@@
 \begin{figure*}
+\vspace{-2mm}
 \begin{minipage}{\linewidth}
@@
                         & \text{if } \{\caseval{x}{e}\} \in h \\
+									& & & & \text{and } k \neq [] \\
 
@@
                         & \configa{e}{\env}{\ostacka{([],(h,\env)) \kcons k}{\cstack}} \\
-      \textsc{ExHn}     & (v, \_, (\ffun{\exnval{l}}::\_,(h,\env)) \kcons k', \cstack) & \ostep
+      \textsc{ExnHn}     & (v, \_, (\ffun{\exnval{l}}::\_,(h,\env)) \kcons k', \cstack) & \ostep
                         & \configa{e}{\envext{\env}{x}{v}}{\ostacka{k'}{\cstack}}
                         & \text{if } \{\caseexn{l}{x}{e}\} \in h \\
-      \textsc{ExFwdC}   & (v, \env, [\ffun{\exnval{l}}::\_,(h,\_)], \cstacka{\fl'}{\ostack}) & \ostep
+      \textsc{ExnFwdC}   & (v, \env, [\ffun{\exnval{l}}::\_,(h,\_)], \cstacka{\fl'}{\ostack}) & \ostep
                         & \configa{v}{\env}{\cstacka{\ffun{\exnval{l}}::\fl'}{\ostack}}
                         & \text{if } \{\caseexn{l}{\_}{\_}\} \notin h \\
-      \textsc{ExFwdFib} & (v, \env, (\ffun{\exnval{l}}::\_,(h,\_)) \kcons (\fl',\hc') \kcons k', \cstack) & \ostep
+      \textsc{ExnFwdFib} & (v, \env, (\ffun{\exnval{l}}::\_,(h,\_)) \kcons (\fl',\hc') \kcons k', \cstack) & \ostep
                         & \configa{v}{\env}{\ostacka{(\ffun{\exnval{l}}::\fl',\hc') \kcons k'}{\cstack}}
@@
 
-\textbf{\textit{Top-level reductions.}} The initial configuration for an expression
-$e$ is $\configa{e}{\emptyset}{\cstacka{[]}{\bullet}}$, where the environment
-and the stack are empty. The top-level reductions (Figure ~\ref{sem:toplevel})
-can be performed by either by taking a C step $\cstep$ or an OCaml step
-$\ostep$.
-
-\textbf{\textit{C reductions.}} We can take a C step (Figure~\ref{sem:cstep}) by
-taking an administrative reduction step $\rightsquigarrow$. The administrative
-reductions are common to both C and OCaml. The rules \textsc{Var},
-\textsc{Arith1}, \textsc{Arith2}, \textsc{App1}, \textsc{App2} and
-\textsc{App3} are standard. \textsc{Arith3} performs the arithmetic operation
-on the integers ($\llbracket n_1 \odot n_2 \rrbracket$). \textsc{Raise} pushes
-an function frame with exception value to indicate that an exception is being
-raised. Similarly, \textsc{Perform} pushes a function frame with an effect
-value with an empty continuation $[[],(\{\caseval{x}{x}\},\emptyset)]$ with no
-captured frames and an empty handler with an identity return case alone. We
-shall return to \textsc{Resume1} and \textsc{Resume2} later.
+\subsubsection{Top-level reductions}
+
+The initial configuration for an expression $e$ is
+$\configa{e}{\emptyset}{\cstacka{[]}{\bullet}}$, where the environment and the
+stack are empty. The top-level reductions (Figure ~\ref{sem:toplevel}) can be
+performed by either by taking a C step $\cstep$ or an OCaml step $\ostep$.
+
+\subsubsection{C reductions}
+
+We can take a C step (Figure~\ref{sem:cstep}) by taking an administrative
+reduction step $\rightsquigarrow$. The administrative reductions are common to
+both C and OCaml. The rules \textsc{Var}, \textsc{Arith1}, \textsc{Arith2},
+\textsc{App1}, \textsc{App2} and \textsc{App3} are standard. \textsc{Arith3}
+performs the arithmetic operation on the integers ($\llbracket n_1 \odot n_2
+\rrbracket$). \textsc{Raise} pushes an function frame with exception value to
+indicate that an exception is being raised. Similarly, \textsc{Perform} pushes
+a function frame with an effect value with an empty continuation
+$[[],(\{\caseval{x}{x}\},\emptyset)]$ with no captured frames and an empty
+handler with an identity return case alone. We shall return to \textsc{Resume1}
+and \textsc{Resume2} later.
 
@@
 
-\textbf{\textit{OCaml reductions.}} In OCaml (Figure~\ref{sem:ostep}),
-reductions always occur on the top-most fiber in the current stack.
-\textsc{AdminO} performs administrative reductions. \textsc{CallO} evaluates an
-OCaml function on the current stack. \textsc{ExtCall} captures the behaviour of
-external calls, which are evaluated on an empty C stack with the current OCaml
-stack as its tail. \textsc{RetToC} returns a value to the enclosing C stack. In
-this case, we have exactly one fiber on the stack, and this was created in the
-rule \textsc{Callback}, whose handler has identity return case alone and the
-environment is empty. \textsc{RetFib} returns value from a fiber to the
-previous one, evaluating the body of the return case.
+\subsubsection{OCaml reductions}
+
+In OCaml (Figure~\ref{sem:ostep}), reductions always occur on the top-most
+fiber in the current stack. \textsc{AdminO} performs administrative reductions.
+\textsc{CallO} evaluates an OCaml function on the current stack.
+\textsc{ExtCall} captures the behaviour of external calls, which are evaluated
+on an empty C stack with the current OCaml stack as its tail. \textsc{RetToC}
+returns a value to the enclosing C stack. In this case, we have exactly one
+fiber on the stack, and this was created in the rule \textsc{Callback}, whose
+handler has identity return case alone and the environment is empty.
+\textsc{RetFib} returns value from a fiber to the previous one, evaluating the
+body of the return case.
 
 The rule \textsc{Handle} installs a handler by pushing a fiber with no frames
-and the given handler. The rule \textsc{ExHn} handles an exception, if the
+and the given handler. The rule \textsc{ExnHn} handles an exception, if the
 current handler has a matching exception case, unwinding the current fiber. The
-rule \textsc{ExFwdC} forwards the exception to C. Here, there is exactly one
+rule \textsc{ExnFwdC} forwards the exception to C. Here, there is exactly one
 fiber on the current stack, and the handler does not have a matching exception
 case, which we know is the case (see \textsc{Callback} rule). The rule
-\textsc{ExFwdFib} forwards the exception to the next fiber if the current
+\textsc{ExnFwdFib} forwards the exception to the next fiber if the current
 handler does not handle it.

### 2020-11-17 retro-concurrency.tex
edits

@@
 
-The rest of the paper is organized as follows. The next section describes the
-details of stock OCaml program stack. In \S\ref{sec:refine}, we refine the
-design further focussing on retrofitting effect handlers into a mainstream
-\emph{systems} language. \S\ref{sec:semantics} presents the compilation scheme
-and operational semantics semantics for Multicore OCaml effect handlers.
-\S\ref{sec:impl} discusses the compiler and the runtime system support for
-implementing effect handlers. \S\ref{sec:eval} presents extensive performance
-evaluation of effect handlers against the goals set in \S\ref{sec:req}.
-\S\ref{sec:related} and \S\ref{sec:conc} present the related work and the
-conclusions.
+The remainder of the paper continues with a description of the stock OCaml program stack (\S\ref{sec:stack}). We then refine the
+design further by focussing on retrofitting effect handlers into a mainstream
+\emph{systems} language (\S\ref{sec:refine}), followed by the compilation scheme
+and operational semantics semantics for Multicore OCaml effect handlers (\S\ref{sec:semantics}).  We then discuss the compiler and runtime system support for
+implementing effect handlers (\S\ref{sec:impl}), and present extensive performance
+evaluation of effect handlers (\S\ref{sec:eval}) against our design goals (\S\ref{sec:req}).
+Finally, we discuss the related work (\S\ref{sec:related}) and conclude (\S\ref{sec:conc}).
 
@@
 
-Continuations $k$ are either empty $[]$ or non-empty sequence of \emph{fibers}.
+Continuations $k$ are either empty $[]$ or a non-empty sequence of \emph{fibers}.
 A fiber $\fiber \coloneqq (\fl,\hc)$ is a list of frames $\fl$ and a handler

### 2020-11-17 retro-concurrency.tex
edits

@@
 We introduce an extensible variant type~\cite{ExtVariants} |'a eff| of effect
-values, which when performed using the |perform| primitive returns a |'a|
+values, which when performed using the |perform| primitive returns an |'a|
 value. Constructors for the value of type |'a eff| are declared using the
@@
 |type _ eff += E : string -> int eff|. We introduce the type
-|('a,'b) continuation| of delimited continuations which expect a |'a| value for
-resumption and return |'b| value. The continuations may be continued with a
+|('a,'b) continuation| of delimited continuations which expect an |'a| value for
+resumption and return a |'b| value. The continuations may be continued with a
 suitably typed value or discontinued with an exception.
@@
 effect to the outer handler and returns the value and exceptions back to the
-original performer. In the implementation, reperfom is implemented as a
+original performer. In the implementation, reperform is implemented as a
 primitive to avoid executing code on the resumption path.
@@
 handler has identity return case alone and the environment is empty.
-\textsc{RetFib} returns value from a fiber to the previous one, evaluating the
+\textsc{RetFib} returns the value from a fiber to the previous one, evaluating the
 body of the return case.
@@
 frames as part of a continuation. To this end, \textsc{EffUnHn} models
-unhandled effect. If the effect bubbles up to the top fiber, which we know does
-not have an effect case (see \textsc{Callback} rule), we raise
+unhandled effect. If the effect bubbles up to the top fiber --- which we know does
+not have an effect case (see \textsc{Callback} rule) --- we raise
 \textsc{Unhandled} exception at the point where the corresponding effect was

### 2020-11-17 retro-concurrency.tex
edits

@@
 
-In this section, we present the implementation details of effect handlers in
-Multicore OCaml. Multicore OCaml currently only supports the x86-64 backend on
-Linux, but our design does not preclude other backends. We assume x86-64
-architecture in the sequel.
+We now present the implementation details of effect handlers in
+Multicore OCaml. We assume an x86\_64 architecture for the remainder of this paper
+-- although  Multicore OCaml currently only supports the x86\_64 backend on
+Linux, our design is intended to support other CPUs and OSs.
 

### 2020-11-17 retro-concurrency.tex
edits

@@
 \emph{exactly once}. This can be enforced by installing a finaliser on every
-captured continuation that |discontinue|s the continuation during finalisation
+captured continuation that calls |discontinue| on the continuation during finalisation
 and ignores the result:

### 2020-11-17 retro-concurrency.tex
edits

@@
 aware of fiber layout and how to follow the parent pointers. The stack map
-generation and scanning each frame for roots remains the same. Sivaramakrishnan
-et al.~\cite{Sivaramakrishnan20} present the challenges and solutions for
-integrating fibers with the concurrent mark-and-sweep GC of Multicore OCaml.
+generation and scanning each frame for roots remains the same. The
+challenges and solutions for integrating fibers with the concurrent
+mark-and-sweep GC of Multicore OCaml have been discussed previously~\cite{Sivaramakrishnan20}.
 
@@
 against the performance requirements set in \S\ref{sec:req}. Multicore OCaml is
-an extension of 4.10.0 OCaml compiler with support for shared memory
+an extension of the 4.10.0 OCaml compiler with support for shared memory
 parallelism and effect handlers. These results were obtained on 2-socket,

### 2020-11-17 retro-concurrency.tex
edits

@@
 necessary. We do not use a \emph{stack cache} of recently freed stacks for
-speeding up fiber allocation~\cite{Farvardin20}. In our experiments, we
-observed that directly using |mimalloc|~\cite{Leijen19} outperforms a stack
-cache.
+speeding up fiber allocation~\cite{Farvardin20} as we
+observed in our evaluation (\S\ref{sec:eval}) that directly using |mimalloc|~\cite{Leijen19} outperforms a stack cache.
 
@@
 
-Next, we have the variable-sized area for the OCaml frames. In order to keep
-fibers small, this area is initially 16 words in length. When the stack pointer
-|rsp| becomes less than the stack threshold (maintained in the |Caml_state|
-table), the stack is said to have overflown. On stack overflow, we copy the
+Next, we have the variable-sized area for the OCaml frames, known as the red zone.
+In order to keep fibers small, this area is initially 16 words in length.
+When the stack pointer |rsp| becomes less than the stack threshold
+(maintained in the |Caml_state| table), the stack is said to have overflown.
+On stack overflow, we copy the
 whole fiber to a new area with double the size. In Multicore OCaml, we
 introduce stack overflow checks into the function prologue of OCaml functions.
-In our evaluation of real world OCaml programs, we observed that most function
-calls are to leaf functions with small frame size. In order to take advantage
-of this fact, we introduce a small red zone at the top of the stack. The
+In our evaluation of real world OCaml programs ((\S\ref{sec:eval}) we observed that most function
+calls are to leaf functions with a small frame size. We introduce a small red zone at the top of the stack to take advantage of this fact. The
 compiler elides the stack overflow check for leaf functions whose frame size is
@@
 we observed that padding tightloops with a few |nop| instructions, which
-changes the loop alignment, makes the code upto 15\% faster. Hence, the
+changes the loop alignment, makes the code up to 15\% faster. Hence, the
 difference in running times under 15\% may not be statistically significant.

### 2020-11-17 retro-concurrency.tex
edits

@@
 \subsection{Heap-allocated fibers}
+\label{sec:hafibers}
 
@@
 Multicore OCaml (|MC|), we also ran the benchmarks on Multicore OCaml with no
-red zone (|MC+RedZone0|) in the fiber (all OCaml functions will have stack
+red zone (|MC+RedZone0|) in the fiber (all OCaml functions will have a stack
 overflow check) and a red zone size of 32 words (|MC+RedZone32|). Recall that
-the default red zone size in Multicore OCaml is 16 words.
+the default red zone size in Multicore OCaml is 16 words (\S\ref{sec:hafibers}).
 

### 2020-11-17 retro-concurrency.tex
evaledits

@@
 using effect handlers (concurrency monad) is 13.77$\times$ (67.09$\times$) more
-expensive than the idiomatic implementation using non-tail calls. Concurrency
+expensive than the idiomatic implementation using non-tail calls. The concurrency
 monad suffers due to the heap allocation of continuation frames (which need to
-be GCed), where as effect handlers benefit from stack allocation of the frames.
-Our concurrency monad, and indeed monadic concurrency libraries such as
-Lwt~\cite{lwt} and Async~\cite{async}, also have other downsides; due to the
-lack of stack, exceptions, backtraces, and DWARF unwinding are no longer
+be GCed), whereas effect handlers benefit from stack allocation of the frames.
+Our concurrency monad (and other monadic concurrency libraries such as
+Lwt~\cite{lwt} and Async~\cite{async}) also have other downsides -- due to the
+lack of a stack, exceptions, backtraces, and DWARF unwinding are no longer
 useful.

### 2020-11-17 retro-concurrency.tex
edits

@@
 full-fledged HTTP/1.1 web server using effect handlers (|MC|). We use effect
-handlers to implement a user-level threading library, which spawns a thread per
-request. We use \texttt{httpaf} library~\cite{httpaf} for HTTP handling, and
+handlers to implement a user-level threading library that spawns a thread per
+request. We use \texttt{httpaf}~\cite{httpaf} for HTTP handling, and
 \texttt{libev}~\cite{libev} for the eventloop. We compare our implementation
 against an Lwt version (|lwt|) which also uses \texttt{httpaf} and
-\texttt{libev}. Unlike using effect handlers, Lwt version is written in monadic
-style, and does not have a notion of thread per request. For comparison, we
-also include a Go 1.13 version (|go|) that uses
+\texttt{libev}. Unlike using effect handlers, the Lwt version is written in monadic
+style and does not have a notion of thread per request. For comparison, we
+also include a Go 1.13 version (|go|) that uses the
 \texttt{net/http}~\cite{nethttp} package. As both the OCaml versions are single

### 2020-11-17 retro-concurrency.tex
edits

@@
 There are several strategies for implementing effect handlers. Eff~\cite{Eff},
-Helium~\cite{Biernacki20}, Frank~\cite{Frank}, Links server
+Helium~\cite{Biernacki20}, Frank~\cite{Frank} and the Links server
 backend~\cite{Hillerstrom20} use an interpreter similar to our operational
@@
 
-Similar to~\cite{Leijen14, Kiselyov12}, Multicore OCaml uses the call stack for
+Similarly to~\cite{Leijen14, Kiselyov12}, Multicore OCaml uses the call stack for
 implementing continuations, but our continuations are one-shot. Bruggeman et
@@
 on modern hardware. Multicore OCaml stacks do not neatly fit the description of
-one of these implementation strategies. They best described as using the resize
+one of these implementation strategies -- they are best described as using the resize
 strategy from Farvardin et al. for each of the fibers, which are in turn linked
@@
 uses its own stack similar to how each handled computation runs in its own
-fiber in Multicore OCaml. Similar to Multicore OCaml, Lua does not allow
-coroutines to include C frames. It is useful to note that Lua is an interpreted
+fiber in Multicore OCaml. Lua does not allow coroutines to include C frames, as in
+Multicore OCaml. It is useful to note that Lua is an interpreted
 language, unlike OCaml, and DWARF unwind tables are not a concern in Lua.

### 2020-11-17 retro-concurrency.tex
edits

@@
 fiber in Multicore OCaml. Lua does not allow coroutines to include C frames, as in
-Multicore OCaml. It is useful to note that Lua is an interpreted
-language, unlike OCaml, and DWARF unwind tables are not a concern in Lua.
+Multicore OCaml. 
+%It is useful to note that Lua is an interpreted
+%language, unlike OCaml, and DWARF unwind tables are not a concern in Lua.
 
@@
 re-allocation) of resources every time the non-local control leaves and enters
-back into a context. |dynamic-wind| is not quite what one wants for resource
-clean up~\cite{Kiselyov,Sitaram03}. The main challenge is distinguishing
-returning exits from non-returning ones. Resource clean up needs to occur only
-during non-returning exit.
+back into a context. |dynamic-wind| is not quite sufficient for resource
+clean up~\cite{Kiselyov,Sitaram03} as returning exits must be distinguished
+from non-returning ones, and resource cleanup is only necessary during
+non-returning exits.
 

### 2020-11-17 retro-concurrency.tex
edits

@@
 
-Similarly to~\cite{Leijen14, Kiselyov12}, Multicore OCaml uses the call stack for
-implementing continuations, but our continuations are one-shot. Bruggeman et
+Multicore OCaml uses the call stack for implementing continuations (as do~\cite{Leijen14, Kiselyov12}), but with one-shot continuations. Bruggeman et
 al.~\cite{Bruggeman96} show how to implement one-shot continuations efficiently
@@
 non-returning exits.
-
-In Multicore OCaml, we build on the existing defensive coding practices against
+Multicore OCaml builds on the existing defensive coding practices against
 exceptions to clean up resources on non-returning exits. Our continuations are
@@
 
-We presented the motivation, design and implementation of effect handlers
-specifically aimed at retrofitting them onto OCaml with tool compatibility,
-negligible overheads for existing OCaml programs and good performance for those
-that use effect handlers. Our evaluation shows that this is indeed the case.
+We have presented the design and implementation of effect handlers
+aimed at retrofitting them onto the large body of existing OCaml
+code.  Our design walks the tightrope of maintaining
+compatibility (e.g. for profiling and debuggers and negligible overheads for existing 
+programs) and also demonstrating good performance and usability for new OCaml
+code that takes advantage of effect handlers. Our implementation in Multicore OCaml
+deviates from previous categorisations~\cite{Farvardin20} and our evaluation shows
+that we have achieved our goal of balancing compatibility with the introduction
+of effect handlers for mainstream OCaml programming.
 

### 2020-11-17 retro-concurrency.tex
edits

@@
 uses its own stack similar to how each handled computation runs in its own
-fiber in Multicore OCaml. Lua does not allow coroutines to include C frames, as in
-Multicore OCaml. 
+fiber in Multicore OCaml.
+%Lua does not allow coroutines to include C frames, as in
+%Multicore OCaml.  - avsm/kc not quite, see https://coco.luajit.org/index.html
 %It is useful to note that Lua is an interpreted
@@
 handlers to the language on code that does not use effect handlers. The |MC| is
-Multicore OCaml with default red zone size (16 words). |MC+RedZone0| uses red
+Multicore OCaml with the default red zone size (16 words). |MC+RedZone0| uses red
 zone size of 0; every OCaml function has a stack overflow check. |MC+RedZone32|
@@
 than once to enable use cases such as backtracking search and probabilistic
-programming (Uber's Pyro langauge~\cite{Pyro}). Multicore OCaml has an
+programming (Uber's Pyro language~\cite{Pyro}). Multicore OCaml has an
 experimental feature for explicitly cloning continuations, which makes a copy
-of each of its constituent fibers. Multi-shot continuations are difficult to
+of each of its constituent fibers.
+
+Multi-shot continuations are difficult to
 reason about in the presence of linear resources such as file descriptors.

### 2020-11-18 retro-concurrency.tex
updates

@@
 	\item \textbf{Backwards compatibility} A well-behaved OCaml program does not
-    break under an OCaml extended with effect handlers. OCaml code that does not
+		break under OCaml extended with effect handlers. OCaml code that does not
 		use effect handlers will pay negligible performance cost.
@@
                         & \text{if } (\term,\env,\fl) \rightsquigarrow (\term',\env',\fl') \\
-      \textsc{CallC}    & (v, \_, \cclos{x}{e}{\env}::\fl, \ostack) & \cstep
+			\textsc{CallC}    & (v, \_, \ffun{\cclos{x}{e}{\env}}::\fl, \ostack) & \cstep
                         & \configa{e}{\envext{\env}{x}{v}}{\cstacka{\fl}{\ostack}} \\
-      \textsc{Callback} & (v, \_, \oclos{x}{e}{\env}::\fl, \ostack) & \cstep
+			\textsc{Callback} & (v, \_, \ffun{\oclos{x}{e}{\env}}::\fl, \ostack) & \cstep
                         & \configa{e}{\envext{\env}{x}{v}}{\ostacka{k}{\cstacka{\fl}{\ostack}}}
@@
                         & \text{if } (\term,\env,\fl) \rightsquigarrow (\term',\env',\fl') \\
-      \textsc{CallO}    & (v, \_, (\oclos{x}{e}{\env}::\fl,\hc) \kcons k, \cstack) & \ostep
+			\textsc{CallO}    & (v, \_, (\ffun{\oclos{x}{e}{\env}}::\fl,\hc) \kcons k, \cstack) & \ostep
                         & \configa{e}{\envext{\env}{x}{v}}{\ostacka{(\fl,\hc) \kcons k}{\cstack}} \\
-      \textsc{ExtCall}  & (v, \_, (\cclos{x}{e}{\env}::\fl,\hc) \kcons k, \cstack) & \ostep
+			\textsc{ExtCall}  & (v, \_, (\ffun{\cclos{x}{e}{\env}}::\fl,\hc) \kcons k, \cstack) & \ostep
                         & \configa{e}{\envext{\env}{x}{v}}{\cstacka{[]}{\ostacka{(\fl,\hc) \kcons k}{\cstack}}} \\
@@
 
-\textbf{\textit{Syntax.}} Our expressions (Figure~\ref{sem:syntax}) consists of
+\subsubsection{Syntax} Our expressions (Figure~\ref{sem:syntax}) consists of
 integer constants ($n$), variables ($x$), abstraction ($\lam{x}{x}$),
@@
 
-Continuations $k$ are either empty $[]$ or a non-empty sequence of \emph{fibers}.
-A fiber $\fiber \coloneqq (\fl,\hc)$ is a list of frames $\fl$ and a handler
+Continuations $k$ are either empty $[]$ or a non-empty list of \emph{fibers}. A
+fiber $\fiber \coloneqq (\fl,\hc)$ is a list of frames $\fl$ and a handler
 closure $\hc \coloneqq (h,\env)$, which is a pair of handler $h$ and its
-environment $\env$.
-
+environment $\env$. We use the infix operator $@$ for appending two lists.
+
+\vspace{-1mm}
 \subsubsection{Top-level reductions}
@@
 
+\vspace{-1mm}
 \subsubsection{C reductions}
@@
 
+\vspace{-1mm}
 \subsubsection{OCaml reductions}
@@
 
-\vspace{-2mm}
+%\vspace{-2mm}
 \subsection{Exceptions}
@@
 
-\vspace{-2mm}
+%\vspace{-2mm}
 \subsection{Heap-allocated fibers}
@@
 
-\vspace{-2mm}
+%\vspace{-2mm}
 \subsection{External calls and callbacks}
@@
 
-\vspace{-2mm}
+%\vspace{-2mm}
 \subsection{Effect handlers}
@@
 
-Even with correct DWARF unwind information, using DWARF to record call stack
+Despite the correct DWARF unwind information, using DWARF to record call stack
 information in |perf| only captures the call stack of the current fiber in
@@
 
-\vspace{-3mm}
+%\vspace{-3mm}
 \subsection{No effects benchmarks}
@@
 
-\vspace{-3mm}
+%\vspace{-3mm}
 \subsection{No perform benchmarks}
@@
 
-\vspace{-3mm}
+%\vspace{-3mm}
 \subsection{Concurrent benchmarks}
@@
 
-\vspace{-3.5mm}
+%\vspace{-3.5mm}
 \section{Related Work}
@@
 
-Multicore OCaml uses the call stack for implementing continuations (as do~\cite{Leijen14, Kiselyov12}), but with one-shot continuations. Bruggeman et
+Multicore OCaml uses the call stack for implementing continuations (as
+do~\cite{Leijen14, Kiselyov12}), but with one-shot continuations. Bruggeman et
 al.~\cite{Bruggeman96} show how to implement one-shot continuations efficiently
@@
 on modern hardware. Multicore OCaml stacks do not neatly fit the description of
-one of these implementation strategies -- they are best described as using the resize
-strategy from Farvardin et al. for each of the fibers, which are in turn linked
-to represent the current stack and the captured continuations.
+one of these implementation strategies -- they are best described as using the
+resize strategy from Farvardin et al. for each of the fibers, which are in turn
+linked to represent the current stack and the captured continuations.
 
@@
 \begin{figure*}
-	\begin{minipage}{\linewidth}
-	\includegraphics[width=\linewidth]{sandmark-notebook/sandmark_time}
-	\subcaption{Normalized time of macro benchmarks. Baseline is Stock OCaml,
+	\includegraphics[angle=90,scale=0.44]{sandmark-notebook/sandmark_time}
+	\caption{Normalized time of macro benchmarks. Baseline is Stock OCaml,
 		whose running time in seconds in given in parenthesis.}
-	\includegraphics[width=\linewidth]{sandmark-notebook/sandmark_codesize}
-	\subcaption{Normalized OCaml code size of macro benchmarks. Baseline is Stock
+	\label{res:macro_time}
+\end{figure*}
+
+\begin{figure*}
+	\includegraphics[angle=90,scale=0.44]{sandmark-notebook/sandmark_codesize}
+	\caption{Normalized OCaml code size of macro benchmarks. Baseline is Stock
 		OCaml, whose size in bytes given in parenthesis.}
-	\end{minipage}
-	\caption{Macro benchmark performance on code that does not use effects.}
-	\label{res:macro}
+	\label{res:macro_size}
 \end{figure*}
 
-Figure~\ref{res:macro} presents the performance of macro benchmarks that do not
-use effects. The aim is to quantify the impact of the addition of effect
-handlers to the language on code that does not use effect handlers. The |MC| is
-Multicore OCaml with the default red zone size (16 words). |MC+RedZone0| uses red
-zone size of 0; every OCaml function has a stack overflow check. |MC+RedZone32|
-uses a red zone size of 32 words.
+Figures~\ref{res:macro_time} and \ref{res:macro_size} present the performance
+of macro benchmarks that do not use effects. The aim is to quantify the impact
+of the addition of effect handlers to the language on code that does not use
+effect handlers. The |MC| is Multicore OCaml with the default red zone size (16
+words). |MC+RedZone0| uses red zone size of 0; every OCaml function has a stack
+overflow check. |MC+RedZone32| uses a red zone size of 32 words.
 

### 2020-11-18 retro-concurrency.tex
Updates

@@
 
-The remainder of the paper continues with a description of the stock OCaml program stack (\S\ref{sec:stack}). We then refine the
-design further by focussing on retrofitting effect handlers into a mainstream
-\emph{systems} language (\S\ref{sec:refine}), followed by the compilation scheme
-and operational semantics semantics for Multicore OCaml effect handlers (\S\ref{sec:semantics}).  We then discuss the compiler and runtime system support for
-implementing effect handlers (\S\ref{sec:impl}), and present extensive performance
-evaluation of effect handlers (\S\ref{sec:eval}) against our design goals (\S\ref{sec:req}).
-Finally, we discuss the related work (\S\ref{sec:related}) and conclude (\S\ref{sec:conc}).
+The remainder of the paper continues with a description of the stock OCaml
+program stack (\S\ref{sec:stack}). We then refine the design further by
+focussing on retrofitting effect handlers into a mainstream \emph{systems}
+language (\S\ref{sec:refine}), followed by the compilation scheme and
+operational semantics for Multicore OCaml effect handlers
+(\S\ref{sec:semantics}).  We then discuss the compiler and runtime system
+support for implementing effect handlers (\S\ref{sec:impl}), and present
+extensive performance evaluation of effect handlers (\S\ref{sec:eval}) against
+our design goals (\S\ref{sec:req}). Finally, we discuss the related work
+(\S\ref{sec:related}) and conclude (\S\ref{sec:conc}).
 
@@
 
-Consider the layout of the stock OCaml stack for the
-program in Figures~\ref{code:meander_ml} and~\ref{code:meander_c}. The OCaml
-main function |omain| installs two exception handlers |h1| and |h2| to handle
-the exceptions |E1| and |E2|. |omain| calls the external C function |ocaml_to_c|,
-which in turn calls back into the OCaml function |c_to_ocaml|, which raises the
+Consider the layout of the stock OCaml stack for the program shown in
+Figures~\ref{code:meander_ml} and~\ref{code:meander_c}. The OCaml main function
+|omain| installs two exception handlers |h1| and |h2| to handle the exceptions
+|E1| and |E2|. |omain| calls the external C function |ocaml_to_c|, which in
+turn calls back into the OCaml function |c_to_ocaml|, which raises the
 exception |E1|. OCaml supports raising exceptions in C as well as throwing
 exceptions across external calls. Hence, the exception |E1| gets caught in the
-handler |h1|, and |omain| returns |42|. The layout of the stack in the native code
-backend just before raising the exception in |c_to_ocaml| is illustrated in
-Figure~\ref{fig:stock_stack}. Note that the stack grows downwards.
+handler |h1|, and |omain| returns |42|. The layout of the stack in the native
+code backend just before raising the exception in |c_to_ocaml| is illustrated
+in Figure~\ref{fig:stock_stack}. Note that the stack grows downwards.
 
@@
 until a matching one is found, the time taken to handle an effect is linear in
-the number of handlers. We observe that the handler stack is shallow in real
-programs.
+the number of handlers. We observed that the handler stack is shallow in
+real programs.
 
@@
 impact was the increase in the code size for OCaml code due to stack overflow
-checks. Compared to |stock|, the code size for OCaml code 19\% (30\%) more for
-|MC| and |MC+RedZone32| (|MC+RedZone0|). The result shows that our red zone is
-effective at reducing the code size, but further work is required to bring it
-closer to |stock|.
+checks. Compared to |stock|, the code size for OCaml code is 19\% more for |MC|
+and |MC+RedZone32|, and 30\% more for |MC+RedZone0|. The result shows that our
+red zone is effective at reducing the code size, but further work is required
+to bring it closer to |stock|.
 
@@
 Next we look at benchmarks that utilize non-local control flow using effect
-handlers. First, we quantify the overhead individual operations in effect
+handlers. First, we quantify the overhead of individual operations in effect
 handling. Consider the following annotated code:

### 2020-11-18 retro-concurrency.tex
Updates

@@
 
-%\vspace{-2mm}
+\vspace{-2mm}
 \subsection{Exceptions}
@@
 
-%\vspace{-2mm}
+\vspace{-2mm}
 \subsection{Heap-allocated fibers}
@@
 necessary. We do not use a \emph{stack cache} of recently freed stacks for
-speeding up fiber allocation~\cite{Farvardin20} as we
-observed in our evaluation (\S\ref{sec:eval}) that directly using |mimalloc|~\cite{Leijen19} outperforms a stack cache.
+speeding up fiber allocation~\cite{Farvardin20} as we observed in our
+evaluation (\S\ref{sec:eval}) that directly using |mimalloc|~\cite{Leijen19}
+outperforms a stack cache.
 
@@
 
-Next, we have the variable-sized area for the OCaml frames, known as the red zone.
-In order to keep fibers small, this area is initially 16 words in length.
-When the stack pointer |rsp| becomes less than the stack threshold
-(maintained in the |Caml_state| table), the stack is said to have overflown.
-On stack overflow, we copy the
-whole fiber to a new area with double the size. In Multicore OCaml, we
+Next, we have the variable-sized area for the OCaml frames. In order to keep
+fibers small, this area is initially 16 words in length. When the stack pointer
+|rsp| becomes less than the \emph{stack threshold} (maintained in the
+|Caml_state| table), the stack is said to have overflown. On stack overflow, we
+copy the whole fiber to a new area with double the size. In Multicore OCaml, we
 introduce stack overflow checks into the function prologue of OCaml functions.
-In our evaluation of real world OCaml programs ((\S\ref{sec:eval}) we observed that most function
-calls are to leaf functions with a small frame size. We introduce a small red zone at the top of the stack to take advantage of this fact. The
-compiler elides the stack overflow check for leaf functions whose frame size is
-less than the size of the red zone.
+
+In our evaluation of real world OCaml programs (\S\ref{sec:eval}), we observed
+that most function calls are to leaf functions with small frame sizes. Can we
+eliminate the stack overflow checks for these functions? To this end, we
+introduce a small, fixed-sized red zone at the top of the stack. The compiler
+elides the stack overflow check for leaf functions whose frame size is less
+than the size of the red zone. The default size of the red zone in Multicore
+OCaml is 16 words.
 
@@
 
-%\vspace{-2mm}
+\vspace{-2mm}
 \subsection{External calls and callbacks}
@@
 
-%\vspace{-2mm}
+\vspace{-2mm}
 \subsection{Effect handlers}
@@
 
+\vspace{-3mm}
 \subsection{Stack unwinding}
@@
 
+\vspace{-2mm}
 \subsection{Garbage collection}
@@
 
+\vspace{-2mm}
 \section{Evaluation}
@@
 
-%\vspace{-3mm}
+\vspace{-3mm}
 \subsection{No effects benchmarks}
@@
 slower} than |stock|. The outliers (on either ends) were due to the difference
-in the allocator and the GC between stock and Multicore OCaml. The biggest
-impact was the increase in the code size for OCaml code due to stack overflow
-checks. Compared to |stock|, the code size for OCaml code is 19\% more for |MC|
-and |MC+RedZone32|, and 30\% more for |MC+RedZone0|. The result shows that our
-red zone is effective at reducing the code size, but further work is required
-to bring it closer to |stock|.
+in the allocator and the GC between stock and Multicore OCaml.
+
+The biggest impact was the increase in the code size due to stack overflow
+checks. We define code size as the sum of the sizes of all the OCaml text
+sections in the compiled binary file. The code size metric excludes data, debug
+symbols, the runtime (implemented in C) and other C libraries. Compared to
+|stock|, the code size for OCaml code is 19\% more for |MC| and |MC+RedZone32|,
+and 30\% more for |MC+RedZone0|. The result shows that our 16-word red zone is
+effective at reducing the code size compared to having no red zone, whereas the
+32-word red zone does not noticeably reduce the code size further. Further work
+is required to bring the code size closer to |stock|.
 
@@
 
-%\vspace{-3mm}
+\vspace{-3mm}
 \subsection{No perform benchmarks}
@@
 
-%\vspace{-3mm}
+\vspace{-3mm}
 \subsection{Concurrent benchmarks}
@@
 
-%\vspace{-3.5mm}
+\vspace{-2mm}
 \section{Related Work}
@@
 non-returning exits.
+
 Multicore OCaml builds on the existing defensive coding practices against
@@
 extends effect handlers with |initially| and |finally| clauses in Koka for
-resource safety.
-
-Sivaramakrishnan et al.~\cite{TFP17} describe the interaction of effect
-handlers and asynchronous exceptions. This is orthogonal to the contributions
-of this paper. Our focus is the compiler and runtime system support for
-implementing effect handlers.
-
+resource safety. Sivaramakrishnan et al.~\cite{TFP17} describe the interaction
+of effect handlers and asynchronous exceptions. This is orthogonal to the
+contributions of this paper. Our focus is the compiler and runtime system
+support for implementing effect handlers.
+
+\vspace{-2mm}
 \section{Conclusions}
@@
 
-We have presented the design and implementation of effect handlers
-aimed at retrofitting them onto the large body of existing OCaml
-code.  Our design walks the tightrope of maintaining
-compatibility (e.g. for profiling and debuggers and negligible overheads for existing
-programs) and also demonstrating good performance and usability for new OCaml
-code that takes advantage of effect handlers. Our implementation in Multicore OCaml
-deviates from previous categorisations~\cite{Farvardin20} and our evaluation shows
-that we have achieved our goal of balancing compatibility with the introduction
-of effect handlers for mainstream OCaml programming.
+We have presented the design and implementation of effect handlers aimed at
+retrofitting them onto the large body of existing OCaml code.  Our design walks
+the tightrope of maintaining compatibility (e.g. for profiling and debuggers
+and negligible overheads for existing programs) and also demonstrating good
+performance and usability for new OCaml code that takes advantage of effect
+handlers. Our evaluation shows that we have achieved our goal of balancing
+compatibility with the introduction of effect handlers for mainstream OCaml
+programming.
 
@@
 
-Figures~\ref{res:macro_time} and \ref{res:macro_size} present the performance
-of macro benchmarks that do not use effects. The aim is to quantify the impact
-of the addition of effect handlers to the language on code that does not use
-effect handlers. The |MC| is Multicore OCaml with the default red zone size (16
-words). |MC+RedZone0| uses red zone size of 0; every OCaml function has a stack
-overflow check. |MC+RedZone32| uses a red zone size of 32 words.
+Figures~\ref{res:macro_time} and \ref{res:macro_size} present the impact on
+macro benchmarks that do not use effects using the Multicore OCaml compiler
+which supports effect handlers. The aim is to quantify the impact of the
+addition of effect handlers to the language on code that does not use effect
+handlers. We define code size as the sum of the sizes of all the text sections
+in the binary generated by compiling the OCaml code. Thus, the code size metric
+excludes data, debug symbols, the runtime (implemented in C) and other
+statically linked C libraries. The |MC| is Multicore OCaml with the default red
+zone size (16 words). |MC+RedZone0| uses red zone size of 0; every OCaml
+function has a stack overflow check. |MC+RedZone32| uses a red zone size of 32
+words.
 

### 2020-11-19 retro-concurrency.tex
Updates

@@
 
-The biggest impact was the increase in the code size due to stack overflow
-checks. We define code size as the sum of the sizes of all the OCaml text
-sections in the compiled binary file. The code size metric excludes data, debug
-symbols, the runtime (implemented in C) and other C libraries. Compared to
-|stock|, the code size for OCaml code is 19\% more for |MC| and |MC+RedZone32|,
-and 30\% more for |MC+RedZone0|. The result shows that our 16-word red zone is
-effective at reducing the code size compared to having no red zone, whereas the
-32-word red zone does not noticeably reduce the code size further. Further work
-is required to bring the code size closer to |stock|.
+The biggest impact was the increase in the OCaml text section size (OTSS) due
+to the stack overflow checks. We define OTSS as the sum of the sizes of all the
+OCaml text sections in the compiled binary file ignoring the data sections, the
+debug symbols, the text sections associated with OCaml runtime and other
+statically linked C libraries. Compared to |stock|, OTSS is 19\% more for |MC|
+and |MC+RedZone32|, and 30\% more for |MC+RedZone0|. The result shows that our
+16-word red zone is effective at reducing OTSS compared to having no red zone,
+whereas the 32-word red zone does not noticeably reduce OTSS further. Further
+work is required to bring OTSS closer to |stock|.
 
@@
 |stock|. In the other programs, |MC| executes more instructions due to stack
-overflow checks. External calls and callbacks are more expensive due to
-additional bookkeeping. Callbacks are significantly more expensive in |MC|.
-Since we reuse the current fiber stack for callbacks, we need to ensure it has
-enough room for inserting additional frames, while |stock| does not need to do
-this. Callback performance is less important than external calls, which are far
-more numerous. As seen previously, these results however do not affect macro
-benchmark performance.
+overflow checks. The performance impact on callbacks is more significant than
+external calls. For callbacks, since we reuse the current fiber stack, we need
+to ensure it has enough room for inserting additional frames, while |stock|
+does not need to do this. Callback performance is less important than external
+calls, which are far more numerous. As seen previously, these results however
+do not affect macro benchmark performance.
 
@@
 handlers compared to a non-tail function call. To this end, we surround the
-non-tail calls in the recursive micro benchmarks with an effect handler. The
-code does not |perform| effects. We also include a version of the same
+non-tail calls in the recursive micro benchmarks with an effect handler. These
+programs do not |perform| effects. We also include a version of the same
 benchmarks implemented using concurrency monad~\cite{Claessen99} (|monad|) as a
@@
 	\includegraphics[angle=90,scale=0.44]{sandmark-notebook/sandmark_codesize}
-	\caption{Normalized OCaml code size of macro benchmarks. Baseline is Stock
-		OCaml, whose size in bytes given in parenthesis.}
+	\caption{Normalized OCaml text section size (OTSS) of macro benchmarks.
+	Baseline is Stock OCaml, whose size in bytes given in parenthesis.}
 	\label{res:macro_size}
@@
 addition of effect handlers to the language on code that does not use effect
-handlers. We define code size as the sum of the sizes of all the text sections
-in the binary generated by compiling the OCaml code. Thus, the code size metric
-excludes data, debug symbols, the runtime (implemented in C) and other
-statically linked C libraries. The |MC| is Multicore OCaml with the default red
-zone size (16 words). |MC+RedZone0| uses red zone size of 0; every OCaml
-function has a stack overflow check. |MC+RedZone32| uses a red zone size of 32
-words.
+handlers. The |MC| is Multicore OCaml with the default red zone size (16
+words). |MC+RedZone0| uses red zone size of 0; every OCaml function has a stack
+overflow check. |MC+RedZone32| uses a red zone size of 32 words.
 

### 2020-11-19 retro-concurrency.tex
Update results based on stack cache

@@
 Fibers are allocated on the heap using |malloc| and are |free|d when no longer
-necessary. We do not use a \emph{stack cache} of recently freed stacks for
-speeding up fiber allocation~\cite{Farvardin20} as we observed in our
-evaluation (\S\ref{sec:eval}) that directly using |mimalloc|~\cite{Leijen19}
-outperforms a stack cache.
+necessary. Similar to Farvardin et al.~\cite{Farvardin20}, we use a \emph{stack
+cache} of recently freed stacks in order to speed up allocation.
 
@@
 \emph{exactly once}. This can be enforced by installing a finaliser on every
-captured continuation that calls |discontinue| on the continuation during finalisation
-and ignores the result:
+captured continuation that calls |discontinue| on the continuation during
+finalisation and ignores the result:
 \begin{lstlisting}
@@
 	& \textbf{ack} & \textbf{fib} & \textbf{motzkin} & \textbf{sudan} & \textbf{tak} \\ \hline
-	\texttt{MC} 	 	& 10.57 & 15.76 & 16.96 & 12.88 & 13.64 \\
-	\texttt{monad} 	& 42.79 & 69.77 & 348.69 & 33.29 & 39.24 \\ \hline
+	\texttt{MC} 	 	& 12.25 & 12.05 & 11.44 & 6.74 & 8.9 \\
+	\texttt{monad} 	& 348.69 & 69.77 & 39.24 & 33.29 & 42.79 \\ \hline
 \end{tabular}
@@
 The results are presented in Table~\ref{tab:micro_noperform}. They show that
-using effect handlers (concurrency monad) is 13.77$\times$ (67.09$\times$) more
+using effect handlers (concurrency monad) is 10.02$\times$ (67.09$\times$) more
 expensive than the idiomatic implementation using non-tail calls. The concurrency
@@
 sequences \texttt{a-b}, \texttt{b-c}, \texttt{c-d} and \texttt{d-e} took
-\textbf{48ns, 10ns, 13ns and 19ns}, respectively. For reference, the idle
+\textbf{23ns, 5ns, 11ns and 7ns}, respectively. For reference, the idle
 memory load latency for the local NUMA node is 93.2ns as measured using the

### 2020-11-19 retro-concurrency.tex
proof-reading edits

@@
 bottom of the stack, we have the |handler_info|, which contains the pointer to
-parent fiber, and the closures for the value, exception and effect cases. The
-closures are created by the translation described in
+the parent fiber, and the closures for the value, exception and effect cases.
+The closures are created by the translation described in
 \S\ref{sec:static_semantics}; Multicore OCaml supports exception patterns in
@@
 
-When the effect |E| performed, we allocate a continuation object |ke| in the
+When the effect |E| is performed, we allocate a continuation object |ke| in the
 heap that points to the current fiber |f|, set fiber |f|'s parent pointer to
@@
 an extension of the 4.10.0 OCaml compiler with support for shared memory
-parallelism and effect handlers. These results were obtained on 2-socket,
+parallelism and effect handlers. These results were obtained on a 2-socket,
 Intel\textregistered Xeon\textregistered Gold 5120 x86-64 server, with 28
@@
 Multicore OCaml (|MC|), we also ran the benchmarks on Multicore OCaml with no
-red zone (|MC+RedZone0|) in the fiber (all OCaml functions will have a stack
+red zone (|MC+RedZone0|) in the fibers (all OCaml functions will have a stack
 overflow check) and a red zone size of 32 words (|MC+RedZone32|). Recall that
@@
 report the number of instructions executed (obtained using |perf|) along with
-time. |exnval| perform 100 million iterations of installing exception handlers
-and returning with value. |exnraise| is similar, but raises an exception in
+time. |exnval| performs 100 million iterations of installing exception handlers
+and returning with a value. |exnraise| is similar, but raises an exception in
 each iteration. |extcall| and |callback| perform 100 million external calls and
@@
 recursive programs and were taken from ~\cite{Farvardin20}. For micro benchmarks,
-we observed that padding tightloops with a few |nop| instructions, which
-changes the loop alignment, makes the code up to 15\% faster. Hence, the
+we observed that padding tight loops with a few |nop| instructions, which
+changes the loop alignment, can make the code up to 15\% faster. Hence, the
 difference in running times under 15\% may not be statistically significant.
@@
 benchmarks implemented using concurrency monad~\cite{Claessen99} (|monad|) as a
-proxy for CPS versions. Recall that OCaml compiler does not use CPS in its IR.
+proxy for CPS versions. Recall that the OCaml compiler does not use CPS in its IR.
 In the |monad| version, we use a |fork| to invoke the non-tail call and use an
@@
 monad suffers due to the heap allocation of continuation frames (which need to
-be GCed), whereas effect handlers benefit from stack allocation of the frames.
-Our concurrency monad (and other monadic concurrency libraries such as
-Lwt~\cite{lwt} and Async~\cite{async}) also have other downsides -- due to the
-lack of a stack, exceptions, backtraces, and DWARF unwinding are no longer
+be garbage collected), whereas effect handlers benefit from stack allocation of
+the frames. Our concurrency monad (and other monadic concurrency libraries such
+as Lwt~\cite{lwt} and Async~\cite{async}) also have other downsides -- due to
+the lack of a stack, exceptions, backtraces, and DWARF unwinding are no longer
 useful.
@@
 \texttt{libev}. Unlike using effect handlers, the Lwt version is written in monadic
-style and does not have a notion of thread per request. For comparison, we
+style and does not have the notion of a thread per request. For comparison, we
 also include a Go 1.13 version (|go|) that uses the
@@
 use an implementation of multi-prompt delimited continuations as an OCaml
-library~\cite{Kiselyov12} to embed Eff language in OCaml. Indeed, Forester et
+library~\cite{Kiselyov12} to embed the Eff language in OCaml. Indeed, Forester et
 al.~\cite{Forster19} showed that in an untyped setting, effect handlers,

### 2020-11-20 retro-concurrency.tex
Updates

@@
 when performed returns the result of division-by-zero. If the denominator of
-the division is 0, then we perform the |DivBy0| effect with the numerator value
-(line 12) so that the context may decide the return value. We obtain the |iOS|
-semantics by wrapping the evaluator with this handler:
+the division is |0|, then we perform the |DivBy0| effect with the numerator
+value (line 12) so that the context may decide the return value. We obtain the
+|iOS| semantics by wrapping the evaluator with this handler:
 \begin{lstlisting}
@@
 delimited continuations necessary for effect handlers without sacrificing the
-desirable properties of OCaml. One standard way of implementing continuations is
-to use continuation-passing style (CPS) in the compiler's intermediate
-representation (IR)~\cite{Leijen14}.
-%If the compiler were to use CPS, then
-%reifying continuations in the source language is relatively straight-forward.
-OCaml does not use a CPS IR, and changing the compiler to utilise a CPS IR
-would be an enormous undertaking that would affect the
-performance profile of existing OCaml applications due to increased memory allocations as
-continuation closures get allocated on the heap~\cite{Farvardin20}. Moreover,
-with CPS, stack unwinding is meaningless and hence, we would lose compatibility
-with tools that inspect the program stack. Hence, we choose not to use CPS
-translation and represent continuations as call stacks.
+desirable properties of OCaml. One standard way of implementing continuations
+is to use continuation-passing style (CPS) in the compiler's intermediate
+representation (IR)~\cite{Leijen14}. OCaml does not use a CPS IR, and changing
+the compiler to utilise a CPS IR would be an enormous undertaking that would
+affect the performance profile of existing OCaml applications due to increased
+memory allocations as continuation closures get allocated on the
+heap~\cite{Farvardin20}. Moreover, with CPS, stack unwinding is meaningless and
+hence, we would lose compatibility with tools that inspect the program stack.
+Hence, we choose not to use CPS translation and represent continuations as call
+stacks.
 
@@
 Loom~\cite{loom} is an OpenJDK project that adds virtual threads and delimited
-continuations to Java. The Swift roadmap~\cite{swift} includes
-direct-style asynchronous programming and structured concurrency as milestones.
-We believe that our design choices will inform similar choices to be made in the
-other industrial-strength languages.
+continuations to Java. The Swift roadmap~\cite{swift} includes direct-style
+asynchronous programming and structured concurrency as milestones. We believe
+that our design choices will inform similar choices to be made in other
+industrial-strength languages.
 
@@
 register |r15| holds the \emph{allocation pointer} into the minor heap used for
-bump pointer allocation, and |r14| holds a reference to the |Caml_state| table
-of global variables used by the runtime. This makes external calls extremely
-fast in OCaml. If the external function does not allocate in the OCaml heap,
-then it can be called directly and no bookkeeping is necessary. For external
-functions which allocate in the OCaml heap, the cached allocation pointer is
-saved to |Caml_state| before the external call and it is restored on return.
-Similarly, callbacks into OCaml are also cheap: these involve loading the
-arguments in the right registers and calling the OCaml function. OCaml
+bump pointer allocation, and |r14| holds a reference to the |Caml_state|, a
+table of global variables used by the runtime. This makes external calls
+extremely fast in OCaml. If the external function does not allocate in the
+OCaml heap, then it can be called directly and no bookkeeping is necessary. For
+external functions which allocate in the OCaml heap, the cached allocation
+pointer is saved to |Caml_state| before the external call and it is restored on
+return. Similarly, callbacks into OCaml are also cheap: these involve loading
+the arguments in the right registers and calling the OCaml function. OCaml
 callbacks are relatively common as the garbage collector (GC) finalisation
@@
 
-\subsubsection{Syntax} Our expressions (Figure~\ref{sem:syntax}) consists of
+\subsubsection{Syntax} Our expressions (Figure~\ref{sem:syntax}) consist of
 integer constants ($n$), variables ($x$), abstraction ($\lam{x}{x}$),
@@
 unnecessary and inefficient. Instead, Multicore OCaml optimises fibers for
-one-shot continuations.
-
-Fibers are allocated on the heap using |malloc| and are |free|d when no longer
-necessary. Similar to Farvardin et al.~\cite{Farvardin20}, we use a \emph{stack
-cache} of recently freed stacks in order to speed up allocation.
+one-shot continuations. Fibers are allocated on the heap using |malloc| and are
+|free|d when no longer necessary. Similar to Farvardin et
+al.~\cite{Farvardin20}, we use a \emph{stack cache} of recently freed stacks in
+order to speed up allocation.
 
@@
 
+%It is useful to note that stack overflow is a rare operation
+%and the CPU branch predictor will correctly predict the branch with high
+%accuracy.
+
 Finally, we have the saved exception pointer, which points to the top-most
@@
 
-\vspace{-2mm}
+\vspace{-3mm}
 \subsection{External calls and callbacks}
@@
 
-\vspace{-2mm}
+\vspace{-3mm}
 \subsection{Effect handlers}
@@
 callbacks to identity functions. The other micro benchmarks are highly
-recursive programs and were taken from ~\cite{Farvardin20}. For micro benchmarks,
-we observed that padding tight loops with a few |nop| instructions, which
-changes the loop alignment, can make the code up to 15\% faster. Hence, the
+recursive programs and were taken from ~\cite{Farvardin20}. For micro
+benchmarks, we observed that padding tight loops with a few |nop| instructions,
+which changes the loop alignment, makes the code up to 15\% faster. Hence, the
 difference in running times under 15\% may not be statistically significant.
@@
 not using the stack for function calls. We observed that the |cps| version was
-the fastest thanks to specialisation and hand optimisation. |MC| version was
+the fastest, thanks to specialisation and hand optimisation. |MC| version was
 only 2.76$\times$ slower than |cps| while being a generic

### 2020-11-20 retro-concurrency.tex
Udpates
Updates

@@
 
-We now present the implementation details of effect handlers in
-Multicore OCaml. We assume an x86\_64 architecture for the remainder of this paper
--- although  Multicore OCaml currently only supports the x86\_64 backend on
-Linux, our design is intended to support other CPUs and OSs.
+We now present the implementation details of effect handlers in Multicore
+OCaml. We assume an x86\_64 architecture for the remainder of this paper.
+Although Multicore OCaml currently only supports the x86\_64 backend on Linux,
+our design does not preclude other CPUs and OSs.
 
@@
 OCaml is 16 words.
-
-%It is useful to note that stack overflow is a rare operation
-%and the CPU branch predictor will correctly predict the branch with high
-%accuracy.
 
@@
 sequences \texttt{a-b}, \texttt{b-c}, \texttt{c-d} and \texttt{d-e} took
-\textbf{23ns, 5ns, 11ns and 7ns}, respectively. For reference, the idle
-memory load latency for the local NUMA node is 93.2ns as measured using the
+\textbf{23 ns, 5 ns, 11 ns and 7 ns}, respectively. For reference, the idle
+memory load latency for the local NUMA node is 93.2 ns as measured using the
 Intel MLC tool~\cite{mlc}.
@@
 		\includegraphics[width=0.9\linewidth]{http-benchmarks/throughput.pdf}
+		\vspace{-2mm}
 		\subcaption{Throughput}
@@
 		\includegraphics[width=0.9\linewidth]{http-benchmarks/latency.pdf}
+		\vspace{-2mm}
 		\subcaption{Tail latency}
@@
 extends effect handlers with |initially| and |finally| clauses in Koka for
-resource safety. Sivaramakrishnan et al.~\cite{TFP17} describe the interaction
+resource safety. Dolan et al.~\cite{TFP17} describe the interaction
 of effect handlers and asynchronous exceptions. This is orthogonal to the
@@
 
-\vspace{-2mm}
+\vspace{-3mm}
 \section{Conclusions}
@@
 
+\if{0}
+\subsection*{Dynamic Semantics}
+
+\begin{definition}[Initial configuration]
+	Given an expression $e$, we define the initial configuration $\textsf{init}(e)$
+	to be $\configa{e}{\emptyset}{\cstacka{[]}{\bullet}}$.
+\end{definition}
+
+Intuitively, the initial environment is empty and the initial stack is an empty
+C stack.
+
+\begin{definition}[Final value configuration]
+	Given a configuration $\config$, we say that the configuration is the final
+	value configuration (written $\textsf{finalval}(\config)$) if ~$\config =
+	\configa{v}{\env}{\cstacka{[]}{\bullet}}$~ for some value $v$ and environment
+	$\env$.
+\end{definition}
+
+\begin{definition}[Final exception configuration]
+	Given a configuration $\config$, we say that the configuration is the final
+	exception configuration (written $\textsf{finalexn}(\config)$) if ~$\config
+	= \configa{v}{\env}{\cstacka{\ffun{\exnval{l}}::\fl}{\bullet}}$~ for some
+	value $v$, label $l$, frame list $\fl$ and environment $\env$.
+\end{definition}
+
+Given a configuration $\config = \configa{\term}{\env}{\stack}$, we define
+$\textsf{term}(\config)$ to be $\tau$. We define the relation $\multistep$ as
+the transitive reflexive closure of the top-level reduction relation $\step$.
+
+\subsubsection*{Example 1}
+
+Let
+
+\begin{smathpar}
+	\begin{array}{rcll}
+		e & = & (\olam{x}{ & \kw{match} ~\perform{E}{~0}~ \kw{with} \\
+			&   &            & \mid \caseexn{\textsf{Unhandled}}{x}{42} \\
+			&   &            & \mid \caseval{v}{v}})~0
+	\end{array}
+\end{smathpar}
+
+then $init(e) \multistep ~\config \wedge final(\config) \wedge term(\config) =
+42$.
+
+\begin{proof}
+\end{proof}
+\fi
+
 \subsection*{No effect macro benchmarks}
@@
 than once to enable use cases such as backtracking search and probabilistic
-programming (Uber's Pyro language~\cite{Pyro}). Multicore OCaml has an
-experimental feature for explicitly cloning continuations, which makes a copy
-of each of its constituent fibers.
+programming (for example, see Uber's Pyro language~\cite{Pyro}). Multicore
+OCaml has an experimental feature for explicitly cloning continuations, which
+makes a copy of each of its constituent fibers.
 

### 2020-11-20 retro-concurrency.tex
Less detail on core count and cache for machine spec in evaluation section; put in reference link for the processor

@@
 an extension of the 4.10.0 OCaml compiler with support for shared memory
-parallelism and effect handlers. These results were obtained on a 2-socket,
-Intel\textregistered Xeon\textregistered Gold 5120 x86-64 server, with 28
-physical cores (14 cores on each socket), and 2 hardware threads per core. Each
-core runs at 2.20GHz and has 32 KB of L1 data cache, 32 KB of L1 instruction
-cache and 1MB of L2 cache. The cores on a socket share a 19.25 MB L3 cache. The
-server has 64GB of main memory and runs Ubuntu 18.04.
+parallelism and effect handlers. These results were obtained on a 2-socket
+Intel\textregistered Xeon\textregistered Gold 5120
+x86-64~\cite{IntelXeonGold5120Spec} server running Ubuntu 18.04 with 64GB of main
+memory.
 

### 2020-11-20 retro-concurrency.tex
addition of branch predictor and stack overflow check sentence to heap-allocated fibers section
Some rewording to free up some space and make things tighter

@@
 
-\noindent \textbf{R3: Effect handler efficiency.} The program must accommodate millions of
-continuations at the same time to support
+\noindent \textbf{R3: Effect handler efficiency.} The program must accommodate
+millions of continuations at the same time to support
 highly-concurrent applications. Installing effect handlers, capturing and
@@
 We have implemented effect handlers in a multicore extension of the OCaml
-programming language. We call our implementation \textbf{Multicore OCaml} to
-distinguish it from stock OCaml.
+programming language which we call \textbf{Multicore OCaml} to distinguish
+it from stock OCaml.
 
@@
 (\S\ref{sec:semantics}).  We then discuss the compiler and runtime system
-support for implementing effect handlers (\S\ref{sec:impl}), and present
+support for implementing effect handlers (\S\ref{sec:impl}), and present an
 extensive performance evaluation of effect handlers (\S\ref{sec:eval}) against
@@
 introduce stack overflow checks into the function prologue of OCaml functions.
+These stack overflow operations are rare and so will be correctly predicted by
+the CPU branch predictor.
 
@@
 (|monad|) version of the generator for the tree. Both |cps| and |monad|
-versions are specialised to the binary tree, and come with the usual caveats of
+versions are specialised to the binary tree with the usual caveats of
 not using the stack for function calls. We observed that the |cps| version was
@@
 
-Using effect handlers in the web server also helps with debugging the system as
-a whole. Multicore OCaml supports getting the backtrace of a continuation,
-which is a minor extension to getting the current backtrace. We can use this
-feature to dump the backtrace of all the user-level threads in the web server.
-Such a feature is provided by Go~\cite{gopprof}, but not possible with OCaml
-concurrency libraries such as Async and Lwt which lack the notion of a thread.
+Multicore OCaml supports backtraces for continuations in addition to backtraces
+of the current stack as in stock OCaml. Using effect handlers in a system such
+as a web server aides debugging and profiling because it is possible to get a
+backtrace snapshot of all current requests. Such a feature is available in
+Go~\cite{gopprof}, but not in OCaml concurrency libraries such as Async and
+Lwt which lack the notion of a thread.
 
@@
 Leijen~\cite{Leijen14} implements effect handlers as a library in C using stack
-copying. Since C allows pointers into the stack, care is taken to ensure that
+copying. C allows pointers into the stack, so care is taken to ensure that
 when continuations are resumed, the constituent frames are restored to the same
@@
 one of these implementation strategies -- they are best described as using the
-resize strategy from Farvardin et al. for each of the fibers, which are in turn
+resize strategy from Farvardin et al. for each of the fibers, which are
 linked to represent the current stack and the captured continuations.
@@
 Base~\cite{BaseProtect} and Core~\cite{CoreProtect} provide mechanisms
-analogous to |unwind-protect|, which are in turn implemented using exceptions.
+analogous to |unwind-protect|, which are implemented using exceptions.
 Thus, the linear use of continuations enabled by the |discontinue| primitive

### 2020-11-20 retro-concurrency.tex
squeeze in numbers of major collections for the micro benchmarks with no perform

@@
 non-tail calls in the recursive micro benchmarks with an effect handler. These
-programs do not |perform| effects. We also include a version of the same
-benchmarks implemented using concurrency monad~\cite{Claessen99} (|monad|) as a
-proxy for CPS versions. Recall that the OCaml compiler does not use CPS in its IR.
-In the |monad| version, we use a |fork| to invoke the non-tail call and use an
-|MVar| to collect its result.
+programs do not |perform| effects. We also implemented the same benchmarks using
+a concurrency monad~\cite{Claessen99} (|monad|) as a proxy for CPS versions.
+Recall that the OCaml compiler does not use CPS in its IR. In the |monad|
+version, we use a |fork| to invoke the non-tail call and use an |MVar| to
+collect its result.
 
@@
 using effect handlers (concurrency monad) is 10.02$\times$ (67.09$\times$) more
-expensive than the idiomatic implementation using non-tail calls. The concurrency
-monad suffers due to the heap allocation of continuation frames (which need to
-be garbage collected), whereas effect handlers benefit from stack allocation of
-the frames. Our concurrency monad (and other monadic concurrency libraries such
-as Lwt~\cite{lwt} and Async~\cite{async}) also have other downsides -- due to
-the lack of a stack, exceptions, backtraces, and DWARF unwinding are no longer
-useful.
+expensive than the idiomatic implementation using non-tail calls. The
+concurrency monad suffers due to the heap allocation of continuation frames
+(which need to be garbage collected), whereas effect handlers benefit from stack
+allocation of the frames. For example the number of major collections for the
+\texttt{ack} benchmark is 0 for stock OCaml, 1 for \texttt{MC} and 112 for
+\texttt{monad}. Our concurrency monad (and other monadic concurrency libraries
+such as Lwt~\cite{lwt} and Async~\cite{async}) also have other downsides --
+exceptions, backtraces, and DWARF unwinding are no longer useful due to the lack
+of a stack.
 
@@
 traversing a complete binary tree of depth 25 using this generator. This
-involves $2^{26}$ stack switches in total. For comparison, we also implemented
+involves $2^{26}$ stack switches in total. For comparison, we implemented
 a hand-written, selective CPSed~\cite{Nielson01},
@@
 only 2.76$\times$ slower than |cps| while being a generic
-solution, whereas the |monad| version was 8.69$\times$ slower than |cps|.
+solution, and the |monad| version was 8.69$\times$ slower than |cps|.
 
@@
 style and does not have the notion of a thread per request. For comparison, we
-also include a Go 1.13 version (|go|) that uses the
+include a Go 1.13 version (|go|) that uses the
 \texttt{net/http}~\cite{nethttp} package. As both the OCaml versions are single

### 2020-11-20 retro-concurrency.tex
Reworked conclusion with avsm

@@
 
-We have presented the design and implementation of effect handlers aimed at
-retrofitting them onto the large body of existing OCaml code. Our design walks
-the tightrope of maintaining compatibility (e.g. for profiling and debuggers
-and negligible overheads for existing programs) and also demonstrating good
-performance and usability for new OCaml code that takes advantage of effect
-handlers. Our evaluation shows that we have achieved our goal of balancing
-compatibility with the introduction of effect handlers for mainstream OCaml
-programming.
+Our design for effect handlers in OCaml walks the tightrope of maintaining
+compatibility (for profiling, debuggers and negligible overheads for existing
+programs), while unlocking the full power of non-local control flow constructs.
+Our evaluation shows that we have achieved our goal: we retain compatibility
+with a surprisingly low performance overhead for sequential code that preserves
+the spirit of ``fast exceptions'' that has always characterised OCaml
+programming. We believe that the introduction of first-class fibres and effects
+into OCaml -- along with a parallel runtime~\cite{Sivaramakrishnan20} -- will
+open OCaml to highly task-parallel applications with minimal hit to sequential
+performance.
 

### 2020-11-20 retro-concurrency.tex
Minor edits

@@
 introduce stack overflow checks into the function prologue of OCaml functions.
-These stack overflow operations are rare and so will be correctly predicted by
-the CPU branch predictor.
+These stack overflows are rare and so the overflow checks will be correctly predicted
+by the CPU branch predictor.
 
@@
 
-\vspace{-2mm}
+%\vspace{-2mm}
 \section{Evaluation}
@@
 concurrency monad suffers due to the heap allocation of continuation frames
-(which need to be garbage collected), whereas effect handlers benefit from stack
-allocation of the frames. For example the number of major collections for the
-\texttt{ack} benchmark is 0 for stock OCaml, 1 for \texttt{MC} and 112 for
-\texttt{monad}. Our concurrency monad (and other monadic concurrency libraries
-such as Lwt~\cite{lwt} and Async~\cite{async}) also have other downsides --
-exceptions, backtraces, and DWARF unwinding are no longer useful due to the lack
-of a stack.
+(which need to be garbage collected), whereas effect handlers benefit from
+stack allocation of the frames. For example, the number of major collections
+for the |ack| benchmark is 0 for stock OCaml, 1 for |MC| and 112 for |monad|.
+Our concurrency monad (and other monadic concurrency libraries such as
+Lwt~\cite{lwt} and Async~\cite{async}) also have other downsides -- exceptions,
+backtraces, and DWARF unwinding are no longer useful due to the lack of a
+stack.
 
@@
 
-\vspace{-3mm}
+%\vspace{-3mm}
 \section{Conclusions}
@@
 the spirit of ``fast exceptions'' that has always characterised OCaml
-programming. We believe that the introduction of first-class fibres and effects
-into OCaml -- along with a parallel runtime~\cite{Sivaramakrishnan20} -- will
-open OCaml to highly task-parallel applications with minimal hit to sequential
-performance.
+programming. We believe that the introduction of lightweight fibers to support
+effect handlers into OCaml -- along with a parallel
+runtime~\cite{Sivaramakrishnan20} -- will open OCaml to highly task-parallel
+applications with minimal hit to sequential performance.
 

### 2020-11-20 retro-concurrency.tex
Conclusions

--- 2020-11-20 retro-concurrency.tex
@@
 
-%\vspace{-3mm}
+\vspace{-3mm}
 \section{Conclusions}
@@
 the spirit of ``fast exceptions'' that has always characterised OCaml
-programming. We believe that the introduction of lightweight fibers to support
-effect handlers into OCaml -- along with a parallel
-runtime~\cite{Sivaramakrishnan20} -- will open OCaml to highly task-parallel
-applications with minimal hit to sequential performance.
+programming. We believe that the introduction of effect handlers into OCaml
+implemented using lightweight fibers, along with a parallel
+runtime~\cite{Sivaramakrishnan20}, will open OCaml to highly scalable
+concurrent and task-parallel applications with minimal hit to sequential
+performance.
 

### 2020-11-21 retro-concurrency.tex
unblind for distribution

@@
-\documentclass[sigplan,10pt,review,anonymous]{acmart}\settopmatter{printfolios=true,printccs=false,printacmref=false}
+\documentclass[sigplan,10pt,review]{acmart}\settopmatter{printfolios=true,printccs=false,printacmref=false}
 \settopmatter{printacmref=false}
@@
 \title{Retrofitting Effect Handlers onto OCaml}
+
+\author{KC Sivaramakrishnan, Stephen Dolan, Leo White, Tom Kelly, Sadiq Jaffer and Anil Madhavapeddy}
+\acmConference[Draft Article]{}{21 Nov 2020}{Under Blind Review}%
 

### 2020-11-23 retro-concurrency.tex
Minor edits

@@
 user-interface continuation so that the application remains responsive. Observe
-that neither of the handlers needs to be aware of the other.
-
+that neither the division-by-zero handler nor the I/O handler needs to be aware
+of the other.
+
+\vspace{-1.5mm}
 \subsection{Retrofitting effect handlers}
@@
 the compiler to utilise a CPS IR would be an enormous undertaking that would
-affect the performance profile of existing OCaml applications due to increased
-memory allocations as continuation closures get allocated on the
+affect the performance profile of existing OCaml applications due to the increased
+memory allocations as the continuation closures get allocated on the
 heap~\cite{Farvardin20}. Moreover, with CPS, stack unwinding is meaningless and
 hence, we would lose compatibility with tools that inspect the program stack.
-Hence, we choose not to use CPS translation and represent continuations as call
-stacks.
+Hence, we choose not to use CPS translation and represent the continuations as
+call stacks.
 
@@
 
-\noindent \textbf{R1: Backwards compatibility.} A well-behaved OCaml program does
-not break under an OCaml that is extended with effect handlers. OCaml code that does not
-use effect handlers will pay negligible performance and memory cost.
-
-\noindent \textbf{R2: Tool compatibility.} OCaml programs with effect handlers produce
-well-formed backtraces and remain compatible with program analysis tools such
-as debuggers and profilers that inspect the stack using DWARF unwind
+\noindent \textbf{R1: Backwards compatibility.} A well-behaved OCaml program
+does not break under OCaml extended with effect handlers. OCaml code that does
+not use effect handlers will pay negligible performance and memory cost.
+
+\noindent \textbf{R2: Tool compatibility.} OCaml programs with effect handlers
+produce well-formed backtraces and remain compatible with program analysis
+tools such as debuggers and profilers that inspect the stack using DWARF unwind
 tables.
@@
 \noindent \textbf{R3: Effect handler efficiency.} The program must accommodate
-millions of continuations at the same time to support
-highly-concurrent applications. Installing effect handlers, capturing and
-resuming continuations must be fast.
-
-\noindent \textbf{R4: Forwards compatibility.} While this work does not present an
-effect system, the design must be compatible with a future effect system for
+millions of continuations at the same time to support highly-concurrent
+applications. Installing effect handlers, capturing and resuming continuations
+must be fast.
+
+\noindent \textbf{R4: Forwards compatibility.} While this work does not present
+an effect system, the design must be compatible with a future effect system for
 OCaml. As a cornerstone of modularity, we also want blocking I/O code to
@@
 We have implemented effect handlers in a multicore extension of the OCaml
-programming language which we call \textbf{Multicore OCaml} to distinguish
-it from stock OCaml.
+programming language which we call \textbf{Multicore OCaml} to distinguish it
+from stock OCaml.
 
@@
 operational semantics for Multicore OCaml effect handlers
-(\S\ref{sec:semantics}).  We then discuss the compiler and runtime system
+(\S\ref{sec:semantics}).  We then discuss the compiler and the runtime system
 support for implementing effect handlers (\S\ref{sec:impl}), and present an
@@
 The interaction of non-local control flow with systems programming is quite
-subtle~\cite{TFP17}. Consider the following function that uses blocking I/O
-functions from the OCaml standard library to copy data from the input channel
-|ic| to the output channel |oc|:
+subtle~\cite{TFP17, Leijen18}. Consider the following function that uses
+blocking I/O functions from the OCaml standard library to copy data from the
+input channel |ic| to the output channel |oc|:
 \begin{lstlisting}
@@
 
-We can then use a top-level I/O handler similar to the introductory example
-(\S\ref{sec:aio}) to discharge the I/O operations asynchronously and resumes
+We can then use a top-level I/O handler, similar to the introductory example
+(\S\ref{sec:aio}), to discharge the I/O operations asynchronously and resume
 with the result. This handles value return cases, but what about the
-exceptional cases |End_of_file| and |Sys_error|? We introduce a |discontinue|
-primitive to resume a continuation by raising an exception. In this example, on
-reaching the end of file, we would discontinue the continuation of the |copy|
-function with |discontinue k End_of_file|, which raises the exception at
-|input_line|, and the open channels will be closed.
+exceptional cases |End_of_file| and |Sys_error|? To this end, we introduce a
+|discontinue| primitive to resume a continuation by raising an exception. In
+this example, on reaching the end of file, we would discontinue the captured
+continuation of the |input_line| function with |discontinue k End_of_file|,
+which raises the exception at |input_line|, and the open channels will be
+closed.
 
@@
 handlers are well-typed Multicore OCaml programs. As a result, our static
-semantics is simpler than languages that ensure effect
-safety~\cite{Eff,Hillerstrom20,Leijen14,Effekt,Frank,Biernacki19}. This is
-important for backwards compatibility as our goal is to retrofit effect
-handlers to a language with large legacy codebases. Programs that do not use
-effects remain well-typed, and those that do compose well with those that
-don't.
+semantics is simpler than languages that ensure effect safety~\cite{Eff,
+Hillerstrom20, Leijen14, Effekt, Frank, Biernacki19}. This is important for
+backwards compatibility as our goal is to retrofit effect handlers to a
+language with large legacy codebases; programs that do not use effects remain
+well-typed, and those that do compose well with those that don't.
 
@@
 |type _ eff += E : string -> int eff|. We introduce the type
-|('a,'b) continuation| of delimited continuations which expect an |'a| value for
-resumption and return a |'b| value. The continuations may be continued with a
+|('a,'b) continuation| of delimited continuations which expects a |'a| value for
+resumption and returns a |'b| value. The continuations may be continued with a
 suitably typed value or discontinued with an exception.
@@
 faithfully captures the semantics of the Multicore OCaml implementation. An
-executable version of the semantics implemented as an OCaml interpreter along
-with examples is included in the supplementary material.
+executable version of the semantics, implemented as an OCaml interpreter, along
+with examples, is included in the supplementary material.
 
@@
 
-The operational semantics is an extension of a CEK machine
+The operational semantics is an extension of the CEK machine
 semantics~\cite{Felleisen86} for effect handlers, following the abstract
@@
 The stack $\stack$ is either a C stack ($\cstack$) or an OCaml stack
-($\ostack$). Every C stack $\cstacka{\fl}{\ostack}$ consists of a list of
-frames $\fl$, and the OCaml stack $\ostack$ under it. Every OCaml stack is
+($\ostack$). The C stack $\cstacka{\fl}{\ostack}$ consists of a list of
+frames $\fl$, and the OCaml stack $\ostack$ under it. The OCaml stack is
 either empty $\ostackemp$ or non-empty $\ostacka{k}{\cstack}$ with the current
@@
 
-Continuations $k$ are either empty $[]$ or a non-empty list of \emph{fibers}. A
+A continuation $k$ is either empty $[]$ or a non-empty list of \emph{fibers}. A
 fiber $\fiber \coloneqq (\fl,\hc)$ is a list of frames $\fl$ and a handler

### 2020-12-10 retro-concurrency.tex
Make all the fixes pointed out by Francois

@@
 \begin{lstlisting}
-let rec copy ic oc =
-  try output_string oc ((input_line ic) ^ "\n");
-      copy ic oc
-  with End_of_file -> close_in ic; close_out oc
+let copy ic oc =
+  let rec loop () =
+	  output_string oc ((input_line ic) ^ "\n");
+		loop ()
+	in
+	try loop () with
+  | End_of_file -> close_in ic; close_out oc
 	| e -> close_in ic; close_out oc; raise e
@@
 effect In_line : in_channel -> string
-effect Out_str : out_channel -> string -> unit
+effect Out_str : out_channel * string -> unit
 let input_line ic = perform (In_line ic)
-let output_string oc = perform (Out_str oc)
+let output_string oc s = perform (Out_str (oc, s))
 \end{lstlisting}
@@
 unnecessary and inefficient. Instead, Multicore OCaml optimises fibers for
-one-shot continuations. Fibers are allocated on the heap using |malloc| and are
-|free|d when no longer necessary. Similar to Farvardin et
-al.~\cite{Farvardin20}, we use a \emph{stack cache} of recently freed stacks in
-order to speed up allocation.
+one-shot continuations. Fibers are allocated on the C heap using |malloc| and
+are |free|d when the handled computation returns with a value or an exception.
+Similar to Farvardin et al.~\cite{Farvardin20}, we use a \emph{stack cache} of
+recently freed stacks in order to speed up allocation.
 
@@
 When the effect |E| is performed, we allocate a continuation object |ke| in the
-heap that points to the current fiber |f|, set fiber |f|'s parent pointer to
-|NULL|, and evaluate the continuation closure |clos_heffect| on the parent
+OCaml heap that points to the current fiber |f|, set fiber |f|'s parent pointer
+to |NULL|, and evaluate the continuation closure |clos_heffect| on the parent
 fiber |e| with the effect |E| and the continuation |ke| as arguments. Since the
@@
 until a matching one is found, the time taken to handle an effect is linear in
-the number of handlers. We observed that the handler stack is shallow in
-real programs.
+the number of handlers. We observed that the handler stack is shallow in real
+programs.
 
@@
 appropriate closure to either |continue| or |discontinue| the continuation
-(position |p3|). We perform standard tail-call optimisation so that resumptions
-at tail positions do not build up stack.
+(position |p3|). We perform tail-call optimisation so that resumptions at tail
+positions do not build up stack.
 
@@
 use an implementation of multi-prompt delimited continuations as an OCaml
-library~\cite{Kiselyov12} to embed the Eff language in OCaml. Indeed, Forester et
+library~\cite{Kiselyov12} to embed the Eff language in OCaml. Indeed, Forster et
 al.~\cite{Forster19} showed that in an untyped setting, effect handlers,

### 2020-12-11 retro-concurrency.tex
Updates

@@
 division-by-zero in a calculator application. Different calculators handle
-division-by-zero differently. In the iOS 13 Calculator (|iOS|), |0/0| returns
-|Error|. In the Google search calculator (|gsearch|) |0/0| returns |Error|, but
-for |n!=0| the |n/0| returns |Infinity| with the same sign as |n|. How do we
-capture all these different behaviours without changing the evaluator?
+division-by-zero differently. In the iOS 13 Calculator (|iOS|),
+division-by-zero returns |Error|. In the Google search calculator (|gsearch|)
+|0/0| returns |Error|, but for |n!=0| the |n/0| returns |Infinity| with the
+same sign as |n|. How do we capture all these different behaviours without
+changing the evaluator?
 
@@
 There are many research languages and libraries built around effect
-handlers~\cite{Leijen14,Hillerstrom20,Pyro,Frank,Eff}. Unlike these efforts, our
-goal is to retrofit effect handlers onto the OCaml programming language, which
-has been in continuous use for the past 25 years in large codebases including
-verification tools~\cite{everest,Coq}, mission critical software
+handlers~\cite{Leijen14,Hillerstrom20,Pyro,Frank,Eff}. Unlike these efforts,
+our goal is to retrofit effect handlers onto the OCaml programming language,
+which has been in continuous use for the past 25 years in large codebases
+including verification tools~\cite{everest,Coq}, mission critical software
 systems~\cite{astree} and latency sensitive networked
@@
 competitive yet predictable performance, with a fast foreign-function interface
-(FFI). It has excellent compatibility with program analysis tools such as debuggers
-and profilers that utilise DWARF stack unwind tables~\cite{DWARF} to obtain a
-backtrace. Our primary motivation for adding effect handlers to OCaml is to
-support: {\em (i)} highly concurrent applications such as web servers using a
-natural \emph{direct-style} as opposed to using callbacks; and {\em (ii)} efficient
-nested parallel programs in the vein of Cilk~\cite{Cilk}, OpenMP~\cite{OpenMP},
-X10~\cite{Charles05}, Intel TBB~\cite{IntelTBB}, and variants of Parallel
-ML~\cite{MaPLe, Fluet10, Sivaramakrishnan14}.
+(FFI). It has excellent compatibility with program analysis tools such as
+debuggers and profilers that utilise DWARF stack unwind tables~\cite{DWARF} to
+obtain a backtrace. Our primary motivation for adding effect handlers to OCaml
+is to support: {\em (i)} highly concurrent applications such as web servers
+using a natural \emph{direct-style} as opposed to using callbacks; and {\em
+(ii)} efficient nested parallel programs in the vein of Cilk~\cite{Cilk},
+OpenMP~\cite{OpenMP}, X10~\cite{Charles05}, Intel TBB~\cite{IntelTBB}, and
+variants of Parallel ML~\cite{MaPLe, Fluet10, Sivaramakrishnan14}.
 
@@
 the compiler to utilise a CPS IR would be an enormous undertaking that would
-affect the performance profile of existing OCaml applications due to the increased
-memory allocations as the continuation closures get allocated on the
+affect the performance profile of existing OCaml applications due to the
+increased memory allocations as the continuation closures get allocated on the
 heap~\cite{Farvardin20}. Moreover, with CPS, stack unwinding is meaningless and
@@
 requirements:
-
-\if{0}
-\begin{enumerate}[label=R\arabic*]
-	\item \textbf{Backwards compatibility} A well-behaved OCaml program does not
-		break under OCaml extended with effect handlers. OCaml code that does not
-		use effect handlers will pay negligible performance cost.
-  \item \textbf{Tool compatibility} OCaml with effect handlers produce
-    well-formed backtraces, and remains compatible with program analysis tools
-    such as debuggers and profilers that inspect the stack using DWARF stack
-    unwind tables.
-  \item \textbf{Effect handler efficiency} In order to support
-    highly-concurrent applications, the program must accommodate millions of
-    continuations at the same time. Installing effect handlers, capturing and
-    resuming continuations must be fast.
-  \item \textbf{Forwards compatibility} While this work does not present an
-    effect system, the design must be compatible with a future effect system
-    for OCaml. As a cornerstone of modularity, we also want blocking I/O code
-    to \emph{transparently} be made asynchronous with the help of effect
-    handlers.
-\end{enumerate}
-\fi
 

### 2021-02-10 PLDI_response/response.md
Added response for the first 3 reviews

We thank the reviewers for their constructive and high-quality reviews. We shall
take into account all of the minor comments in the final draft.

# Review A

We shall take a close look at the opportunistic optimisation from [24]. We have
experimental support for multi-shot continuations (see supplement). However,
multi-shot continuations currently break compiler optimisations that convert
heap allocations into stack allocation. The details are found in the supplement.
In the future, we plan to fully support multi-shot continuations in Multicore
OCaml.

We shall include the complete results in the evaluation section in the final
draft, so that the distribution of overheads become evident.

# Review B

As the reviewer has rightly suggested, we shall introduce citing [54] in section
1.5. We shall drop the reference to effect system from R4 and retain the
modularity arguments. Section 3 (line 437+) refers to the modularity requirement
and motivates the need for `discontinue` primitive.

Fibers are not necessary for parallelism, which is the main focus of [54].
Effect handlers (supported by fibers in the runtime) bring in native support for
concurrency to OCaml (the focus of this paper). The only reason to discuss
fibers in [54] was to highlight the interaction between the concurrent
mark-and-sweep GC and objects on the fiber stack. Indeed, we plan to retrofit
the OCaml language with parallelism without fibers, and then add fibers+effect
handlers.

The no-effect benchmarks, on which the outliers were reported, ever only have
one fiber. In these benchmarks, the roots on this only fiber stack are marked at
the beginning of the cycle, which is the same as what stock OCaml does. Hence,
the integration of fibers with concurrent mark-and-sweep GC does not contribute
to the observed differences. The observed differences are due to the difference
in the allocator designs between stock and multicore; multicore allocator has
less fragmentation and this leads to differences in the pacing of the
incremental GC.

We shall add comparison to domains (units of parallelism in Multicore OCaml) in
the microbenchmarks.

# Review C

We shall take into account the useful suggestions for improvements to the
introduction. We shall augment the example to better bring out the benefits of
dynamic scoping of handlers.

As the reviewer has pointed out, the modularity that we refer to is that the
change for asynchronous I/O is localised to the point at which the effect is
performed (`eval`) and where it is handled (`run`), and not the intervening code
(`gsearch`). Modularity ensures that neither the handler in `run` nor the one in
`gsearch` need to be aware of the other (lines 129-131). We shall improve the
presentation here.

The semantics aims to faithfully model the implementation modulo the one-shot
continuations and the specialised exception handler in the implementation. The
executable semantics has been useful for quickly prototyping different variants
of the effect handler (deep vs shallow, or making both available).

> MAP_GROWSDOWN

XXX TODO

> Relation to aspect-oriented programming

Kammar et al. "Handlers in Action", ICFP 13, makes connections between effect
handlers and aspect-oriented programming.

> line 291: "For external functions which allocate in the OCaml heap" -- how
  does the compiler know whether a native callee does or doesn't allocate?

Non-allocating external functions may optionally be annotated by the programmer.

> line 313: "In order to forward exceptions..." -- perhaps a silly question but
why is forwarding necessary? Also, can C frames install exception handlers?

C frames cannot install exception handlers, but there may be exception handlers
in the outer OCaml frames to which the exception must be forwarded.

> (This is for me one of the weaker points of the design... are finalisers
  *that* expensive? Is there experimental data on that?)

Thanks for this suggestion. We shall include a thorough evaluation of the
overheads of installing a finaliser for every captured continuation on the micro
and macro benchmarks.

### 2021-02-10 PLDI_response/response.md
update

@@
 
-> MAP_GROWSDOWN
+> Using MAP_GROWSDOWN
 
-XXX TODO
+The use of mmap with MAP_GROWSDOWN for fibers would mean that the fibers are at
+least 2 4k pages in size. Currently, our fibers start at 32 words. Secondly,
+stack overflow checks are not that expensive (as shown in our results). Lastly,
+we do want effect handlers to be supported on 32-bit platforms where OCaml runs.
 

### 2021-02-11 PLDI_response/response.md
Reorder review C responses to match order in reviewer's comments
Merge updates

--- 2021-02-10 PLDI_response/response.md
@@
 
-> MAP_GROWSDOWN
-
-XXX TODO
-
 > Relation to aspect-oriented programming
@@
 
+> Using MAP_GROWSDOWN
+
+The use of mmap with MAP_GROWSDOWN for fibers would mean that the fibers are at
+least two 4k pages in size. Currently, our fibers start at 32 words. Secondly,
+stack overflow checks are not that expensive (as shown in our results). Last, we
+do want effect handlers to be supported on 32-bit platforms where OCaml runs,
+even if parallelism may not be supported. Stack switching is portable across the
+platforms supported by OCaml. That said, MAP_GROWSDOWN solution may be
+appropriate where the stacks cannot be moved (which we understand is the
+situation with Web Assembly).
+
+> Relation to aspect-oriented programming
+
+Kammar et al. "Handlers in Action", ICFP 13, makes some connections between
+effect handlers and aspect-oriented programming.
+
 > line 291: "For external functions which allocate in the OCaml heap" -- how
@@
 
-Non-allocating external functions may optionally be annotated by the programmer.
+Non-allocating external functions may optionally be annotated by the programmer
+as `@noalloc`. The compiler does not ensure that these annotations are correct.
 
@@
 
-Thanks for this suggestion. We shall include a thorough evaluation of the
+Thanks for this suggestion. As other reviewers have pointed out, this is one of
+the weaknesses of our solution. We shall include a thorough evaluation of the
 overheads of installing a finaliser for every captured continuation on the micro
-and macro benchmarks.
+and macro benchmarks to justify our trade off.
+
+# Review 4
+
+We shall go for a more illustrative example in the introduction (as suggested by
+other reviewers as well). We shall add the main results in the body of the paper
+for the final version.
+
+Multicore OCaml clearly separates out concurrency from parallelism (running on
+2+ cores) using distinct mechanisms to express them -- domains and effect
+handlers. We can have one without the other. [54] discusses retrofitting
+parallelism to OCaml; fibers are not necessary for parallelism. The focus of
+[54] is the concurrent garbage collector. [54] briefly touches upon fibers only
+to describe the interaction with the concurrent GC. None of the benchmarks in
+[54] use effect handlers. The focus of this paper is retrofitting concurrency to
+OCaml. None of the benchmarks in this paper involve parallelism.
+
+Our aim with the semantics was to formally capture the semantics of stack
+manipulation, making explicit the interaction between C and OCaml stacks, which
+other presentations typically avoid. We chose not to model one-shot
+continuations since they are well understood. In the implementation,
+continuations are fresh heap objects that refer to the fiber. This reference
+gets overwritten with NULL when a continuation is resumed. Continuation
+resumption raises exception when the reference to the fiber in a continuation is
+NULL. Modelling this in the semantics would be straight-forward (by adding a
+heap to the configuration), but tedious. It does not bring in novel insights but
+would obscure the presentation. Hence, we avoided encoding one-shot semantics.
+
+Since the exactly-once use of continuations is an expectation not enforced by
+the compiler, the semantics would not change. In the final version, we shall
+state upfront (in Section 4.2) that the dynamic semantics aims to model stack
+manipulation.
+
+Hillerstrom et al. [29] do not model C and OCaml stacks explicitly. [29] is also
+in a setting with an effect system. Hence, programs with unhandled effects would
+be statically rejected. In this work, unhandled effects raise exceptions at the
+point of perform to clean up resources. [29] models both deep and shallow
+handlers. Our handlers are deep. [29] uses fine-grained call-by-value (FGCBV),
+but we use call-by-value (CBV). While FGCBV would avoid a few administrative
+reduxes by having a separate syntactic category of "interesting" operations, we
+chose CBV so that the examples in the executable semantics (see supplement)
+remain syntactically similar to their analogues in Multicore OCaml.

### 2021-02-11 PLDI_response/response.md
update

@@
 
-As the reviewer has rightly suggested, we shall introduce citing [54] in section
-1.5. We shall drop the reference to effect system from R4 and retain the
-modularity arguments. Section 3 (line 437+) refers to the modularity requirement
-and motivates the need for `discontinue` primitive.
+As the reviewer has rightly suggested, we shall introduce Multicore OCaml citing 
+[54] in section 1.5. We shall drop the reference to effect system from R4 and 
+retain the modularity arguments. Section 3 (line 437+) refers to the modularity 
+requirement and motivates the need for `discontinue` primitive.
 
@@
 to the observed differences. The observed differences are due to the difference
-in the allocator designs between stock and multicore; multicore allocator has
-less fragmentation and this leads to differences in the pacing of the
-incremental GC.
+in the allocator designs between stock and multicore; Multicore OCaml's 
+size-segmented free-list allocator has less fragmentation that the stock OCaml's 
+next fit allocator and this leads to differences in the pacing of the incremental 
+GC.
 
@@
 We shall take into account the useful suggestions for improvements to the
-introduction. We shall augment the example to better bring out the benefits of
-dynamic scoping of handlers.
+introduction. We shall choose more illustrative example to better bring out 
+the benefits of dynamic scoping of handlers.
 
@@
 (`gsearch`). Modularity ensures that neither the handler in `run` nor the one in
-`gsearch` need to be aware of the other (lines 129-131). We shall improve the
-presentation here.
+`gsearch` need to be aware of the other (lines 129-131). We shall make this
+explicit in the text.
 
@@
 of the effect handler (deep vs shallow, or making both available).
-
-> Relation to aspect-oriented programming
-
-Kammar et al. "Handlers in Action", ICFP 13, makes connections between effect
-handlers and aspect-oriented programming.
 
@@
 do want effect handlers to be supported on 32-bit platforms where OCaml runs,
-even if parallelism may not be supported. Stack switching is portable across the
-platforms supported by OCaml. That said, MAP_GROWSDOWN solution may be
-appropriate where the stacks cannot be moved (which we understand is the
+even if parallelism may not be supported. The current solutions is portable 
+across the platforms supported by OCaml. That said, MAP_GROWSDOWN solution 
+may be appropriate where the stacks cannot be moved (which we understand is the
 situation with Web Assembly).
@@
 
-Multicore OCaml clearly separates out concurrency from parallelism (running on
+Multicore OCaml separates out concurrency from parallelism (running on
 2+ cores) using distinct mechanisms to express them -- domains and effect
@@
 Since the exactly-once use of continuations is an expectation not enforced by
-the compiler, the semantics would not change. In the final version, we shall
-state upfront (in Section 4.2) that the dynamic semantics aims to model stack
-manipulation.
+the compiler, the semantics would not change due to this requirement. In the 
+final version, we shall state upfront (in Section 4.2) that the dynamic semantics 
+aims to model stack manipulation.
 

### 2021-03-10 retro-concurrency.tex
Edits

@@
 
-\vspace{-1.5mm}
+%\vspace{-1.5mm}
 \subsection{Retrofitting effect handlers}
@@
 
-\noindent \textbf{R4: Forwards compatibility.} While this work does not present
-an effect system, the design must be compatible with a future effect system for
-OCaml. As a cornerstone of modularity, we also want blocking I/O code to
-\emph{transparently} be made asynchronous with the help of effect handlers.
+\noindent \textbf{R4: Forwards compatibility.} As a cornerstone of modularity,
+we also want blocking I/O code to \emph{transparently} be made asynchronous
+with the help of effect handlers.
 
@@
 programming language which we call \textbf{Multicore OCaml} to distinguish it
-from stock OCaml.
+from stock OCaml. Multicore OCaml clearly delineates concurrency (overlapped
+execution of tasks) from parallelism (simultaneous execution of tasks) with
+separate mechanisms for expressing them. Sivaramakrishnan et
+al.~\cite{Sivaramakrishnan20} describe the parallelism support in Multicore
+OCaml enabled by \emph{domains}. The focus of this paper is the concurrency
+support enabled by effect handlers.
 
@@
     \end{lstlisting}
-    \vspace{-2mm}
+    %\vspace{-2mm}
     \subcaption{\texttt{meander.ml}}
@@
     \end{lstlisting}
-    \vspace{-4mm}
+    %\vspace{-4mm}
     \subcaption{\texttt{meander.c}}
@@
 
-For stack walking during GC, the only change necessary was to make the runtime
-aware of fiber layout and how to follow the parent pointers. The stack map
-generation and scanning each frame for roots remains the same. The
+For stack walking during GC, the only change necessary was to make the
+runtime aware of fiber layout and how to follow the parent pointers. The
+stack map generation and scanning each frame for roots remains the same. The
 challenges and solutions for integrating fibers with the concurrent
-mark-and-sweep GC of Multicore OCaml have been discussed previously~\cite{Sivaramakrishnan20}.
+mark-and-sweep GC of Multicore OCaml have been discussed
+previously~\cite{Sivaramakrishnan20}.
 
@@
 \begin{figure*}
-	\includegraphics[angle=90,scale=0.44]{sandmark-notebook/sandmark_time}
+	\includegraphics[scale=0.35]{sandmark-notebook/sandmark_time}
 	\caption{Normalized time of macro benchmarks. Baseline is Stock OCaml,
@@
 \begin{figure*}
-	\includegraphics[angle=90,scale=0.44]{sandmark-notebook/sandmark_codesize}
+	\includegraphics[scale=0.35]{sandmark-notebook/sandmark_codesize}
 	\caption{Normalized OCaml text section size (OTSS) of macro benchmarks.

### 2021-03-11 retro-concurrency.tex
Landscape mode supplement figures

@@
 \end{lstlisting}
-\vspace{-2mm}
+%\vspace{-2mm}
 \subcaption{\texttt{gdb} backtrace before raise \texttt{E1}.}
@@
 \end{minipage}
-\vspace{-2mm}
+%\vspace{-2mm}
 \caption{Program stack on stock OCaml.}
-\vspace{-2mm}
+%\vspace{-2mm}
 \end{figure*}
@@
 \begin{figure*}
-\vspace{-2mm}
+%\vspace{-2mm}
 \begin{minipage}{\linewidth}
@@
 
-\vspace{-1mm}
+%\vspace{-1mm}
 \subsubsection{Top-level reductions}
@@
 
-\vspace{-1mm}
+%\vspace{-1mm}
 \subsubsection{C reductions}
@@
 
-\vspace{-1mm}
+%\vspace{-1mm}
 \subsubsection{OCaml reductions}
@@
     \label{fig:mcstack}
-    \vspace{2mm}
+    %\vspace{2mm}
   \end{minipage}
@@
   \begin{minipage}{0.55\linewidth}
+		\vspace{1mm}
     \begin{lstlisting}
@@
 effect F : unit;;
+
 match (* comp_e *)
@@
   \begin{minipage}{0.44\linewidth}
+		\vspace{2mm}
     \centering
@@
 \end{minipage}
-\vspace{-2mm}
+%\vspace{-2mm}
 \caption{Layout of Multicore OCaml effect handlers.}
-\vspace{-3mm}
+%\vspace{-3mm}
 \end{figure*}
@@
 
-\vspace{-3mm}
+%\vspace{-3mm}
 \subsection{External calls and callbacks}
@@
 
-\vspace{-3mm}
+%\vspace{-3mm}
 \subsection{Effect handlers}
@@
 
-\vspace{-3mm}
+%\vspace{-3mm}
 \subsection{Stack unwinding}
@@
 
-\vspace{-2mm}
+%\vspace{-2mm}
 \subsection{Garbage collection}
@@
 
-The complete results are presented in the supplementary material, which we
-summarise here. On average (geometric mean of the normalized values against
-|stock| as the baseline), the multicore variants were \textbf{less than 1\%
-slower} than |stock|. The outliers (on either ends) were due to the difference
-in the allocator and the GC between stock and Multicore OCaml.
+\begin{figure*}
+	\includegraphics[scale=0.35]{sandmark-notebook/sandmark_time}
+	\caption{Normalized time of macro benchmarks. Baseline is Stock OCaml,
+		whose running time in seconds in given in parenthesis.}
+	\label{res:macro_time}
+\end{figure*}
+
+\begin{figure*}
+	\includegraphics[scale=0.35]{sandmark-notebook/sandmark_codesize}
+	\caption{Normalized OCaml text section size (OTSS) of macro benchmarks.
+	Baseline is Stock OCaml, whose size in kilobytes given in parenthesis.}
+	\label{res:macro_size}
+\end{figure*}
+
+Figure~\ref{res:macro_time} presents the running time of the different
+multicore variants normalized against the sequential baseline |stock|. On
+average (geometric mean of the normalized values against |stock| as the
+baseline), the multicore variants were \textbf{less than 1\% slower} than
+|stock|. The outliers (on either ends) were due to the difference in the
+allocator and the GC between stock and Multicore OCaml.
 
@@
 debug symbols, the text sections associated with OCaml runtime and other
-statically linked C libraries. Compared to |stock|, OTSS is 19\% more for |MC|
-and |MC+RedZone32|, and 30\% more for |MC+RedZone0|. The result shows that our
-16-word red zone is effective at reducing OTSS compared to having no red zone,
-whereas the 32-word red zone does not noticeably reduce OTSS further. Further
-work is required to bring OTSS closer to |stock|.
+statically linked C libraries. Figure~\ref{res:macro_size} presents the OTSS of
+the multicore variants normalized against the sequential baseline |stock|.
+Compared to |stock|, OTSS is 19\% more for |MC| and |MC+RedZone32|, and 30\%
+more for |MC+RedZone0|. The result shows that our 16-word red zone is effective
+at reducing OTSS compared to having no red zone, whereas the 32-word red zone
+does not noticeably reduce OTSS further. Further work is required to bring OTSS
+closer to |stock|.
 
@@
 	difference for Multicore OCaml over stock OCaml.}
-\vspace{-7mm}
+%\vspace{-7mm}
 {
@@
 \end{tabular}
-\vspace{-5mm}
+%\vspace{-5mm}
 }
@@
 	slowdown factor ($\times$ times) over its idiomatic implementation in stock OCaml.}
-\vspace{-3mm}
+%\vspace{-3mm}
 {
@@
 \label{tab:micro_noperform}
-	\vspace{-5mm}
+	%\vspace{-5mm}
 \end{table}
@@
 		\includegraphics[width=0.9\linewidth]{http-benchmarks/throughput.pdf}
-		\vspace{-2mm}
+		%\vspace{-2mm}
 		\subcaption{Throughput}
@@
 		\includegraphics[width=0.9\linewidth]{http-benchmarks/latency.pdf}
-		\vspace{-2mm}
+		%\vspace{-2mm}
 		\subcaption{Tail latency}
@@
 	\end{minipage}
-	\vspace{-3mm}
+	%\vspace{-3mm}
 	\caption{Web server performance.}
-	\vspace{-5mm}
+	%\vspace{-5mm}
 \end{figure}
@@
 
-\vspace{-2mm}
+%\vspace{-2mm}
 \section{Related Work}
@@
 
-\clearpage
+%\clearpage
 
@@
 \fi
-
-\subsection*{No effect macro benchmarks}
-
-\begin{figure*}
-	\includegraphics[scale=0.35]{sandmark-notebook/sandmark_time}
-	\caption{Normalized time of macro benchmarks. Baseline is Stock OCaml,
-		whose running time in seconds in given in parenthesis.}
-	\label{res:macro_time}
-\end{figure*}
-
-\begin{figure*}
-	\includegraphics[scale=0.35]{sandmark-notebook/sandmark_codesize}
-	\caption{Normalized OCaml text section size (OTSS) of macro benchmarks.
-	Baseline is Stock OCaml, whose size in bytes given in parenthesis.}
-	\label{res:macro_size}
-\end{figure*}
-
-Figures~\ref{res:macro_time} and \ref{res:macro_size} present the impact on
-macro benchmarks that do not use effects using the Multicore OCaml compiler
-which supports effect handlers. The aim is to quantify the impact of the
-addition of effect handlers to the language on code that does not use effect
-handlers. The |MC| is Multicore OCaml with the default red zone size (16
-words). |MC+RedZone0| uses red zone size of 0; every OCaml function has a stack
-overflow check. |MC+RedZone32| uses a red zone size of 32 words.
 

### 2021-03-11 retro-concurrency.tex
updates

@@
-\documentclass[sigplan,10pt,review]{acmart}\settopmatter{printfolios=true,printccs=false,printacmref=false}
-\settopmatter{printacmref=false}
+\documentclass[sigplan,screen]{acmart}
 \setcopyright{none}
-\renewcommand\footnotetextcopyrightpermission[1]{}
-\pagestyle{plain}
+\acmPrice{}
+\acmDOI{10.1145/3453483.3454039}
+\acmYear{2021}
+\copyrightyear{2021}
+\acmSubmissionID{pldi21main-p82-p}
+\acmISBN{978-1-4503-8391-2/21/06}
+\acmConference[PLDI '21]{Proceedings of the 42nd ACM SIGPLAN International Conference on Programming Language Design and Implementation}{June 20--25, 2021}{Virtual, UK}
+\acmBooktitle{Proceedings of the 42nd ACM SIGPLAN International Conference on Programming Language Design and Implementation (PLDI '21), June 20--25, 2021, Virtual, UK}
 
@@
 
-\author{KC Sivaramakrishnan, Stephen Dolan, Leo White, Tom Kelly, Sadiq Jaffer and Anil Madhavapeddy}
-\acmConference[Draft Article]{}{21 Nov 2020}{Under Blind Review}%
+\author{KC Sivaramakrishnan}
+\affiliation{
+  \institution{IIT Madras}            %% \institution is required
+	\country{India}
+}
+\email{kcsrk@cse.iitm.ac.in}          %% \email is recommended
+
+\author{Stephen Dolan}
+\affiliation{
+  \institution{OCaml Labs}            %% \institution is required
+	\country{UK}
+}
+\email{stephen.dolan@cl.cam.ac.uk}
+
+\author{Leo White}
+\affiliation{
+  \institution{Jane Street}            %% \institution is required
+	\country{UK}
+}
+\email{leo@lpw25.net}
+
+\author{Sadiq Jaffer}
+\affiliation{
+  \institution{Opsian and OCaml Labs} %% \institution is required
+	\country{UK}
+}
+\email{sadiq@toao.com}
+
+\author{Tom Kelly}
+\affiliation{
+  \institution{OCaml Labs}            %% \institution is required
+	\country{UK}
+}
+\email{tom.kelly@cantab.net}
+
+\author{Anil Madhavapeddy}
+\affiliation{
+  \institution{University of Cambridge Computer Laboratory and OCaml Labs}            %% \institution is required
+	\country{UK}
+}
+\email{avsm2@cl.cam.ac.uk}          %% \email is recommended
 
@@
 is implemented. This computation may be enclosed by different handlers that
-interpret |Read| differently. For example, |Read| may be implemented by
-performing a blocking read on the input channel or performing the read
-asynchronously by offloading it to an event loop such as |libuv|. Thanks to the
-separation of effectful operations from their interpretations, effect handlers
-enable new approaches to modular programming.
-
-\subsection{An effectful calculator}
+handle |Read| differently. For example, |Read| may be implemented by performing
+a blocking read on the input channel or performing the read asynchronously by
+offloading it to an event loop such as |libuv|. Thanks to the separation of
+effectful operations from their interpretations, effect handlers enable new
+approaches to modular programming.
+
+\subsection{Asynchronous I/O}
 
@@
 continuations~\cite{Danvy90} that enables non-local control-flow mechanisms
-such as resumable exceptions, coroutines, generators and asynchronous I/O to be
-composably expressed. For example, consider the task of handling
-division-by-zero in a calculator application. Different calculators handle
-division-by-zero differently. In the iOS 13 Calculator (|iOS|),
-division-by-zero returns |Error|. In the Google search calculator (|gsearch|)
-|0/0| returns |Error|, but for |n!=0| the |n/0| returns |Infinity| with the
-same sign as |n|. How do we capture all these different behaviours without
-changing the evaluator?
-
-Common Lisp Conditions~\cite{Steele90} are an elegant way to solve this
-problem. Unlike exceptions where the library decides whether an error is fatal
-or not, conditions allow the client of the library to make that decision.
-Conditions provide a \emph{restart} mechanism for non-fatal errors to resume
-execution from the point where the condition was signalled. On
-division-by-zero, the calculator signals the handler and lets the handler
-decide the course of action based on the desired semantics. We now implement
-this using effect handlers in OCaml syntax:
+such as resumable exceptions, lightweight threads, coroutines, generators and
+asynchronous I/O to be composably expressed. Let us illustrate the power of
+effect handlers in OCaml by implementing a simple lightweight thread library
+with support for forking new threads and yielding control to other threads. We
+will then extend this scheduler with support for synchronously reading from
+channels and subsequently make it asynchronous without changing the client code
+for asynchrony.
+
+Many programming languages, including OCaml, require non-local change to the
+code in order to support asynchronous I/O, often leading to a dichotomy
+synchronous and asynchronous code~\cite{colour}. Since OCaml does not have
+primitive support for concurrency, users will have to use monadic I/O libraries
+such as Lwt~\cite{lwt} and Async~\cite{async}. Languages such as GHC Haskell
+and Go provide lightweight threads but bake their implementation into the
+runtime system. With effect handlers, asynchronous I/O can be implemented
+directly in OCaml.
+
+In order to support forking and yielding threads, we declare effects the
+following effects:
+
+\begin{lstlisting}
+effect Fork : (unit -> unit) -> unit
+effect Yield : unit
+\end{lstlisting}
+
+The |Fork| effect takes a thunk which is spawned as a concurrent thread, and
+|Yield| yields control to another thread in the scheduler queue. We can define
+helper functions to perform these effects:
+
+\begin{lstlisting}
+let fork f = perform (Fork f)
+let yield () = perform Yield
+\end{lstlisting}
+
+The implementation of the scheduler queue is defined in the |run| function,
+which handles the effects appropriately:
 
 \begin{lstlisting}[numbers=left]
-type sign = P | N
-type val = Z of int | Error | Inf of sign
-type exp = Val of val | ... | Div of exp*exp
-effect DivBy0 : int -> val
-let rec eval e =
-  match e with
-  | Val v -> v
-  | ...
-  | Div (e1,e2) ->
-    match eval e1, eval e2 with
-    | Z v1, Z v2 ->
-        if v2 = 0 then perform (DivBy0 v1)
-        else Z (v1/v2)
-    | Error, _ -> Error | Inf s, _ -> Inf s
-    | ...
-let (/) n d = Div (Val (Z n), Val (Z d))
+let run main =
+  let runq = Queue.create () in
+  let suspend k = Queue.push k runq in
+  let rec run_next () =
+    match Queue.take_opt runq with
+    | None -> ()
+    | Some k -> continue k ()
+  in
+  let rec spawn f =
+    match f () with
+		| () -> run_next () (* value case *)
+    | effect Yield k -> suspend k; run_next ()
+    | effect (Fork f') k -> suspend k; spawn f'
+  in
+  spawn main
 \end{lstlisting}
 
-Line 4 declares an effect |DivBy0| that is parameterised by an integer, which
-when performed returns the result of division-by-zero. If the denominator of
-the division is |0|, then we perform the |DivBy0| effect with the numerator
-value (line 12) so that the context may decide the return value. We obtain the
-|iOS| semantics by wrapping the evaluator with this handler:
+The function |spawn| (line 9) evaluates the computation |f| in an effect
+handler. The computation |f| may return normally with a value, or perform
+effects |Fork f'| and |Yield|. The pattern |effect Yield k| handles the effect
+|Yield| and binds |k| to the continuation of the corresponding |perform|
+delimited by this handler. The scheduler queue |runq| maintains a queue of
+these continuations. |suspend| pushes continuations into the queue, |run_next|
+pops continuations from the queue and resumes them with |()| value using the
+|continue| primitive. In the case of |Yield| effect, we suspend the current
+continuation |k| and resume the next available continuation. In the case of
+|Fork f'| effect, we suspend the current continuation and recursively call
+|spawn| on |f'| in order to run |f'| concurrently.
+
+Effect handlers provide modularity. Observe that we can change the scheduling
+algorithm from FIFO to LIFO by changing the scheduler queue to a stack. We can
+implement support for synchronous read from channels by adding the following
+case to the effect handler in |spawn|:
+
 \begin{lstlisting}
-let ios e = match eval e with
-| v -> v (* value case *)
-| effect (DivBy0 _) k -> continue k Error (* effect case *)
+let rec spawn f =
+  match f () with
+	...
+  | effect (Read ic) k -> continue k (input_line ic)
 \end{lstlisting}
-If the evaluator returns a value, we simply return that. If the evaluator
-performs the |DivBy0| effect, we catch the effect in the pattern match along with
-the continuation |k| of the corresponding |perform| \emph{delimited} by this
-handler. For |iOS|, we |continue| the suspended computation with an
-|Error| value and the computation |ios(1/0)| returns |Error|.
-We can get the |gsearch| semantics by enclosing the evaluator in this handler:
-\begin{lstlisting}
-let gsearch e = match eval e with
-| v -> v
-| effect (DivBy0 v) k ->
-		if v = 0 then continue k Error
-		else if v > 0 then continue k (Inf P)
-		else continue k (Inf N)
+%
+which uses OCaml standard library's |input_line| function to read a line from
+the channel |ic| and resume the continuation |k| with the resultant string.
+However, performing reads synchronously blocks the entire scheduler, preventing
+other threads to run until the I/O is completed.
+
+We can make the I/O asynchronous by modifying the |run| function as follows:
+
+\begin{lstlisting}[numbers=left]
+let run main =
+  let runq = Queue.create () in
+	let suspend k =
+	  Queue.push (fun () -> continue k ()) runq
+	in
+  let pending_reads = ref [] in
+  let rec run_next () =
+    match Queue.take_opt runq with
+    | None ->
+      begin match !pending_reads with
+      | [] -> () (* no pending reads *)
+      | todo ->
+        let (done_,todo) = do_reads todo in
+        List.iter (fun (str,k) ->
+          Queue.push (fun () -> continue k str) runq) done_;
+        pending_reads := todo;
+        run_next ()
+      end
+    | Some f -> f ()
+  in
+  let rec spawn f =
+    match f () with
+		...
+    | effect (Read ic) k ->
+        pending_reads := (ic,k)::!pending_reads;
+        run_next ()
+  in
+  ...
 \end{lstlisting}
-Executing |gsearch(1/0)| returns positive infinity (|Inf P|).
-
-\subsection{Asynchronous I/O}
-\label{sec:aio}
-
-We can modularly extend the calculator to read user input using the |Read|
-effect defined earlier as follows:
-\begin{lstlisting}
-type exp = ... | Inp
-effect Read : in_channel -> string
-let rec eval e = match e with
-| ...
-| Inp -> int_of_string (perform (Read stdin))
-\end{lstlisting}
-The |Read| effect can then be handled by another handler enclosing the
-division-by-zero handler:
-\begin{lstlisting}
-let run e = match gsearch e with
-| v -> v
-| effect (Read chan) k ->
-		register_async_input_line (chan, k);
-		continue (get_ui_k ()) ()
-\end{lstlisting}
-\noindent which registers the read operation to be performed asynchronously
-with the continuation as the callback. Subsequently, it resumes the
-user-interface continuation so that the application remains responsive. Observe
-that neither the division-by-zero handler nor the I/O handler needs to be aware
-of the other.
+
+The scheduler queue |runq| now holds thunks instead of continuations. The value
+|pending_reads| maintains a list of pending reads and the associated
+continuations (line 6). We handle the |Read| effect by pushing the pair of
+input channel |ic| and continuation |k| to |pending_reads| (line 25), allowing
+other threads in the scheduler to run.
+
+When the scheduler queue is empty, the |run_next| function performs the pending
+reads. We abstract away the details of the event-based I/O using the |do_reads|
+function (line 13). |do_reads| takes a list of pending reads and blocks until
+at least one of the reads succeeds. It returns a pair of lists |done_| and
+|todo|. |done_| contains the result strings from successful reads and
+corresponding continuations. These continuations are arranged to be resumed
+with the read result and pushed into the scheduler queue. |todo| contains the
+channels on which input is still pending and their corresponding continuations.
+|pending_reads| is updated to point to the |todo| list so that they may be
+attempted later. Observe that all of the changes to add asynchrony are
+localised to the |run| function, and the computation that performs these
+effects can remain in \emph{direct-style} (as opposed to the monadic-style in
+Lwt and Async).
 
@@
 
-\noindent \textbf{R1: Backwards compatibility.} A well-behaved OCaml program
-does not break under OCaml extended with effect handlers. OCaml code that does
+\begin{enumerate}[label=R\arabic*]
+	\item \textbf{Backwards compatibility.} A well-behaved OCaml program does
+		not break under OCaml extended with effect handlers. OCaml code that does
 		not use effect handlers will pay negligible performance and memory cost.
 
-\noindent \textbf{R2: Tool compatibility.} OCaml programs with effect handlers
+	\item \textbf{Tool compatibility.} OCaml programs with effect handlers
 		produce well-formed backtraces and remain compatible with program analysis
-tools such as debuggers and profilers that inspect the stack using DWARF unwind
-tables.
-
-\noindent \textbf{R3: Effect handler efficiency.} The program must accommodate
+		tools such as debuggers and profilers that inspect the stack using DWARF
+		unwind tables.
+
+	\item \textbf{Effect handler efficiency.} The program must accommodate
 		millions of continuations at the same time to support highly-concurrent
-applications. Installing effect handlers, capturing and resuming continuations
-must be fast.
-
-\noindent \textbf{R4: Forwards compatibility.} As a cornerstone of modularity,
-we also want blocking I/O code to \emph{transparently} be made asynchronous
+		applications. Installing effect handlers, capturing and resuming
+		continuations must be fast.
+
+	\item \textbf{Forwards compatibility.} As a cornerstone of modularity, we
+		also want blocking I/O code to \emph{transparently} be made asynchronous
 		with the help of effect handlers.
+\end{enumerate}
 
@@
 
-\textbf{\textit{Generators.}} Generators allow data structures to be traversed on
-demand. Many languages including JavaScript and Python provide generators as a
-built-in primitive. Using effect handlers (|MC|), given any data structure
-(|'a t|) and its iterator (|val iter: 'a t -> ('a -> unit) -> unit|), we can
-derive its generator function (|val next : unit -> 'a option|). The code is
-available in the supplementary material. We evaluate the performance of
-traversing a complete binary tree of depth 25 using this generator. This
-involves $2^{26}$ stack switches in total. For comparison, we implemented
-a hand-written, selective CPSed~\cite{Nielson01},
-defunctionalised~\cite{Danvy01} version (|cps|) and a concurrency monad
-(|monad|) version of the generator for the tree. Both |cps| and |monad|
-versions are specialised to the binary tree with the usual caveats of
-not using the stack for function calls. We observed that the |cps| version was
-the fastest, thanks to specialisation and hand optimisation. |MC| version was
-only 2.76$\times$ slower than |cps| while being a generic
-solution, and the |monad| version was 8.69$\times$ slower than |cps|.
-
-\textbf{\textit{Chameneos.}} Chameneos~\cite{Chameneos} is a concurrency game
-aimed at measuring context switching and synchronization overheads. Our
-implementation uses |MVars| for synchronization. We compare effect handler
-(|MC|), concurrency monad (|monad|) and Lwt, a widely used concurrency
-programming library for OCaml (|lwt|) versions. We observed that |MC| was the
-fastest, and |monad| (|lwt|) was 1.67$\times$ (4.29$\times$) slower than |MC|.
-
-\textbf{\textit{Web server.}} As one of our primary goals of adding effect
-handlers to OCaml is to support direct-style concurrency, we implement a
-full-fledged HTTP/1.1 web server using effect handlers (|MC|). We use effect
-handlers to implement a user-level threading library that spawns a thread per
-request. We use \texttt{httpaf}~\cite{httpaf} for HTTP handling, and
+\subsubsection{Generators}
+
+Generators allow data structures to be traversed on demand. Many languages
+including JavaScript and Python provide generators as a built-in primitive.
+Using effect handlers (|MC|), given any data structure (|'a t|) and its
+iterator (|val iter: 'a t -> ('a -> unit) -> unit|), we can derive its
+generator function (|val next : unit -> 'a option|). The code is available in
+the supplementary material. We evaluate the performance of traversing a
+complete binary tree of depth 25 using this generator. This involves $2^{26}$
+stack switches in total. For comparison, we implemented a hand-written,
+selective CPSed~\cite{Nielson01}, defunctionalised~\cite{Danvy01} version
+(|cps|) and a concurrency monad (|monad|) version of the generator for the
+tree. Both |cps| and |monad| versions are specialised to the binary tree with
+the usual caveats of not using the stack for function calls. We observed that
+the |cps| version was the fastest, thanks to specialisation and hand
+optimisation. |MC| version was only 2.76$\times$ slower than |cps| while being
+a generic solution, and the |monad| version was 8.69$\times$ slower than |cps|.
+
+\subsubsection{Chameneos}
+
+Chameneos~\cite{Chameneos} is a concurrency game aimed at measuring context
+switching and synchronization overheads. Our implementation uses |MVars| for
+synchronization. We compare effect handler (|MC|), concurrency monad (|monad|)
+and Lwt, a widely used concurrency programming library for OCaml (|lwt|)
+versions. We observed that |MC| was the fastest, and |monad| (|lwt|) was
+1.67$\times$ (4.29$\times$) slower than |MC|.
+
+\subsubsection{Webserver}
+
+As one of our primary goals of adding effect handlers to OCaml is to support
+direct-style concurrency, we implement a full-fledged HTTP/1.1 web server using
+effect handlers (|MC|). We use effect handlers to implement a user-level
+threading library that spawns a thread per request. We use
+\texttt{httpaf}~\cite{httpaf} for HTTP handling, and
 \texttt{libev}~\cite{libev} for the eventloop. We compare our implementation
 against an Lwt version (|lwt|) which also uses \texttt{httpaf} and
-\texttt{libev}. Unlike using effect handlers, the Lwt version is written in monadic
-style and does not have the notion of a thread per request. For comparison, we
-include a Go 1.13 version (|go|) that uses the
+\texttt{libev}. Unlike using effect handlers, the Lwt version is written in
+monadic style and does not have the notion of a thread per request. For
+comparison, we include a Go 1.13 version (|go|) that uses the
 \texttt{net/http}~\cite{nethttp} package. As both the OCaml versions are single
@@
 		\centering
-		\includegraphics[width=0.9\linewidth]{http-benchmarks/throughput.pdf}
+		\includegraphics[width=\linewidth]{http-benchmarks/throughput.pdf}
 		%\vspace{-2mm}
@@
 	\begin{minipage}{0.49\linewidth}
-		\includegraphics[width=0.9\linewidth]{http-benchmarks/latency.pdf}
+		\includegraphics[width=\linewidth]{http-benchmarks/latency.pdf}
 		%\vspace{-2mm}

### 2021-03-12 retro-concurrency.tex
updates

@@
 
-\newcommand{\farg}[2]{\langle #1 ~#2 \rangle_a}
-\newcommand{\ffun}[1]{\langle #1 \rangle_f}
-\newcommand{\faritha}[3]{\langle #1 ~#2 ~#3 \rangle_{b1}}
-\newcommand{\farithb}[2]{\langle #1 ~#2 \rangle_{b2}}
+\newcommand{\farg}[2]{\langle #1 ~#2 \rangle_{\text a}}
+\newcommand{\ffun}[1]{\langle #1 \rangle_{\text f}}
+\newcommand{\faritha}[3]{\langle #1 ~#2 ~#3 \rangle_{\text b1}}
+\newcommand{\farithb}[2]{\langle #1 ~#2 \rangle_{\text b2}}
 
@@
 \newcommand{\ostack}{\omega} % OCaml Stack
-\newcommand{\cstacka}[2]{\big \lceil #1, #2 \big \rceil_c} % C stack with arguments
-\newcommand{\ostacka}[2]{\big \lceil #1, #2 \big \rceil_o} % OCaml stack with arguments
+\newcommand{\cstacka}[2]{\big \lceil #1, #2 \big \rceil_{\text c}} % C stack with arguments
+\newcommand{\ostacka}[2]{\big \lceil #1, #2 \big \rceil_{\text o}} % OCaml stack with arguments
 \newcommand{\ostackemp}{\bullet}
@@
 
-\newcommand{\ostep}{\xrightarrow{o}}
-\newcommand{\cstep}{\xrightarrow{c}}
+\newcommand{\ostep}{\xrightarrow{\text o}}
+\newcommand{\cstep}{\xrightarrow{\text c}}
 \newcommand{\step}{\rightarrow}
@@
   \institution{IIT Madras}            %% \institution is required
+	\city{Chennai}
 	\country{India}
@@
   \institution{OCaml Labs}            %% \institution is required
+	\city{Cambridge}
 	\country{UK}
@@
   \institution{Jane Street}            %% \institution is required
+	\city{London}
 	\country{UK}
@@
   \institution{Opsian and OCaml Labs} %% \institution is required
+	\city{Cambridge}
 	\country{UK}
@@
   \institution{University of Cambridge Computer Laboratory and OCaml Labs}            %% \institution is required
+	\city{Cambridge}
 	\country{UK}
@@
 \subsection{Asynchronous I/O}
+\label{sec:aio}
 
@@
 such as resumable exceptions, lightweight threads, coroutines, generators and
-asynchronous I/O to be composably expressed. Let us illustrate the power of
-effect handlers in OCaml by implementing a simple lightweight thread library
-with support for forking new threads and yielding control to other threads. We
-will then extend this scheduler with support for synchronously reading from
-channels and subsequently make it asynchronous without changing the client code
-for asynchrony.
+asynchronous I/O to be composably expressed. One of the primary motivations to
+extend OCaml with effect handlers is to support concurrent programming. To
+illustrate this, let us implement a simple lightweight thread library with
+support for forking new threads and yielding control to other threads. We will
+then extend this scheduler with support for synchronously reading from channels
+and subsequently make it asynchronous without changing the client code for
+asynchrony.
 
@@
 primitive support for concurrency, users will have to use monadic I/O libraries
-such as Lwt~\cite{lwt} and Async~\cite{async}. Languages such as GHC Haskell
-and Go provide lightweight threads but bake their implementation into the
+such as Lwt~\cite{lwt} and Async~\cite{async}, where all asynchronous code must
+use monad syntax. Languages such as GHC Haskell and Go provide lightweight
+threads, which avoids the dichotomy between synchronous and asynchronous code.
+However, these languages bake-in the lightweight thread implementation into the
 runtime system. With effect handlers, asynchronous I/O can be implemented
-directly in OCaml.
+directly in OCaml as a library.
 
@@
 The interaction of non-local control flow and resources has been studied
-extensively. Scheme uses |dynamic-wind|~\cite{R5RS}, which is a generalisation of
-Common Lisp |unwind-protect|~\cite{Steele90}, which ensures de-allocation (and
-re-allocation) of resources every time the non-local control leaves and enters
-back into a context. |dynamic-wind| is not quite sufficient for resource
+extensively. Scheme uses |dynamic-wind|~\cite{R5RS}, which is a generalisation
+of Common Lisp |unwind-protect|~\cite{Steele90}, which ensures de-allocation
+(and re-allocation) of resources every time the non-local control leaves and
+enters back into a context. |dynamic-wind| is not quite sufficient for resource
 clean up~\cite{Kiselyov,Sitaram03} as returning exits must be distinguished
@@
 Multicore OCaml builds on the existing defensive coding practices against
-exceptions to clean up resources on non-returning exits. Our continuations are
-assumed to be resumed exactly once using |continue| or |discontinue|. As a
-result, when a computation performs an effect, we are guaranteed that the
-control will return. For the non-returning cases (value and exceptional
-return), the code already handles resource cleanup. OCaml does not have a
-\texttt{\footnotesize try/finally} construct commonly used for resource cleanup
-in many programming languages. Alternative standard libraries such as
+exceptions to clean up resources on non-returning exits. We assume that the
+continuations are resumed exactly once using |continue| or |discontinue|. Under
+this assumption, when a computation performs an effect, we expect the control
+to return. For the non-returning cases (value and exceptional return), the code
+already handles resource cleanup.
+
+OCaml does not have a \texttt{\footnotesize try/finally} construct commonly
+used for resource cleanup in many programming languages. The OCaml standard
+library~\cite{FunProtect} as well as alternative standard libraries such as
 Base~\cite{BaseProtect} and Core~\cite{CoreProtect} provide mechanisms
-analogous to |unwind-protect|, which are implemented using exceptions.
-Thus, the linear use of continuations enabled by the |discontinue| primitive
-ensures backwards compatibility of legacy OCaml systems code under non-local
-control flow introduced by effect handlers. Leijen~\cite{Leijen18} explicitly
-extends effect handlers with |initially| and |finally| clauses in Koka for
-resource safety. Dolan et al.~\cite{TFP17} describe the interaction
-of effect handlers and asynchronous exceptions. This is orthogonal to the
-contributions of this paper. Our focus is the compiler and runtime system
-support for implementing effect handlers.
+analogous to |unwind-protect|, which are in turn implemented using exception
+handlers. Thus, the linear use of continuations enabled by the |discontinue|
+primitive ensures backwards compatibility of legacy OCaml systems code under
+non-local control flow introduced by effect handlers.
+
+Leijen~\cite{Leijen18} explicitly extends effect handlers with |initially| and
+|finally| clauses in Koka for resource safety. Dolan et al.~\cite{TFP17}
+describe the interaction of effect handlers and asynchronous exceptions. This
+is orthogonal to the contributions of this paper. Our focus is the compiler and
+runtime system support for implementing effect handlers.
 

### 2021-03-12 retro-concurrency.tex
updates

@@
 \documentclass[sigplan,screen]{acmart}
-\setcopyright{none}
+\setcopyright{rightsretained}
 \acmPrice{}
@@
 \affiliation{
-  \institution{IIT Madras}            %% \institution is required
-	\city{Chennai}
-	\country{India}
+  \institution{\small IIT Madras}            %% \institution is required
+	\city{\small Chennai}
+	\country{\small India}
 }
@@
 \affiliation{
-  \institution{OCaml Labs}            %% \institution is required
-	\city{Cambridge}
-	\country{UK}
+  \institution{\small OCaml Labs}            %% \institution is required
+	\city{\small Cambridge}
+	\country{\small UK}
 }
@@
 \affiliation{
-  \institution{Jane Street}            %% \institution is required
-	\city{London}
-	\country{UK}
+  \institution{\small Jane Street}            %% \institution is required
+	\city{\small London}
+	\country{\small UK}
 }
@@
 \affiliation{
-  \institution{Opsian and OCaml Labs} %% \institution is required
-	\city{Cambridge}
-	\country{UK}
+  \institution{\small Opsian and OCaml Labs} %% \institution is required
+	\city{\small Cambridge}
+	\country{\small UK}
 }
@@
 \affiliation{
-  \institution{University of Cambridge Computer Laboratory and OCaml Labs}            %% \institution is required
-	\city{Cambridge}
-	\country{UK}
+  \institution{\small University of Cambridge Computer Laboratory and OCaml Labs}            %% \institution is required
+	\city{\small Cambridge}
+	\country{\small UK}
 }
@@
 \end{abstract}
+
+\begin{CCSXML}
+<ccs2012>
+   <concept>
+       <concept_id>10011007.10011006.10011041.10011048</concept_id>
+       <concept_desc>Software and its engineering~Runtime environments</concept_desc>
+       <concept_significance>500</concept_significance>
+       </concept>
+   <concept>
+       <concept_id>10011007.10011006.10011008.10011024.10011034</concept_id>
+       <concept_desc>Software and its engineering~Concurrent programming structures</concept_desc>
+       <concept_significance>500</concept_significance>
+       </concept>
+   <concept>
+       <concept_id>10011007.10011006.10011008.10011024.10011027</concept_id>
+       <concept_desc>Software and its engineering~Control structures</concept_desc>
+       <concept_significance>500</concept_significance>
+       </concept>
+   <concept>
+       <concept_id>10011007.10011006.10011008.10011009.10010175</concept_id>
+       <concept_desc>Software and its engineering~Parallel programming languages</concept_desc>
+       <concept_significance>300</concept_significance>
+       </concept>
+   <concept>
+       <concept_id>10011007.10011006.10011008.10011009.10011014</concept_id>
+       <concept_desc>Software and its engineering~Concurrent programming languages</concept_desc>
+       <concept_significance>500</concept_significance>
+       </concept>
+ </ccs2012>
+\end{CCSXML}
+
+\ccsdesc[500]{Software and its engineering~Runtime environments}
+\ccsdesc[500]{Software and its engineering~Concurrent programming structures}
+\ccsdesc[500]{Software and its engineering~Control structures}
+\ccsdesc[300]{Software and its engineering~Parallel programming languages}
+\ccsdesc[500]{Software and its engineering~Concurrent programming languages}
+
+\keywords{Effect handlers, Backwards compatibility, Fibers, Continuations,
+Backtraces}
 

### 2021-03-12 retro-concurrency.tex
Added file to track changes promised in response

@@
 Captured continuations are copies of the original fibers and resuming the
-continuation copies the fibers and leaves the continuation as it is. In our use
-cases, continuations will be resumed at-most once, and copying fibers is
-unnecessary and inefficient. Instead, Multicore OCaml optimises fibers for
-one-shot continuations. Fibers are allocated on the C heap using |malloc| and
-are |free|d when the handled computation returns with a value or an exception.
-Similar to Farvardin et al.~\cite{Farvardin20}, we use a \emph{stack cache} of
-recently freed stacks in order to speed up allocation.
+continuation copies the fibers and leaves the continuation as it is. Since our
+primary use case is concurrency, continuations will be resumed at-most once, and
+copying fibers is unnecessary and inefficient. Instead, Multicore OCaml
+optimises fibers for one-shot continuations. Fibers are allocated on the C heap
+using |malloc| and are |free|d when the handled computation returns with a value
+or an exception. Similar to Farvardin et al.~\cite{Farvardin20}, we use a
+\emph{stack cache} of recently freed stacks in order to speed up allocation.
 

### 2021-03-13 retro-concurrency.tex
very minor proof-reading edits to the introduction

@@
 Many programming languages, including OCaml, require non-local change to the
-code in order to support asynchronous I/O, often leading to a dichotomy
+code in order to support asynchronous I/O, often leading to a dichotomy between
 synchronous and asynchronous code~\cite{colour}. Since OCaml does not have
@@
 
-In order to support forking and yielding threads, we declare effects the
+In order to support forking and yielding threads, we declare the
 following effects:
@@
 
-The |Fork| effect takes a thunk which is spawned as a concurrent thread, and
-|Yield| yields control to another thread in the scheduler queue. We can define
+The |Fork| effect takes a thunk which is spawned as a concurrent thread, and the
+|Yield| effect yields control to another thread in the scheduler queue. We can define
 helper functions to perform these effects:

### 2021-03-14 retro-concurrency.tex
avoid repeating effects in the abstract

@@
 \begin{abstract}
-Effect handlers have been gathering momentum as a mechanism for modular
-	programming with user-defined effects. Effect handlers allow for non-local
+Effect handlers have been gathering momentum as a mechanism for non-local
 	control flow mechanisms such as generators, async/await, lightweight threads

### 2021-03-14 retro-concurrency.tex
done_ -> compl

@@
       | todo ->
-        let (done_,todo) = do_reads todo in
+        let compl,todo = do_reads todo in
         List.iter (fun (str,k) ->
-          Queue.push (fun () -> continue k str) runq) done_;
+          Queue.push (fun () -> continue k str) runq) compl;
         pending_reads := todo;
@@
     | effect (Read ic) k ->
-        pending_reads := (ic,k)::!pending_reads;
+        pending_reads := (ic,k) :: !pending_reads;
         run_next ()
@@
 function (line 13). |do_reads| takes a list of pending reads and blocks until
-at least one of the reads succeeds. It returns a pair of lists |done_| and
-|todo|. |done_| contains the result strings from successful reads and
+at least one of the reads succeeds. It returns a pair of lists |compl| and
+|todo|. |compl| contains the result strings from successful reads and
 corresponding continuations. These continuations are arranged to be resumed

### 2021-03-14 retro-concurrency.tex
intro

@@
 
-Many programming languages, including OCaml, require non-local change to the
+Many programming languages, including OCaml, require non-local changes to source
 code in order to support asynchronous I/O, often leading to a dichotomy between
-synchronous and asynchronous code~\cite{colour}. Since OCaml does not have
-primitive support for concurrency, users will have to use monadic I/O libraries
-such as Lwt~\cite{lwt} and Async~\cite{async}, where all asynchronous code must
-use monad syntax. Languages such as GHC Haskell and Go provide lightweight
+synchronous and asynchronous code~\cite{colour}.
+Since OCaml only provides preemptive system threads as support for concurrency,
+users usually use cooperative I/O libraries such as Lwt~\cite{lwt} and
+Async~\cite{async}, where all asynchronous code must
+use a monadic style. Languages such as GHC Haskell and Go provide lightweight
 threads, which avoids the dichotomy between synchronous and asynchronous code.
@@
 runtime system. With effect handlers, asynchronous I/O can be implemented
-directly in OCaml as a library.
+directly in OCaml as a library without imposing a monadic form on users of it.
 

### 2021-03-14 retro-concurrency.tex
show exceptions in action alongside effects

@@
   let rec run_next () =
-    match Queue.take_opt runq with
-    | None -> ()
-    | Some k -> continue k ()
+    match Queue.pop runq with
+    | k -> continue k ()
+    | exception Queue.Empty -> ()
   in

### 2021-03-14 retro-concurrency.tex
cite pottier 21 on separation logic for effect handlers
Update code to use intro exceptions

@@
   let rec run_next () =
-    match Queue.take_opt runq with
-    | None ->
+    match Queue.pop runq with
+    | f -> f ()
+    | exception Queue.empty ->
       begin match !pending_reads with
@@
       end
-    | Some f -> f ()
   in
@@
 Multi-shot continuations are difficult to
-reason about in the presence of linear resources such as file descriptors.
+reason about in the presence of linear resources such as file descriptors~\cite{de-vilhena-pottier-21}.
 Recall that OCaml programs are written with the expectation that function calls

### 2021-03-14 retro-concurrency.tex
Revert abstract

@@
 \begin{abstract}
-Effect handlers have been gathering momentum as a mechanism for non-local
+Effect handlers have been gathering momentum as a mechanism for modular
+	programming with user-defined effects. Effect handlers allow for non-local
 	control flow mechanisms such as generators, async/await, lightweight threads

### 2021-03-14 retro-concurrency.tex
make the example standalone (Queue.empty -> Queue.Empty as well)

@@
     | f -> f ()
-    | exception Queue.empty ->
-      begin match !pending_reads with
+    | exception Queue.Empty ->
+      match !pending_reads with
       | [] -> () (* no pending reads *)
@@
         run_next ()
-      end
   in
@@
     match f () with
-    ...
+    | () -> run_next () (* value case *)
+    | effect Yield k -> suspend k; run_next ()
+    | effect (Fork f') k -> suspend k; spawn f'
     | effect (Read ic) k ->
@@
   in
-  ...
+  spawn main
 \end{lstlisting}

### 2021-03-14 retro-concurrency.tex
space
edits

@@
 \affiliation{
-  \institution{\small University of Cambridge Computer Laboratory and OCaml Labs}            %% \institution is required
+  \institution{\small University of Cambridge and OCaml Labs}            %% \institution is required
 	\city{\small Cambridge}
@@
 
+\vspace{-1em}
 \subsection{Asynchronous I/O}
@@
 %
-which uses OCaml standard library's |input_line| function to read a line from
+This uses OCaml standard library's |input_line| function to read a line from
 the channel |ic| and resume the continuation |k| with the resultant string.
 However, performing reads synchronously blocks the entire scheduler, preventing
-other threads to run until the I/O is completed.
+other threads from running until the I/O is completed.
 

### 2021-03-14 retro-concurrency.tex
edits
Minor edit

@@
 reads. We abstract away the details of the event-based I/O using the |do_reads|
-function (line 13). |do_reads| takes a list of pending reads and blocks until
+function (line 14). |do_reads| takes a list of pending reads and blocks until
 at least one of the reads succeeds. It returns a pair of lists |compl| and
@@
 focussing on retrofitting effect handlers into a mainstream \emph{systems}
-language (\S\ref{sec:refine}), followed by the compilation scheme and
-operational semantics for Multicore OCaml effect handlers
-(\S\ref{sec:semantics}).  We then discuss the compiler and the runtime system
-support for implementing effect handlers (\S\ref{sec:impl}), and present an
-extensive performance evaluation of effect handlers (\S\ref{sec:eval}) against
-our design goals (\S\ref{sec:req}). Finally, we discuss the related work
-(\S\ref{sec:related}) and conclude (\S\ref{sec:conc}).
+language (\S\ref{sec:refine}), followed by the static and dynamic semantics for
+Multicore OCaml effect handlers (\S\ref{sec:semantics}).  We then discuss the
+compiler and the runtime system support for implementing effect handlers
+(\S\ref{sec:impl}), and present an extensive performance evaluation of effect
+handlers (\S\ref{sec:eval}) against our design goals (\S\ref{sec:req}).
+Finally, we discuss the related work (\S\ref{sec:related}) and conclude
+(\S\ref{sec:conc}).
 

### 2021-03-14 retro-concurrency.tex
minor intro code edit, make contributions explicit

@@
 
-\vspace{-1em}
+%\vspace{-1em}
 \subsection{Asynchronous I/O}
@@
   let runq = Queue.create () in
-  let suspend k =
-    Queue.push (fun () -> continue k ()) runq
-  in
+  let suspend k = Queue.push (continue k) runq in
   let pending_reads = ref [] in
@@
 |pending_reads| maintains a list of pending reads and the associated
-continuations (line 6). We handle the |Read| effect by pushing the pair of
-input channel |ic| and continuation |k| to |pending_reads| (line 25), allowing
+continuations (line 4). We handle the |Read| effect by pushing the pair of
+input channel |ic| and continuation |k| to |pending_reads| (line 24), allowing
 other threads in the scheduler to run.
@@
 reads. We abstract away the details of the event-based I/O using the |do_reads|
-function (line 14). |do_reads| takes a list of pending reads and blocks until
+function (line 12). |do_reads| takes a list of pending reads and blocks until
 at least one of the reads succeeds. It returns a pair of lists |compl| and
@@
 	\item a formal operational semantics for the effect handler implementation in
-    OCaml.
+		OCaml. Our formalism explicitly models the subtle interactions with the C
+		stack, which is generally overlooked by other formal models, but which the
+		implementations must handle.
 	\item extensive evaluation which shows that our implementation has negligible
@@
 		whose running time in seconds in given in parenthesis.}
+	%\vspace{-2mm}
 	\label{res:macro_time}
@@
 	Baseline is Stock OCaml, whose size in kilobytes given in parenthesis.}
+	%\vspace{-2mm}
 	\label{res:macro_size}
@@
 	difference for Multicore OCaml over stock OCaml.}
-%\vspace{-7mm}
+\vspace{-5mm}
 {
@@
 \end{tabular}
-%\vspace{-5mm}
+\vspace{-5mm}
 }
@@
 	slowdown factor ($\times$ times) over its idiomatic implementation in stock OCaml.}
-%\vspace{-3mm}
+\vspace{-3mm}
 {
@@
 \label{tab:micro_noperform}
-	%\vspace{-5mm}
+\vspace{-3mm}
 \end{table}
@@
 	\end{minipage}
-	%\vspace{-3mm}
+	\vspace{-3mm}
 	\caption{Web server performance.}
-	%\vspace{-5mm}
+	\vspace{-5mm}
 \end{figure}

### 2021-03-15 retro-concurrency.tex
Updates

@@
 return exactly once (\S\ref{sec:refine}). Consider the scenario when a
-continuation is neither continued nor discontinued. Since fibers are
-|malloc|ed, and are |free|d when the computation being handled returns a value
-or raises an exception, not resuming continuations leaks memory. In addition,
-unresumed continuations may also leak other resource such as open file
-descriptors.
-
-We consider well-behaved Multicore OCaml programs to resume their continuations
-\emph{exactly once}. This can be enforced by installing a finaliser on every
-captured continuation that calls |discontinue| on the continuation during
-finalisation and ignores the result:
+continuation is never resumed. Since fibers allocate memory for the stack using
+|malloc|, which are |free|d when the computation returns, not resuming
+continuations leaks memory. In addition, unresumed continuations may also leak
+other system resources such as sockets and open file descriptors.
+
+We make a pragmatic trade-off and expect the user code to resume captured
+continuations \emph{exactly once}. One can use the GC support to free up
+resources by installing a finaliser that discontinues the continuation and
+ignores the result:
+
 \begin{lstlisting}
@@
 \end{lstlisting}
-However, Multicore OCaml does not enforce this behaviour in order to avoid the
-costs associated with installing a finaliser every time a continuation is
-captured.
-
-For stack walking during GC, the only change necessary was to make the
-runtime aware of fiber layout and how to follow the parent pointers. The
-stack map generation and scanning each frame for roots remains the same. The
-challenges and solutions for integrating fibers with the concurrent
-mark-and-sweep GC of Multicore OCaml have been discussed
+
+This frees up both the memory allocated for the fiber stack as well as other
+system resources, assuming that user code does not handle |Unwind| exception
+and fails to re-raise it. But installing a finaliser on every captured
+continuation by default introduces significant overhead to continuation
+capture. Hence, we choose not to implement such a finaliser by default. It is
+also useful to note that even if the memory for the fiber stack is managed by
+the GC rather than being allocated in the C heap as we do now, we would still
+need a finalisation mechanism to unwind the stack and release other system
+resources that may be held by the continuation.
+
+For stack walking during GC in order to find the roots, the only change
+necessary was to make the runtime aware of the fiber layout and how to follow
+the parent pointers. The stack map generation and scanning each frame for roots
+remains the same. The challenges and the solutions for integrating fibers with
+the concurrent mark-and-sweep GC of Multicore OCaml have been discussed
 previously~\cite{Sivaramakrishnan20}.
@@
 an extension of the 4.10.0 OCaml compiler with support for shared memory
-parallelism and effect handlers. These results were obtained on a 2-socket
-Intel\textregistered Xeon\textregistered Gold 5120
-x86-64~\cite{IntelXeonGold5120Spec} server running Ubuntu 18.04 with 64GB of main
-memory.
+parallelism and effect handlers. Since our objective is to evaluate the impact
+of effect handlers, none of our benchmarks utilise parallelism. These results
+were obtained on a 2-socket Intel\textregistered Xeon\textregistered Gold 5120
+x86-64~\cite{IntelXeonGold5120Spec} server running Ubuntu 18.04 with 64GB of
+main memory.
 
@@
 Next we look at benchmarks that utilize non-local control flow using effect
-handlers. First, we quantify the overhead of individual operations in effect
+handlers. First, we quantify the cost of individual operations in effect
 handling. Consider the following annotated code:
@@
 |perf| support for cycle-accurate tracing on modern Intel processors. We
-executed 10 iterations of the code, with 3 warm-up runs. We observed that the
-sequences \texttt{a-b}, \texttt{b-c}, \texttt{c-d} and \texttt{d-e} took
-\textbf{23 ns, 5 ns, 11 ns and 7 ns}, respectively. For reference, the idle
-memory load latency for the local NUMA node is 93.2 ns as measured using the
-Intel MLC tool~\cite{mlc}.
+executed 10 iterations of the code, with 3 warm-up runs. For calibration, the
+idle memory load latency for the local NUMA node is 93.2 ns as measured using
+the Intel MLC tool~\cite{mlc}. We observed that the sequences \texttt{a-b},
+\texttt{b-c}, \texttt{c-d} and \texttt{d-e} took \textbf{23 ns, 5 ns, 11 ns and
+7 ns}, respectively. The time in the sequence \texttt{a-b} is dominated by the
+memory allocation. Thus, the individual operations in effect handling are fast.
 
@@
 iterator (|val iter: 'a t -> ('a -> unit) -> unit|), we can derive its
-generator function (|val next : unit -> 'a option|). The code is available in
-the supplementary material. We evaluate the performance of traversing a
-complete binary tree of depth 25 using this generator. This involves $2^{26}$
-stack switches in total. For comparison, we implemented a hand-written,
-selective CPSed~\cite{Nielson01}, defunctionalised~\cite{Danvy01} version
-(|cps|) and a concurrency monad (|monad|) version of the generator for the
-tree. Both |cps| and |monad| versions are specialised to the binary tree with
-the usual caveats of not using the stack for function calls. We observed that
-the |cps| version was the fastest, thanks to specialisation and hand
-optimisation. |MC| version was only 2.76$\times$ slower than |cps| while being
-a generic solution, and the |monad| version was 8.69$\times$ slower than |cps|.
+generator function (|val next : unit -> 'a option|)\footnote{https://gist.github.com/kayceesrk/eb0ab496c22861f21b1d9484772e982d}.
+We evaluate the performance of traversing a complete binary tree of depth 25
+using this generator. This involves $2^{26}$ stack switches in total. For
+comparison, we implemented a hand-written, selective CPSed~\cite{Nielson01},
+defunctionalised~\cite{Danvy01} version (|cps|) and a concurrency monad
+(|monad|) version of the generator for the tree. Both |cps| and |monad|
+versions are specialised to the binary tree with the usual caveats of not using
+the stack for function calls. We observed that the |cps| version was the
+fastest, thanks to specialisation and hand optimisation. |MC| version was only
+2.76$\times$ slower than |cps| while being a generic solution, and the |monad|
+version was 8.69$\times$ slower than |cps|.
 

### 2021-03-23 retro-concurrency.tex
Address more comments
Add points about callee-saved registers

@@
 
-The lack of callee-saved registers also makes exception handling fast as there
-are no callee-saved registers to be restored when the exception handler runs.
-Installing an exception handler simply pushes the program counter (|pc|) of the
-handler and the current exception pointer (|exn_ptr| -- a field in
-|Caml_state|) to the stack. After this, the current exception pointer is
+The lack of callee-saved registers also makes exception handling fast. In the
+absence of callee-saved registers, no registers need to be saved when entering
+a |try| block. Similarly, no registers need to be restored when handling an
+exception. Installing an exception handler simply pushes the program counter
+(|pc|) of the handler and the current exception pointer (|exn_ptr| -- a field
+in |Caml_state|) to the stack. After this, the current exception pointer is
 updated to be the current stack pointer (|rsp|). This creates a linked-list of
@@
 and jumps to the |pc| of the handler.
-
-% Note |caml_call_ocaml| is |caml_start_program| in ocaml-multicore source
 
@@
 
-In order to illustrate the implementation of handling effects, consider the
+Similar to exception handlers, the lack of callee-saved registers in OCaml
+benefits effect handlers. There is no register state to save when entering an
+effect handler or performing an effect. Similarly, there is no register state
+to restore when handling an effect or resuming a continuation. This fortuitous
+design choice in stock OCaml has a significant impact towards fast switching
+between fibers in Multicore OCaml.
+
+In order to illustrate the runtime support for handling effects, consider the
 example presented in Figure~\ref{code:effimpl}. The layout of the program state
@@
 callbacks to identity functions. The other micro benchmarks are highly
-recursive programs and were taken from ~\cite{Farvardin20}. For micro
-benchmarks, we observed that padding tight loops with a few |nop| instructions,
-which changes the loop alignment, makes the code up to 15\% faster. Hence, the
-difference in running times under 15\% may not be statistically significant.
+recursive programs and were taken from Farvardin et al.~\cite{Farvardin20}. For
+micro benchmarks, we observed that padding tight loops with a few |nop|
+instructions, which changes the loop alignment, makes the code up to 15\%
+faster. Hence, the difference in running times under 15\% may not be
+statistically significant.
 
@@
 
+\if{0}
 \clearpage
@@
 
-\if{0}
 \subsection*{Dynamic Semantics}
@@
 \end{proof}
-\fi
 
@@
 continuations to future work.
+\fi
 

### 2021-03-23 retro-concurrency.tex
updates

@@
 
-%\vspace{-1em}
-\subsection{Asynchronous I/O}
-\label{sec:aio}
-
 Effect handlers provide a structured form of delimited
@@
 asynchronous I/O to be composably expressed. One of the primary motivations to
-extend OCaml with effect handlers is to support concurrent programming. To
-illustrate this, let us implement a simple lightweight thread library with
-support for forking new threads and yielding control to other threads. We will
-then extend this scheduler with support for synchronously reading from channels
-and subsequently make it asynchronous without changing the client code for
-asynchrony.
-
-Many programming languages, including OCaml, require non-local changes to source
-code in order to support asynchronous I/O, often leading to a dichotomy between
-synchronous and asynchronous code~\cite{colour}.
-Since OCaml only provides preemptive system threads as support for concurrency,
-users usually use cooperative I/O libraries such as Lwt~\cite{lwt} and
-Async~\cite{async}, where all asynchronous code must
-use a monadic style. Languages such as GHC Haskell and Go provide lightweight
-threads, which avoids the dichotomy between synchronous and asynchronous code.
-However, these languages bake-in the lightweight thread implementation into the
-runtime system. With effect handlers, asynchronous I/O can be implemented
-directly in OCaml as a library without imposing a monadic form on users of it.
+extend OCaml with effect handlers is to natively support composable
+asynchronous I/O. Many programming languages, including OCaml, require
+non-local changes to source code in order to support asynchronous I/O, often
+leading to a dichotomy between synchronous and asynchronous code~\cite{colour}.
+In the absence of native support for concurrent programming, OCaml developers
+use monadic I/O libraries such as Lwt~\cite{lwt} and Async~\cite{async}, where
+asynchronous functions are represented as monadic computations. In these
+libraries, while asynchronous functions can call synchronous functions
+directly, the converse is not true. In particular, any function that calls an
+asynchronous function will also have to be marked as asynchronous. Languages
+such as GHC Haskell and Go provide lightweight threads, which avoids the
+dichotomy between synchronous and asynchronous code. However, these languages
+bake-in the lightweight thread implementation into the runtime system. With
+effect handlers, asynchronous I/O can be implemented directly in OCaml as a
+library without imposing a monadic form on the users.
+
+There are many research languages and libraries built around effect
+handlers~\cite{Leijen14,Hillerstrom20,Pyro,Frank,Eff}. Unlike these efforts,
+our goal is to retrofit effect handlers onto the OCaml programming language,
+which has been in continuous use for the past 25 years in large codebases
+including verification tools~\cite{everest,Coq}, mission critical software
+systems~\cite{astree} and latency sensitive networked
+applications~\cite{Madhavapeddy13}. OCaml is particularly favoured for its
+competitive yet predictable performance, with a fast foreign-function interface
+(FFI). It has excellent compatibility with program analysis tools such as
+debuggers and profilers that utilise DWARF stack unwind tables~\cite{DWARF} to
+obtain a backtrace.
+
+OCaml currently does not support any non-local control flow mechanisms other
+than exceptions. This makes it particularly challenging to implement the
+delimited continuations necessary for effect handlers without sacrificing the
+desirable properties of OCaml. One standard way of implementing continuations
+is to use continuation-passing style (CPS) in the compiler's intermediate
+representation (IR)~\cite{Leijen14}. OCaml does not use a CPS IR, and changing
+the compiler to utilise a CPS IR would be an enormous undertaking that would
+affect the performance profile of existing OCaml applications due to the
+increased memory allocations as the continuation closures get allocated on the
+heap~\cite{Farvardin20}. Moreover, with CPS, stack unwinding is meaningless,
+and hence, we would lose compatibility with tools that inspect the program
+stack. Hence, we choose not to use CPS translation and represent the
+continuations as call stacks.
+
+The search for an expressive effect system that guarantees that all the effects
+performed in the program are handled (\emph{effect safety}) in the presence of
+advanced features such as polymorphism, modularity and generativity is an
+active area of research~\cite{Leijen14, Biernacki19, Biernacki20,
+Hillerstrom20}. We do not focus on this question in this paper, and our
+implementation of effect handlers in OCaml does not guarantee effect safety. We
+leave the question of effect safety for future work.
+
+\subsection{Requirements}
+\label{sec:req}
+
+We motivate our effect handler design based on the following ideal
+requirements:
+
+\begin{enumerate}[label=R\arabic*]
+	\item \textbf{Backwards compatibility.} A well-behaved OCaml program does
+		not break under OCaml extended with effect handlers. OCaml code that does
+		not use effect handlers will pay negligible performance and memory cost.
+
+	\item \textbf{Tool compatibility.} OCaml programs with effect handlers
+		produce well-formed backtraces and remain compatible with program analysis
+		tools such as debuggers and profilers that inspect the stack using DWARF
+		unwind tables.
+
+	\item \textbf{Effect handler efficiency.} The program must accommodate
+		millions of continuations at the same time to support highly-concurrent
+		applications. Installing effect handlers, capturing and resuming
+		continuations must be fast.
+
+	\item \textbf{Forwards compatibility.} As a cornerstone of modularity, we
+		also want blocking I/O code to \emph{transparently} be made asynchronous
+		with the help of effect handlers.
+\end{enumerate}
+
+The need to host millions of continuations at the same time rules out the use
+of a large contiguous stack space as in C for continuations. Instead, we resort
+to using small initial stacks and growing the stacks on demand. As a result,
+OCaml functions, irrespective of whether they use effect handlers, need to
+perform stack overflow checks, and external C functions (which do not have stack
+overflow checks) must be performed on a separate system stack. Additionally, we
+must generate DWARF stack unwind tables for stacks that may be non-contiguous.
+In this work, we develop the compiler and runtime support for implementing
+efficient effect handlers for OCaml that satisfy these requirements.
+
+Our work is also timely. The WebAssembly~\cite{Hass17} community group is
+considering effect handlers as one of the mechanisms for supporting
+concurrency, asynchronous I/O and generators~\cite{WasmProposal}. Project
+Loom~\cite{loom} is an OpenJDK project that adds virtual threads and delimited
+continuations to Java. The Swift roadmap~\cite{swift} includes direct-style
+asynchronous programming and structured concurrency as milestones. We believe
+that our design choices will inform similar choices to be made in other
+industrial-strength languages.
+
+\subsection{Contributions}
+
+Our contributions are to present:
+
+\begin{itemize}
+  \item the design and implementation of one-shot effect handlers for OCaml.
+    Our design retains OCaml's compatibility with program analysis tools that
+    inspect the stack using DWARF unwind tables. We have validated our DWARF
+    unwind tables with the assistance of an automated validator
+    tool~\cite{Bastian19}.
+	\item a formal operational semantics for the effect handler implementation in
+		OCaml. Our formalism explicitly models the subtle interactions with the C
+		stack, which is generally overlooked by other formal models, but which the
+		implementations must handle.
+	\item extensive evaluation which shows that our implementation has negligible
+		impact on code that does not use effect handlers, and serves as an
+		efficient foundation for scalable concurrent programming.
+\end{itemize}
+
+We have implemented effect handlers in a multicore extension of the OCaml
+programming language which we call \textbf{Multicore OCaml} to distinguish it
+from stock OCaml. Multicore OCaml clearly delineates concurrency (overlapped
+execution of tasks) from parallelism (simultaneous execution of tasks) with
+separate mechanisms for expressing them. Sivaramakrishnan et
+al.~\cite{Sivaramakrishnan20} describe the parallelism support in Multicore
+OCaml enabled by \emph{domains}. The focus of this paper is the concurrency
+support enabled by effect handlers.
+
+The remainder of the paper continues with a description of the stock OCaml
+program stack (\S\ref{sec:stack}). We then refine the design further by
+focussing on retrofitting effect handlers into a mainstream \emph{systems}
+language (\S\ref{sec:refine}), followed by the static and dynamic semantics for
+Multicore OCaml effect handlers (\S\ref{sec:semantics}).  We then discuss the
+compiler and the runtime system support for implementing effect handlers
+(\S\ref{sec:impl}), and present an extensive performance evaluation of effect
+handlers (\S\ref{sec:eval}) against our design goals (\S\ref{sec:req}).
+Finally, we discuss the related work (\S\ref{sec:related}) and conclude
+(\S\ref{sec:conc}).
+
+\section{Background: OCaml Stacks}
+\label{sec:stack}
+
+\begin{figure*}
+\begin{minipage}{0.30\linewidth}
+  \begin{minipage}{\linewidth}
+    \begin{lstlisting}[numbers=left]
+external ocaml_to_c
+  : unit -> int = "ocaml_to_c"
+exception E1
+exception E2
+let c_to_ocaml () = raise E1
+let _ = Callback.register
+    "c_to_ocaml" c_to_ocaml
+let omain () =
+  try (* h1 *)
+    try (* h2 *) ocaml_to_c ()
+    | with E2 -> 0
+  | with E1 -> 42;;
+let _ = assert (omain () = 42)
+    \end{lstlisting}
+    %\vspace{-2mm}
+    \subcaption{\texttt{meander.ml}}
+    \label{code:meander_ml}
+  \end{minipage}
+  \begin{minipage}{\linewidth}
+    \begin{lstlisting}[language=c,numbers=left]
+#include <caml/mlvalues.h>
+#include <caml/callback.h>
+
+value ocaml_to_c (value unit) {
+  caml_callback(*caml_named_value
+    ("c_to_ocaml"), Val_unit);
+  return Val_int(0);
+}
+    \end{lstlisting}
+    %\vspace{-4mm}
+    \subcaption{\texttt{meander.c}}
+    \label{code:meander_c}
+  \end{minipage}
+\end{minipage}
+%
+\begin{minipage}{0.30\linewidth}
+  \centering
+  \includegraphics[scale=0.44]{figures/stock_stack}
+  \subcaption{Stack layout before raise \texttt{E1}.}
+  \label{fig:stock_stack}
+\end{minipage}
+%
+\begin{minipage}{0.39\linewidth}
+\begin{lstlisting}[language=c,basicstyle=\ttfamily\footnotesize]
+#0  0x925dc in caml_raise_exn ()
+#1  0x6fd3e in camlMeander__c_to_ocaml_83 () at meander.ml:5
+#2  0x925a4 in caml_call_ocaml ()
+#3  0x8a84a in caml_callback_exn (...) at callback.c:145
+#4  caml_callback (...) at callback.c:199
+#5  0x76e0a in ocaml_to_c (unit=1) at meander.c:5
+#6  0x6fd77 in camlMeander__omain_88 () at meander.ml:10
+#7  0x6fe92 in camlMeander__entry () at meander.ml:13
+#8  0x6f719 in caml_program ()
+#9  0x925a4 in caml_call_ocaml ()
+#10 0x92e4c in caml_startup_common (...) at startup_nat.c:162
+#11 0x92eab in caml_startup_exn (...) at startup_nat.c:167
+#12 caml_startup (...) at startup_nat.c:172
+#13 0x6f55c in main (...) at main.c:44
+\end{lstlisting}
+%\vspace{-2mm}
+\subcaption{\texttt{gdb} backtrace before raise \texttt{E1}.}
+\label{code:gdb_backtrace}
+\end{minipage}
+%\vspace{-2mm}
+\caption{Program stack on stock OCaml.}
+%\vspace{-2mm}
+\end{figure*}
+
+The main challenge in implementing effect handlers in Multicore OCaml is
+managing the program stack and preserving its desirable properties. In this
+section, we provide an overview of the program stack and related mechanisms in
+stock OCaml.
+
+Consider the layout of the stock OCaml stack for the program shown in
+Figures~\ref{code:meander_ml} and~\ref{code:meander_c}. The OCaml main function
+|omain| installs two exception handlers |h1| and |h2| to handle the exceptions
+|E1| and |E2|. |omain| calls the external C function |ocaml_to_c|, which in
+turn calls back into the OCaml function |c_to_ocaml|, which raises the
+exception |E1|. OCaml supports raising exceptions in C as well as throwing
+exceptions across external calls. Hence, the exception |E1| gets caught in the
+handler |h1|, and |omain| returns |42|. The layout of the stack in the native
+code backend just before raising the exception in |c_to_ocaml| is illustrated
+in Figure~\ref{fig:stock_stack}. Note that the stack grows downwards.
+
+OCaml uses the same program stack as C, and hence the stack has
+alternating sequences of C and OCaml frames. However, unlike C, OCaml does not
+create pointers into OCaml frames. OCaml uses the hardware support for |call|
+and |return| instructions for function calls and returns. OCaml does not
+perform explicit stack overflow checks in code, and, just like C, relies on the
+guard page at the end of the stack region to detect stack overflow. Stack
+overflow is detected by a memory fault and a |Stack_overflow| exception is
+raised to unwind the stack.
+
+\subsection{External calls and callbacks}
+\label{sec:external}
+
+OCaml does not use the C calling convention. In particular, there are no
+callee-saved registers in OCaml. In the |x86-64| backend, the OCaml runtime
+makes use of two C callee-saved registers for supporting OCaml execution. The
+register |r15| holds the \emph{allocation pointer} into the minor heap used for
+bump pointer allocation, and |r14| holds a reference to the |Caml_state|, a
+table of global variables used by the runtime. This makes external calls
+extremely fast in OCaml. If the external function does not allocate in the
+OCaml heap, then it can be called directly and no bookkeeping is necessary. For
+external functions which allocate in the OCaml heap, the cached allocation
+pointer is saved to |Caml_state| before the external call and it is restored on
+return. Similarly, callbacks into OCaml are also cheap: these involve loading
+the arguments in the right registers and calling the OCaml function. OCaml
+callbacks are relatively common as the garbage collector (GC) finalisation
+functions are executed as OCaml callbacks.
+
+\subsection{Exception handlers}
+\label{sec:exn_handlers}
+
+The lack of callee-saved registers also makes exception handling fast. In the
+absence of callee-saved registers, no registers need to be saved when entering
+a |try| block. Similarly, no registers need to be restored when handling an
+exception. Installing an exception handler simply pushes the program counter
+(|pc|) of the handler and the current exception pointer (|exn_ptr| -- a field
+in |Caml_state|) to the stack. After this, the current exception pointer is
+updated to be the current stack pointer (|rsp|). This creates a linked-list of
+exception handler frames on the stack as shown in Figure~\ref{fig:stock_stack}.
+Raising an exception simply sets |rsp| to |exn_ptr|, loads the saved |exn_ptr|,
+and jumps to the |pc| of the handler.
+
+In order to forward exceptions across C frames, the C stub function
+|caml_call_ocaml|, pushes an exception handler frame that either forwards the
+exception to the innermost OCaml exception handler (|raise_exn_c| in
+Figure~\ref{fig:stock_stack}) or prints a fatal error (|fatal_uncaught|) if
+there are no enclosing handlers. Exceptions are so cheap in OCaml that it is
+common to use them for \emph{local} control flow.
+
+\subsection{Stack unwinding}
+\label{sec:unwind}
+
+OCaml generates \emph{stack maps} in order to accurately identify roots on the
+stack for assisting the GC. For every call point in the program, the OCaml
+compiler emits the size of the frame and the set of all live registers in the
+frame containing pointers into the heap. During a GC, the OCaml stack is walked
+and the roots are marked, skipping over the C frames.
+
+OCaml also generates precise DWARF unwind information for OCaml, thanks to
+which debuggers such as |gdb| and |lldb|, and profilers such as |perf| work
+out-of-the-box. For example, for the program in Figures~\ref{code:meander_ml}
+and~\ref{code:meander_c}, one could set a break point in |gdb| at
+|caml_raise_exn| to get the backtrace in Figure~\ref{code:gdb_backtrace} which
+corresponds to the stack in Figure~\ref{fig:stock_stack}.
+
+The same backtrace can also be obtained by using \emph{frame pointers} instead
+of DWARF unwind tables. OCaml allows compiling code with frame pointers, but
+they are not enabled by default. The OCaml stack tends to be deep with small frames
+due to the pervasive use of recursive functions, not all of which are
+tail-recursive. Hence, the addition of frame pointers can significantly
+increase the size of the
+stack\footnote{https://github.com/ocaml/ocaml/issues/5721\#issuecomment-472965549}.
+Moreover, not using frame pointers saves two instructions in the function
+prologue and epilogue, and makes an extra register (|rbp| on x86\_64) available.
+Note that the DWARF unwind information is complementary to the information used
+by OCaml to walk the stack for GC.
+
+\section{Refining the design}
+\label{sec:refine}
 
@@
 Lwt and Async).
-
-%\vspace{-1.5mm}
-\subsection{Retrofitting effect handlers}
-
-There are many research languages and libraries built around effect
-handlers~\cite{Leijen14,Hillerstrom20,Pyro,Frank,Eff}. Unlike these efforts,
-our goal is to retrofit effect handlers onto the OCaml programming language,
-which has been in continuous use for the past 25 years in large codebases
-including verification tools~\cite{everest,Coq}, mission critical software
-systems~\cite{astree} and latency sensitive networked
-applications~\cite{Madhavapeddy13}. OCaml is particularly favoured for its
-competitive yet predictable performance, with a fast foreign-function interface
-(FFI). It has excellent compatibility with program analysis tools such as
-debuggers and profilers that utilise DWARF stack unwind tables~\cite{DWARF} to
-obtain a backtrace. Our primary motivation for adding effect handlers to OCaml
-is to support: {\em (i)} highly concurrent applications such as web servers
-using a natural \emph{direct-style} as opposed to using callbacks; and {\em
-(ii)} efficient nested parallel programs in the vein of Cilk~\cite{Cilk},
-OpenMP~\cite{OpenMP}, X10~\cite{Charles05}, Intel TBB~\cite{IntelTBB}, and
-variants of Parallel ML~\cite{MaPLe, Fluet10, Sivaramakrishnan14}.
-
-OCaml currently does not support any non-local control flow mechanisms other
-than exceptions. This makes it particularly challenging to implement the
-delimited continuations necessary for effect handlers without sacrificing the
-desirable properties of OCaml. One standard way of implementing continuations
-is to use continuation-passing style (CPS) in the compiler's intermediate
-representation (IR)~\cite{Leijen14}. OCaml does not use a CPS IR, and changing
-the compiler to utilise a CPS IR would be an enormous undertaking that would
-affect the performance profile of existing OCaml applications due to the
-increased memory allocations as the continuation closures get allocated on the
-heap~\cite{Farvardin20}. Moreover, with CPS, stack unwinding is meaningless and
-hence, we would lose compatibility with tools that inspect the program stack.
-Hence, we choose not to use CPS translation and represent the continuations as
-call stacks.
-
-The search for an expressive effect system that guarantees that all the effects
-performed in the program are handled (\emph{effect safety}) in the presence of
-advanced features such as polymorphism, modularity and generativity is an
-active area of research~\cite{Leijen14, Biernacki19, Biernacki20,
-Hillerstrom20}. We do not focus on this question in this paper, and our
-implementation of effect handlers in OCaml does not guarantee effect safety.
-For example, running |perform (Read stdin)| without an enclosing handler raises
-|Unhandled| exception at the point of |perform|. We leave the question of
-effect safety for future work.
-
-\subsection{Requirements}
-\label{sec:req}
-
-We motivate our effect handler design based on the following ideal
-requirements:
-
-\begin{enumerate}[label=R\arabic*]
-	\item \textbf{Backwards compatibility.} A well-behaved OCaml program does
-		not break under OCaml extended with effect handlers. OCaml code that does
-		not use effect handlers will pay negligible performance and memory cost.
-
-	\item \textbf{Tool compatibility.} OCaml programs with effect handlers
-		produce well-formed backtraces and remain compatible with program analysis
-		tools such as debuggers and profilers that inspect the stack using DWARF
-		unwind tables.
-
-	\item \textbf{Effect handler efficiency.} The program must accommodate
-		millions of continuations at the same time to support highly-concurrent
-		applications. Installing effect handlers, capturing and resuming
-		continuations must be fast.
-
-	\item \textbf{Forwards compatibility.} As a cornerstone of modularity, we
-		also want blocking I/O code to \emph{transparently} be made asynchronous
-		with the help of effect handlers.
-\end{enumerate}
-
-The need to host millions of continuations at the same time rules out the use
-of a large contiguous stack space as in C for continuations. Instead, we resort
-to using small initial stacks and growing the stacks on demand. As a result,
-OCaml functions, irrespective of whether they use effect handlers, need to
-perform stack overflow checks, and external C functions (which do not have stack
-overflow checks) must be performed on a separate system stack. Additionally, we
-must generate DWARF stack unwind tables for stacks that may be non-contiguous.
-In this work, we develop the compiler and runtime support for implementing
-efficient effect handlers for OCaml that satisfy these requirements.
-
-Our work is also timely. The WebAssembly~\cite{Hass17} community group is
-considering effect handlers as one of the mechanisms for supporting
-concurrency, asynchronous I/O and generators~\cite{WasmProposal}. Project
-Loom~\cite{loom} is an OpenJDK project that adds virtual threads and delimited
-continuations to Java. The Swift roadmap~\cite{swift} includes direct-style
-asynchronous programming and structured concurrency as milestones. We believe
-that our design choices will inform similar choices to be made in other
-industrial-strength languages.
-
-\subsection{Contributions}
-
-Our contributions are to present:
-
-\begin{itemize}
-  \item the design and implementation of one-shot effect handlers for OCaml.
-    Our design retains OCaml's compatibility with program analysis tools that
-    inspect the stack using DWARF unwind tables. We have validated our DWARF
-    unwind tables with the assistance of an automated validator
-    tool~\cite{Bastian19}.
-	\item a formal operational semantics for the effect handler implementation in
-		OCaml. Our formalism explicitly models the subtle interactions with the C
-		stack, which is generally overlooked by other formal models, but which the
-		implementations must handle.
-	\item extensive evaluation which shows that our implementation has negligible
-		impact on code that does not use effect handlers, and serves as an
-		efficient foundation for scalable concurrent programming.
-\end{itemize}
-
-We have implemented effect handlers in a multicore extension of the OCaml
-programming language which we call \textbf{Multicore OCaml} to distinguish it
-from stock OCaml. Multicore OCaml clearly delineates concurrency (overlapped
-execution of tasks) from parallelism (simultaneous execution of tasks) with
-separate mechanisms for expressing them. Sivaramakrishnan et
-al.~\cite{Sivaramakrishnan20} describe the parallelism support in Multicore
-OCaml enabled by \emph{domains}. The focus of this paper is the concurrency
-support enabled by effect handlers.
-
-The remainder of the paper continues with a description of the stock OCaml
-program stack (\S\ref{sec:stack}). We then refine the design further by
-focussing on retrofitting effect handlers into a mainstream \emph{systems}
-language (\S\ref{sec:refine}), followed by the static and dynamic semantics for
-Multicore OCaml effect handlers (\S\ref{sec:semantics}).  We then discuss the
-compiler and the runtime system support for implementing effect handlers
-(\S\ref{sec:impl}), and present an extensive performance evaluation of effect
-handlers (\S\ref{sec:eval}) against our design goals (\S\ref{sec:req}).
-Finally, we discuss the related work (\S\ref{sec:related}) and conclude
-(\S\ref{sec:conc}).
-
-\section{Background: OCaml Stacks}
-\label{sec:stack}
-
-\begin{figure*}
-\begin{minipage}{0.30\linewidth}
-  \begin{minipage}{\linewidth}
-    \begin{lstlisting}[numbers=left]
-external ocaml_to_c
-  : unit -> int = "ocaml_to_c"
-exception E1
-exception E2
-let c_to_ocaml () = raise E1
-let _ = Callback.register
-    "c_to_ocaml" c_to_ocaml
-let omain () =
-  try (* h1 *)
-    try (* h2 *) ocaml_to_c ()
-    | with E2 -> 0
-  | with E1 -> 42;;
-let _ = assert (omain () = 42)
-    \end{lstlisting}
-    %\vspace{-2mm}
-    \subcaption{\texttt{meander.ml}}
-    \label{code:meander_ml}
-  \end{minipage}
-  \begin{minipage}{\linewidth}
-    \begin{lstlisting}[language=c,numbers=left]
-#include <caml/mlvalues.h>
-#include <caml/callback.h>
-
-value ocaml_to_c (value unit) {
-  caml_callback(*caml_named_value
-    ("c_to_ocaml"), Val_unit);
-  return Val_int(0);
-}
-    \end{lstlisting}
-    %\vspace{-4mm}
-    \subcaption{\texttt{meander.c}}
-    \label{code:meander_c}
-  \end{minipage}
-\end{minipage}
-%
-\begin{minipage}{0.30\linewidth}
-  \centering
-  \includegraphics[scale=0.44]{figures/stock_stack}
-  \subcaption{Stack layout before raise \texttt{E1}.}
-  \label{fig:stock_stack}
-\end{minipage}
-%
-\begin{minipage}{0.39\linewidth}
-\begin{lstlisting}[language=c,basicstyle=\ttfamily\footnotesize]
-#0  0x925dc in caml_raise_exn ()
-#1  0x6fd3e in camlMeander__c_to_ocaml_83 () at meander.ml:5
-#2  0x925a4 in caml_call_ocaml ()
-#3  0x8a84a in caml_callback_exn (...) at callback.c:145
-#4  caml_callback (...) at callback.c:199
-#5  0x76e0a in ocaml_to_c (unit=1) at meander.c:5
-#6  0x6fd77 in camlMeander__omain_88 () at meander.ml:10
-#7  0x6fe92 in camlMeander__entry () at meander.ml:13
-#8  0x6f719 in caml_program ()
-#9  0x925a4 in caml_call_ocaml ()
-#10 0x92e4c in caml_startup_common (...) at startup_nat.c:162
-#11 0x92eab in caml_startup_exn (...) at startup_nat.c:167
-#12 caml_startup (...) at startup_nat.c:172
-#13 0x6f55c in main (...) at main.c:44
-\end{lstlisting}
-%\vspace{-2mm}
-\subcaption{\texttt{gdb} backtrace before raise \texttt{E1}.}
-\label{code:gdb_backtrace}
-\end{minipage}
-%\vspace{-2mm}
-\caption{Program stack on stock OCaml.}
-%\vspace{-2mm}
-\end{figure*}
-
-The main challenge in implementing effect handlers in Multicore OCaml is
-managing the program stack and preserving its desirable properties. In this
-section, we provide an overview of the program stack and related mechanisms in
-stock OCaml.
-
-Consider the layout of the stock OCaml stack for the program shown in
-Figures~\ref{code:meander_ml} and~\ref{code:meander_c}. The OCaml main function
-|omain| installs two exception handlers |h1| and |h2| to handle the exceptions
-|E1| and |E2|. |omain| calls the external C function |ocaml_to_c|, which in
-turn calls back into the OCaml function |c_to_ocaml|, which raises the
-exception |E1|. OCaml supports raising exceptions in C as well as throwing
-exceptions across external calls. Hence, the exception |E1| gets caught in the
-handler |h1|, and |omain| returns |42|. The layout of the stack in the native
-code backend just before raising the exception in |c_to_ocaml| is illustrated
-in Figure~\ref{fig:stock_stack}. Note that the stack grows downwards.
-
-OCaml uses the same program stack as C, and hence the stack has
-alternating sequences of C and OCaml frames. However, unlike C, OCaml does not
-create pointers into OCaml frames. OCaml uses the hardware support for |call|
-and |return| instructions for function calls and returns. OCaml does not
-perform explicit stack overflow checks in code, and, just like C, relies on the
-guard page at the end of the stack region to detect stack overflow. Stack
-overflow is detected by a memory fault and a |Stack_overflow| exception is
-raised to unwind the stack.
-
-\subsection{External calls and callbacks}
-\label{sec:external}
-
-OCaml does not use the C calling convention. In particular, there are no
-callee-saved registers in OCaml. In the |x86-64| backend, the OCaml runtime
-makes use of two C callee-saved registers for supporting OCaml execution. The
-register |r15| holds the \emph{allocation pointer} into the minor heap used for
-bump pointer allocation, and |r14| holds a reference to the |Caml_state|, a
-table of global variables used by the runtime. This makes external calls
-extremely fast in OCaml. If the external function does not allocate in the
-OCaml heap, then it can be called directly and no bookkeeping is necessary. For
-external functions which allocate in the OCaml heap, the cached allocation
-pointer is saved to |Caml_state| before the external call and it is restored on
-return. Similarly, callbacks into OCaml are also cheap: these involve loading
-the arguments in the right registers and calling the OCaml function. OCaml
-callbacks are relatively common as the garbage collector (GC) finalisation
-functions are executed as OCaml callbacks.
-
-\subsection{Exception handlers}
-\label{sec:exn_handlers}
-
-The lack of callee-saved registers also makes exception handling fast. In the
-absence of callee-saved registers, no registers need to be saved when entering
-a |try| block. Similarly, no registers need to be restored when handling an
-exception. Installing an exception handler simply pushes the program counter
-(|pc|) of the handler and the current exception pointer (|exn_ptr| -- a field
-in |Caml_state|) to the stack. After this, the current exception pointer is
-updated to be the current stack pointer (|rsp|). This creates a linked-list of
-exception handler frames on the stack as shown in Figure~\ref{fig:stock_stack}.
-Raising an exception simply sets |rsp| to |exn_ptr|, loads the saved |exn_ptr|,
-and jumps to the |pc| of the handler.
-
-In order to forward exceptions across C frames, the C stub function
-|caml_call_ocaml|, pushes an exception handler frame that either forwards the
-exception to the innermost OCaml exception handler (|raise_exn_c| in
-Figure~\ref{fig:stock_stack}) or prints a fatal error (|fatal_uncaught|) if
-there are no enclosing handlers. Exceptions are so cheap in OCaml that it is
-common to use them for \emph{local} control flow.
-
-\subsection{Stack unwinding}
-\label{sec:unwind}
-
-OCaml generates \emph{stack maps} in order to accurately identify roots on the
-stack for assisting the GC. For every call point in the program, the OCaml
-compiler emits the size of the frame and the set of all live registers in the
-frame containing pointers into the heap. During a GC, the OCaml stack is walked
-and the roots are marked, skipping over the C frames.
-
-OCaml also generates precise DWARF unwind information for OCaml, thanks to
-which debuggers such as |gdb| and |lldb|, and profilers such as |perf| work
-out-of-the-box. For example, for the program in Figures~\ref{code:meander_ml}
-and~\ref{code:meander_c}, one could set a break point in |gdb| at
-|caml_raise_exn| to get the backtrace in Figure~\ref{code:gdb_backtrace} which
-corresponds to the stack in Figure~\ref{fig:stock_stack}.
-
-The same backtrace can also be obtained by using \emph{frame pointers} instead
-of DWARF unwind tables. OCaml allows compiling code with frame pointers, but
-they are not enabled by default. The OCaml stack tends to be deep with small frames
-due to the pervasive use of recursive functions, not all of which are
-tail-recursive. Hence, the addition of frame pointers can significantly
-increase the size of the
-stack\footnote{https://github.com/ocaml/ocaml/issues/5721\#issuecomment-472965549}.
-Moreover, not using frame pointers saves two instructions in the function
-prologue and epilogue, and makes an extra register (|rbp| on x86\_64) available.
-Note that the DWARF unwind information is complementary to the information used
-by OCaml to walk the stack for GC.
-
-\section{Refining the design}
-\label{sec:refine}
 

### 2021-03-24 retro-concurrency.tex
Move example to section 3

@@
 \begin{lstlisting}
-effect Read : in_channel -> string
+effect In_line : in_channel -> string
 \end{lstlisting}
 %
-declares an \emph{effect} |Read|, which is parameterised with an input channel
-of type |in_channel|, which when \emph{performed} returns a |string| value. A
-computation can perform the |Read| effect without knowing how the |Read| effect
-is implemented. This computation may be enclosed by different handlers that
-handle |Read| differently. For example, |Read| may be implemented by performing
-a blocking read on the input channel or performing the read asynchronously by
-offloading it to an event loop such as |libuv|. Thanks to the separation of
-effectful operations from their interpretations, effect handlers enable new
-approaches to modular programming.
-
-Effect handlers provide a structured form of delimited
+declares an \emph{effect} |In_line|, which is parameterised with an input
+channel of type |in_channel|, which when \emph{performed} returns a |string|
+value. A computation can perform the |In_line| effect without knowing how the
+|In_line| effect is implemented. This computation may be enclosed by different
+handlers that handle |In_line| differently. For example, |In_line| may be
+implemented by performing a blocking read on the input channel or performing
+the read asynchronously by offloading it to an event loop such as |libuv|.
+Thanks to the separation of effectful operations from their interpretations,
+effect handlers enable new approaches to modular programming.
+
+Effect handlers expose a structured form of delimited
 continuations~\cite{Danvy90} that enables non-local control-flow mechanisms
@@
 extend OCaml with effect handlers is to natively support composable
-asynchronous I/O. Many programming languages, including OCaml, require
-non-local changes to source code in order to support asynchronous I/O, often
-leading to a dichotomy between synchronous and asynchronous code~\cite{colour}.
-In the absence of native support for concurrent programming, OCaml developers
-use monadic I/O libraries such as Lwt~\cite{lwt} and Async~\cite{async}, where
+asynchronous I/O to enable highly scalable concurrent applications such as web
+servers. Many programming languages, including OCaml, require non-local changes
+to source code in order to support asynchronous I/O, often leading to a
+dichotomy between synchronous and asynchronous code~\cite{colour}. In the
+absence of native support for concurrent programming, OCaml developers use
+monadic I/O libraries such as Lwt~\cite{lwt} and Async~\cite{async}, where
 asynchronous functions are represented as monadic computations. In these
@@
 directly, the converse is not true. In particular, any function that calls an
-asynchronous function will also have to be marked as asynchronous. Languages
-such as GHC Haskell and Go provide lightweight threads, which avoids the
-dichotomy between synchronous and asynchronous code. However, these languages
-bake-in the lightweight thread implementation into the runtime system. With
-effect handlers, asynchronous I/O can be implemented directly in OCaml as a
-library without imposing a monadic form on the users.
+asynchronous function will also have to be marked as asynchronous. As a result,
+large parts of the applications using these libraries tend to be in monadic
+form. Languages such as GHC Haskell and Go provide lightweight threads, which
+avoids the dichotomy between synchronous and asynchronous code. However, these
+languages bake-in the lightweight thread implementation into the runtime
+system. With effect handlers, asynchronous I/O can be implemented directly in
+OCaml as a library without imposing a monadic form on the users.
 
@@
 delimited continuations necessary for effect handlers without sacrificing the
-desirable properties of OCaml. One standard way of implementing continuations
-is to use continuation-passing style (CPS) in the compiler's intermediate
+desirable properties of OCaml. A standard way of implementing continuations is
+to use continuation-passing style (CPS) in the compiler's intermediate
 representation (IR)~\cite{Leijen14}. OCaml does not use a CPS IR, and changing
@@
 increased memory allocations as the continuation closures get allocated on the
-heap~\cite{Farvardin20}. Moreover, with CPS, stack unwinding is meaningless,
+heap~\cite{Farvardin20}. Moreover, with CPS, stack unwinding is not meaningful,
 and hence, we would lose compatibility with tools that inspect the program
@@
 programming language which we call \textbf{Multicore OCaml} to distinguish it
-from stock OCaml. Multicore OCaml clearly delineates concurrency (overlapped
+from \emph{stock OCaml}. Multicore OCaml delineates concurrency (overlapped
 execution of tasks) from parallelism (simultaneous execution of tasks) with
-separate mechanisms for expressing them. Sivaramakrishnan et
+distinct mechanisms for expressing them. Sivaramakrishnan et
 al.~\cite{Sivaramakrishnan20} describe the parallelism support in Multicore
@@
 The remainder of the paper continues with a description of the stock OCaml
-program stack (\S\ref{sec:stack}). We then refine the design further by
-focussing on retrofitting effect handlers into a mainstream \emph{systems}
-language (\S\ref{sec:refine}), followed by the static and dynamic semantics for
-Multicore OCaml effect handlers (\S\ref{sec:semantics}).  We then discuss the
-compiler and the runtime system support for implementing effect handlers
-(\S\ref{sec:impl}), and present an extensive performance evaluation of effect
-handlers (\S\ref{sec:eval}) against our design goals (\S\ref{sec:req}).
-Finally, we discuss the related work (\S\ref{sec:related}) and conclude
-(\S\ref{sec:conc}).
+program stack (\S\ref{sec:stack}). We describe effect handlers in Multicore
+OCaml focussing on the challenges in retrofitting them into a mainstream
+\emph{systems} language (\S\ref{sec:refine}), followed by the static and
+dynamic semantics for Multicore OCaml effect handlers (\S\ref{sec:semantics}).
+We then discuss the compiler and the runtime system support for implementing
+effect handlers (\S\ref{sec:impl}), and present an extensive performance
+evaluation of effect handlers (\S\ref{sec:eval}) against our design goals
+(\S\ref{sec:req}). Finally, we discuss the related work (\S\ref{sec:related})
+and conclude (\S\ref{sec:conc}).
 
@@
 of DWARF unwind tables. OCaml allows compiling code with frame pointers, but
-they are not enabled by default. The OCaml stack tends to be deep with small frames
-due to the pervasive use of recursive functions, not all of which are
+they are not enabled by default. The OCaml stack tends to be deep with small
+frames due to the pervasive use of recursive functions, not all of which are
 tail-recursive. Hence, the addition of frame pointers can significantly
@@
 Moreover, not using frame pointers saves two instructions in the function
-prologue and epilogue, and makes an extra register (|rbp| on x86\_64) available.
-Note that the DWARF unwind information is complementary to the information used
-by OCaml to walk the stack for GC.
-
-\section{Refining the design}
+prologue and epilogue, and makes an extra register (|rbp| on x86\_64)
+available. Note that the DWARF unwind information is complementary to the
+information used by OCaml to walk the stack for GC.
+
+\section{Effect Handlers}
 \label{sec:refine}
 
-In order to support forking and yielding threads, we declare the
-following effects:
+In this section, we describe the effect handlers in Multicore OCaml, and refine
+the design to retrofit them onto OCaml.
+
+\subsection{Asynchronous I/O}
+
+Since our primary motivation is to enable composable asynchronous I/O, let us
+implement a simple lightweight thread library with support for forking new
+threads and yielding control to other threads. We will then extend this library
+with support for synchronously reading from channels and subsequently make it
+asynchronous \textit{without changing the client code for asynchrony}. In order
+to support forking and yielding threads, we declare the following effects:
 
@@
 
-The |Fork| effect takes a thunk which is spawned as a concurrent thread, and the
-|Yield| effect yields control to another thread in the scheduler queue. We can define
-helper functions to perform these effects:
+The |Fork| effect takes a thunk which is spawned as a concurrent thread, and
+the |Yield| effect yields control to another thread in the scheduler queue. We
+can define helper functions to perform these effects:
 
@@
 |Fork f'| effect, we suspend the current continuation and recursively call
-|spawn| on |f'| in order to run |f'| concurrently.
-
-Effect handlers provide modularity. Observe that we can change the scheduling
-algorithm from FIFO to LIFO by changing the scheduler queue to a stack. We can
-implement support for synchronous read from channels by adding the following
-case to the effect handler in |spawn|:
+|spawn| on |f'| in order to run |f'| concurrently. Observe that we can change
+the scheduling algorithm from FIFO to LIFO by changing the scheduler queue to a
+stack.
+
+We can implement support for synchronous read from channels by adding the
+following case to the effect handler in |spawn|:
 
@@
 	...
-  | effect (Read ic) k -> continue k (input_line ic)
+  | effect (In_line ic) k -> continue k (input_line ic)
 \end{lstlisting}
 %
-This uses OCaml standard library's |input_line| function to read a line from
-the channel |ic| and resume the continuation |k| with the resultant string.
-However, performing reads synchronously blocks the entire scheduler, preventing
-other threads from running until the I/O is completed.
+This uses OCaml standard library's |input_line| function to read a line
+synchronously from the channel |ic| and resume the continuation |k| with the
+resultant string. However, performing reads synchronously blocks the entire
+scheduler, preventing other threads from running until the I/O is completed.
 
@@
     | effect (Fork f') k -> suspend k; spawn f'
-    | effect (Read ic) k ->
+    | effect (In_line ic) k ->
         pending_reads := (ic,k) :: !pending_reads;
@@
 |pending_reads| maintains a list of pending reads and the associated
-continuations (line 4). We handle the |Read| effect by pushing the pair of
+continuations (line 4). We handle the |In_line| effect by pushing the pair of
 input channel |ic| and continuation |k| to |pending_reads| (line 24), allowing
@@
 
-In this section, we refine the effect handler design for the OCaml use case. The
-primary motivation for adding effect handlers to OCaml is to support a non-local
-control flow mechanism to implement asynchronous I/O in direct style, generators
-and enable task parallelism. These use cases do not resume the continuations
-more than once. Hence, our continuations are one-shot, and resuming the
-continuation more than once raises an |Invalid_argument| exception. It is
-well-known that one-shot continuations can be implemented
-efficiently~\cite{Bruggeman96}.
+Observe that this example does not resume a continuation more than once. This
+also holds true for other use cases such as generators and coroutines. Hence,
+our continuations are one-shot, and resuming the continuation more than once
+raises an |Invalid_argument| exception. It is well-known that one-shot
+continuations can be implemented efficiently~\cite{Bruggeman96}.
 
@@
 setting.
+
+\subsection{Resource cleanup}
 
@@
 
-We can then use a top-level I/O handler, similar to the introductory example
-(\S\ref{sec:aio}), to discharge the I/O operations asynchronously and resume
-with the result. This handles value return cases, but what about the
-exceptional cases |End_of_file| and |Sys_error|? To this end, we introduce a
-|discontinue| primitive to resume a continuation by raising an exception. In
-this example, on reaching the end of file, we would discontinue the captured
-continuation of the |input_line| function with |discontinue k End_of_file|,
-which raises the exception at |input_line|, and the open channels will be
-closed.
-
-OCaml programs that use resources such as channels are usually written defensively
-with the assumption that calling a function will return \emph{exactly once},
-either normally or exceptionally. Since effect handlers in Multicore OCaml do
-not ensure that all the effects are handled, if the function performs an effect
-with no matching handler, then the function \emph{will not return at all}. To
-remedy this, when such an effect bubbles up to the top-level, we discontinue the
-continuation with an |Unhandled| exception so that the exception handlers may
-run and clean up the resources.
+We can then use the |run| function that we defined earlier, to discharge the
+I/O operations asynchronously and resume with the result. While this handles
+value return cases, what about the exceptional cases |End_of_file| and
+|Sys_error|? To this end, we introduce a |discontinue| primitive to resume a
+continuation by raising an exception. In this example, on reaching the end of
+file, we would discontinue the captured continuation of the |input_line|
+function with |discontinue k End_of_file|, which raises the exception at
+|input_line| call site, and the open channels will be closed.
+
+OCaml programs that use resources such as channels are usually written
+defensively with the assumption that calling a function will return
+\emph{exactly once}, either normally or exceptionally. Since effect handlers in
+Multicore OCaml do not ensure that all the effects are handled, if the function
+performs an effect with no matching handler, then the function \emph{will not
+return at all}. To remedy this, when such an effect bubbles up to the
+top-level, we discontinue the continuation with an |Unhandled| exception so
+that the exception handlers may run and clean up the resources.
 

### 2021-03-24 retro-concurrency.tex
updates

@@
 implemented by performing a blocking read on the input channel or performing
-the read asynchronously by offloading it to an event loop such as |libuv|.
-Thanks to the separation of effectful operations from their interpretations,
-effect handlers enable new approaches to modular programming.
-
-Effect handlers expose a structured form of delimited
-continuations~\cite{Danvy90} that enables non-local control-flow mechanisms
-such as resumable exceptions, lightweight threads, coroutines, generators and
-asynchronous I/O to be composably expressed. One of the primary motivations to
-extend OCaml with effect handlers is to natively support composable
-asynchronous I/O to enable highly scalable concurrent applications such as web
-servers. Many programming languages, including OCaml, require non-local changes
-to source code in order to support asynchronous I/O, often leading to a
-dichotomy between synchronous and asynchronous code~\cite{colour}. In the
-absence of native support for concurrent programming, OCaml developers use
-monadic I/O libraries such as Lwt~\cite{lwt} and Async~\cite{async}, where
-asynchronous functions are represented as monadic computations. In these
-libraries, while asynchronous functions can call synchronous functions
-directly, the converse is not true. In particular, any function that calls an
-asynchronous function will also have to be marked as asynchronous. As a result,
-large parts of the applications using these libraries tend to be in monadic
-form. Languages such as GHC Haskell and Go provide lightweight threads, which
-avoids the dichotomy between synchronous and asynchronous code. However, these
-languages bake-in the lightweight thread implementation into the runtime
-system. With effect handlers, asynchronous I/O can be implemented directly in
-OCaml as a library without imposing a monadic form on the users.
+the read asynchronously by offloading it to an event loop such as |libuv|,
+without changing the computation. Thanks to the separation of effectful
+operations from their implementation, effect handlers enable new approaches to
+modular programming. Effect handlers are a generalization of exception
+handlers, where, in addition to the effect being handled, the handler is
+provided with the delimited continuation~\cite{Danvy90} of the perform site,
+which may be resumed later. This enables non-local control-flow mechanisms such
+as resumable exceptions, lightweight threads, coroutines, generators and
+asynchronous I/O to be composably expressed.
+
+One of the primary motivations to extend OCaml with effect handlers is to
+natively support asynchronous I/O in order to express highly scalable
+concurrent applications such as web servers in \emph{direct style} (as opposed
+to using \emph{callbacks}). Many programming languages, including OCaml,
+require non-local changes to source code in order to support asynchronous I/O,
+often leading to a dichotomy between synchronous and asynchronous
+code~\cite{colour}. For asynchronous I/O, OCaml developers typically use
+libraries such as Lwt~\cite{lwt} and Async~\cite{async}, where asynchronous
+functions are represented as monadic computations. In these libraries, while
+asynchronous functions can call synchronous functions directly, the converse is
+not true. In particular, any function that calls an asynchronous function will
+also have to be marked as asynchronous. As a result, large parts of the
+applications using these libraries end up being in monadic form. Languages such
+as GHC Haskell and Go provide lightweight threads, which avoids the dichotomy
+between synchronous and asynchronous code. However, these languages bake-in the
+lightweight thread implementation into the runtime system. With effect
+handlers, asynchronous I/O can be implemented directly in OCaml as a library
+without imposing a monadic form on the users.
 
@@
 increased memory allocations as the continuation closures get allocated on the
-heap~\cite{Farvardin20}. Moreover, with CPS, stack unwinding is not meaningful,
-and hence, we would lose compatibility with tools that inspect the program
-stack. Hence, we choose not to use CPS translation and represent the
-continuations as call stacks.
+heap~\cite{Farvardin20}. Moreover, with CPS, an explicit stack is absent, and
+hence, we would lose compatibility with tools that inspect the program stack.
+Hence, we choose not to use CPS translation and represent the continuations as
+call stacks.
 
@@
 \begin{enumerate}[label=R\arabic*]
-	\item \textbf{Backwards compatibility.} A well-behaved OCaml program does
-		not break under OCaml extended with effect handlers. OCaml code that does
-		not use effect handlers will pay negligible performance and memory cost.
+	\item \textbf{Backwards compatibility.} Extant OCaml programs \kc{Extant
+		better than legacy?} do not break under OCaml extended with effect
+		handlers. OCaml code that does not use effect handlers will pay negligible
+		performance and memory cost.
 
@@
 Loom~\cite{loom} is an OpenJDK project that adds virtual threads and delimited
-continuations to Java. The Swift roadmap~\cite{swift} includes direct-style
+continuations to Java. The Swift roadmap~\cite{swift} includes direct style
 asynchronous programming and structured concurrency as milestones. We believe
@@
 \begin{itemize}
-  \item the design and implementation of one-shot effect handlers for OCaml.
-    Our design retains OCaml's compatibility with program analysis tools that
-    inspect the stack using DWARF unwind tables. We have validated our DWARF
-    unwind tables with the assistance of an automated validator
-    tool~\cite{Bastian19}.
+	\item the design and implementation of effect handlers for OCaml. Our design
+		retains OCaml's compatibility with program analysis tools that inspect the
+		stack using DWARF unwind tables. We have validated our DWARF unwind tables
+		with the assistance of an automated validator tool~\cite{Bastian19}.
 	\item a formal operational semantics for the effect handler implementation in
@@
 pointer is saved to |Caml_state| before the external call and it is restored on
-return. Similarly, callbacks into OCaml are also cheap: these involve loading
-the arguments in the right registers and calling the OCaml function. OCaml
-callbacks are relatively common as the garbage collector (GC) finalisation
-functions are executed as OCaml callbacks.
+return. Similarly, callbacks into OCaml from C are also cheap: these involve
+loading the arguments in the right registers and calling the OCaml function.
+OCaml callbacks are relatively common as the garbage collector (GC), which is
+implemented in C, executes OCaml finalisation functions as callbacks.
 
@@
 (|pc|) of the handler and the current exception pointer (|exn_ptr| -- a field
-in |Caml_state|) to the stack. After this, the current exception pointer is
+in |Caml_state|) onto the stack. After this, the current exception pointer is
 updated to be the current stack pointer (|rsp|). This creates a linked-list of
@@
 compiler emits the size of the frame and the set of all live registers in the
-frame containing pointers into the heap. During a GC, the OCaml stack is walked
-and the roots are marked, skipping over the C frames.
+frame that point to the heap. During a GC, the OCaml stack is walked and the
+roots are marked, skipping over the C frames.
 
@@
 localised to the |run| function, and the computation that performs these
-effects can remain in \emph{direct-style} (as opposed to the monadic-style in
+effects can remain in direct style (as opposed to the monadic-style in
 Lwt and Async).
@@
 handles a |'a comp| that returns a |'a| value, and itself returns a |'b| value.
-The handler has a return case |retc| of type |'a -> 'b|. The effect case |effc|
-handles effects of type |'c eff| with |('c,'b) continuation| and returns a
-value of type |'b|. The last case in |effc| \emph{reperforms} any unmatched
+The handler has a return field |retc| of type |'a -> 'b|. The effect field
+|effc| handles effects of type |'c eff| with |('c,'b) continuation| and returns
+a value of type |'b|. The last case in |effc| \emph{reperforms} any unmatched
 effect to the outer handler and returns the value and exceptions back to the
@@
 \subsubsection{Syntax} Our expressions (Figure~\ref{sem:syntax}) consist of
-integer constants ($n$), variables ($x$), abstraction ($\lam{x}{x}$),
+integer constants ($n$), variables ($x$), abstraction ($\lam{x}{e}$),
 application ($e~e$), arithmetic expressions ($e \odot e$) where $\odot$ ranges
@@
 handler with an identity return case alone. We shall return to \textsc{Resume1}
-and \textsc{Resume2} later.
+and \textsc{Resume2} in the next subsection.
 
@@
 continuation copies the fibers and leaves the continuation as it is. Since our
-primary use case is concurrency, continuations will be resumed at-most once, and
+primary use case is concurrency, continuations will be resumed at most once, and
 copying fibers is unnecessary and inefficient. Instead, Multicore OCaml
@@
 |rsp| becomes less than the \emph{stack threshold} (maintained in the
-|Caml_state| table), the stack is said to have overflown. On stack overflow, we
-copy the whole fiber to a new area with double the size. In Multicore OCaml, we
-introduce stack overflow checks into the function prologue of OCaml functions.
-These stack overflows are rare and so the overflow checks will be correctly predicted
-by the CPU branch predictor.
+|Caml_state| table), the stack is said to have overflowed. On stack overflow,
+we copy the whole fiber to a new area with double the size. In Multicore OCaml,
+we introduce stack overflow checks into the function prologue of OCaml
+functions. These stack overflows are rare and so the overflow checks will be
+correctly predicted by the CPU branch predictor.
 
@@
 As one of our primary goals of adding effect handlers to OCaml is to support
-direct-style concurrency, we implement a full-fledged HTTP/1.1 web server using
+direct style concurrency, we implement a full-fledged HTTP/1.1 web server using
 effect handlers (|MC|). We use effect handlers to implement a user-level
@@
 selective CPS translation. These language are equipped with an effect system,
-which allows compiling pure code in direct-style and effectful code in CPS.
+which allows compiling pure code in direct style and effectful code in CPS.
 Leijen~\cite{Leijen14} implements effect handlers as a library in C using stack
@@
 implemented using lightweight fibers, along with a parallel
-runtime~\cite{Sivaramakrishnan20}, will open OCaml to highly scalable
-concurrent and task-parallel applications with minimal hit to sequential
-performance.
+runtime~\cite{Sivaramakrishnan20}, as has been demonstrated in our work, will
+open OCaml to highly scalable concurrent and task-parallel applications with
+minimal hit to sequential performance.
 

### 2021-03-24 retro-concurrency.tex
minor edit
updates

@@
 Although Multicore OCaml currently only supports the x86\_64 backend on Linux,
-our design does not preclude other CPUs and OSs.
+our design does not preclude other architectures and OSs.
 
@@
 registers are stored. To avoid reifying this large table, DWARF directives
-represent the table using a compact bytecode representation. This representation
-is interpreted on demand for building the stack unwind table. Within each
+represent the table using a compact bytecode representation. Each block of such
+instructions usually describe the call-frame information for a single function,
+where each instruction describes the table as a delta of the state at the
+previous instruction. While this leads to a compact representation, computing
+the return address at a given instruction requires interpreting the bytecode
+from the start of the function, which may be potentially very slow. Within each
 function, DWARF maintains a \emph{canonical frame address} (CFA) and is
-traditionally the stack pointer before entering this function. Hence, on x86-64,
-where the return address is pushed on the stack on call, the return address is
-at |CFA - 8|.
-
-The goal is to encode the CFA using DWARF directives when stacks are switched.
+traditionally the stack pointer before entering this function. Hence, on
+x86-64, where the return address is pushed on the stack on call, the return
+address is at |CFA - 8|.
+
+Our goal is to encode the CFA using DWARF directives when stacks are switched.
 Recall that stack switching occurs in effect handlers, external calls and
@@
 sampled~\cite{Bastian19}. This only includes the frames from the current fiber.
-This is a limitation of |perf| and not of our stack layout.
+This is a limitation of |perf| and not of our stack layout. Bastian et
+al.~\cite{Bastian19} report on a technique to pre-compile the unwind table to
+assembly which speeds up DWARF-based unwinding by upto 25$\times$. With this
+fast unwinding, |perf| would no longer needs to dump the call stack, and hence,
+we can get the backtrace of the entire call stack and not just the current
+fiber.
 
@@
 one of these implementation strategies -- they are best described as using the
-resize strategy from Farvardin et al. for each of the fibers, which are
-linked to represent the current stack and the captured continuations.
-
-Kawahara et al.~\cite{Kawahara20} implement one-shot effect handlers using
-coroutines as a macro-expressible translation, and present an embedding in Lua
-and Ruby. Lua provides asymmetric coroutines~\cite{Lua} where each coroutine
-uses its own stack similar to how each handled computation runs in its own
-fiber in Multicore OCaml.
-%Lua does not allow coroutines to include C frames, as in
-%Multicore OCaml.  - avsm/kc not quite, see https://coco.luajit.org/index.html
-%It is useful to note that Lua is an interpreted
-%language, unlike OCaml, and DWARF unwind tables are not a concern in Lua.
+resize strategy from Farvardin et al. for each of the fibers, which are linked
+to represent the current stack and the captured continuations. Kawahara et
+al.~\cite{Kawahara20} implement one-shot effect handlers using coroutines as a
+macro-expressible translation, and present an embedding in Lua and Ruby. Lua
+provides asymmetric coroutines~\cite{Lua} where each coroutine uses its own
+stack similar to how each handled computation runs in its own fiber in
+Multicore OCaml.
 

### 2021-03-24 retro-concurrency.tex
Minor proofread edits

@@
 	\item \textbf{Backwards compatibility.} Extant OCaml programs \kc{Extant
-		better than legacy?} do not break under OCaml extended with effect
-		handlers. OCaml code that does not use effect handlers will pay negligible
-		performance and memory cost.
+		better than legacy?} \tk{Existing might be better?} do not break under
+    OCaml extended with effect handlers. OCaml code that does not use effect
+    handlers will pay negligible performance and memory cost.
 
@@
 pops continuations from the queue and resumes them with |()| value using the
-|continue| primitive. In the case of |Yield| effect, we suspend the current
-continuation |k| and resume the next available continuation. In the case of
+|continue| primitive. In the case of the |Yield| effect, we suspend the current
+continuation |k| and resume the next available continuation. In the case of the
 |Fork f'| effect, we suspend the current continuation and recursively call
@@
 channels being closed externally. Both |input_line| and |output_string| raise
-|Sys_error| exception if the channel is closed. In this case, the catch-all
+a |Sys_error| exception if the channel is closed. In this case, the catch-all
 exception handler closes the channels and reraises the exception to communicate

### 2021-03-24 retro-concurrency.tex
updates

@@
 \subsection{Garbage collection}
+\label{sec:gc}
 
@@
 and fails to re-raise it. But installing a finaliser on every captured
-continuation by default introduces significant overhead to continuation
-capture. Hence, we choose not to implement such a finaliser by default. It is
-also useful to note that even if the memory for the fiber stack is managed by
-the GC rather than being allocated in the C heap as we do now, we would still
-need a finalisation mechanism to unwind the stack and release other system
-resources that may be held by the continuation.
+continuation by default introduces significant overhead to continuation capture
+(\S\ref{sec:res_final_k}). Hence, we choose not to implement such a finaliser by
+default. It is also useful to note that even if the memory for the fiber stack
+is managed by the GC rather than being allocated in the C heap as we do now, we
+would still need a finalisation mechanism to unwind the stack and release other
+system resources that may be held by the continuation.
 
@@
 
+\subsubsection{Finalised continuations}
+\label{sec:res_final_k}
+
+In \S\ref{sec:gc}, we described how continuation resources can be cleaned up by
+attaching a finaliser. Attaching this finaliser to every captured continuation
+slows down generator (chameneos) benchmark by 4.1$\times$ (2.1$\times$)
+compared to not attaching a finaliser. Hence, Multicore OCaml does not attach
+such finaliser to every continuation by default.
+
 \subsubsection{Webserver}

### 2021-03-24 retro-concurrency.tex
updates
More updates

@@
 	control flow mechanisms such as generators, async/await, lightweight threads
-	and coroutines to be composably expressed.  We present a design and evaluate a
-	full-fledged efficient implementation of effect handlers for OCaml, an
+	and coroutines to be composably expressed.  We present a design and evaluate
+	a full-fledged efficient implementation of effect handlers for OCaml, an
 	industrial-strength multi-paradigm programming language. Our implementation
@@
 	other than exceptions. Our implementation of effect handlers for OCaml: {\em
-	(i)} imposes negligible overhead on code that does not use effect handlers;
-	{\em (ii)} remains compatible with program analysis tools that inspect the
-	stack; and {\em (iii)} is efficient for new code that makes use of effect
-	handlers.
+	(i)} imposes a mean 1\% overhead on a comprehensive macro benchmark suite
+	that does not use effect handlers; {\em (ii)} remains compatible with program
+	analysis tools that inspect the stack; and {\em (iii)} is efficient for new
+	code that makes use of effect handlers.
 \end{abstract}
@@
 \begin{enumerate}[label=R\arabic*]
-	\item \textbf{Backwards compatibility.} Extant OCaml programs \kc{Extant
-		better than legacy?} \tk{Existing might be better?} do not break under
-    OCaml extended with effect handlers. OCaml code that does not use effect
-    handlers will pay negligible performance and memory cost.
+	\item \textbf{Backwards compatibility.} Existing OCaml programs do not break
+		under OCaml extended with effect handlers. OCaml code that does not use
+		effect handlers will pay minimal performance and memory cost.
 
@@
 		implementations must handle.
-	\item extensive evaluation which shows that our implementation has negligible
+	\item extensive evaluation which shows that our implementation has minimal
 		impact on code that does not use effect handlers, and serves as an
@@
 average (geometric mean of the normalized values against |stock| as the
-baseline), the multicore variants were \textbf{less than 1\% slower} than
-|stock|. The outliers (on either ends) were due to the difference in the
-allocator and the GC between stock and Multicore OCaml.
+baseline), the multicore variants were less than 1\% slower than |stock|. The
+outliers (on either ends) were due to the difference in the allocator and the
+GC between stock and Multicore OCaml. Of the 54 programs in the benchmark
+suite, 32 programs had an overhead of 5\% or lower, and 8 programs had more
+than 10\% overhead.
 
@@
 \end{tabular}
-\vspace{-5mm}
 }
 \label{tab:micro_noeffect}
+\vspace{-3mm}
 \end{table}
@@
 Our design for effect handlers in OCaml walks the tightrope of maintaining
-compatibility (for profiling, debuggers and negligible overheads for existing
+compatibility (for profiling, debuggers and minimal overheads for existing
 programs), while unlocking the full power of non-local control flow constructs.
@@
 
-%\clearpage
+\section*{Acknowledgements}
+
+This research was funded via a Royal Commission for the Exhibition
+of 1851 and Darwin College Research Fellowships, and by grants from Jane Street
+and Tezos Foundation.
 

### 2021-03-24 retro-concurrency.tex
Updates

@@
 The remainder of the paper continues with a description of the stock OCaml
-program stack (\S\ref{sec:stack}). We describe effect handlers in Multicore
-OCaml focussing on the challenges in retrofitting them into a mainstream
-\emph{systems} language (\S\ref{sec:refine}), followed by the static and
-dynamic semantics for Multicore OCaml effect handlers (\S\ref{sec:semantics}).
-We then discuss the compiler and the runtime system support for implementing
-effect handlers (\S\ref{sec:impl}), and present an extensive performance
-evaluation of effect handlers (\S\ref{sec:eval}) against our design goals
-(\S\ref{sec:req}). Finally, we discuss the related work (\S\ref{sec:related})
-and conclude (\S\ref{sec:conc}).
+program stack (\S\ref{sec:stack}). We then describe effect handlers in
+Multicore OCaml focussing on the challenges in retrofitting them into a
+mainstream \emph{systems} language (\S\ref{sec:refine}), followed by the static
+and dynamic semantics for Multicore OCaml effect handlers
+(\S\ref{sec:semantics}). We then discuss the compiler and the runtime system
+support for implementing effect handlers (\S\ref{sec:impl}), and present an
+extensive performance evaluation of effect handlers (\S\ref{sec:eval}) against
+our design goals (\S\ref{sec:req}). Finally, we discuss the related work
+(\S\ref{sec:related}) and conclude (\S\ref{sec:conc}).
 
@@
 the return address at a given instruction requires interpreting the bytecode
-from the start of the function, which may be potentially very slow. Within each
-function, DWARF maintains a \emph{canonical frame address} (CFA) and is
-traditionally the stack pointer before entering this function. Hence, on
+from the start of the function, which may potentially be very slow.
+
+Within each function, DWARF maintains a \emph{canonical frame address} (CFA)
+and is traditionally the stack pointer before entering this function. Hence, on
 x86-64, where the return address is pushed on the stack on call, the return
-address is at |CFA - 8|.
-
-Our goal is to encode the CFA using DWARF directives when stacks are switched.
-Recall that stack switching occurs in effect handlers, external calls and
-callbacks. When starting to run on a new fiber, we insert DWARF bytecode to
-follow the |parent_fiber| pointer and dereference the |saved_sp| to get the CFA
-(|saved_sp + 8|). During callbacks into OCaml, we save the current system stack
-pointer in the |context block| in Figure~\ref{fig:fiber} to identify the CFA in
-the C stack. DWARF unwinding for external calls is implemented by following a
-link to the current OCaml stack pointer. With these changes, we get the same
-backtrace for the \texttt{meander} program from \S\ref{sec:unwind}, modulo
-runtime system functions due to effect handlers.
+address is at |CFA - 8|. Our goal is to encode the CFA using DWARF directives
+when stacks are switched. Recall that stack switching occurs in effect
+handlers, external calls and callbacks. When starting to run on a new fiber, we
+insert DWARF bytecode to follow the |parent_fiber| pointer and dereference the
+|saved_sp| to get the CFA (|saved_sp + 8|). During callbacks into OCaml, we
+save the current system stack pointer in the |context block| in
+Figure~\ref{fig:fiber} to identify the CFA in the C stack. DWARF unwinding for
+external calls is implemented by following a link to the current OCaml stack
+pointer. With these changes, we get the same backtrace for the \texttt{meander}
+program from \S\ref{sec:unwind}, modulo runtime system functions due to effect
+handlers.
 
@@
 fast unwinding, |perf| would no longer needs to dump the call stack, and hence,
-we can get the backtrace of the entire call stack and not just the current
+we could get the backtrace of the entire call stack and not just the current
 fiber.
@@
 
-This research was funded via a Royal Commission for the Exhibition
-of 1851 and Darwin College Research Fellowships, and by grants from Jane Street
-and Tezos Foundation.
+We thank our shepherd, Matthew Fluet, whose comments substantially improved the
+presentation. This research was funded via a Royal Commission for the
+Exhibition of 1851 and Darwin College Research Fellowships, and by grants from
+Jane Street and Tezos Foundation.
 

### 2021-03-24 retro-concurrency.tex
avoid breaking on bullet numbers

@@
 	other than exceptions. Our implementation of effect handlers for OCaml: {\em
-	(i)} imposes a mean 1\% overhead on a comprehensive macro benchmark suite
-	that does not use effect handlers; {\em (ii)} remains compatible with program
-	analysis tools that inspect the stack; and {\em (iii)} is efficient for new
+	(i)}~imposes a mean 1\% overhead on a comprehensive macro benchmark suite
+	that does not use effect handlers; {\em (ii)}~remains compatible with program
+	analysis tools that inspect the stack; and {\em (iii)}~is efficient for new
 	code that makes use of effect handlers.

### 2021-03-24 retro-concurrency.tex
use perma citations for Lwt and Async

@@
 code~\cite{colour}. For asynchronous I/O, OCaml developers typically use
-libraries such as Lwt~\cite{lwt} and Async~\cite{async}, where asynchronous
+libraries such as Lwt~\cite{lwt} and Async~\cite[\S18]{rwo}, where asynchronous
 functions are represented as monadic computations. In these libraries, while

### 2021-03-24 retro-concurrency.tex
edits

@@
 Since our primary motivation is to enable composable asynchronous I/O, let us
-implement a simple lightweight thread library with support for forking new
+implement a cooperative lightweight thread library with support for forking new
 threads and yielding control to other threads. We will then extend this library
@@
 %
-This uses OCaml standard library's |input_line| function to read a line
+This uses OCaml's standard |input_line| function to read a line
 synchronously from the channel |ic| and resume the continuation |k| with the
@@
 
-Observe that this example does not resume a continuation more than once. This
+This example does not resume a continuation more than once. This
 also holds true for other use cases such as generators and coroutines. Hence,

### 2021-03-25 retro-concurrency.tex
right shepherd :)
14 pages

@@
     \end{lstlisting}
-    %\vspace{-2mm}
+    \vspace{-2mm}
     \subcaption{\texttt{meander.ml}}
@@
     \end{lstlisting}
-    %\vspace{-4mm}
+    \vspace{-4mm}
     \subcaption{\texttt{meander.c}}
@@
 \end{minipage}
-%\vspace{-2mm}
+\vspace{-2mm}
 \caption{Program stack on stock OCaml.}
-%\vspace{-2mm}
+\vspace{-2mm}
 \end{figure*}
@@
 
+\vspace{-1mm}
 \subsection{External calls and callbacks}
@@
 
+\vspace{-1mm}
 \subsection{Exception handlers}
@@
 
+\vspace{-1mm}
 \subsection{Stack unwinding}
@@
 
+\vspace{-1mm}
 \section{Effect Handlers}
@@
 
+\vspace{-1mm}
 \subsection{Asynchronous I/O}
+\label{sec:async_io}
 
@@
     | effect (In_line ic) k ->
-        pending_reads := (ic,k) :: !pending_reads;
-        run_next ()
+      pending_reads := (ic,k)::!pending_reads; run_next ()
   in
@@
 |pending_reads| maintains a list of pending reads and the associated
-continuations (line 4). We handle the |In_line| effect by pushing the pair of
-input channel |ic| and continuation |k| to |pending_reads| (line 24), allowing
-other threads in the scheduler to run.
+continuations (line 4). At line 24, we handle the |In_line| effect by pushing
+the pair of input channel |ic| and continuation |k| to |pending_reads|,
+allowing other threads in the scheduler to run.
 
@@
 
-This example does not resume a continuation more than once. This
-also holds true for other use cases such as generators and coroutines. Hence,
-our continuations are one-shot, and resuming the continuation more than once
-raises an |Invalid_argument| exception. It is well-known that one-shot
-continuations can be implemented efficiently~\cite{Bruggeman96}.
+This example does not resume a continuation more than once. This also holds
+true for other use cases such as generators and coroutines. Hence, our
+continuations are one-shot, and resuming the continuation more than once raises
+an |Invalid_argument| exception. It is well-known that one-shot continuations
+can be implemented efficiently~\cite{Bruggeman96}.
 
@@
   let rec loop () =
-	  output_string oc ((input_line ic) ^ "\n");
-		loop ()
-	in
+	  output_string oc ((input_line ic) ^ "\n"); loop () in
 	try loop () with
@@
 
+\vspace{-1mm}
 \subsection{Static semantics}
@@
 
+\vspace{-1mm}
 \subsection{Dynamic semantics}
@@
 
-\subsubsection{Syntax} Our expressions (Figure~\ref{sem:syntax}) consist of
-integer constants ($n$), variables ($x$), abstraction ($\lam{x}{e}$),
-application ($e~e$), arithmetic expressions ($e \odot e$) where $\odot$ ranges
-over $\{+,-,*,/\}$, raising exceptions ($\throw{l}{e}$), performing effects
-($\perform{l}{e}$), and handling effects ($\handle{e}{h}$). Abstractions come
-in two forms: OCaml abstractions ($\lambda^o$) and C abstractions
-($\lambda^c$). The handler consists of a return case ($\caseval{x}{e}$), zero
-or more exception cases ($\caseexn{l}{x}{e}$) with label $l$, parameter $x$ and
-body $e$, and zero or more effect cases ($\caseeff{l}{x}{k}{e}$) with label
-$l$, parameter $x$, continuation $k$ and body $e$.
+\vspace{-1mm}
+\subsubsection{Syntax}
+
+Our expressions (Figure~\ref{sem:syntax}) consist of integer constants ($n$),
+variables ($x$), abstraction ($\lam{x}{e}$), application ($e~e$), arithmetic
+expressions ($e \odot e$) where $\odot$ ranges over $\{+,-,*,/\}$, raising
+exceptions ($\throw{l}{e}$), performing effects ($\perform{l}{e}$), and
+handling effects ($\handle{e}{h}$). Abstractions come in two forms: OCaml
+abstractions ($\lambda^o$) and C abstractions ($\lambda^c$). The handler
+consists of a return case ($\caseval{x}{e}$), zero or more exception cases
+($\caseexn{l}{x}{e}$) with label $l$, parameter $x$ and body $e$, and zero or
+more effect cases ($\caseeff{l}{x}{k}{e}$) with label $l$, parameter $x$,
+continuation $k$ and body $e$.
 
@@
 
-%\vspace{-1mm}
+\vspace{-1mm}
 \subsubsection{Top-level reductions}
@@
 
-%\vspace{-1mm}
+\vspace{-1mm}
 \subsubsection{C reductions}
@@
 
-%\vspace{-1mm}
+\vspace{-1mm}
 \subsubsection{OCaml reductions}
@@
 Observe that |continue| and |discontinue| are not part of the expressions. They
-are encoded as follows: $\textsf{continue} ~k ~e = (k ~(\olam{x}{x})) ~e$ and
+are encoded as $\textsf{continue} ~k ~e = (k ~(\olam{x}{x})) ~e$ and
 $\textsf{discontinue} ~k ~l ~e = (k ~(\olam{x}{\throw{l}{x}})) ~e$.
@@
 
+\vspace{-2mm}
 \section{Implementation}
@@
 We now present the implementation details of effect handlers in Multicore
-OCaml. We assume an x86\_64 architecture for the remainder of this paper.
-Although Multicore OCaml currently only supports the x86\_64 backend on Linux,
+OCaml. While we assume an x86\_64 architecture for the remainder of this paper,
 our design does not preclude other architectures and OSs.
@@
 \end{minipage}
-%\vspace{-2mm}
+\vspace{-2mm}
 \caption{Layout of Multicore OCaml effect handlers.}
-%\vspace{-3mm}
+\vspace{-3mm}
 \end{figure*}
@@
 
-%\vspace{-3mm}
+\vspace{-2mm}
 \subsection{External calls and callbacks}
@@
 
-%\vspace{-3mm}
+\vspace{-2mm}
 \subsection{Effect handlers}
@@
 
-%\vspace{-3mm}
+\vspace{-2mm}
 \subsection{Stack unwinding}
@@
 registers are stored. To avoid reifying this large table, DWARF directives
-represent the table using a compact bytecode representation. Each block of such
-instructions usually describe the call-frame information for a single function,
-where each instruction describes the table as a delta of the state at the
-previous instruction. While this leads to a compact representation, computing
-the return address at a given instruction requires interpreting the bytecode
-from the start of the function, which may potentially be very slow.
-
-Within each function, DWARF maintains a \emph{canonical frame address} (CFA)
-and is traditionally the stack pointer before entering this function. Hence, on
+represent the table using a compact bytecode representation. This bytecode must
+be interpreted on demand from the start of the function definition to compute
+the call-frame information at any given instruction within the function. For
+each function, DWARF maintains a \emph{canonical frame address} (CFA) and is
+traditionally the stack pointer before entering this function. Hence, on
 x86-64, where the return address is pushed on the stack on call, the return
-address is at |CFA - 8|. Our goal is to encode the CFA using DWARF directives
-when stacks are switched. Recall that stack switching occurs in effect
-handlers, external calls and callbacks. When starting to run on a new fiber, we
-insert DWARF bytecode to follow the |parent_fiber| pointer and dereference the
-|saved_sp| to get the CFA (|saved_sp + 8|). During callbacks into OCaml, we
-save the current system stack pointer in the |context block| in
-Figure~\ref{fig:fiber} to identify the CFA in the C stack. DWARF unwinding for
-external calls is implemented by following a link to the current OCaml stack
-pointer. With these changes, we get the same backtrace for the \texttt{meander}
-program from \S\ref{sec:unwind}, modulo runtime system functions due to effect
-handlers.
-
-We have utilized the DWARF verification tool from~\cite{Bastian19} to flag
-possible errors in the hand written DWARF directives where we have altered the
-OCaml assembler to implement effect handlers. In three places additional manual
-verification was required; two were where the stack is intentionally switched
-between fibers and one was where the control returns to the parent fiber which
-happened to be resized due to effect handler invocation, which the tool could
-not track.
+address is at |CFA - 8|.
+
+Our goal is to encode the CFA using DWARF directives when stacks are switched.
+Recall that stack switching occurs in effect handlers, external calls and
+callbacks. At the entry to a effect handler block, we insert DWARF bytecode to
+follow the |parent_fiber| pointer and dereference the |saved_sp| to get the CFA
+(|saved_sp + 8|). During callbacks into OCaml, we save the current system stack
+pointer in the |context block| in Figure~\ref{fig:fiber} to identify the CFA in
+the C stack. DWARF unwinding for external calls is implemented by following a
+link to the current OCaml stack pointer. With these changes, we get the same
+backtrace for the \texttt{meander} program from \S\ref{sec:unwind}, modulo
+runtime system functions due to effect handlers. We have verified the
+correctness of our DWARF directives using the verification tool from Bastian et
+al.~\cite{Bastian19}.
 
@@
 al.~\cite{Bastian19} report on a technique to pre-compile the unwind table to
-assembly which speeds up DWARF-based unwinding by upto 25$\times$. With this
-fast unwinding, |perf| would no longer needs to dump the call stack, and hence,
-we could get the backtrace of the entire call stack and not just the current
+assembly, which speeds up DWARF-based unwinding by up to 25$\times$. With this
+technique, |perf| can unwind the stack at sample points rather than dumping the
+call stack, which captures the complete backtrace rather than just the current
 fiber.
 
-%\vspace{-2mm}
+\vspace{-2mm}
 \subsection{Garbage collection}
@@
 system resources, assuming that user code does not handle |Unwind| exception
-and fails to re-raise it. But installing a finaliser on every captured
-continuation by default introduces significant overhead to continuation capture
-(\S\ref{sec:res_final_k}). Hence, we choose not to implement such a finaliser by
-default. It is also useful to note that even if the memory for the fiber stack
-is managed by the GC rather than being allocated in the C heap as we do now, we
-would still need a finalisation mechanism to unwind the stack and release other
-system resources that may be held by the continuation.
+and fails to re-raise it. Since installing a finaliser on every captured
+continuation introduces significant overhead (\S\ref{sec:res_final_k}), we
+choose not to do it by default. It is also useful to note that even if the
+memory for the fiber stack is managed by the GC, we would still need a
+finalisation mechanism to unwind the stack and release other system resources
+that may be held by the continuation.
 
@@
 
-%\vspace{-2mm}
+\vspace{-2mm}
 \section{Evaluation}
@@
 
-%\vspace{-3mm}
+\vspace{-2mm}
 \subsection{No effects benchmarks}
@@
 	\includegraphics[scale=0.35]{sandmark-notebook/sandmark_time}
+	\vspace{-4mm}
 	\caption{Normalized time of macro benchmarks. Baseline is Stock OCaml,
 		whose running time in seconds in given in parenthesis.}
-	%\vspace{-2mm}
+	\vspace{-2mm}
 	\label{res:macro_time}
@@
 	\includegraphics[scale=0.35]{sandmark-notebook/sandmark_codesize}
+	\vspace{-4mm}
 	\caption{Normalized OCaml text section size (OTSS) of macro benchmarks.
 	Baseline is Stock OCaml, whose size in kilobytes given in parenthesis.}
-	%\vspace{-2mm}
+	\vspace{-2mm}
 	\label{res:macro_size}
@@
 \label{tab:micro_noeffect}
-\vspace{-3mm}
+\vspace{-5mm}
 \end{table}
@@
 does not need to do this. Callback performance is less important than external
-calls, which are far more numerous. As seen previously, these results however
-do not affect macro benchmark performance.
-
-%\vspace{-3mm}
+calls, which are far more numerous.
+
+\vspace{-2mm}
 \subsection{No perform benchmarks}
@@
 
-As one of our primary goals of adding effect handlers to OCaml is to support
-direct style concurrency, we implement a full-fledged HTTP/1.1 web server using
-effect handlers (|MC|). We use effect handlers to implement a user-level
-threading library that spawns a thread per request. We use
-\texttt{httpaf}~\cite{httpaf} for HTTP handling, and
-\texttt{libev}~\cite{libev} for the eventloop. We compare our implementation
-against an Lwt version (|lwt|) which also uses \texttt{httpaf} and
-\texttt{libev}. Unlike using effect handlers, the Lwt version is written in
-monadic style and does not have the notion of a thread per request. For
-comparison, we include a Go 1.13 version (|go|) that uses the
+Using effect handlers, we have implemented a full-fledged HTTP/1.1 web server
+by extending the example from \S\ref{sec:async_io} (|MC|). The web server
+spawns a lightweight thread per request. We use \texttt{httpaf}~\cite{httpaf}
+for HTTP handling, and \texttt{libev}~\cite{libev} for the eventloop. We
+compare our implementation against an Lwt version (|lwt|) which also uses
+\texttt{httpaf} and \texttt{libev}. Unlike using effect handlers, the Lwt
+version is written in monadic style and does not have the notion of a thread
+per request. For comparison, we include a Go 1.13 version (|go|) that uses the
 \texttt{net/http}~\cite{nethttp} package. As both the OCaml versions are single
@@
 
-%\vspace{-2mm}
+\vspace{-2mm}
 \section{Related Work}
@@
 
-%\vspace{-3mm}
+\vspace{-2mm}
 \section{Conclusions}
@@
 
-We thank our shepherd, Matthew Fluet, whose comments substantially improved the
-presentation. This research was funded via a Royal Commission for the
-Exhibition of 1851 and Darwin College Research Fellowships, and by grants from
-Jane Street and Tezos Foundation.
+We thank the PLDI reviewers and our shepherd, Matthew Flatt, whose comments
+substantially improved the presentation. This research was funded via a Royal
+Commission for the Exhibition of 1851 and Darwin College Research Fellowships,
+and by grants from Jane Street and Tezos Foundation.
 

### 2021-03-25 retro-concurrency.tex
british english to be consistent and expand acronym

@@
 operations from their implementation, effect handlers enable new approaches to
-modular programming. Effect handlers are a generalization of exception
+modular programming. Effect handlers are a generalisation of exception
 handlers, where, in addition to the effect being handled, the handler is
@@
 OCaml. While we assume an x86\_64 architecture for the remainder of this paper,
-our design does not preclude other architectures and OSs.
+our design does not preclude other architectures and operating systems.
 

### 2021-03-25 retro-concurrency.tex
tweak
typo and add acks

@@
 must generate DWARF stack unwind tables for stacks that may be non-contiguous.
-In this work, we develop the compiler and runtime support for implementing
+In this work, we develop the compiler and runtime support required for implementing
 efficient effect handlers for OCaml that satisfy these requirements.
@@
 of the current stack as in stock OCaml. Using effect handlers in a system such
-as a web server aides debugging and profiling because it is possible to get a
-backtrace snapshot of all current requests. Such a feature is available in
+as a web server aids debugging and profiling because it is possible to get a
+backtrace snapshot of all current requests. This feature is available in
 Go~\cite{gopprof}, but not in OCaml concurrency libraries such as Async and
@@
 
-We thank the PLDI reviewers and our shepherd, Matthew Flatt, whose comments
-substantially improved the presentation. This research was funded via a Royal
-Commission for the Exhibition of 1851 and Darwin College Research Fellowships,
-and by grants from Jane Street and Tezos Foundation.
+We thank Sam Lindley, Franc\'{o}is Pottier, the PLDI reviewers and our shepherd, Matthew Flatt, whose comments substantially improved the presentation.
+This research was funded via Royal Commission for the Exhibition of 1851 and Darwin College Research Fellowships,
+and by grants from Jane Street and the Tezos Foundation.
 

### 2021-03-26 retro-concurrency.tex
tweak fig 1 layout to make (a) (b) divide more obvious
Minor edits

@@
     \begin{lstlisting}[language=c,numbers=left]
+
 #include <caml/mlvalues.h>
@@
 of Common Lisp |unwind-protect|~\cite{Steele90}, which ensures de-allocation
-(and re-allocation) of resources every time the non-local control leaves and
-enters back into a context. |dynamic-wind| is not quite sufficient for resource
-clean up~\cite{Kiselyov,Sitaram03} as returning exits must be distinguished
-from non-returning ones, and resource cleanup is only necessary during
-non-returning exits.
+and re-allocation of resources every time the non-local control leaves and
+enters back into a context. |dynamic-wind| is not quite the right abstraction
+as resources need to be cleaned up only on non-returning
+exists~\cite{Kiselyov,Sitaram03}. This requires distinguishing returning exits
+from non-returning ones.
 
@@
 
-We thank Sam Lindley, Franc\'{o}is Pottier, the PLDI reviewers and our shepherd, Matthew Flatt, whose comments substantially improved the presentation.
-This research was funded via Royal Commission for the Exhibition of 1851 and Darwin College Research Fellowships,
-and by grants from Jane Street and the Tezos Foundation.
+We thank Sam Lindley, Franc\'{o}is Pottier, the PLDI reviewers and our
+shepherd, Matthew Flatt, whose comments substantially helped improve the
+presentation. This research was funded via Royal Commission for the Exhibition
+of 1851 and Darwin College Research Fellowships, and by grants from Jane Street
+and the Tezos Foundation.
 

### 2021-03-30 retro-concurrency.tex
updates
cleanup

@@
 handlers, where, in addition to the effect being handled, the handler is
-provided with the delimited continuation~\cite{Danvy90} of the perform site,
-which may be resumed later. This enables non-local control-flow mechanisms such
-as resumable exceptions, lightweight threads, coroutines, generators and
-asynchronous I/O to be composably expressed.
+provided with the delimited continuation~\cite{Danvy90} of the perform site.
+This continuation may be used to resume the suspended computation later. This
+enables non-local control-flow mechanisms such as resumable exceptions,
+lightweight threads, coroutines, generators and asynchronous I/O to be
+composably expressed.
 
@@
 There are many research languages and libraries built around effect
-handlers~\cite{Leijen14,Hillerstrom20,Pyro,Frank,Eff}. Unlike these efforts,
-our goal is to retrofit effect handlers onto the OCaml programming language,
-which has been in continuous use for the past 25 years in large codebases
-including verification tools~\cite{everest,Coq}, mission critical software
-systems~\cite{astree} and latency sensitive networked
+handlers~\cite{Leijen14,Biernacki20,Hillerstrom20,Pyro,Frank,Eff}. Unlike these
+efforts, our goal is to retrofit effect handlers onto the OCaml programming
+language, which has been in continuous use for the past 25 years in large
+codebases including verification tools~\cite{everest,Coq}, mission critical
+software systems~\cite{astree} and latency sensitive networked
 applications~\cite{Madhavapeddy13}. OCaml is particularly favoured for its
@@
 
+\vspace{-1mm}
 \subsection{Requirements}
@@
 OCaml functions, irrespective of whether they use effect handlers, need to
-perform stack overflow checks, and external C functions (which do not have stack
-overflow checks) must be performed on a separate system stack. Additionally, we
-must generate DWARF stack unwind tables for stacks that may be non-contiguous.
-In this work, we develop the compiler and runtime support required for implementing
-efficient effect handlers for OCaml that satisfy these requirements.
+perform stack overflow checks, and external C functions (which do not have
+stack overflow checks) must be performed on a separate system stack.
+Additionally, we must generate DWARF stack unwind tables for stacks that may be
+non-contiguous. In this work, we develop the compiler and runtime support
+required for implementing efficient effect handlers for OCaml that satisfy
+these requirements.
 
@@
 	\item a formal operational semantics for the effect handler implementation in
-		OCaml. Our formalism explicitly models the subtle interactions with the C
-		stack, which is generally overlooked by other formal models, but which the
+		OCaml. Our formalism explicitly models the interactions with the C stack,
+		which is generally overlooked by other formal models, but which the
 		implementations must handle.
@@
 Hence, we retain the linked exception handler frame implementation of stock
-OCaml exception handlers in Multicore OCaml for performance backwards
-compatibility. This differs from other research languages with effect
+OCaml in Multicore OCaml to ensure performance backwards compatibility. This
+differs from other research languages with effect
 handlers~\cite{Hillerstrom20,Frank,Eff}, which implement exceptions using
@@
 to restore when handling an effect or resuming a continuation. This fortuitous
-design choice in stock OCaml has a significant impact towards fast switching
-between fibers in Multicore OCaml.
+design choice in stock OCaml has a significant impact in enabling fast
+switching between fibers in Multicore OCaml.
 
@@
 registers are stored. To avoid reifying this large table, DWARF directives
-represent the table using a compact bytecode representation. This bytecode must
-be interpreted on demand from the start of the function definition to compute
-the call-frame information at any given instruction within the function. For
-each function, DWARF maintains a \emph{canonical frame address} (CFA) and is
-traditionally the stack pointer before entering this function. Hence, on
-x86-64, where the return address is pushed on the stack on call, the return
-address is at |CFA - 8|.
-
-Our goal is to encode the CFA using DWARF directives when stacks are switched.
-Recall that stack switching occurs in effect handlers, external calls and
-callbacks. At the entry to a effect handler block, we insert DWARF bytecode to
-follow the |parent_fiber| pointer and dereference the |saved_sp| to get the CFA
-(|saved_sp + 8|). During callbacks into OCaml, we save the current system stack
-pointer in the |context block| in Figure~\ref{fig:fiber} to identify the CFA in
-the C stack. DWARF unwinding for external calls is implemented by following a
-link to the current OCaml stack pointer. With these changes, we get the same
-backtrace for the \texttt{meander} program from \S\ref{sec:unwind}, modulo
-runtime system functions due to effect handlers. We have verified the
-correctness of our DWARF directives using the verification tool from Bastian et
-al.~\cite{Bastian19}.
+represent the table using a compact bytecode representation that describes the
+unwind table as a sequence of edits from the start of the function. In order to
+compute the call-frame information at any given instruction within a function,
+the DWARF bytecode from the start of that function must be interpreted on
+demand. For each function, DWARF maintains a \emph{canonical frame address}
+(CFA) and is traditionally the stack pointer before entering this function.
+Hence, on x86-64, where the return address is pushed on the stack on call, the
+return address is at |CFA - 8|.
+
+Our goal is to compute the CFA of the caller when stacks are switched using the
+DWARF directives. Recall that stack switching occurs in effect handlers,
+external calls and callbacks. At the entry to an effect handler block, we insert
+DWARF bytecode to follow the |parent_fiber| pointer and dereference the
+|saved_sp| to get the CFA (|saved_sp + 8|). During callbacks into OCaml, we
+save the current system stack pointer in the |context block| in
+Figure~\ref{fig:fiber} to identify the CFA in the C stack. DWARF unwinding for
+external calls is implemented by following a link to the current OCaml stack
+pointer. With these changes, we get the same backtrace for the \texttt{meander}
+program from \S\ref{sec:unwind}, modulo runtime system functions due to effect
+handlers. We have verified the correctness of our DWARF directives using the
+verification tool from Bastian et al.~\cite{Bastian19}.
 
@@
 technique, |perf| can unwind the stack at sample points rather than dumping the
-call stack, which captures the complete backtrace rather than just the current
-fiber.
-
-\vspace{-2mm}
+call stack, which would capture the complete backtrace rather than just the
+current fiber.
+
+\vspace{-1mm}
 \subsection{Garbage collection}
@@
 
-For stack walking during GC in order to find the roots, the only change
-necessary was to make the runtime aware of the fiber layout and how to follow
-the parent pointers. The stack map generation and scanning each frame for roots
-remains the same. The challenges and the solutions for integrating fibers with
-the concurrent mark-and-sweep GC of Multicore OCaml have been discussed
+The challenges and the solutions for integrating fibers with the concurrent
+mark-and-sweep GC of Multicore OCaml have been discussed
 previously~\cite{Sivaramakrishnan20}.
 
-\vspace{-2mm}
+%\vspace{-1mm}
 \section{Evaluation}
@@
 against the performance requirements set in \S\ref{sec:req}. Multicore OCaml is
-an extension of the 4.10.0 OCaml compiler with support for shared memory
+an extension of the OCaml 4.10.0 compiler with support for shared memory
 parallelism and effect handlers. Since our objective is to evaluate the impact
@@
 
-\vspace{-2mm}
+\vspace{-1mm}
 \subsection{No effects benchmarks}
@@
 as resources need to be cleaned up only on non-returning
-exists~\cite{Kiselyov,Sitaram03}. This requires distinguishing returning exits
+exits~\cite{Kiselyov,Sitaram03}. This requires distinguishing returning exits
 from non-returning ones.

### 2021-03-30 retro-concurrency.tex
use the right async citation later in the paper too

@@
 Our concurrency monad (and other monadic concurrency libraries such as
-Lwt~\cite{lwt} and Async~\cite{async}) also have other downsides -- exceptions,
+Lwt~\cite{lwt} and Async~\cite[\S18]{rwo}) also have other downsides -- exceptions,
 backtraces, and DWARF unwinding are no longer useful due to the lack of a

### 2021-03-31 retro-concurrency.tex
Update for camera ready

@@
 \newcommand{\caseval}[2]{\kw{return}~#1 \mapsto #2}
-\newcommand{\caseexn}[3]{\kw{exception} ~#1 ~#2 \mapsto #3}
+\newcommand{\caseexn}[3]{\kw{exception} ~#1 ~#2 \allowbreak \mapsto #3}
 \newcommand{\caseeff}[4]{\kw{effect} ~#1 ~#2 ~#3 \mapsto #4}
@@
   \begin{minipage}{\linewidth}
-    \begin{lstlisting}[numbers=left]
+		\begin{lstlisting}[xleftmargin=0.05\linewidth]
 external ocaml_to_c
@@
   \begin{minipage}{\linewidth}
-    \begin{lstlisting}[language=c,numbers=left]
-
+		\begin{lstlisting}[language=c, xleftmargin=0.05\linewidth]
 #include <caml/mlvalues.h>
@@
 
-\begin{lstlisting}[numbers=left]
+\begin{lstlisting}[numbers=left, xleftmargin=0.025\textwidth]
 let run main =
@@
 
-\begin{lstlisting}[numbers=left]
+\begin{lstlisting}[numbers=left, xleftmargin=.025\textwidth]
 let run main =
@@
         let compl,todo = do_reads todo in
-        List.iter (fun (str,k) ->
-          Queue.push (fun () -> continue k str) runq) compl;
+        List.iter (fun (s,k) ->
+         Queue.push (fun () -> continue k s) runq) compl;
         pending_reads := todo;
@@
 
+\vspace{-1mm}
 \section{Semantics}
@@
 As mentioned earlier, effect handlers in Multicore OCaml do not guarantee
-effect safety, but only guarantee type safety. Programs without matching effect
-handlers are well-typed Multicore OCaml programs. As a result, our static
-semantics is simpler than languages that ensure effect safety~\cite{Eff,
-Hillerstrom20, Leijen14, Effekt, Frank, Biernacki19}. This is important for
-backwards compatibility as our goal is to retrofit effect handlers to a
-language with large legacy codebases; programs that do not use effects remain
-well-typed, and those that do compose well with those that don't.
+effect safety. Programs without matching effect handlers are well-typed
+Multicore OCaml programs. As a result, our static semantics is simpler than
+languages that ensure effect safety~\cite{Eff, Hillerstrom20, Leijen14, Effekt,
+Frank, Biernacki19}. This is important for backwards compatibility as our goal
+is to retrofit effect handlers to a language with large legacy codebases;
+programs that do not use effects remain well-typed, and those that do compose
+well with those that don't.
 
@@
 
-For handling the effects, our implementation extends OCaml's |match ... with|
-syntax with effect patterns. The expression
+For effect handling, we extend OCaml's |match ... with| syntax with effect
+patterns. The expression
 \begin{lstlisting}
@@
 %
-\begin{minipage}{\linewidth}
+\begin{minipage}{0.25\linewidth}
 	\begin{smathpar}
-		\begin{array}{cc}
+		\begin{array}{c}
 			\textsc{StepC} ~
 			\inferrule{(\term, \env, \fl, \ostack) \cstep \config}
-								{\configa{\term}{\env}{\cstacka{\fl}{\ostack}} \step \config} &
+								{\configa{\term}{\env}{\cstacka{\fl}{\ostack}} \step \config} \\ \\
 			\textsc{StepO} ~
@@
 %
-\begin{minipage}{\linewidth}
+\begin{minipage}{0.69\linewidth}
   \begin{smathpar}
@@
 
-\vspace{-2mm}
+%\vspace{-1mm}
 \subsection{External calls and callbacks}
@@
 
-\vspace{-2mm}
+%\vspace{-1mm}
 \subsection{Effect handlers}
@@
 \begin{figure*}
-	\includegraphics[scale=0.35]{sandmark-notebook/sandmark_time}
+	\includegraphics[scale=0.34]{sandmark-notebook/sandmark_time}
 	\vspace{-4mm}
@@
 \begin{figure*}
-	\includegraphics[scale=0.35]{sandmark-notebook/sandmark_codesize}
+	\includegraphics[scale=0.34]{sandmark-notebook/sandmark_codesize}
 	\vspace{-4mm}
@@
 \bibliographystyle{ACM-Reference-Format}
+\balance
 \bibliography{retro-concurrency}
 
-\if{0}
-\clearpage
-
-\section*{Supplementary Material}
-\setcounter{figure}{0}
-
-\subsection*{Dynamic Semantics}
-
-\begin{definition}[Initial configuration]
-	Given an expression $e$, we define the initial configuration $\textsf{init}(e)$
-	to be $\configa{e}{\emptyset}{\cstacka{[]}{\bullet}}$.
-\end{definition}
-
-Intuitively, the initial environment is empty and the initial stack is an empty
-C stack.
-
-\begin{definition}[Final value configuration]
-	Given a configuration $\config$, we say that the configuration is the final
-	value configuration (written $\textsf{finalval}(\config)$) if ~$\config =
-	\configa{v}{\env}{\cstacka{[]}{\bullet}}$~ for some value $v$ and environment
-	$\env$.
-\end{definition}
-
-\begin{definition}[Final exception configuration]
-	Given a configuration $\config$, we say that the configuration is the final
-	exception configuration (written $\textsf{finalexn}(\config)$) if ~$\config
-	= \configa{v}{\env}{\cstacka{\ffun{\exnval{l}}::\fl}{\bullet}}$~ for some
-	value $v$, label $l$, frame list $\fl$ and environment $\env$.
-\end{definition}
-
-Given a configuration $\config = \configa{\term}{\env}{\stack}$, we define
-$\textsf{term}(\config)$ to be $\tau$. We define the relation $\multistep$ as
-the transitive reflexive closure of the top-level reduction relation $\step$.
-
-\subsubsection*{Example 1}
-
-Let
-
-\begin{smathpar}
-	\begin{array}{rcll}
-		e & = & (\olam{x}{ & \kw{match} ~\perform{E}{~0}~ \kw{with} \\
-			&   &            & \mid \caseexn{\textsf{Unhandled}}{x}{42} \\
-			&   &            & \mid \caseval{v}{v}})~0
-	\end{array}
-\end{smathpar}
-
-then $init(e) \multistep ~\config \wedge final(\config) \wedge term(\config) =
-42$.
-
-\begin{proof}
-\end{proof}
-
-\subsection*{Experimental support for multi-shot continuations}
-
-Languages with effect handlers typically allow continuations to be resumed more
-than once to enable use cases such as backtracking search and probabilistic
-programming (for example, see Uber's Pyro language~\cite{Pyro}). Multicore
-OCaml has an experimental feature for explicitly cloning continuations, which
-makes a copy of each of its constituent fibers.
-
-Multi-shot continuations are difficult to
-reason about in the presence of linear resources such as file descriptors~\cite{de-vilhena-pottier-21}.
-Recall that OCaml programs are written with the expectation that function calls
-return \emph{exactly-once} either normally or exceptionally. With multi-shot
-continuations, they may return \emph{more than once}, leading to errors such as
-double-freeing malloced memory. Since our goal is to retrofit effect handlers
-onto OCaml, we relegate continuation cloning to the OCaml |Obj| module, which
-contains unsafe features.
-
-In addition, multi-shot continuations currently are not safe under compiler
-optimisations. For example consider the following program:
-
-\begin{minipage}{\linewidth}
-\begin{lstlisting}
-effect E : unit
-let foo () = perform E
-let bar f =
-  let r = ref 0 in
-	f ();
-  r := !r + 1;
-  !r
-;;
-match bar foo with
-| v -> v
-| effect E k ->
-    let k' = Obj.clone_continuation k in
-    let r1 = continue k () in
-    let r2 = continue k' () in
-    Printf.printf "%d %d\n" r1 r2;
-		0
-\end{lstlisting}
-\end{minipage}
-
-This program is expected to print |1 2|. However, OCaml optimizes |bar| by
-converting the heap allocation of |r| to a stack allocation. As a result, each
-continuation ends up with a copy of |r|, and the program prints |1 1|. Thus, in
-the presence of multi-shot continuations, compiler optimisations need to be
-made aware of the effects of a function. We leave proper support for multi-shot
-continuations to future work.
-\fi
-
 \end{document}

### 2021-03-31 retro-concurrency.tex
Type 1 fonts in the webserver benchmarks
Virtually in Canada not the UK

@@
 \acmISBN{978-1-4503-8391-2/21/06}
-\acmConference[PLDI '21]{Proceedings of the 42nd ACM SIGPLAN International Conference on Programming Language Design and Implementation}{June 20--25, 2021}{Virtual, UK}
-\acmBooktitle{Proceedings of the 42nd ACM SIGPLAN International Conference on Programming Language Design and Implementation (PLDI '21), June 20--25, 2021, Virtual, UK}
+\acmConference[PLDI '21]{Proceedings of the 42nd ACM SIGPLAN International Conference on Programming Language Design and Implementation}{June 20--25, 2021}{Virtual, Canada}
+\acmBooktitle{Proceedings of the 42nd ACM SIGPLAN International Conference on Programming Language Design and Implementation (PLDI '21), June 20--25, 2021, Virtual, Canada}
 
@@
 
+\author{Tom Kelly}
+\affiliation{
+  \institution{\small OCaml Labs}            %% \institution is required
+	\city{\small Cambridge}
+	\country{\small UK}
+}
+\email{tom.kelly@cantab.net}
+
 \author{Sadiq Jaffer}
@@
 \email{sadiq@toao.com}
-
-\author{Tom Kelly}
-\affiliation{
-  \institution{\small OCaml Labs}            %% \institution is required
-	\city{\small Cambridge}
-	\country{\small UK}
-}
-\email{tom.kelly@cantab.net}
 
